### 2020-05-14 README.Rmd
modified the DESCRIPTION, added several other files

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Estimating total escapment past Lower Granite Dam using all possible data"
```

# ManuscriptSTADEM

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/KevinSee/ManuscriptSTADEM/master?urlpath=rstudio)

This repository contains the data and code for our paper:

> See, K.E., R.N. Kinzer and M.W. Ackerman, (2020). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> See, K.E., R.N. Kinzer and M.W. Ackerman, (2020). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> See, K.E., R.N. Kinzer and M.W. Ackerman, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to download or install

You can download the compendium as a zip from from this URL: <https://github.com/KevinSee/ManuscriptSTADEM/archive/master.zip>

Or you can install this compendium as an R package, ManuscriptSTADEM, from GitHub with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
remotes::install_github("KevinSee/ManuscriptSTADEM")
```

To

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2020-08-07 analysis/paper/appendices/Appendix_STADEM_descrip.Rmd
updated manuscript Rmd to match latest Word version from Mike.

# Appendix A - STADEM Model Description {#append1}

## Total and Weekly Escapement

Escapement at LGD is estimated by combining two independent observations, trap catches and window counts, of the true number of fish crossing the dam in a state-space model with a weekly time-step. Both are assumed to be corrupted observations of the true unknown number of fish crossing LGD each week. The log of the true number of fish crossing ($X_t$), is modeled as a random walk process [@Shumway2010].

$$
\begin{aligned}
  \log \left( X_{t} \right) &= \log \left( X_{t-1} \right) + e_t \\
  e_t &\sim \mathcal{N}(0, \sigma^2_X)
\end{aligned}
$$

The number of fish caught in the trap, $Y^T_t$, for week $t$ is modeled as a binomial process based on the unknown true trap rate that week, $\nu_t$, and the unknown true number of fish crossing the dam that week, $X_t$. The true trap rate is estimated from a beta distribution with previously estimated parameters, $\hat{\alpha_t}$ and $\hat{\beta_t}$, informed by the mark-recapture estimate of the trap rate.

$$
\begin{aligned}
  \nu_t &\sim \text{Beta} \left(\hat{\alpha_t}, \hat{\beta_t} \right) \\
  Y^T_t &\sim \text{Bin} \left( \nu_t, X_t \right) \\
\end{aligned}
$$

The estimate of the weekly trap rate is derived based on previously PIT-tagged spring/summer Chinook and steelhead who are crossing LGD that week. The fish caught in the trap that week are considered the "mark" group, and all the PIT tagged fish who are detected at the upper end of the LGD fish ladder that week are considered the second capture group (which includes recaptures of the the "marked" fish). From this data, we can estimate the total number of previoiusly PIT tagged fish crossing LGD that week, and the proportion of those that are caught in the trap is the weekly trap rate. The uncertainty in estimates of total fish is translated into uncertainty in the trap rate, and this is then summarised with a beta distribution ($\text{Beta} \left(\hat{\alpha_t}, \hat{\beta_t} \right)$) that becomes an input into STADEM. Although the group of previously PIT tagged fish is not assumed to be representative of the overall run, the rate at which they are caught in the trap should be the same rate that the overall run experiences.

The number of fish counted at the window, $Y^W_t$, is modeled as a (potentially) over-dispersed negative binomial process, with an expected value of $X_t^{day}$, the number of fish crossing the dam while the window is open. This is simply the total number of fish crossing that week, $X_t$, multiplied by the propoortion of fish crossing while the window is open for counting, $\theta_t$, calculated on a weekly basis. In the formula below, $p_t$ is the proportion of fish observed at the window and $r$ is the shape parameter. If $r$ is estimated to be small provides evidence for over-dispersion, and as it grows very large, the negative binomial distribution behaves like a Poisson distribution.

$$
\begin{aligned}
  X_t^{day} &= X_t * \theta_t \\
  p_t &= \frac{r}{\left( r + X_t^{day} \right)} \\
  Y^W_t &\sim \text{NegBin} \left( p_t, r \right) \\
\end{aligned}
$$

Thus, the unknown true number of fish crossing LGD each week, $X_t$, is estimated from two different data source: window counts and fish sampled in the trap. The window counts provide an estimate (with some potential observer error) of the fish crossing during daytime hours, while the fish in the trap, when expanded by the estimated true trap rate, provide an estimate of the total fish crossing that week. For weeks when we have a more precise estimate of the trap rate (i.e. weeks when lots of previously PIT tagged fish are crossing LGD), STADEM will tend to favor the estimate of total escapement based on the trap data, whereas when that trap rate is more uncertain (e.g. fewer PIT tagged fish to use in estimating the trap rate), STADEM will rely more on the window counts to estimate total escapement. During peak run times, when lots of fish are crossing LGD, estimates based on trap data and trap rates will be more precise, while estimates from the window counts may have more observation error due to so many fish passing the window. For weeks when the trap is down, STADEM relies exclusively on the window counts and night-time passage data, but there will be more uncertainty in the estimates.

## Day-time Passage and Re-ascension Rates

There are two other processes that must be accounted for, first, the proportion of fish that cross the dam while the window is closed for counting (night-time passage rate), and the second, the proportion of fish that are crossing the dam multiple times (re-ascension rate) and therefore potentially double-counted. Both rates can be estimated from previously PIT tagged fish that are crossing the dam each week.

The proportion of fish passing the window during non-operational hours, night-time passage rate, is just the complement of the rate of fish passing during the day when the window is operating. The daytime passage rate for week $t$, $\theta_t$, is modeled as a random walk process and estimated from a binomial distribution based on the number of PIT tags observed to cross the dam during operational hours, $y^{day}_t$, and the total number of PIT tags observed to cross the dam at any point that week, $N_t$ [@Shumway2010].

$$
\begin{aligned}
  y^{day}_t &\sim \text{Bin} \left(\theta_t, N_t \right) \\
  \text{logit} (\theta_{t}) &= \text{logit} \left( \theta_{t-1} \right) + g_t \\
  g_t &\sim \mathcal{N}(0, \sigma^2_\theta) \\
\end{aligned}
$$

The number of total fish crossing Lower Granite differs from the number of unique fish crossing Lower Granite because some fish fall back and re-ascend the dam. These fish are potentially double-counted at the window, and have the potential to be caught in the fish trap more than once. The number of tags known to be re-ascending the dam each week, $y^{reasc}_t$, is modeled as a binomial process based on the estimated re-ascension rate, $\eta_t$, and the total number of tags crossing the dam that week, $N_t$. The logit of the re-ascension rate is modeled as a random walk process similar to day-time passage [@Shumway2010].

$$
\begin{aligned}
  y^{reasc}_t &\sim \text{Bin} \left(\eta_t, N_t \right) \\
  \text{logit} \left( \eta_{t} \right) &= \text{logit} \left(\eta_{t-1} \right) + f_t \\
  f_t &\sim \mathcal{N}(0, \sigma^2_\eta)
\end{aligned}
$$

## Origin Proportions

After estimating the total number of fish to have crossed Lower Granite each week, $X_t$, the total must be further refined into the number of wild fish, $X_{w,t}$, hatchery fish, $X_{h,t}$ and hatchery no-clip fish, $X_{hnc, t}$. This is done by estimating a weekly origin proportion vector, $\omega_t$ based on the random sample of fish caught in trap that week, $Y^T_t$. The observed number of wild, $Y^T_{w,t}$, hatchery, $Y^T_{h,t}$, and hatchery no-clip, $Y^T_{hnc,t}$, fish caught in the trap that week is assumed to come from a multinomial distribution with probability vector $\omega_t$. The log-odds ratio of the proportions in $\omega_t$, in relation to the proportion of hatchery fish, $\omega_{h,t}$ is modeled as a random walk, so it can change through time. This allows the proportions of wild, hatchery and hatchery no-clip fish to shift throughout the season, based on the data available from the fish trap.

$$
\begin{aligned}
  \left( Y^T_{w,t}, Y^T_{h,t}, Y^T_{hnc,t} \right) &\sim \text{Multinom} \left( \omega_t, Y^T_t \right) \\
  \omega_t &= \frac{\exp(\phi_t)}{\sum{\exp(\phi_t})} \\
  \phi_{w,t} &= \log \left( \frac{\omega_{w,t}}{\omega_{h,t}} \right) \\
  \phi_{hnc,t} &= \log \left( \frac{\omega_{hnc,t}}{\omega_{h,t}} \right) \\
  \phi_{h,t} &= 0 \\
  \phi_{w,t} &= \phi_{w, t-1} + d_{w,t} \\
  \phi_{hnc,t} &= \phi_{hnc, t-1} + d_{hnc,t} \\
  d_t &\sim \mathcal{N}(0, \sigma^2_\omega) \\
\end{aligned}
$$

Finally, the number of unique fish crossing Lower Granite each week, $X_{w,t}$, is the product of the total fish crossing that week, $X_t$ multiplied by one minus the re-ascension rate, $(1 - \eta_t)$, and the origin proportion vector, $\omega_t$.

$$
\begin{aligned}
  \begin{bmatrix} X_{w,t} \\ X_{h,t} \\ X_{hnc,t} \end{bmatrix}
  &= X_t * (1 - \eta_t) * \begin{bmatrix} \omega_{w,t} \\ \omega_{h,t} \\ \omega_{hnc,t} \end{bmatrix}
\end{aligned}
$$

## Model Fitting

The model was fit using the JAGS program [@R-rjags], run with R software [@R-Core-Team2020]. Uninformative priors were used for $\sigma_X, \sigma_\eta, \sigma_\theta, \sigma_\omega$ and $\log(X_1)$ (Uniform(0,10)), as well as logit($\eta_1$) and logit($\theta_1$) ($\mathcal{N}(0, 1000)$), and finally $\phi_{w,1}$ and $\phi_{hnc,1}$ ($\mathcal{N}(0, 100)$).

### 2020-08-07 analysis/paper/appendices/Appendix_simulation_details.Rmd
updated manuscript Rmd to match latest Word version from Mike.

# Appendix B - Simulation Details {#append2}

To simulate fish passing a dam, an $\mathbb{R}$ software function was developed [@R-Core-Team2020]. The function randomly samples observations from assumed probability distribution functions ($\mathcal{PDF}$) with known parameters. Total unique fish, $N$, and a vector, $\omega$, containing the proportions of wild ($w$), hatchery ($h$) and hatchery no-clip ($hnc$) fish passing the dam is set to establish known "truths" of escapement by origin.

$$
\begin{aligned}
\left[N_{w}, N_{h}, N_{hnc} \right] &= N * \left[\omega_{w}, \omega_{h}, \omega_{hnc} \right]
\end{aligned}
$$

Escapement is then randomly divided across a set number of populations, $n$, by randomly drawing proportions, $\phi_{j,p}$, of origin group $j$ in each poplulation $p$ using a dirichlet $\mathcal{PDF}$. The dirichlet function is parameterized from a vector, $\eta_j$, containing 1's and 0's designating populations with origin $j$ fish returning. $\eta_j$ originates from a bernoulli $\mathcal{PDF}$ and the proportion of populations with each origin, $\tau_j$. Wild fish are assumed to be in all populations; $\tau_w = 1.0$. The product of sampled population proportions $\phi_{j,p}$ and fixed $N_j$ yields a random variable of abundance for each origin in each population, $N_{j,p}$. Summing across origin abundances then gives a random total population abundance, $N_p$, crossing the dam.

$$
\begin{aligned}
  \left[\eta_{j,p=1},  ... , \eta_{j,p=n}\right] &\sim \text{Bern}(\tau_j) \\
  \left[\phi_{j,p=1},  ... , \phi_{j,p=n}\right] &\sim \text{Dir}(\eta_{j,p=1},  ... , \eta_{j,p=n}) \\
  N_{j,p} &= N_j * \phi_{j,p} \\
  N_p &= \sum_{j=w, h, hnc} N_{j,p}
\end{aligned}
$$

Mean arrival date, $\bar{a}_p$, for each population returning to the dam is drawn from a normal $\mathcal{PDF}$ with hyper-parameters $\mu_a$ and $\sigma^2_a$. Similarly, the variance or spread in run-timing within populations is the absolute value of random variables drawn from a normal $\mathcal{PDF}$ with hyper-parameters $\mu_s$ and $\sigma^2_s$.

$$
\begin{aligned}
 \left[\bar{a}_p, ..., \bar{a}_n\right] &\sim \text{Norm}(\mu_a, \sigma^2_a) \\
 \left[s_p, ..., s_n\right] &\sim \text{Norm}(\mu_s, \sigma^2_s)
\end{aligned}
$$

After sampling the mean date of arrival and variances for each population, the date of arrival, $a_{i,p}$, for individual fish, $i$, within each population are drawn from a normal $\mathcal{PDF}$ with population parameters $\bar{a}_p$ and $s^2_p$. This simulates a random arrival day that is similar for all fish returning to the same population, regardless of origin.

$$
\begin{aligned}
 date_{i,p} &\sim \text{Norm}(\bar{a}_p, s^2_p) \\
\end{aligned}
$$

To examine the sensitivites of models to different fish behavior and dam operational scenarios, seven addtional attributes are randomly assigned to each individual fish. Each attribute is randomly assigned a TRUE/FALSE using a bernoulli $\mathcal{PDF}$ and a fixed probability parameter. Fish passage during the day-time (i.e., during periods of window operation) is modeled using one minus the night-time passage rate ($1 - \nu$). Window observations are conditioned on fish passing during the day and being observed at a set rate, $\gamma$. Whether fish $i$ is sampled by the adult trap is modeled on the weekly set trap rate, $\delta_t$. The rate of previously PIT-tagged fish is determined by $\lambda$, and their subsequent detection at the ladder PIT antenna is governed by $\kappa$. Fallback behavior is modeled with a common rate across all populations, $\psi$. Re-ascension occurs with probability $\rho$, conditioned on fish $i$ falling back. If fish $i$ falls back and re-ascends, the entire process described above is repeated, with some time-lag between initial ascension and re-ascension that is governed by a Poisson $\mathcal{PDF}$ with mean = 2 days. Fish may fall-back and re-ascend up to 3 times, allowing for the possibility of the same fish being counted or trapped multiple times.

$$
\begin{aligned}
  day_{i} &\sim \text{Bern}(1-\nu) \\
  window_{i|d = TRUE} &\sim \text{Bern}(\gamma) \\
  trapped_{i} &\sim \text{Bern}(\delta_t) \\
  tagged_{i} &\sim \text{Bern}(\lambda) \\
  ladder_{i|m = TRUE} &\sim \text{Bern}(\kappa) \\  
  fallback_{i} &\sim \text{Bern}(\psi) \\
  re-ascend_{i|f = TRUE} &\sim \text{Bern}(\rho)
\end{aligned}
$$

Simulation parameters for model evaluations were set to mimic typical ecapement of spring/summer Chinook to LGD with similar origin proportions, marking rates and run timing as those observed from return years 2010 - 2015. Escapement of each origin ($N_j$) was set at 25,000 wild, 70,000 hatchery and 5,000 hatchery no-clips spread randomly across 25 populations ($n$). Of the 25 populations, each had a 1.0 probability of containing wild fish, 0.50 probability of having hatchery fish and 0.15 probability of receiving hatchery no-clip ($\tau_j$); resulting in an expected 25 wild, 12.5 hatchery and 3.75 hatchery no-clip populations. Mean arrival dates and variability were estimated from PIT-tag detection data queried from the Columbia Basin Research Data Access in Real Time ([DART](http://www.cbr.washington.edu/dart/query/pit_adult_window)) website and organized by release subbasin. Mean arrival date across all subbasins and 2010 - 2015 return years was June $19^{th}$ ($\mu_a = 171$) with a standard deviation of 13 days ($\sigma_a$). While the observed spread (i.e., variance) of arrival dates within subbasins was determined to have a mean ($\mu_s$) of 22 days and a standard deviation of 7 days ($\sigma_s$).

For the specific simulated scenarios, we were interested in STADEM model estimates of origin specific escapment from the combinations of two seperate trapping rates, two fallback, re-ascension and night-passage combinations and three window count error rates; resulting in twelve different scenarios. First, trapping rates were set static at 0.15 across all weeks for six scenarios to mimic an optimium trap operation for an expected return of 25,000 wild fish (i.e., trap $\approx$ 4,000 wild fish). For the remaining six scenarios, trapping rates for weeks 30, 31 and 32 (i.e., July $22^{nd}$ to August $11^{th}$) were changed to 0.00 to test STADEM sensitivities to potential trap shut downs similar to those observed in 2013, 2014 and 2015 [@Ogden2014; @Ogden2016; @Ogden2016b]. To simulate and control the number of re-ascending and night-time passing fish to model response, we altered fallback and night-time passage rates while holding the re-ascension rate constant at $\rho = 1.0$. Altering fallback rates and holding re-ascension constant allowed for a more simple control of the number of fish re-ascending; because the number of re-ascending fish is a function of the number of fallbacks and the re-ascension rate. Six scenarios had equal rates of fallback and night-time passage set at $\psi = \nu = 0.06$ [@Boggs2004] which means other estimator assumptions [@Schrader2013]. The other six scenarios set fallback at $\psi = 0.10$ and night-time passage at $\nu = 0.05$ to create a 5.0% positive bias of unique fish at the window. A potential 5.0% weekly bias was determined from PIT-tag data and within the range of observed weekly difference for return years 2010 - 2015 (Figure \@ref(fig:night-reasc-diff-fig)). Finally, we desired to test the sensitivites of STADEM to potential rates of window count error; 0%, 5% and 10% [@Hatch1994]. Error at the window was simulated as unbiased (i.e., expect varying high and low counts to cancel each other) and is introduced to the sum of daily window counts using a normal $\mathcal{PDF}$.

### 2020-08-07 analysis/paper/paper.Rmd
updated manuscript Rmd to match latest Word version from Mike.

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Mike W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
  - Katie Lawry:
      email: katie.lawry@idfg.idaho.gov
      institute: [IDFG]
      correspondence: false
institute:
  - BM: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
  - IDFG: Idaho Department of Fish and Game, 1414 East Locust Lane, Nampa, Idaho, 83686, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  This is the abstract
keywords: |
  escapement; Chinook; steelhead; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 600
)
```

```{r load_libraries}
# setwd('analysis/paper')

library(tidyverse)
library(lubridate)
library(msm)
library(HDInterval)
library(ggpubr)
library(knitr)
library(kableExtra)

library(STADEM)

# set default theme for ggplot2
theme_set(theme_bw())

# default options for tables
options(knitr.kable.NA = '-')

# when knitting to Word, use this
# what kind of document is being created?
doc.type <- knitr::opts_knit$get('rmarkdown.pandoc.to')

if(doc.type == 'docx') {
  options(knitr.table.format = "pandoc")
}

```

```{r package-bibtex, eval = F}
knitr::write_bib(c("lubridate",
                   "msm",
                   "HDInterval",
                   "ggpubr",
                   "knitr", 
                   "rjags",
                   "rmarkdown"),
                 file = 'packages.bib')
```

```{r load_sim_results, cache = F}

result_nms <- list.files('../data/derived_data/SimulationFits')
result_nms <- result_nms[!grepl('other', result_nms)]
scenario_nms <- gsub('^Sim_', '', result_nms)
scenario_nms <- gsub('\\.rda$', '', scenario_nms)
# length(scenario_nms)

# compile all results into single dataframe
res_df <- scenario_nms %>%
  as.list() %>%
  rlang::set_names() %>%
  map_df(.id = 'Scenario',
         .f = function(nm) {
    load(paste0('../data/derived_data/SimulationFits/Sim_', nm, '.rda'))
    res %>%
      map_df(.id = 'sim',
             .f = identity)
  }) %>%
  mutate(Variable = factor(Variable, levels = c('Unique.Wild.Fish',
                                                'Unique.Hatch.Fish',
                                                'Unique.HNC.Fish',
                                                'All.Fish',
                                                'Daytime.Fish',
                                                'Night.Fish',
                                                'Reascent.Fish'))) %>%
  mutate(Scenario = recode(Scenario,
                           'Baseline_WinErr_high' = 'Baseline_Err_H',
                           'Baseline_WinErr_low' = 'Baseline_Err_L',
                           'NightReasc' = 'N-R',
                           'NightReasc_WinErr_high' = 'N-R_Err_H',
                           'NightReasc_WinErr_low' = 'N-R_Err_L',
                           'TrapShutdown' = 'TrapDown',
                           'TrapShutdown_WinErr_high' = 'TrapDown_Err_H',
                           'TrapShutdown_WinErr_low' = 'TrapDown_Err_L',
                           'NightReasc_TrapShutdown' = 'N-R_TrapDown',
                           'NightReasc_TrapShutdown_WinErr_high' = 'N-R_TrapDown_Err_H',
                           'NightReasc_TrapShutdown_WinErr_low' = 'N-R_TrapDown_Err_L'),
         Scenario = factor(Scenario, 
                           levels = paste0(rep(c('Baseline', 'TrapDown', 'N-R', 'N-R_TrapDown'), each = 3), 
                                           c('', '_Err_L', '_Err_H'))))

#------------------------------------------------------
# Re-format res_df into long format of point ests.,
# CI's and CIin for both Models : STADEM, SCOBI
#------------------------------------------------------
res_long_df <- res_df %>%
  pivot_longer(cols = -(Scenario:Truth),
               names_to = "key",
               values_to = "point") %>%
  separate(key, into = c("Model","est"), sep = "_") %>%
  pivot_wider(names_from = "est",
              values_from = "point") %>%
  mutate(Model = recode(Model, 
                        "ISEMP" = "STADEM")) %>%
  mutate(se = (uppCI - lowCI) / (2 * qnorm(0.975)),
         cv = se / est,
         bias = est - Truth,
         rel_bias = bias / Truth,
         MSE = se^2 + bias^2,
         inCI = ifelse(inCI == 1, T, ifelse(inCI == 0, F, NA)),
         Window = ifelse(grepl("H",Scenario),"Window_Err_High",
                         ifelse(grepl("L",Scenario),"Window_Err_Low","No Window Error")),
         Night = ifelse(grepl("N-R",Scenario),"Yes","No"))
```

```{r load-lgr-results}
lgr_res = list.files('../data/derived_data/STADEM_results') %>%
  as.list() %>%
  map(.f = function(x) {
    load(paste('../data/derived_data/STADEM_results', x, sep = '/'))

    tot_est = stadem_mod$summary %>%
      as_tibble(rownames = 'variable') %>%
      filter(grepl("X.tot.new", variable)) %>%
      mutate(cv = sd / mean) %>%
      mutate(Year = str_extract(x, "[:digit:]+") %>%
               as.numeric()) %>%
      select(Year, variable, mean, cv) %>%
      bind_rows(stadem_list$weeklyData %>%
                  group_by(Species) %>%
                  summarise_at(vars(mean = win_cnt),
                               list(sum),
                               na.rm = T) %>%
                  mutate(variable = "win_cnt")) %>%
      tidyr::fill(Year, Species,
                  .direction = "downup") %>%
      select(Species, Year, everything())

    rate_df = stadem_list$weeklyData %>%
      mutate(Year = str_extract(x, "[:digit:]+") %>%
               as.numeric()) %>%
      select(Species, Year, Start_Date:reascent_tags) %>%
      mutate(night_rate = 1 - day_tags / tot_tags,
             reasc_rate = reascent_tags / tot_tags)

    return(list(tot_est = tot_est,
                rate_df = rate_df))

  })

lgr_est = map_df(lgr_res,
                 .f = "tot_est")

rate_comp = map_df(lgr_res,
                   .f = "rate_df")

```

**Possible journals:**

First Choice:

* **[North American Journal of Fisheries](http://www.tandfonline.com/action/journalInformation?show=aimsScope&journalCode=ujfm20)**

Second Choice:

* [Transactions of the American Fisheries Society](https://afspubs.onlinelibrary.wiley.com/journal/15488659)

Others:

* [Canadian Journal of Fisheries and Aquatic Sciences](http://www.nrcresearchpress.com/page/cjfas/editors)

* [Fisheries Research](https://www.journals.elsevier.com/fisheries-research)

* [Fishery Bulletin](http://fishbull.noaa.gov/)

* [Fish and Fisheries](http://onlinelibrary.wiley.com/journal/10.1111/(ISSN)1467-2979/homepage/ProductInformation.html)

# Introduction

Salmon escapement is the number of adults that escape harvest and return to their freshwater habitat to potentially spawn (e.g. @Bue1998). For anadromous fishes, escapement is often estimated at a given point in a river system prior to fish reaching their spawning area. Accurate escapement estimates facilitate effective fisheries management, and particularly, estimates of escapement by groups (e.g., stock, population, origin [hatchery, wild]) provide valuable information that fisheries managers can use to achieve sustainable harvest of larger groups, while protecting vulnerable populations. Accurate escapement estimates are increasingly important for depleted populations as they facilitate assessments of population viability and extinction risk. In this study, we describe a novel method to estimate escapement, with uncertainty, using multiple sources of data and demonstrate that the method is essentially unbiased. Moreover, the method allows escapement to be parsed by groups (e.g., wild, hatchery) which is important towards effective conservation and management decision making.

Populations of Chinook salmon Oncorhynchus tshawytscha and steelhead trout O. mykiss in the Snake River basin of Pacific Northwest, USA, have been depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric dams on the Snake and Columbia rivers) and tributary habitats. As a result, Snake River spring-summer run Chinook salmon (hereafter Chinook salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA); Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later. Snake River Chinook salmon and steelhead have substantial recreational, commercial, and subsistence value both within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries. The aggregate escapement of Snake River Chinook salmon and steelhead populations, with the exception of Tucannon River, is measured at Lower Granite Dam located in southeast Washington, the final dam on the Snake River that returning adults must ascend prior to heading to tributary habitats to spawn. Many fisheries management and conservation actions are made based on estimates of escapement at Lower Granite Dam parsed by species and origin (hatchery, wild). Harvest openings and closures, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia River fisheries, can be predicated on escapements at Lower Granite Dam. Moreover, escapement of wild populations is used to inform viable salmonid population (VSP) analyses used for ESA status reviews.

Chinook salmon and steelhead returning to a majority of populations in the Snake River must ascend a fish ladder on Lower Granite Dam before migrating to their natal tributary habitats to spawn. Previously, management agencies used counts of fish, by species, passing an observation window located on the side of the fish ladder as a census of fish migrating past the dam (i.e., Lower Granite escapement). For both species, total escapement was then parsed to groups (wild, hatchery clipped, hatchery unclipped) using mark and genetic data from a sample of fish captured at an adult trap located on the fish ladder upstream of the observation window [@Camacho2017]. Window counts as a census proved beneficial as being an easy, straight-forward method that could be ascertained in near real-time. Moreover, window count estimates have been used in downriver fisheries management arenas for several years and consistency in methods is often desirable for management decision making.

However, using window counts as a census of Chinook salmon and steelhead passing Lower Granite Dam can be problematic as it avoids multiple sources of uncertainty and ignores known biological processes. First, live window counts only occur from April through October each year. The live window counts only occur for 16 hours each day; fish counters working at the observation window look directly into the fish ladders to count fish, by species, passing by for 50 minutes of each hour. Counts are then multiplied to provide an estimate for the hour. From November through March, the remainder of the year, video tape fish counting is used and only occurs for 10 hours each day; fish counters then reach the tapes and submit daily fish counts [@Hatch1994]. Typically, the error rates of live and video window counts are unknown. Additionally, two biological processes are unaccounted for: 1) fish that cross the dam while the window is closed for counting (i.e., nighttime passage), and 2) fish that cross the dam multiple times (i.e., re-ascension) and therefore potentially double-counted. Uncounted nighttime passage potentially results in an underestimate of escapement. Finally, Chinook salmon and steelhead that migrate through the ladder and past the dam may fallback over the dam (e.g., over spillway, through navigation locks), and later, may or may not re-ascend the fish ladder, again [@Boggs2004]. Both fallback with no re-ascension and fallback with re-ascension potentially result in an overestimate of escapement; re-ascension can result in double-counting of fish [@Dauble2000]. Previously, it was assumed that nighttime passage rates and fallback/re-ascension rates cancelled each other out resulting in window counts providing an unbiased estimate of escapement; we further evaluate that assumption here.

Here, we present a novel method for estimating escapement, by species, past Lower Granite Dam that incorporates window counts, data from the adult fish trap, and observations of PIT-tagged fish in the adult ladder and that explicitly models nighttime passage, re-ascension and error from both window and trap estimates. To meet desired management and conservation objectives, total escapement estimates include estimates of uncertainty and are parsed into weekly strata. Further, weekly estimates are parsed into three origin groups: wild, hatchery, and hatchery no-clip. Estimates of escapement account for fish that migrate through the ladder at night outside of observation hours and account for fish that may ascend the ladder multiple times. Our methods is implemented in the **ST**ate-space **A**dult **D**am **E**scapement **M**odel (STADEM). The STADEM model is available as a package for the statistical software R [@R-Core-Team2020], and can be downloaded from a [GitHub site](https://github.com/KevinSee/STADEM). To validate the model, we simulate 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates. The STADEM model combines multiple imperfect sources of data to reduce bias in escapement estimates and provide more reasonable estimates of uncertainty. This STADEM framework could be applied elsewhere to any fish passage barrier with a counting mechanism like a window, an adult trap that samples a portion of the run and contains a PIT tag antenna (or similar).

# Materials and Methods

## Data Requirements

We used STADEM and three sources of data to estimate Chinook salmon and steelhead escapement at Lower Granite Dam. Data sources included 1) counts of fish migrating pas an observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of previously PIT tagged fish detected in the adult fish ladder. Following, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; however, similar data could also likely be obtained from other fish passage facilities.

### Window Counts

Daily counts of adult Chinook salmon and steelhead passing an observation window located on the Lower Granite fish ladder are the first source of data. Daily counts are made and provided by the US Army Corps of Engineers (COE) and, when summed, provide an estimate of the number of fish ascending and passing (i.e., escapement) Lower Granite Dam each season. Window counts are made for each species using both video monitoring and direct visual monitoring during daytime hours [@Hatch1994]. Video monitoring occurs during the beginning and tail ends of the adult runs (March 1 – March 31 and November 1 – December) and for 10 hours per day (0600 – 1600 hours). Direct visual monitoring occurs during peak run times (April 1 – October 31) and for 16 hours per day (0400 – 2000 hours) [@USACE2015]. During direct visual monitoring, observers record each adult (≥ 30cm), by species, passing the window for 50 minutes of each hour of operation. Salmonids under 30cm in length are not identified to species. The sum of the daily 50-minute counts are then multiplied by 1.2 to account for the 10 minutes where fish are not counted. These daytime window counts are not expanded for fish that may ascend the ladder outside of operational hours (i.e., nighttime) [@USACE2015]. Window counts are accessed through the Columbia Basin Research Data Access in Real Time (DART) [website](www.cbr.washington.edu/dart/query/adult_daily), using their window count query. Counts are provided for each day the fish ladder is open to passage. Although window counts are assumed to be a census of every fish passing Lower Granite Dam, nighttime passage and re-ascending fish are not accounted for. Further, there is no estimation of daily or seasonal observation error.

### Adult Fish Trap Data

The second source of data comes from a trap that captures a sample of the adults as they migrate past Lower Granite Dam [@Ogden2016b]. The trap, also located within the adult fish ladder and upstream of the observation window, provides biological data (e.g., origin [wild, hatchery], genetic stock, length, age, sex) for captured adults that allows decomposition of the escapement into specific groups (e.g., @Camacho2019, @Steinhorst2017). The trap randomly samples the daily run by opening four times per hour for a length of time determined by a set daily trapping rate. The trap is operational 24 hours per day. The trap rate is determined by a committee of collaborating management agencies with a goal of capturing a target number of wild fish but also balancing fish handling concerns. Trap sample rates are typically 10-25%, but may change throughout the season due to high water temperatures, decreased flows, trap malfunctions and/or closures, fish handling logistics, etc. All captured fish are anesthetized, identified to species, examined for marks/tags, and measured for fork length (FL). Additionally, each fish has scale and genetic tissue samples taken; scale samples are used to estimate age [@Wright2015], tissue samples are used to determine sex and for genetic stock identification analysis (e.g., @Hargrove2019). Prior to release, all non-PIT tagged fish with an intact adipose fin (i.e., putatively wild) receive a PIT tag. Final determination of wild, hatchery, and hatchery no-clip origins are assigned using a post-hoc analysis of marks, tags, and genetic information. Data from the adult trap are collected and managed by multiple agencies and are made available by the Idaho Department of Fish and Game (IDFG) in a comma separated value (.csv) file [@Camacho2019].

### PIT-tag Data

The last source of data is observations of PIT tagged adult Chinook salmon and steelhead at detection sites located in the Lower Granite Dam fish ladder. These observations provide estimates of 1) a trapping rate, 2) the proportion of fish passing during nighttime hours (i.e., outside of window observation hours), and 3) the proportion of fish that ascend the fish ladder multiple times (i.e., re-ascension rate). Detections used in the model include all fish that were previously PIT tagged as juveniles or adults prior to reaching Lower Granite Dam (i.e., not those tagged at the dam as adults) and detected at adult detection sites in the dam passage system. Data are provided through DART and the [adult ladder PIT tag query](http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail).

A trap rate estimate is derived using mark-recapture methods and PIT tag observations of both Chinook salmon and steelhead at Lower Granite Dam adult detection sites. The "mark" group includes all tags detected in the adult trap and the "capture" group includes tags observed to cross the weir at the upstream end of the fish ladder as adults leave the passage system. Using a mark-recapture model with differing capture probabilities, we can estimate the trap rate on a weekly basis. Those estimates, with associated uncertainties, are then incorporated into the model as informed priors, while the model estimates the true trap rate based on all the data, including trap and window counts. The true trapping rate is estimated as the recorded time of the trap being open does not always reflect the true proportion of fish that are captured in the trap, due to trap malfunctions, sort-by-code fish opening the trap more frequently than expected, or process error (Rick Orme, *per. comm.*). Therefore, we use the mark-recapture estimate of trap rate.

The nighttime passage rate is estimated simply as the proportion of tags that migrate through the fish ladder during "nighttime" hours outside of the window observation hours divided by the total number of tags passing the fish ladder and is calculated on a weekly basis. The re-ascension rate is the number of tags observed passing the upstream most detection sites in the adult fish ladder (i.e., passing the dam) and later detected re-entering the downstream end of the fish ladder at a later time divided by the total number of tags leaving the fish ladder. Previously, we examined for differences in nigh passage and re-ascension rates estimated using wild fish, versus combining hatchery and wild fish together, and found no difference. Therefore, we combine wild and hatchery PIT tagged fish observations to estimate common nighttime passage and re-ascension rates to increase sample sizes.

## Model Framework

STADEM estimates the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for nighttime passage and fallback-and-reascension rates. Using a state-space approach, STADEM assumes that the window counts and the estimates from the trap (fish in the trap divided by trap rate that week) are generated by processes with observation error. In the case of the trap, there is sampling variation and uncertainty in what the true trap rate is. STADEM further accounts for the proportion of fish that pass while the counting window is closed (i.e., night), as well as the potential double-counting (or more) of fish that have fallen back below the dam and later re-ascended the fish ladder. Next, origin data (wild, hatchery, hatchery no-clip) are used to partition the total escapement estimate by origin (Figure \@ref(fig:stadem-examp-fig). Additional model details can be found in [Appendix A](#append1). STADEM is implemented in R software [@R-Core-Team2020] and is available from the primary author at a [GitHub site](https://github.com/KevinSee/STADEM).

## Simulations

We tested the STADEM model on a variety of simulated data sets. These simulated data sets contained a fixed number of unique adult fish of known origin crossing a dam, from a total of 25 fictional populations with differential run-timing (i.e., date of passage at Lower Granite Dam). Each simulated fish is given a date of ascending the ladder, based on its population and the range of run timing for that population. Each fish is also simulated to cross the dam either while the window is open for counting, or not, and is given the change to be "caught" in the simulated fish trap given the week when it ascends the dam, and the known trap rate that week. Fallback and re-ascension behavior was also simulated, with each fish having the possibility of falling back and re-ascending the ladder up to three times.

Our objective was to examine STADEM model estimates of origin-specific (wild, hatchery, hatchery no-clip) escapement from the combinations of two separate trapping rates, two fallback, re-ascension and nighttime passage combinations, and three window count error rates; resulting in twelve different scenarios (Table \@ref(tab:lgr-sim-tab)). The simulation parameters such as proportion of origin, run-timing, nighttime passage rates, fallback and re-ascension rates and trap rates were based on observed values at Lower Granite Dam between 2010-2015. Further details about simulation procedures can be found in [Appendix B](#append2).

## Lower Granite Application

Finally, we applied STADEM to data from Lower Granite Dam for Chinook salmon and steelhead returning to the Snake River to spawn during 2010 to 2019. Window counts for both species were accessed from DART via STADEM. For spring-summer run Chinook salmon, a spawn year refers to adults that migrate pas the dam prior to August 17 each year and spawn that late summer and fall. For steelhead, the spawn year is defined as steelhead that migrate past Lower Granite Dam starting July 1 the previous year and prior to June 30 of the given year (e.g., spawn year 2017 steelhead migrate past Lower Granite between July 1, 2016 and June 30, 2017). Data from the adult trap was made available by IDFG and adult detection data at Lower Granite Dam from the fish passage ladder is available in the regional database PTAGIS ([PIT Tag Information System](https://www.ptagis.org/)).

# Results

## Simulations

Simulation results of observed bias, sampling variation, precision, root mean squared deviation (RMSD), and coverage probabilities were qualitatively similar for hatchery (N = 70,000) and hatchery no-clip (N = 5,000) origins as observed in the wild origin (N = 25,000) comparisons. As such, only diagnostic measures of Lower Granite Dam model fits to a medium sized escapement level (e.g., wild origin escapement) are presented.

STADEM results were very similar across all scenarios (Figure \@ref(fig:rel-bias-lgd)). Estimates were unbiased, with an average relative bias of 0.2 – 0.3%. The CV of the estimates average 2 – 3%, with coverage probabilities that always exceeded 95%. We calculated RMSD as the square root of the sum of the variance of the estimate and the squared expected bias, which accounts for the size of the uncertainty in the estimator as well as its bias. The RMSD was near 500 for each scenario, representing an estimate within 2% of the true value (Table \@ref(tab:summ-tab)).

## Lower Granite Application

We applied STADEM to data from Lower Granite Dam for Chinook salmon and steelhead for spawn years 2010 – 2019. Estimates of total escapement, as well as estimates of wild, hatchery, and hatchery no-clip estimates are presented in Table \@ref(tab:lgr-tab). Coefficients of variation range from 3.5 - 7.5% for wild fish, 3.0 - 6.2% for hatchery fish, 4.0 - 8.4% for hatchery no-clip fish and 2.9 - 5.6% for total unique fish past Lower Granite Dam.

Weekly estimates of total escapement over Lower Granite Dam track the window counts and trap estimates (Figure \@ref(fig:jags-fit-fig)). STADEM point estimates are often between estimates based on window counts, and those based on the number of fish caught in the adult trap. However, when very few fish are caught in the trap, or there is more uncertainty about the trap rate that week, STADEM estimates track the window counts more closely, as seen in the second week of July 2014 in Figure \@ref(fig:jags-fit-fig). That year shows the utility of STADEM in dealing with missing data, as the trap was shut down for several weeks in July and August.

Examining the estimates of nighttime passage and re-ascension rates based on the observed PIT tags crossing over Lower Granite Dam, the two rates often do not match up (Figure \@ref(fig:night-reasc-diff-fig)). In particular, there are several weeks when the window counts are quite high, and the rates differ by as much as 10%. Clearly, those two biological processes do not cancel each other out, and thus, employing a model that accounts for both will result in more accurate estimates of escapement.

# Discussion

We have presented a novel method for estimating adult salmonid escapement past a large hydroelectric facility (e.g., Lower Granite Dam) that incorporates data from window counts, a fish trap, and observations of PIT-tagged fish in the adult passage ladder. Our model explicitly models nighttime passage, re-ascension, and potential error in both window and trap estimates. In doing so, we have demonstrated that at Lower Granite Dam, nighttime passage and re-ascension do not always offset each other, and the assumption that they do will lead to biased estimates in some years. This framework and the STADEM model could be applied elsewhere to any fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run, and contains observation or detection infrastructure (e.g., a PIT tag detection array or similar) and could be applied to similar migratory fishes (e.g., Pacific lamprey Lampetra tridentata). Our state-space model combines multiple imperfect sources of data to reduce bias in adult escapement estimates and provide more reasonable estimates of uncertainty. Accurate population or stock abundance estimates and uncertainty accounting for observation and process error can be particularly important when estimates are used or leveraged for management and conservation decisions such as population viability analyses.

Combining data from the adult fish trap with live and video window counts provides several benefits. First, it allows us to model observer error in the window counts, which is typically unknown. If estimates rely on window counts alone, this is impossible. We believe it necessary to model and estimate multiple sources of uncertainty in escapement estimates. Second, by incorporating both sources of information in a state-space framework, STADEM incorporates missing data from either source seamlessly. At Lower Granite Dam, the adult trap has been closed for brief or extended periods of time (i.e., days, weeks) intermittently over the past several years, often during peak run times. Trap closures are typically associated due to elevated temperatures resulting in potential fish handling stress and/or trap malfunctions. Climate change scenarios for the Pacific Northwest could cause trap closures from high water temperatures more commonplace in the future, amplifying the need for a framework that accounts for those periods of missing data. Additionally, having a framework in place that accounts for missing periods of data, at either the observation window or adult trap, allows for increased logistic flexibility if, for example, maintenance or construction needed to occur at the observational window or similar.

<!-- I can't find the Steele et al. 2013 reference -->

Although not currently set up for this, STADEM could be modified to be run on a weekly basis or in near real time to provide better in-season estimates for fisheries managers. Currently, the only roadblock to this, at Lower Granite Dam, is that the genetic origin data (***Steele et al. 2013***) that is used to identify phenotypically wild fish that are truly hatchery fish (i.e., hatchery no-clip) is completed post hoc after the trapping season. This information typically results in a reduction in wild escapement estimates and an associated increase in hatchery no-clip escapement. However, if in-season management decisions do not rely on this correction, origin calls at the trap could be used instead, certainly as a first approximation. Estimates, parsed by origin, could then be finalized at season’s end. Most other data included in this model (e.g., window counts and PIT observations) are otherwise provided in almost real time by DART. As long as the Lower Granite Dam adult fish trap database was updated and made available frequently, there are minimal obstacles for adapting the STADEM framework to provide in-season estimates of escapement. Results could be provided in a web-accessible application to managers or end users.

The STADEM model is available as a package for the statistical software R [@R-Core-Team2020] and can be downloaded from a [GitHub site](https://github.com/KevinSee/STADEM). STADEM also uses the JAGS software [@R-rjags] for Bayesian inference from the model. Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate escapement of spring/summer Chinook salmon and steelhead past Lower Granite Dam. Estimates of escapement, by species and origin, are then available to parse into sex- or age-structured escapement estimates (e.g., @Camacho2019, @Schrader2013) that are important for fisheries management and productivity monitoring of wild populations. As an example, STADEM is being applied at Lower Granite to estimate the total unique wild fish migrating past; that information is then combined with PIT tag observations at instream PIT tag detection systems throughout the Snake River basin that are used to estimate movement or transition probabilities to populations and locations throughout the basin using the Dam Adult Branch Occupancy Model ([DABOM](https://github.com/KevinSee/DABOM)). Combined, escapement estimates from STADEM and movement probability estimates from DABOM, provide abundance estimates to given tributaries or populations that, joined with sex and age data collected at the adult fish trap, provides necessary information to evaluate productivity and population viability for select Snake River Chinook salmon and steelhead groups.

Our STADEM framework can be applied to any fish passage barrier with a counting mechanism like a window, a trap that samples a subset of the run, and contains a PIT tag antenna. Although STADEM was developed with salmonids in mind, it could be applied to any migrator fish species. Certainly, there is justification for applying a similar framework at Bonneville Dam, as the lowest dam on the Columbia River, to estimate total returns to the Columbia River basin. Lower Granite Dam is important because it is the farthest upstream dam on the Snake River with fish passage, and other methods of estimating escapted to smaller spatial domains (e.g., MGP/DPS, populations) upstream rely on good estimates of total escapement there. Priest Rapids plays a similar role for upstream migrating adult salmonids in the Upper Columbia River, and could also benefit from the STADEM approach.

# Acknowledgements

Funding for this study and development of the STADEM model was partially provided by the Bonneville Power Administration under project 2003-017-00. Special thanks to Darren Ogden and staff at the Lower Granite Dam adult trap for their hard work and diligent data collection. Thank you to personnel at the Idaho Department of Fish and Game, particularly Paul Bunn and Bill Schrader, for providing access to data from the adult fish trap. Special thanks to staff at Columbia Basin Research and the Columbia River Data Access in Real Time (DART) application, and Susannah Iltis in particular. And finally, thank you to all the folks who have contributed to the development of the STADEM framework through productive critique and conversations.

<!-- The following line inserts a page break  -->
\newpage

# References 
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

\newpage

# Tables

```{r lgr-sim-tab}
crossing(`Trap Rate` = c('0.15', 
                         "0.15 and 0.00 3 weeks"),
         `Fallback Rate` = c(0.06,
                             0.10),
         `Re-ascension Rate` = 1,
         `Window Count Error` = factor(c("No Error",
                                         "10% Error",
                                         "5% Error"),
                                       levels = c("No Error",
                                                  "10% Error",
                                                  "5% Error"))) %>%
  mutate(`Nighttime Passage Rate` = if_else(`Fallback Rate` == 0.06,
                                            0.06,
                                            0.05)) %>%
  mutate(Scenario = paste(rep(c('Baseline',
                                "N-R",
                                "N-R Trap Down",
                                "Trap Down"),
                              each = 3),
                          c("",
                            "Err H", 
                            "Err L"))) %>%
  select(Scenario, 
         `Trap Rate`:`Re-ascension Rate`,
         `Nighttime Passage Rate`,
         `Window Count Error`) %>%
  kable(digits = 2,
        booktabs = T,
        caption = "Summary of simulation scenarios including varying adult trapping, fallback rates, re-ascension, nighttime passage, and window count error rates use to evaluate the performance of STADEM.") %>%
  kable_styling()

```

```{r summ-tab}
sum_tab <- res_long_df %>%
  filter(Variable == 'Unique.Wild.Fish',
         Model == 'STADEM') %>%
  group_by(Scenario) %>%
  summarise(inCI = sum(inCI),
            nsims = n(),
            Sample_Mean = mean(est),
            `Relative Bias` = mean(rel_bias),
            SE = sqrt(var(est)),
            `Mean CV` = mean(cv),
            RMSD = sqrt(var(est) + mean(bias)^2),
            Coverage = inCI/nsims) %>%
  mutate(Scenario = str_replace_all(Scenario,
                                    "_", " "),
         Scenario = str_replace(Scenario,
                                "TrapDown",
                                "Trap Down")) %>%
  select(Scenario,
         `Relative Bias`,
         `Mean CV`,
         RMSD,
         Coverage)

# mu_rmse <- sum_tab %>%
#   group_by(Model) %>%
#   summarise(mu = mean(RMSE))

sum_tab %>%
  kable(digits = 3,
        booktabs = T,
        caption = "Summary statistics, including relative bias, mean coefficient of variation (CV), root mean squared deviation (RMSD) and coverage for results from each of the twelve simulation scenarios.") %>%
  kable_styling()
```

```{r lgr-tab}
lgr_est %>%
  mutate(prnt_val = if_else(is.na(cv),
                            as.character(prettyNum(round(mean), big.mark = ',')),
                            paste0(prettyNum(round(mean), big.mark = ","), " (", round(cv, 3), ")"))) %>%
  pivot_wider(id_cols = c("Species",
                          "Year"),
              names_from = "variable",
              values_from = "prnt_val") %>%
  select(Species, 
         Year,
         `Window Counts` = win_cnt,
         Total = X.tot.new.all,
         Wild = X.tot.new.wild,
         Hatchery = X.tot.new.hatch,
         `Hatchery No-Clip` = X.tot.new.hnc) %>%
  kable(booktabs = T,
        caption = "Window counts, and estimates of total, wild, hatchery and hatchery no-clip escapement, with coefficients of variation in parenthesis, for Chinook salmon and steelhead from spawn years 2010 to 2019.") %>%
  kable_styling()
```

\newpage

# Figures

```{r stadem-examp-fig, fig.cap = "Schematic of how the STADEM model works. Panel A shows the posterior of the estimate of fish crossing the dam while the window is open (dashed line shows observed window counts). That estimate is divided by the nighttime passage rate (B). The total fish is then discounted by the reascension rate to estimate unique fish (C). Those unique fish are then multiplied by the proportion of wild fish (D), to estimate unique wild fish."}

arrow_length = unit(0.15, 'inches')
arrow_height = 1.05
my_xlim = c(-3, 7.5)
my_ylim = c(0.06, 1.1)

p1 = ggplot(data.frame(x = c(-5, 8)),
            aes(x)) +
  stat_function(fun = dnorm,
                args = list(mean = 4, sd = 0.4),
                aes(color = 'Daytime Fish'),
                lwd = 2) +
  stat_function(fun = dnorm,
                args = list(mean = 6, sd = 0.55),
                aes(color = 'Total Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = 5, sd = 0.6),
                aes(color = 'Unique Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.75),
                aes(color = 'Unique Wild Fish'),
                geom = 'blank') +
  scale_color_brewer(palette = 'Set1') +
  geom_vline(xintercept = 4,
             linetype = 2) +
  coord_cartesian(xlim = my_xlim,
                  ylim = my_ylim) +
  theme_classic() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = 'bottom') +
  labs(x = 'Escapement',
       y = 'Density',
       color = 'Fish Type',
       title = 'Daytime Fish')

p2 = p1 +
  stat_function(fun = dnorm,
                args = list(mean = 6, sd = 0.55),
                aes(color = 'Total Fish'),
                lwd = 2) +
  labs(title = 'Including Night Passage')

p2a = p2 +
  geom_segment(x = 4,
               y = arrow_height,
               xend = 6,
               yend = arrow_height,
               arrow = arrow(length = arrow_length))

p3 = p2 +
  stat_function(fun = dnorm,
                args = list(mean = 5, sd = 0.6),
                aes(color = 'Unique Fish'),
                lwd = 2) +
  labs(title = 'Account for Reascension')

p3a = p3 +
  geom_segment(x = 6,
               y = arrow_height,
               xend = 5,
               yend = arrow_height,
               arrow = arrow(length = arrow_length))

p4 = p3 +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.75),
                aes(color = 'Unique Wild Fish'),
                lwd = 2) +
  geom_segment(x = 5,
               y = arrow_height,
               xend = -1,
               yend = arrow_height,
               arrow = arrow(length = arrow_length)) +
  labs(title = 'Proportion Wild')

examp_p = ggarrange(p1, p2a, p3a, p4,
                    labels = 'AUTO',
                    ncol = 2,
                    nrow = 2,
                    common.legend = T,
                    legend = 'bottom')

examp_p
```

```{r rel-bias-lgd, fig.cap = "'Boxplots show relative bias of STADEM estimates for wild escapement across various scenarios. The boxes contain 50% of the simulations, whiskers contain 95% of the simulations, and points are outliers."}
hdi_box = function(x) {
  r = c(hdi(x, 0.95), hdi(x, 0.5), median(x))
  # r = quantile(x, probs = c(0.025, 0.975, 0.25, 0.75, 0.5))
  names(r) = c('ymin', 'ymax', 'lower', 'upper', 'middle')
  r
}

hdi_out = function(x) {
  subset(x, x < hdi(x, 0.95)[1] | x > hdi(x, 0.95)[2])
  # subset(x, x < quantile(x, probs = c(0.025)) | x > quantile(x, probs = c(0.975)))
}

rel_bias_p = res_long_df %>%
  filter(grepl('Unique', Variable),
         Model == 'STADEM') %>%
  mutate(Variable = fct_recode(Variable,
                               Wild = 'Unique.Wild.Fish',
                               Hatchery = 'Unique.Hatch.Fish',
                               HNC = 'Unique.HNC.Fish')) %>%
  mutate(grp = paste(Variable, Model, sep = ' / ')) %>%
  # mutate(Scenario = factor(Scenario, 
  #                          levels = sort(levels(res_long_df$Scenario), decreasing = F))) %>%
  filter(Variable == "Wild") %>%
  ggplot(aes(x = Scenario,
             y = rel_bias,
             fill = grp)) +
  stat_summary(fun.data = hdi_box, 
               geom = 'boxplot') +
  stat_summary(fun = hdi_out,
               geom = 'point',
               color = 'gray') +
  geom_hline(yintercept = 0,
             linetype = 2) +
  labs(y = 'Relative Bias',
       fill = 'Fish / Model') +
  facet_wrap(~ Variable) +
  theme(legend.position = 'none',
        axis.text.x = element_text(angle=45, hjust = 1))

rel_bias_p
```

```{r jags-fit-fig, fig.cap = "Time-series plot showing estimates of total escapement for Chinook in 2014, including raw window counts, window counts adjusted for night-time passage, trap estimates and STADEM estimates. Gray ribbon represents the 95% credible interval for STADEM estimates."}

load('../data/derived_data/STADEM_results/LGR_STADEM_Chinook_2014.rda')
plotTimeSeries(stadem_mod,
               weeklyData = stadem_list$weeklyData)
```

```{r night-reasc-diff-fig, fig.cap = "Night-time passage rate plotted against re-ascension rate, calculated from observed PIT tags for each week of spawn years 2010-2019. Colors correspond to different spawn year, while the size of each point is proportional to the window count that week. The dashed line is the 1-1 line."}

rate_comp %>%
  filter(win_cnt > 0) %>%
  filter(tot_tags > 0,
         !night_rate %in% c(1, 0),
         !reasc_rate %in% c(1, 0)) %>%
  # arrange(desc(night_rate))
  ggplot(aes(x = night_rate,
             y = reasc_rate,
             color = as.factor(Year))) +
  # scale_color_viridis_d() +
  scale_color_brewer(palette = "Paired") +
  geom_abline(linetype = 2) +
  geom_point(aes(size = win_cnt)) +
  facet_wrap(~ Species,
             scales = 'free') +
  theme(legend.position = 'bottom') +
  # scale_x_continuous(trans = "log",
  #                    breaks = scales::pretty_breaks()) +
  # scale_y_continuous(trans = "log",
  #                    breaks = scales::pretty_breaks()) +
  guides(size = guide_legend(nrow = 2)) +
  labs(x = "Nighttime Passage Rate",
       y = "Re-ascension Rate",
       color = "Spawn\nYear",
       size = "Window\nCount")
```

\newpage

```{r child = 'appendices/Appendix_STADEM_descrip.Rmd'}
```

\newpage

```{r child = 'appendices/Appendix_simulation_details.Rmd'}
```

### 2020-08-07 inst/scripts/STADEM_model_runs.R
added a script to recreate STADEM results

@@ -1,0 +1,96 @@
+#------------------------------------------------------------------------------
+# Script gathers data and runs the STADEM model.  A loop is built in to cycle
+# through Chinook and steelhead and all years begining in 2010.
+#
+# Author: Ryan Kinzer
+#------------------------------------------------------------------------------
+
+# Download STADEM Package
+if(!require(STADEM)){
+  devtools::install_github('KevinSee/STADEM', ref = 'master', force = TRUE,
+                           upgrade = "always")
+}
+
+
+
+library(tidyverse)
+library(stringr)
+library(lubridate)
+library(magrittr)
+library(forcats)
+library(STADEM)
+
+# set up folder structure
+stademFolder = 'analysis/data/derived_data/STADEM_results'
+if(!dir.exists(stademFolder)) {
+  dir.create(stademFolder)
+}
+
+modelFolder = 'analysis/data/derived_data/ModelFiles'
+if(!dir.exists(modelFolder)) {
+  dir.create(modelFolder)
+}
+
+## Requires a copy of the Lower Granite Dam Trap Database
+trap_dbase <- readLGRtrapDB('analysis/data/raw_data/tblLGDMasterCombineExportJodyW.csv')
+
+# set species and spawn year
+species = c('Chinook', 'Steelhead')  # either Chinook or Steelhead
+year = 2010:2019        # tagging operations started at Lower Granite with spawn year 2009.
+
+# Loop through species and years
+for(i in 1:length(species)){
+  for(j in 1:length(year)){
+
+    spp <- species[i]
+    yr <- year[j]
+
+    if(spp == 'Chinook'){
+      incl_jacks = TRUE
+    } else{
+      incl_jacks = FALSE
+    }
+
+    if(spp == 'Chinook'){
+      start_date = paste0(yr,'0301')
+      end_date = paste0(yr,'0817')
+    } else {
+      start_date = paste0(yr-1,'0701')
+      end_date = paste0(yr,'0630')
+    }
+
+
+    stadem_list = compileGRAdata(spp = spp,
+                                 start_date = start_date,
+                                 end_date = end_date,
+                                 strata_beg = 'Mon',
+                                 trap_dbase = trap_dbase,
+                                 incl_jacks = incl_jacks)
+
+    # creates JAGs data list.
+    jags_data_list = prepJAGS(stadem_list[['weeklyData']])
+
+    # JAGs needs to access a .txt file of the model code
[…]

### 2020-08-07 analysis/paper/paper.Rmd
Removed Katie Lawry as a co-author, and reknit pdf and html files

@@
       correspondence: false
-  - Katie Lawry:
-      email: katie.lawry@idfg.idaho.gov
-      institute: [IDFG]
-      correspondence: false
 institute:
@@
   - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
-  - IDFG: Idaho Department of Fish and Game, 1414 East Locust Lane, Nampa, Idaho, 83686, USA
 date: "`r format(Sys.time(), '%B %d, %Y')`"

### 2020-08-07 analysis/paper/paper.Rmd
changed the link for NAJFM, put in pdf figure width default (single column), and changed table headers to only have first word capitalized (per NAJFM instructions).

@@
       latex_engine: "xelatex"
+      fig_width: 3.5
       pandoc_args:
@@
 
-* **[North American Journal of Fisheries](http://www.tandfonline.com/action/journalInformation?show=aimsScope&journalCode=ujfm20)**
+* **[North American Journal of Fisheries](https://afspubs.onlinelibrary.wiley.com/journal/15488675)**
 
@@
 ```{r lgr-sim-tab}
-crossing(`Trap Rate` = c('0.15', 
+crossing(`Trap rate` = c('0.15', 
                          "0.15 and 0.00 3 weeks"),
-         `Fallback Rate` = c(0.06,
+         `Fallback rate` = c(0.06,
                              0.10),
-         `Re-ascension Rate` = 1,
-         `Window Count Error` = factor(c("No Error",
+         `Re-ascension rate` = 1,
+         `Window count error` = factor(c("No Error",
                                          "10% Error",
@@
                                                   "5% Error"))) %>%
-  mutate(`Nighttime Passage Rate` = if_else(`Fallback Rate` == 0.06,
+  mutate(`Nighttime passage rate` = if_else(`Fallback rate` == 0.06,
                                             0.06,
@@
                                 "N-R",
-                                "N-R Trap Down",
-                                "Trap Down"),
+                                "N-R trap down",
+                                "Trap down"),
                               each = 3),
@@
   select(Scenario, 
-         `Trap Rate`:`Re-ascension Rate`,
-         `Nighttime Passage Rate`,
-         `Window Count Error`) %>%
+         `Trap rate`:`Re-ascension rate`,
+         `Nighttime passage rate`,
+         `Window count error`) %>%
   kable(digits = 2,
@@
             Sample_Mean = mean(est),
-            `Relative Bias` = mean(rel_bias),
+            `Relative bias` = mean(rel_bias),
             SE = sqrt(var(est)),
@@
   select(Scenario,
-         `Relative Bias`,
+         `Relative bias`,
          `Mean CV`,
@@
          Year,
-         `Window Counts` = win_cnt,
+         `Window counts` = win_cnt,
          Total = X.tot.new.all,
@@
          Hatchery = X.tot.new.hatch,
-         `Hatchery No-Clip` = X.tot.new.hnc) %>%
+         `Hatchery no-clip` = X.tot.new.hnc) %>%
   kable(booktabs = T,

### 2020-08-07 analysis/paper/appendices/Appendix_STADEM_descrip.Rmd
changed "log" to "ln" per AFS journal instructions

@@
 \begin{aligned}
-  \log \left( X_{t} \right) &= \log \left( X_{t-1} \right) + e_t \\
+  \ln \left( X_{t} \right) &= \ln \left( X_{t-1} \right) + e_t \\
   e_t &\sim \mathcal{N}(0, \sigma^2_X)
@@
   \omega_t &= \frac{\exp(\phi_t)}{\sum{\exp(\phi_t})} \\
-  \phi_{w,t} &= \log \left( \frac{\omega_{w,t}}{\omega_{h,t}} \right) \\
-  \phi_{hnc,t} &= \log \left( \frac{\omega_{hnc,t}}{\omega_{h,t}} \right) \\
+  \phi_{w,t} &= \ln \left( \frac{\omega_{w,t}}{\omega_{h,t}} \right) \\
+  \phi_{hnc,t} &= \ln \left( \frac{\omega_{hnc,t}}{\omega_{h,t}} \right) \\
   \phi_{h,t} &= 0 \\
@@
 The model was fit using the JAGS program [@R-rjags], run with R software [@R-Core-Team2020].
-Uninformative priors were used for $\sigma_X, \sigma_\eta, \sigma_\theta, \sigma_\omega$ and $\log(X_1)$ (Uniform(0,10)), as well as logit($\eta_1$) and logit($\theta_1$) ($\mathcal{N}(0, 1000)$), and finally $\phi_{w,1}$ and $\phi_{hnc,1}$ ($\mathcal{N}(0, 100)$).
+Uninformative priors were used for $\sigma_X, \sigma_\eta, \sigma_\theta, \sigma_\omega$ and $\ln(X_1)$ (Uniform(0,10)), as well as logit($\eta_1$) and logit($\theta_1$) ($\mathcal{N}(0, 1000)$), and finally $\phi_{w,1}$ and $\phi_{hnc,1}$ ($\mathcal{N}(0, 100)$).

### 2020-08-07 analysis/paper/paper.Rmd
added some default figure width (7 in, 2 columns). Re-knit pdf, html and Word versions, with correct AFS journal formatting.

@@
 output:
-    bookdown::html_document2:
-      fig_caption: yes
-      number_sections: FALSE
-      toc: no
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
     bookdown::pdf_document2:
@@
       latex_engine: "xelatex"
-      fig_width: 3.5
+      fig_width: 7
       pandoc_args:
@@
       - --lua-filter=../templates/pagebreak.lua
+      includes:
+        in_header: ../templates/header_manuscript.tex
     bookdown::word_document2:
@@
       always_allow_html: true
+      fig_width: 7
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
+    bookdown::html_document2:
+      fig_caption: yes
+      number_sections: FALSE
+      toc: no
+      fig_width: 7
       pandoc_args:
@@
 
-# Materials and Methods
+# Methods
 
@@
 
-# Acknowledgements
+## Acknowledgements
 

### 2020-08-18 analysis/paper/paper.Rmd
revised document and added SimFnc.R

--- 2020-08-07 analysis/paper/paper.Rmd
@@
 output:
-    bookdown::html_document2:
+    bookdown::word_document2:
       fig_caption: yes
+      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
       number_sections: FALSE
       toc: no
+      always_allow_html: true
+      fig_width: 7
       pandoc_args:
@@
       - --lua-filter=../templates/pagebreak.lua
-    bookdown::word_document2:
+      includes:
+        in_header: ../templates/header_manuscript.tex
+    bookdown::html_document2:
       fig_caption: yes
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
       number_sections: FALSE
       toc: no
-      always_allow_html: true
+      fig_width: 7
       pandoc_args:
@@
 
-* **[North American Journal of Fisheries](https://afspubs.onlinelibrary.wiley.com/journal/15488675)**
+* **[North American Journal of Fisheries Management](https://afspubs.onlinelibrary.wiley.com/journal/15488675)**
 
@@
 
-Salmon escapement is the number of adults that escape harvest and return to their freshwater habitat to potentially spawn (e.g.
+Fish escapement is the number of adults that survive juvenile and adult rearing, escape harvest and return to their natal habitat to potentially spawn (e.g.
 @Bue1998).
-For anadromous fishes, escapement is often estimated at a given point in a river system prior to fish reaching their spawning area.
-Accurate escapement estimates facilitate effective fisheries management, and particularly, estimates of escapement by groups (e.g., stock, population, origin [hatchery, wild]) provide valuable information that fisheries managers can use to achieve sustainable harvest of larger groups, while protecting vulnerable populations.
-Accurate escapement estimates are increasingly important for depleted populations as they facilitate assessments of population viability and extinction risk.
-In this study, we describe a novel method to estimate escapement, with uncertainty, using multiple sources of data and demonstrate that the method is essentially unbiased.
-Moreover, the method allows escapement to be parsed by groups (e.g., wild, hatchery) which is important towards effective conservation and management decision making.
+For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their spawning area.
+Escapement estimates facilitate effective fisheries management, and particularly, estimates of escapement for each component group (e.g., stock, population, age, origin [wild, hatchery]; @hess2014monitoring; @steinhorst2017abundance) provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations.
+Accurate escapement estimates are increasingly important for depleted populations as they inform viable salmonid metrics [@mcelhany2000viable], and facilitate assessments of population viability and extinction risk [@nwfsc2015; @williams2016viability] and climate change vulerability [@crozier2019climate].
 
-Populations of Chinook salmon Oncorhynchus tshawytscha and steelhead trout O.
-mykiss in the Snake River basin of Pacific Northwest, USA, have been depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric dams on the Snake and Columbia rivers) and tributary habitats.
-As a result, Snake River spring-summer run Chinook salmon (hereafter Chinook salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA); Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later.
-Snake River Chinook salmon and steelhead have substantial recreational, commercial, and subsistence value both within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries.
-The aggregate escapement of Snake River Chinook salmon and steelhead populations, with the exception of Tucannon River, is measured at Lower Granite Dam located in southeast Washington, the final dam on the Snake River that returning adults must ascend prior to heading to tributary habitats to spawn.
-Many fisheries management and conservation actions are made based on estimates of escapement at Lower Granite Dam parsed by species and origin (hatchery, wild).
-Harvest openings and closures, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia River fisheries, can be predicated on escapements at Lower Granite Dam.
-Moreover, escapement of wild populations is used to inform viable salmonid population (VSP) analyses used for ESA status reviews.
+Populations of Chinook Salmon Oncorhynchus tshawytscha and steelhead trout O.
+mykiss in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric dams on the Snake and Columbia rivers) and tributary habitats [@nehlsen1991pacific; @mcclure2003large; @nwfsc2015].
+As a result, Snake River spring-summer run Chinook Salmon (hereafter Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937).
+Snake River Chinook Salmon and steelhead have substantial recreational, commercial, cultural and subsistence value both within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries.
+The aggregate escapement of Snake River Chinook Salmon and steelhead populations, with the exception of Tucannon River, is measured at Lower Granite Dam located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations.
+Many fisheries management and conservation actions are made based on estimates of escapement at Lower Granite Dam parsed by species and origin (**Snake Basin Harvest Plans?**).
+Additionally, harvest openings and closures, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia River fisheries, are predicated on escapements at Lower Granite Dam (**could cite US v OR Management Agreement and Snake Basin Harvest Plans?**).
 
-Chinook salmon and steelhead returning to a majority of populations in the Snake River must ascend a fish ladder on Lower Granite Dam before migrating to their natal tributary habitats to spawn.
-Previously, management agencies used counts of fish, by species, passing an observation window located on the side of the fish ladder as a census of fish migrating past the dam (i.e., Lower Granite escapement).
-For both species, total escapement was then parsed to groups (wild, hatchery clipped, hatchery unclipped) using mark and genetic data from a sample of fish captured at an adult trap located on the fish ladder upstream of the observation window [@Camacho2017].
-Window counts as a census proved beneficial as being an easy, straight-forward method that could be ascertained in near real-time.
-Moreover, window count estimates have been used in downriver fisheries management arenas for several years and consistency in methods is often desirable for management decision making.
+Chinook Salmon and steelhead returning to a majority of populations in the Snake River must ascend a fish ladder on Lower Granite Dam before migrating to their natal tributary spawning locations.
+Fish management agencies previously used counts of fish, by species, passing an observation window located on the side of the fish ladder as a census of fish escaping past the dam.
+For both species, total escapement was then parsed into groups (e.g., wild, hatchery) using observed marks and genetic data from a sample of fish captured at an adult trap located on the fish ladder upstream of the observation window [@Camacho2017, @Steinhorst2017].
+Window counts as a census proved beneficial as being an easy, straight-forward method that was ascertained in near real-time.
+Moreover, downriver fisheries management arenas have used window counts at lower-river dams as escapement estimates for the past several years, and consistency in methods is often desirable for management decision making (**citation?**).
 
-However, using window counts as a census of Chinook salmon and steelhead passing Lower Granite Dam can be problematic as it avoids multiple sources of uncertainty and ignores known biological processes.
-First, live window counts only occur from April through October each year.
-The live window counts only occur for 16 hours each day; fish counters working at the observation window look directly into the fish ladders to count fish, by species, passing by for 50 minutes of each hour.
-Counts are then multiplied to provide an estimate for the hour.
-From November through March, the remainder of the year, video tape fish counting is used and only occurs for 10 hours each day; fish counters then reach the tapes and submit daily fish counts [@Hatch1994].
-Typically, the error rates of live and video window counts are unknown.
-Additionally, two biological processes are unaccounted for: 1) fish that cross the dam while the window is closed for counting (i.e., nighttime passage), and 2) fish that cross the dam multiple times (i.e., re-ascension) and therefore potentially double-counted.
-Uncounted nighttime passage potentially results in an underestimate of escapement.
-Finally, Chinook salmon and steelhead that migrate through the ladder and past the dam may fallback over the dam (e.g., over spillway, through navigation locks), and later, may or may not re-ascend the fish ladder, again [@Boggs2004].
-Both fallback with no re-ascension and fallback with re-ascension potentially result in an overestimate of escapement; re-ascension can result in double-counting of fish [@Dauble2000].
-Previously, it was assumed that nighttime passage rates and fallback/re-ascension rates cancelled each other out resulting in window counts providing an unbiased estimate of escapement; we further evaluate that assumption here.
+Using window counts, however, as a census of Chinook Salmon and steelhead passing Lower Granite Dam can be problematic as it avoids multiple sources of uncertainty and ignores known biological processes.
+First, live (i.e., in-person) window counts only occur from April through October each year.
+Additionally, live window counts only occur for 16-hours each day, and fish counters working at the observation window look directly into the fish ladders to identify and count all passing fish, by species, for only 50 minutes of each hour.
+Counts are then expanded to provide an estimate for the hour [@USACE2015].
+From November through March, the remainder of the year, video tape fish counting is used and only occurs for 10 hours each day; fish counters then read the video tapes and submit daily fish counts [@Hatch1994].
+Typically, the observational error rates of live and video window counts are unknown, and sampling error rates are ignored.
+Additionally, two biological processes are unaccounted for: 1) fish that cross the dam during the 8-hours when the window is closed for counting (i.e., nighttime passage) which may result in an underestimate of escapement, and 2) fish that cross the dam multiple times (i.e., re-ascension), and therefore potentially double-counted.
+Finally, Chinook Salmon and steelhead that migrate through the ladder and past the dam may fallback over the dam (e.g., over spillway, through navigation locks), and later, may or may not re-ascend the fish ladder again [@Boggs2004].
+Both fallback with no re-ascension and fallback with re-ascension potentially result in an overestimate of escapement [@Dauble2000].
+Previously, it was assumed that nighttime passage rates and fallback/re-ascension rates cancelled each other out resulting in window counts providing an unbiased estimate of escapement (**citation?**)
 
-Here, we present a novel method for estimating escapement, by species, past Lower Granite Dam that incorporates window counts, data from the adult fish trap, and observations of PIT-tagged fish in the adult ladder and that explicitly models nighttime passage, re-ascension and error from both window and trap estimates.
-To meet desired management and conservation objectives, total escapement estimates include estimates of uncertainty and are parsed into weekly strata.
-Further, weekly estimates are parsed into three origin groups: wild, hatchery, and hatchery no-clip.
-Estimates of escapement account for fish that migrate through the ladder at night outside of observation hours and account for fish that may ascend the ladder multiple times.
-Our methods is implemented in the **ST**ate-space **A**dult **D**am **E**scapement **M**odel (STADEM).
-The STADEM model is available as a package for the statistical software R [@R-Core-Team2020], and can be downloaded from a [GitHub site](https://github.com/KevinSee/STADEM).
-To validate the model, we simulate 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates.
+In this study, we describe a novel method to estimate aggregate and group escapement which incorporates all sources of known uncertainty, we demonstrate the estimation method is essentially unbiased, thus improving effective conservation and management decision making, and show night-time passage and fallback/re-ascension rates are unequal.
+Our method for estimating escapement, by species, past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of PIT tagged fish in the adult ladder to explicitly model nighttime passage, re-ascension, and error from both window and trap estimates using a state-space approach [@royle2008hierarchical].
+To meet desired management and conservation objectives, modeled escapement includes estimates of uncertainty and is parsed into weekly strata.
+Further, total and weekly estimates are parsed into three origin groups: wild, hatchery, and hatchery no-clip.
+Estimates of escapement account for fish that migrate through the ladder at night outside of observation hours and account for fish that may ascend the ladder multiple times due to fallback and re-ascension.
+Our model is implemented in the **ST**ate-space **A**dult **D**am **E**scapement **M**odel (STADEM) package for the statistical software R [@R-Core-Team2020], and is available for downloaded at [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM).
+To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates.
 The STADEM model combines multiple imperfect sources of data to reduce bias in escapement estimates and provide more reasonable estimates of uncertainty.
-This STADEM framework could be applied elsewhere to any fish passage barrier with a counting mechanism like a window, an adult trap that samples a portion of the run and contains a PIT tag antenna (or similar).
+Additionally, the STADEM model framework can be applied elsewhere to estimate aggregate and group escapement, provided a passage barrier containing multiple counting and observational mechanisms exist, such as, a counting window, an adult trap sampling a portion of the run, and PIT tag antenna infrastructure.
 
-# Materials and Methods
+# Methods
 
@@
 
-We used STADEM and three sources of data to estimate Chinook salmon and steelhead escapement at Lower Granite Dam.
-Data sources included 1) counts of fish migrating pas an observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of previously PIT tagged fish detected in the adult fish ladder.
-Following, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; however, similar data could also likely be obtained from other fish passage facilities.
+We used STADEM and three sources of data to estimate Chinook Salmon and steelhead escapement at Lower Granite Dam from 2010-2019.
+Data sources included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of previously PIT tagged fish detected in the adult fish ladder.
+Following, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; however, similar data could likely be obtained from other fish passage facilities.
 
@@
 
-Daily counts of adult Chinook salmon and steelhead passing an observation window located on the Lower Granite fish ladder are the first source of data.
-Daily counts are made and provided by the US Army Corps of Engineers (COE) and, when summed, provide an estimate of the number of fish ascending and passing (i.e., escapement) Lower Granite Dam each season.
-Window counts are made for each species using both video monitoring and direct visual monitoring during daytime hours [@Hatch1994].
-Video monitoring occurs during the beginning and tail ends of the adult runs (March 1 – March 31 and November 1 – December) and for 10 hours per day (0600 – 1600 hours).
-Direct visual monitoring occurs during peak run times (April 1 – October 31) and for 16 hours per day (0400 – 2000 hours) [@USACE2015].
-During direct visual monitoring, observers record each adult (≥ 30cm), by species, passing the window for 50 minutes of each hour of operation.
-Salmonids under 30cm in length are not identified to species.
-The sum of the daily 50-minute counts are then multiplied by 1.2 to account for the 10 minutes where fish are not counted.
-These daytime window counts are not expanded for fish that may ascend the ladder outside of operational hours (i.e., nighttime) [@USACE2015].
-Window counts are accessed through the Columbia Basin Research Data Access in Real Time (DART) [website](www.cbr.washington.edu/dart/query/adult_daily), using their window count query.
-Counts are provided for each day the fish ladder is open to passage.
-Although window counts are assumed to be a census of every fish passing Lower Granite Dam, nighttime passage and re-ascending fish are not accounted for.
-Further, there is no estimation of daily or seasonal observation error.
+Daily counts of adult Chinook Salmon and steelhead passing an observation window located on the Lower Granite fish ladder were the first source of data.
+Daily counts were made and provided by the US Army Corps of Engineers, and when summed, provide an estimate of the number of fish ascending and passing (i.e., escapement) Lower Granite Dam each season.
+Window counts were made for each species using video monitoring and direct in-person visual monitoring during daytime hours [@Hatch1994].
+Video monitoring occured during the beginning and tail ends of the adult runs (March 1 – March 31 and November 1 – December) for 10 hours per day (0600 – 1600 hours).
+Direct visual monitoring occured during peak run times (April 1 – October 31) for 16 hours per day (0400 – 2000 hours) [@USACE2015].
+During direct visual monitoring, observers recorded each adult (≥ 30cm), by species, passing the window for 50 minutes of each hour of operation.
+Salmonids under 30cm in length were not identified to species.
+The sum of the daily 50-minute counts were then multiplied by 1.2 to account for the 10 minutes when fish were not counted.
+Daytime window counts were not expanded for fish that may have ascended the ladder outside of operational hours (i.e., nighttime) [@USACE2015].
+Window counts were accessed through the Columbia Basin Research Data Access in Real Time (DART) website, [www.cbr.washington.edu/dart/query/adult_daily](www.cbr.washington.edu/dart/query/adult_daily), using their window count query.
+Counts were provided for each day the fish ladder was open to passage.
+Although window counts were assumed to be a census of every fish passing Lower Granite Dam, corrections were not applied for nighttime passage or re-ascending fish.
+Further, there was no estimation of daily or seasonal observation or sampling error.
 
@@
 
-The second source of data comes from a trap that captures a sample of the adults as they migrate past Lower Granite Dam [@Ogden2016b].
-The trap, also located within the adult fish ladder and upstream of the observation window, provides biological data (e.g., origin [wild, hatchery], genetic stock, length, age, sex) for captured adults that allows decomposition of the escapement into specific groups (e.g., @Camacho2019, @Steinhorst2017).
-The trap randomly samples the daily run by opening four times per hour for a length of time determined by a set daily trapping rate.
-The trap is operational 24 hours per day.
-The trap rate is determined by a committee of collaborating management agencies with a goal of capturing a target number of wild fish but also balancing fish handling concerns.
-Trap sample rates are typically 10-25%, but may change throughout the season due to high water temperatures, decreased flows, trap malfunctions and/or closures, fish handling logistics, etc.
-All captured fish are anesthetized, identified to species, examined for marks/tags, and measured for fork length (FL).
-Additionally, each fish has scale and genetic tissue samples taken; scale samples are used to estimate age [@Wright2015], tissue samples are used to determine sex and for genetic stock identification analysis (e.g., @Hargrove2019).
-Prior to release, all non-PIT tagged fish with an intact adipose fin (i.e., putatively wild) receive a PIT tag.
-Final determination of wild, hatchery, and hatchery no-clip origins are assigned using a post-hoc analysis of marks, tags, and genetic information.
-Data from the adult trap are collected and managed by multiple agencies and are made available by the Idaho Department of Fish and Game (IDFG) in a comma separated value (.csv) file [@Camacho2019].
+The second source of data came from a sample of fish collected in the adult trap as they migrated past Lower Granite Dam [@Ogden2016b].
+The trap, also located within the adult fish ladder and upstream of the observation window, provided biological data (e.g., origin [wild, hatchery], genetic stock, length, age, sex) for captured adults that allowed decomposition of the escapement into specific groups (e.g., @Camacho2019, @Steinhorst2017).
+The trap was operational for 24 hours per day, and randomly sampled the daily run by opening four times per hour for a length of time determined by a set daily trapping rate.
+The trap rate was determined by a committee of collaborating management agencies with a goal of capturing a target number of wild fish, but also balancing fish handling concerns.
+Trap sample rates were typically 10-25%, but fluctuated throughout the season due to high water temperatures, decreased flows, trap malfunctions and/or closures, fish handling logistics, etc.
+All captured fish were anesthetized, identified to species, examined for existing marks/tags, and measured for fork length.
+Additionally, each fish had scale and genetic tissue samples taken; scale samples were used to estimate age [@Wright2015], tissue samples were used to determine sex and for genetic stock identification analysis (e.g., @Hargrove2019).
+Prior to release, all non-PIT tagged fish with an intact adipose fin (i.e., putatively wild) received a PIT tag.
+Final determination of wild, hatchery, and hatchery no-clip origins were assigned using a post-hoc analysis of marks, tags, and genetic information.
+Data from the adult trap were collected and managed by multiple agencies and were made available by the Idaho Department of Fish and Game [@Camacho2019].
 
-### PIT-tag Data
+### PIT tag Data
 
-The last source of data is observations of PIT tagged adult Chinook salmon and steelhead at detection sites located in the Lower Granite Dam fish ladder.
+The last source of data was observations of PIT tagged adult Chinook Salmon and steelhead at detection sites located in the Lower Granite Dam fish ladder.
 These observations provide estimates of 1) a trapping rate, 2) the proportion of fish passing during nighttime hours (i.e., outside of window observation hours), and 3) the proportion of fish that ascend the fish ladder multiple times (i.e., re-ascension rate).
-Detections used in the model include all fish that were previously PIT tagged as juveniles or adults prior to reaching Lower Granite Dam (i.e., not those tagged at the dam as adults) and detected at adult detection sites in the dam passage system.
-Data are provided through DART and the [adult ladder PIT tag query](http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail).
+Detections used in the model include all fish that were previously PIT tagged as juveniles or adults prior to reaching Lower Granite Dam (i.e., does not include newly tagged adults at the dam) and detected at adult detection sites in the dam passage system.
+PIT tag data was provided through DART and the adult ladder PIT tag query; [http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail](http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail).
 
-A trap rate estimate is derived using mark-recapture methods and PIT tag observations of both Chinook salmon and steelhead at Lower Granite Dam adult detection sites.
-The "mark" group includes all tags detected in the adult trap and the "capture" group includes tags observed to cross the weir at the upstream end of the fish ladder as adults leave the passage system.
-Using a mark-recapture model with differing capture probabilities, we can estimate the trap rate on a weekly basis.
-Those estimates, with associated uncertainties, are then incorporated into the model as informed priors, while the model estimates the true trap rate based on all the data, including trap and window counts.
-The true trapping rate is estimated as the recorded time of the trap being open does not always reflect the true proportion of fish that are captured in the trap, due to trap malfunctions, sort-by-code fish opening the trap more frequently than expected, or process error (Rick Orme, *per.
+A trap rate estimate was derived using mark-recapture methods and PIT tag observations of both Chinook Salmon and steelhead at Lower Granite Dam adult detection sites.
+The "mark" group included all tags detected in the adult trap and the "capture" group included tags observed to cross the weir at the upstream end of the fish ladder as adults left the passage system.
+Using a mark-recapture model with differing capture probabilities, we estimated the trap rate on a weekly basis.
+Those estimates, with associated uncertainties, were then incorporated into the model as informed priors, while the model estimated the true trap rate based on all the data, including trap and window counts.
+<!--The true trapping rate was estimated as the recorded time of the trap being open does not always reflect the true proportion of fish that are captured in the trap, due to trap malfunctions, sort-by-code fish opening the trap more frequently than expected, or process error (Rick Orme, *per.
 comm.*).
 Therefore, we use the mark-recapture estimate of trap rate.
+-->
 
-The nighttime passage rate is estimated simply as the proportion of tags that migrate through the fish ladder during "nighttime" hours outside of the window observation hours divided by the total number of tags passing the fish ladder and is calculated on a weekly basis.
-The re-ascension rate is the number of tags observed passing the upstream most detection sites in the adult fish ladder (i.e., passing the dam) and later detected re-entering the downstream end of the fish ladder at a later time divided by the total number of tags leaving the fish ladder.
-Previously, we examined for differences in nigh passage and re-ascension rates estimated using wild fish, versus combining hatchery and wild fish together, and found no difference.
+The nighttime passage rate was estimated as the count of tags that migrated through the fish ladder during non-window observation hours divided by the total number of tags passing the fish ladder, and was calculated on a weekly basis.
+The re-ascension rate was the count of tags observed passing the upstream most detection sites in the adult fish ladder (i.e., passing the dam) and later detected re-entering the downstream end of the fish ladder at a later time divided by the total number of tags leaving the fish ladder.
+Previously, we examined for differences in nighttime passage and re-ascension rates as estimated using wild fish only, versus combining hatchery and wild fish together, and found no difference.
 Therefore, we combine wild and hatchery PIT tagged fish observations to estimate common nighttime passage and re-ascension rates to increase sample sizes.
@@
 
-STADEM estimates the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for nighttime passage and fallback-and-reascension rates.
-Using a state-space approach, STADEM assumes that the window counts and the estimates from the trap (fish in the trap divided by trap rate that week) are generated by processes with observation error.
-In the case of the trap, there is sampling variation and uncertainty in what the true trap rate is.
-STADEM further accounts for the proportion of fish that pass while the counting window is closed (i.e., night), as well as the potential double-counting (or more) of fish that have fallen back below the dam and later re-ascended the fish ladder.
-Next, origin data (wild, hatchery, hatchery no-clip) are used to partition the total escapement estimate by origin (Figure \@ref(fig:stadem-examp-fig).
+We estimated the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for nighttime passage, and fallback/re-ascension rates using a state-space modeling approach [@royle2008hierarchical] implemented in the STADEM package for the R statistical software [@R-Core-Team2020].
+We assumed that the window counts and the estimates from the trap (fish in the trap divided by trap rate that week) were generated by processes with observation error.
+In the case of the trap for example, we assumed there was sampling variation and uncertainty around our estimates of the true unknown trap rate.
+STADEM further accounted for the proportion of fish that ascended the ladder while the counting window was closed (i.e., night), as well as for fish potentially double-counted (or more) after falling back below the dam and later re-ascending the fish ladder.
+Last, adult sampling data from the trap (wild, hatchery, hatchery no-clip) were used to partition the total escapement estimate by origin (Figure \@ref(fig:stadem-examp-fig).
 Additional model details can be found in [Appendix A](#append1).
-STADEM is implemented in R software [@R-Core-Team2020] and is available from the primary author at a [GitHub site](https://github.com/KevinSee/STADEM).
+The STADEM package is available from the primary author at a [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM), and requires the use of the JAGS software [@R-rjags] for Bayesian inference.
 
@@
 These simulated data sets contained a fixed number of unique adult fish of known origin crossing a dam, from a total of 25 fictional populations with differential run-timing (i.e., date of passage at Lower Granite Dam).
-Each simulated fish is given a date of ascending the ladder, based on its population and the range of run timing for that population.
-Each fish is also simulated to cross the dam either while the window is open for counting, or not, and is given the change to be "caught" in the simulated fish trap given the week when it ascends the dam, and the known trap rate that week.
+Each simulated fish was given a date of ladder ascension, based on its population and the range of run timing for that population.
+Each fish was also simulated to cross the dam either while the window was open for counting, or not, and was given the chance to be "caught" in the simulated fish trap given the week when it ascended the dam, and the known trap rate that week.
 Fallback and re-ascension behavior was also simulated, with each fish having the possibility of falling back and re-ascending the ladder up to three times.
 
-Our objective was to examine STADEM model estimates of origin-specific (wild, hatchery, hatchery no-clip) escapement from the combinations of two separate trapping rates, two fallback, re-ascension and nighttime passage combinations, and three window count error rates; resulting in twelve different scenarios (Table \@ref(tab:lgr-sim-tab)).
+Our objective was to examine STADEM model estimates of origin-specific (wild, hatchery, hatchery no-clip) escapement from the combinations of two separate trapping rates, two fallback/re-ascension and nighttime passage combinations, and three window count error rates; resulting in twelve different scenarios (Table \@ref(tab:lgr-sim-tab)).
 The simulation parameters such as proportion of origin, run-timing, nighttime passage rates, fallback and re-ascension rates and trap rates were based on observed values at Lower Granite Dam between 2010-2015.
@@
 
-Finally, we applied STADEM to data from Lower Granite Dam for Chinook salmon and steelhead returning to the Snake River to spawn during 2010 to 2019.
-Window counts for both species were accessed from DART via STADEM.
-For spring-summer run Chinook salmon, a spawn year refers to adults that migrate pas the dam prior to August 17 each year and spawn that late summer and fall.
+Finally, we applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead returning to the Snake River to spawn during 2010 to 2019.
+Window counts for both species were accessed from DART via functions within STADEM.
+For Chinook Salmon, a spawn year refers to adults that migrate past the dam prior to August 17 each year and spawn that late summer and fall.
 For steelhead, the spawn year is defined as steelhead that migrate past Lower Granite Dam starting July 1 the previous year and prior to June 30 of the given year (e.g., spawn year 2017 steelhead migrate past Lower Granite between July 1, 2016 and June 30, 2017).
-Data from the adult trap was made available by IDFG and adult detection data at Lower Granite Dam from the fish passage ladder is available in the regional database PTAGIS ([PIT Tag Information System](https://www.ptagis.org/)).
+Data from the adult trap was made available by Idaho Department of Fish and Game, and adult PIT tag detection data within the fish passage ladder at Lower Granite Dam was accessed from the PTAGIS regional database ([https://www.ptagis.org/](https://www.ptagis.org/)).
 
@@
 STADEM results were very similar across all scenarios (Figure \@ref(fig:rel-bias-lgd)).
-Estimates were unbiased, with an average relative bias of 0.2 – 0.3%.
-The CV of the estimates average 2 – 3%, with coverage probabilities that always exceeded 95%.
+Estimates were unbiased, with an average relative bias of 0.2–0.3%.
+The CV of the estimates average 2.0–3.0%, with coverage probabilities that always exceeded 95%.
 We calculated RMSD as the square root of the sum of the variance of the estimate and the squared expected bias, which accounts for the size of the uncertainty in the estimator as well as its bias.
@@
 
-We applied STADEM to data from Lower Granite Dam for Chinook salmon and steelhead for spawn years 2010 – 2019.
+We applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead for spawn years 2010–2019.
 Estimates of total escapement, as well as estimates of wild, hatchery, and hatchery no-clip estimates are presented in Table \@ref(tab:lgr-tab).
-Coefficients of variation range from 3.5 - 7.5% for wild fish, 3.0 - 6.2% for hatchery fish, 4.0 - 8.4% for hatchery no-clip fish and 2.9 - 5.6% for total unique fish past Lower Granite Dam.
+Coefficients of variation ranged from 3.5-7.5% for wild fish, 3.0-6.2% for hatchery fish, 4.0-8.4% for hatchery no-clip fish and 2.9-5.6% for total unique fish past Lower Granite Dam.
 
-Weekly estimates of total escapement over Lower Granite Dam track the window counts and trap estimates (Figure \@ref(fig:jags-fit-fig)).
-STADEM point estimates are often between estimates based on window counts, and those based on the number of fish caught in the adult trap.
-However, when very few fish are caught in the trap, or there is more uncertainty about the trap rate that week, STADEM estimates track the window counts more closely, as seen in the second week of July 2014 in Figure \@ref(fig:jags-fit-fig).
+Weekly estimates of total escapement over Lower Granite Dam tracked the window counts and trap estimates (Figure \@ref(fig:jags-fit-fig)).
+STADEM point estimates were often between estimates based on window counts, and those based on the number of fish caught in the adult trap.
+However, when very few fish were caught in the trap, or there was more uncertainty about the trap rate that week, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \@ref(fig:jags-fit-fig).
 That year shows the utility of STADEM in dealing with missing data, as the trap was shut down for several weeks in July and August.
 
-Examining the estimates of nighttime passage and re-ascension rates based on the observed PIT tags crossing over Lower Granite Dam, the two rates often do not match up (Figure \@ref(fig:night-reasc-diff-fig)).
+Examining the estimates of nighttime passage and re-ascension rates based on the observed PIT tags crossing over Lower Granite Dam, the two rates did not match up (Figure \@ref(fig:night-reasc-diff-fig)).
 In particular, there are several weeks when the window counts are quite high, and the rates differ by as much as 10%.
@@
 
-We have presented a novel method for estimating adult salmonid escapement past a large hydroelectric facility (e.g., Lower Granite Dam) that incorporates data from window counts, a fish trap, and observations of PIT-tagged fish in the adult passage ladder.
+We have presented a novel method for estimating adult salmonid escapement past a large hydroelectric facility (e.g., Lower Granite Dam) that incorporates data from window counts, fish trap, and observations of PIT tagged fish in the adult passage ladder.
 Our model explicitly models nighttime passage, re-ascension, and potential error in both window and trap estimates.
-In doing so, we have demonstrated that at Lower Granite Dam, nighttime passage and re-ascension do not always offset each other, and the assumption that they do will lead to biased estimates in some years.
-This framework and the STADEM model could be applied elsewhere to any fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run, and contains observation or detection infrastructure (e.g., a PIT tag detection array or similar) and could be applied to similar migratory fishes (e.g., Pacific lamprey Lampetra tridentata).
-Our state-space model combines multiple imperfect sources of data to reduce bias in adult escapement estimates and provide more reasonable estimates of uncertainty.
+In doing so, we demonstrated that at Lower Granite Dam, nighttime passage and re-ascension rates do not always offset each other, and assuming they do, will lead to biased escapement estimates in some years.
+With minor adjustments this modelling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall Chinook Salmon, Pacific lamprey Lampetra tridentata), or elsewhere; provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run, and a tag observation or detection infrastructure (e.g., a PIT tag detection array or similar) exists.
+Our state-space model combined multiple imperfect sources of data to reduce bias in adult escapement estimates and provided more reasonable estimates of uncertainty.
 Accurate population or stock abundance estimates and uncertainty accounting for observation and process error can be particularly important when estimates are used or leveraged for management and conservation decisions such as population viability analyses.
@@
 First, it allows us to model observer error in the window counts, which is typically unknown.
-If estimates rely on window counts alone, this is impossible.
-We believe it necessary to model and estimate multiple sources of uncertainty in escapement estimates.
-Second, by incorporating both sources of information in a state-space framework, STADEM incorporates missing data from either source seamlessly.
+If estimates rely on window counts alone, quantifying observer error is impossible, and we believe it's prudent to capture and account for known sources of error to minimize management decision risk.
+Second, by incorporating both sources of information in a state-space framework, STADEM incorporates missing data at either the observation window or adult trap seamlessly.
 At Lower Granite Dam, the adult trap has been closed for brief or extended periods of time (i.e., days, weeks) intermittently over the past several years, often during peak run times.
-Trap closures are typically associated due to elevated temperatures resulting in potential fish handling stress and/or trap malfunctions.
-Climate change scenarios for the Pacific Northwest could cause trap closures from high water temperatures more commonplace in the future, amplifying the need for a framework that accounts for those periods of missing data.
-Additionally, having a framework in place that accounts for missing periods of data, at either the observation window or adult trap, allows for increased logistic flexibility if, for example, maintenance or construction needed to occur at the observational window or similar.
+Trap closures are typically associated with elevated water temperatures resulting in potential fish handling stress and/or trap malfunctions.
+Given predicted Pacific Northwest climate change scenarios trap closures from high water temperatures may become more commonplace in the future; amplifying the need for a modeling framework that accounts for periods of missing data while still capturing estimate uncertainty.
+Additionally, having a framework in place that accounts for missing periods of data will allow for increased logistic flexibility if, for example, maintenance or construction is needed at the observational window or adult trap.
 
@@
 
-Although not currently set up for this, STADEM could be modified to be run on a weekly basis or in near real time to provide better in-season estimates for fisheries managers.
-Currently, the only roadblock to this, at Lower Granite Dam, is that the genetic origin data (***Steele et al.
-2013***) that is used to identify phenotypically wild fish that are truly hatchery fish (i.e., hatchery no-clip) is completed post hoc after the trapping season.
-This information typically results in a reduction in wild escapement estimates and an associated increase in hatchery no-clip escapement.
-However, if in-season management decisions do not rely on this correction, origin calls at the trap could be used instead, certainly as a first approximation.
-Estimates, parsed by origin, could then be finalized at season’s end.
-Most other data included in this model (e.g., window counts and PIT observations) are otherwise provided in almost real time by DART.
-As long as the Lower Granite Dam adult fish trap database was updated and made available frequently, there are minimal obstacles for adapting the STADEM framework to provide in-season estimates of escapement.
-Results could be provided in a web-accessible application to managers or end users.
+Although not currently set up for this, STADEM could be modified and run on a weekly basis or in near real-time to provide better in-season estimates for fisheries managers.
+Currently, the only roadblock to this, at Lower Granite Dam, is the identification of hatchery origin fish from phenotypically wild fish (i.e., hatchery no-clip) using genetic tissues samples (***Steele et al.
+2013***) collected at the adult trap, which currently is completed post hoc after the trapping season.
+The inclusion of genetic information typically results in a reduction in wild escapement estimates and an associated increase in hatchery no-clip escapement.
+However, if in-season management decisions do not require this correction or could accept the potential bias, origin calls at the trap could be used in-season as a first approximation to escapement.
+Final post-hoc estimates, parsed by origin, could then be finalized at season’s end.
+All other data included in this model (e.g., window counts and PIT observations) are otherwise provided in almost real-time by DART.
+Provided the Lower Granite Dam adult fish trap database was updated and made available frequently, there are minimal obstacles for adapting the STADEM framework to provide in-season estimates of escapement.
 
-The STADEM model is available as a package for the statistical software R [@R-Core-Team2020] and can be downloaded from a [GitHub site](https://github.com/KevinSee/STADEM).
-STADEM also uses the JAGS software [@R-rjags] for Bayesian inference from the model.
-Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate escapement of spring/summer Chinook salmon and steelhead past Lower Granite Dam.
-Estimates of escapement, by species and origin, are then available to parse into sex- or age-structured escapement estimates (e.g., @Camacho2019, @Schrader2013) that are important for fisheries management and productivity monitoring of wild populations.
-As an example, STADEM is being applied at Lower Granite to estimate the total unique wild fish migrating past; that information is then combined with PIT tag observations at instream PIT tag detection systems throughout the Snake River basin that are used to estimate movement or transition probabilities to populations and locations throughout the basin using the Dam Adult Branch Occupancy Model ([DABOM](https://github.com/KevinSee/DABOM)).
-Combined, escapement estimates from STADEM and movement probability estimates from DABOM, provide abundance estimates to given tributaries or populations that, joined with sex and age data collected at the adult fish trap, provides necessary information to evaluate productivity and population viability for select Snake River Chinook salmon and steelhead groups.
+Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary spawning areas[@iptds2020report; @kinzer2020report].
+Estimates of escapement at Lower Granite Dam, by species and origin, include known uncertainty, and are available to further parse into sex- or age-structured escapement estimates (e.g., @Camacho2017, @Schrader2013) that are important for fisheries management and productivity monitoring of wild populations.
+As an example, STADEM is being applied at Lower Granite to estimate the total unique wild fish migrating past the dam.
+Estimates of fish passing the dam are then combined with PIT tag observations at instream PIT tag detection systems throughout the Snake River basin and estimated movement or transition probabilities, simialar to @waterhouse2020bayesian, to estimate escapement to Snake River populations and locations throughout the basin [@orme2018population].
+Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations that, joined with sex and age data collected at the adult fish trap [@Hargrove2019], provides necessary information to evaluate productivity and population viability for select Snake River Chinook Salmon and steelhead groups [@iptds2020report].
 
-Our STADEM framework can be applied to any fish passage barrier with a counting mechanism like a window, a trap that samples a subset of the run, and contains a PIT tag antenna.
-Although STADEM was developed with salmonids in mind, it could be applied to any migrator fish species.
-Certainly, there is justification for applying a similar framework at Bonneville Dam, as the lowest dam on the Columbia River, to estimate total returns to the Columbia River basin.
-Lower Granite Dam is important because it is the farthest upstream dam on the Snake River with fish passage, and other methods of estimating escapted to smaller spatial domains (e.g., MGP/DPS, populations) upstream rely on good estimates of total escapement there.
-Priest Rapids plays a similar role for upstream migrating adult salmonids in the Upper Columbia River, and could also benefit from the STADEM approach.
+Although STADEM was developed with salmonid escapement at Lower Granite Dam in mind, it could be applied to any migratory fish species at locations with similar monitoring infrastructure.
+Justification and infrastructure exist for applying a modified STADEM framework for fish passing Bonneville Dam, the lowest dam on the Columbia River, or Priest Rapids Dam in the upper Columbia River.
+Both locations currently trap a sub-sample of passing Chinook Salmon and steelhead for biological information, and use window counts as a surrogate of true escapement, however, similar problems to those observed at Lower Granite Dam, such as, unaccounted observer and sampling error, night-time passage, and re-ascension exists at the sites.
+Certainly, estimating an unbiased total return to the entire Columbia River basin (i.e., Bonneville Dam) and Upper Columbia River with uncertainty would benefit managers and decision making.
 
-# Acknowledgements
+## Acknowledgements
 
@@
 Special thanks to Darren Ogden and staff at the Lower Granite Dam adult trap for their hard work and diligent data collection.
-Thank you to personnel at the Idaho Department of Fish and Game, particularly Paul Bunn and Bill Schrader, for providing access to data from the adult fish trap.
-Special thanks to staff at Columbia Basin Research and the Columbia River Data Access in Real Time (DART) application, and Susannah Iltis in particular.
-And finally, thank you to all the folks who have contributed to the development of the STADEM framework through productive critique and conversations.
+Thank you to personnel at the Idaho Department of Fish and Game, particularly Paul Bunn, Tim Copeland, and Bill Schrader, for providing access to data from the adult fish trap and conceptualizing methods, and the staff at the Eagle Fish Lab for analyzing genetic samples.
+Special thanks to Columbia Basin Research staff and the Columbia River Data Access in Real Time (DART) application, and Susannah Iltis in particular.
+And finally, thank you to Rick Orme for his contributions to the development of STADEM, and all the other folks who have contributed through productive critique and conversations.
 
@@
   kable(booktabs = T,
-        caption = "Window counts, and estimates of total, wild, hatchery and hatchery no-clip escapement, with coefficients of variation in parenthesis, for Chinook salmon and steelhead from spawn years 2010 to 2019.") %>%
+        caption = "Window counts, and estimates of total, wild, hatchery and hatchery no-clip escapement, with coefficients of variation in parenthesis, for Chinook Salmon and steelhead from spawn years 2010 to 2019.") %>%
   kable_styling()

### 2020-08-18 inst/scripts/SimFnc.R

@@ -1,0 +1,436 @@
+# Author: Ryan N. Kinzer
+# Purpose: Simulate data to estimate total escapement over Lower Granite Dam
+# Created: 8/10/2016
+# Last Modified: 9/12/2016
+#-----------------------------------------------------------------
+# LGR Simulation Function
+SimulateLGRdata = function(N.lgr = 100000,
+                           h.prob = .70,
+                           hnc.prob = .05,
+                           n.pops = 25,
+                           hatch.pop.prob = 0.5,
+                           hnc.pop.prob = 0.15,                           
+                           run.mu.mu = 171,
+                           run.mu.sd = 13,
+                           run.sd.mu = 22,
+                           run.sd.sd = 7,
+                           fallback.rate = .06,
+                           reascension.rate = 1,
+                           night.passage.rate = .06,
+                           window.rate = 50/60,
+                           marked.rate = 0.05,
+                           ladder.det = 0.99,
+                           start.date = ymd('20150101'),
+                           trap.rate.df) {
+  # N.lgr: Lower Granite total escapement
+  # n.pops: number of spring pops and summer pops
+  # run.mu.mu, run.mu.sd: Hyper-parameters for mean run-timing across spring and summer populations
+  # run.sd.mu, run.sd.sd: Hyper-parameters for sd of run-timing across spring and summer populations
+  # hatch.pop.prob: percentage of populations with hatchery populations
+  # hnc.pop.prob: percentage of populations with hatchery no-clip populations
+  # fallback.rate: probability of fish fall-back
+  # reascension.rate: conditional probability of fish reascending given they fall-back
+  # need to think about Tucannon R., reascension does not equal 1
+  # possibly set at some weighted mean from previous studies
+  # night.passage.rate: probability of fish passing during non window operation
+  # window.rate: minutes per hour
+  # marked.rate: proportion of fish previously tagged (i.e. marked)
[…]

### 2020-08-18 inst/scripts/RunSims_Baseline.R
added RunSims scripts

@@ -1,0 +1,253 @@
+# Author: Kevin See
+# Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model
+# Created: 8/31/2016
+# Last Modified: 9/12/2016
+# Notes: 
+
+#-----------------------------------------------------------------
+# these functions contains the scripts to run the SCOBI estimator
+source('SCOBI/SCOBIv2.R')
+source('SCOBI/SCSrank.R')
+
+# load needed packages
+library(Hmisc) # needed for SCOBI model
+library(MCMCpack)
+library(FSA)
+library(Rcapture)
+library(boot)
+library(msm)
+library(lubridate)
+library(magrittr)
+library(plyr)
+library(dplyr)
+library(tidyr)
+library(jagsUI)
+
+# this contains the function to simulate data
+source('SimFnc.R')
+
+#-----------------------------------------------------------------
+# set mcmc parameters
+#-----------------------------------------------------------------
+# number of total samples
+mcmc.chainLength = 7000
+
+# number of burn-in samples
+mcmc.burn = 2000
+
+# thinning interval
+mcmc.thin = 10
+
+# number of MCMC chains
+mcmc.chains = 4
+
+# how many poseterior samples will we have?
+data.frame(Each.Chain = (mcmc.chainLength - mcmc.burn) / mcmc.thin, All.Chains = (mcmc.chainLength - mcmc.burn) / mcmc.thin * mcmc.chains)
+
+
+#-----------------------------------------------------------------
+# set up model location and which parameters to track with JAGS
+#-----------------------------------------------------------------
+model.loc = 'LGR_TotalEscape_AllOrigins.txt'
+
+jags.params = c('X.tot.all', 'X.tot.day', 'X.tot.night', 'X.tot.reasc', 'X.tot.new.wild', 'X.tot.new.hatch', 'X.tot.new.hnc', 'X.sigma', 'true.prop', 'win.prop.avg', 'win.prop.true', 'win.prop.sigma', 'reasc.avg', 'reasc.true', 'reasc.sigma', 'acf', 'org.prop', 'org.sigma', 'trap.rate.true', 'r', 'k')
+
+# set initial values
+jags.inits = function(){
+  return(list('X.log.all' = with(jags.data, log(Y.window + 1)),
+              'trap.rate.true' = with(jags.data, trap.rate + 0.0001)))
+}
+
+#-----------------------------------------------------------------
[…]

### 2020-08-18 inst/scripts/RunSims_Baseline_WinErr_high.R

@@ -1,0 +1,258 @@
+# Author: Kevin See
+# Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model
+# Created: 8/31/2016
+# Last Modified: 9/12/2016
+# Notes: 
+
+#-----------------------------------------------------------------
+# these functions contains the scripts to run the SCOBI estimator
+source('SCOBI/SCOBIv2.R')
+source('SCOBI/SCSrank.R')
+
+# load needed packages
+library(Hmisc) # needed for SCOBI model
+library(MCMCpack)
+library(FSA)
+library(Rcapture)
+library(boot)
+library(msm)
+library(lubridate)
+library(magrittr)
+library(plyr)
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(jagsUI)
+
+
+theme_set(theme_bw())
+
+# this contains the function to simulate data
+source('SimFnc.R')
+
+
+#-----------------------------------------------------------------
+# set mcmc parameters
+#-----------------------------------------------------------------
+# number of total samples
+mcmc.chainLength = 7000
+
+# number of burn-in samples
+mcmc.burn = 2000
+
+# thinning interval
+mcmc.thin = 10
+
+# number of MCMC chains
+mcmc.chains = 4
+
+# how many poseterior samples will we have?
+data.frame(Each.Chain = (mcmc.chainLength - mcmc.burn) / mcmc.thin, All.Chains = (mcmc.chainLength - mcmc.burn) / mcmc.thin * mcmc.chains)
+
+
+#-----------------------------------------------------------------
+# set up model location and which parameters to track with JAGS
+#-----------------------------------------------------------------
+model.loc = 'LGR_TotalEscape_AllOrigins.txt'
+
+jags.params = c('X.tot.all', 'X.tot.day', 'X.tot.night', 'X.tot.reasc', 'X.tot.new.wild', 'X.tot.new.hatch', 'X.tot.new.hnc', 'X.sigma', 'true.prop', 'win.prop.avg', 'win.prop.true', 'win.prop.sigma', 'reasc.avg', 'reasc.true', 'reasc.sigma', 'acf', 'org.prop', 'org.sigma', 'trap.rate.true', 'r', 'k')
+
+# set initial values
+jags.inits = function(){
+  return(list('X.log.all' = with(jags.data, log(Y.window + 1)),
+              'trap.rate.true' = with(jags.data, trap.rate + 0.0001)))
+}
+
[…]

### 2020-08-18 inst/scripts/RunSims_Baseline_WinErr_low.R

@@ -1,0 +1,258 @@
+# Author: Kevin See
+# Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model
+# Created: 8/31/2016
+# Last Modified: 9/12/2016
+# Notes: 
+
+#-----------------------------------------------------------------
+# these functions contains the scripts to run the SCOBI estimator
+source('SCOBI/SCOBIv2.R')
+source('SCOBI/SCSrank.R')
+
+# load needed packages
+library(Hmisc) # needed for SCOBI model
+library(MCMCpack)
+library(FSA)
+library(Rcapture)
+library(boot)
+library(msm)
+library(lubridate)
+library(magrittr)
+library(plyr)
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(jagsUI)
+
+
+theme_set(theme_bw())
+
+# this contains the function to simulate data
+source('SimFnc.R')
+
+
+#-----------------------------------------------------------------
+# set mcmc parameters
+#-----------------------------------------------------------------
+# number of total samples
+mcmc.chainLength = 7000
+
+# number of burn-in samples
+mcmc.burn = 2000
+
+# thinning interval
+mcmc.thin = 10
+
+# number of MCMC chains
+mcmc.chains = 4
+
+# how many poseterior samples will we have?
+data.frame(Each.Chain = (mcmc.chainLength - mcmc.burn) / mcmc.thin, All.Chains = (mcmc.chainLength - mcmc.burn) / mcmc.thin * mcmc.chains)
+
+
+#-----------------------------------------------------------------
+# set up model location and which parameters to track with JAGS
+#-----------------------------------------------------------------
+model.loc = 'LGR_TotalEscape_AllOrigins.txt'
+
+jags.params = c('X.tot.all', 'X.tot.day', 'X.tot.night', 'X.tot.reasc', 'X.tot.new.wild', 'X.tot.new.hatch', 'X.tot.new.hnc', 'X.sigma', 'true.prop', 'win.prop.avg', 'win.prop.true', 'win.prop.sigma', 'reasc.avg', 'reasc.true', 'reasc.sigma', 'acf', 'org.prop', 'org.sigma', 'trap.rate.true', 'r', 'k')
+
+# set initial values
+jags.inits = function(){
+  return(list('X.log.all' = with(jags.data, log(Y.window + 1)),
+              'trap.rate.true' = with(jags.data, trap.rate + 0.0001)))
+}
+
[…]

### 2020-08-18 inst/scripts/RunSims_NightReasc.R

@@ -1,0 +1,260 @@
+# Author: Kevin See
+# Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model and TAC model. Night time passage and re-ascension rates are not equal
+# Created: 8/31/2016
+# Last Modified: 9/12/2016
+# Notes: 
+
+#-----------------------------------------------------------------
+# these functions contains the scripts to run the SCOBI estimator
+source('SCOBI/SCOBIv2.R')
+source('SCOBI/SCSrank.R')
+
+# load needed packages
+library(Hmisc) # needed for SCOBI model
+library(MCMCpack)
+library(FSA)
+library(Rcapture)
+library(boot)
+library(msm)
+library(lubridate)
+library(magrittr)
+library(plyr)
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(jagsUI)
+
+
+theme_set(theme_bw())
+
+# this contains the function to simulate data
+source('SimFnc.R')
+
+
+#-----------------------------------------------------------------
+# set mcmc parameters
+#-----------------------------------------------------------------
+# number of total samples
+mcmc.chainLength = 7000
+
+# number of burn-in samples
+mcmc.burn = 2000
+
+# thinning interval
+mcmc.thin = 10
+
+# number of MCMC chains
+mcmc.chains = 4
+
+# how many poseterior samples will we have?
+data.frame(Each.Chain = (mcmc.chainLength - mcmc.burn) / mcmc.thin, All.Chains = (mcmc.chainLength - mcmc.burn) / mcmc.thin * mcmc.chains)
+
+
+#-----------------------------------------------------------------
+# set up model location and which parameters to track with JAGS
+#-----------------------------------------------------------------
+model.loc = 'LGR_TotalEscape_AllOrigins.txt'
+
+jags.params = c('X.tot.all', 'X.tot.day', 'X.tot.night', 'X.tot.reasc', 'X.tot.new.wild', 'X.tot.new.hatch', 'X.tot.new.hnc', 'X.sigma', 'true.prop', 'win.prop.avg', 'win.prop.true', 'win.prop.sigma', 'reasc.avg', 'reasc.true', 'reasc.sigma', 'acf', 'org.prop', 'org.sigma', 'trap.rate.true', 'r', 'k')
+
+# set initial values
+jags.inits = function(){
+  return(list('X.log.all' = with(jags.data, log(Y.window + 1)),
[…]

### 2020-08-18 inst/scripts/RunSims_NightReasc_TrapShutdown.R

@@ -1,0 +1,266 @@
+# Author: Kevin See
+# Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model and TAC model. Night time passage and re-ascension rates are not equal
+# Created: 8/31/2016
+# Last Modified: 9/12/2016
+# Notes: 
+
+#-----------------------------------------------------------------
+# these functions contains the scripts to run the SCOBI estimator
+source('SCOBI/SCOBIv2.R')
+source('SCOBI/SCSrank.R')
+
+# load needed packages
+library(Hmisc) # needed for SCOBI model
+library(MCMCpack)
+library(FSA)
+library(Rcapture)
+library(boot)
+library(msm)
+library(lubridate)
+library(magrittr)
+library(plyr)
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(jagsUI)
+
+
+theme_set(theme_bw())
+
+# this contains the function to simulate data
+source('SimFnc.R')
+
+
+#-----------------------------------------------------------------
+# set mcmc parameters
+#-----------------------------------------------------------------
+# number of total samples
+mcmc.chainLength = 7000
+
+# number of burn-in samples
+mcmc.burn = 2000
+
+# thinning interval
+mcmc.thin = 10
+
+# number of MCMC chains
+mcmc.chains = 4
+
+# how many poseterior samples will we have?
+data.frame(Each.Chain = (mcmc.chainLength - mcmc.burn) / mcmc.thin, All.Chains = (mcmc.chainLength - mcmc.burn) / mcmc.thin * mcmc.chains)
+
+
+#-----------------------------------------------------------------
+# set up model location and which parameters to track with JAGS
+#-----------------------------------------------------------------
+model.loc = 'LGR_TotalEscape_AllOrigins.txt'
+
+jags.params = c('X.tot.all', 'X.tot.day', 'X.tot.night', 'X.tot.reasc', 'X.tot.new.wild', 'X.tot.new.hatch', 'X.tot.new.hnc', 'X.sigma', 'true.prop', 'win.prop.avg', 'win.prop.true', 'win.prop.sigma', 'reasc.avg', 'reasc.true', 'reasc.sigma', 'acf', 'org.prop', 'org.sigma', 'trap.rate.true', 'r', 'k')
+
+# set initial values
+jags.inits = function(){
+  return(list('X.log.all' = with(jags.data, log(Y.window + 1)),
[…]

### 2020-08-18 inst/scripts/RunSims_NightReasc_TrapShutdown_WinErr_high.R

@@ -1,0 +1,266 @@
+# Author: Kevin See
+# Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model and TAC model. Night time passage and re-ascension rates are not equal
+# Created: 8/31/2016
+# Last Modified: 9/12/2016
+# Notes: 
+
+#-----------------------------------------------------------------
+# these functions contains the scripts to run the SCOBI estimator
+source('SCOBI/SCOBIv2.R')
+source('SCOBI/SCSrank.R')
+
+# load needed packages
+library(Hmisc) # needed for SCOBI model
+library(MCMCpack)
+library(FSA)
+library(Rcapture)
+library(boot)
+library(msm)
+library(lubridate)
+library(magrittr)
+library(plyr)
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(jagsUI)
+
+
+theme_set(theme_bw())
+
+# this contains the function to simulate data
+source('SimFnc.R')
+
+
+#-----------------------------------------------------------------
+# set mcmc parameters
+#-----------------------------------------------------------------
+# number of total samples
+mcmc.chainLength = 7000
+
+# number of burn-in samples
+mcmc.burn = 2000
+
+# thinning interval
+mcmc.thin = 10
+
+# number of MCMC chains
+mcmc.chains = 4
+
+# how many poseterior samples will we have?
+data.frame(Each.Chain = (mcmc.chainLength - mcmc.burn) / mcmc.thin, All.Chains = (mcmc.chainLength - mcmc.burn) / mcmc.thin * mcmc.chains)
+
+
+#-----------------------------------------------------------------
+# set up model location and which parameters to track with JAGS
+#-----------------------------------------------------------------
+model.loc = 'LGR_TotalEscape_AllOrigins.txt'
+
+jags.params = c('X.tot.all', 'X.tot.day', 'X.tot.night', 'X.tot.reasc', 'X.tot.new.wild', 'X.tot.new.hatch', 'X.tot.new.hnc', 'X.sigma', 'true.prop', 'win.prop.avg', 'win.prop.true', 'win.prop.sigma', 'reasc.avg', 'reasc.true', 'reasc.sigma', 'acf', 'org.prop', 'org.sigma', 'trap.rate.true', 'r', 'k')
+
+# set initial values
+jags.inits = function(){
+  return(list('X.log.all' = with(jags.data, log(Y.window + 1)),
[…]

### 2020-08-18 inst/scripts/RunSims_NightReasc_TrapShutdown_WinErr_low.R

@@ -1,0 +1,266 @@
+# Author: Kevin See
+# Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model and TAC model. Night time passage and re-ascension rates are not equal
+# Created: 8/31/2016
+# Last Modified: 9/12/2016
+# Notes: 
+
+#-----------------------------------------------------------------
+# these functions contains the scripts to run the SCOBI estimator
+source('SCOBI/SCOBIv2.R')
+source('SCOBI/SCSrank.R')
+
+# load needed packages
+library(Hmisc) # needed for SCOBI model
+library(MCMCpack)
+library(FSA)
+library(Rcapture)
+library(boot)
+library(msm)
+library(lubridate)
+library(magrittr)
+library(plyr)
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(jagsUI)
+
+
+theme_set(theme_bw())
+
+# this contains the function to simulate data
+source('SimFnc.R')
+
+
+#-----------------------------------------------------------------
+# set mcmc parameters
+#-----------------------------------------------------------------
+# number of total samples
+mcmc.chainLength = 7000
+
+# number of burn-in samples
+mcmc.burn = 2000
+
+# thinning interval
+mcmc.thin = 10
+
+# number of MCMC chains
+mcmc.chains = 4
+
+# how many poseterior samples will we have?
+data.frame(Each.Chain = (mcmc.chainLength - mcmc.burn) / mcmc.thin, All.Chains = (mcmc.chainLength - mcmc.burn) / mcmc.thin * mcmc.chains)
+
+
+#-----------------------------------------------------------------
+# set up model location and which parameters to track with JAGS
+#-----------------------------------------------------------------
+model.loc = 'LGR_TotalEscape_AllOrigins.txt'
+
+jags.params = c('X.tot.all', 'X.tot.day', 'X.tot.night', 'X.tot.reasc', 'X.tot.new.wild', 'X.tot.new.hatch', 'X.tot.new.hnc', 'X.sigma', 'true.prop', 'win.prop.avg', 'win.prop.true', 'win.prop.sigma', 'reasc.avg', 'reasc.true', 'reasc.sigma', 'acf', 'org.prop', 'org.sigma', 'trap.rate.true', 'r', 'k')
+
+# set initial values
+jags.inits = function(){
+  return(list('X.log.all' = with(jags.data, log(Y.window + 1)),
[…]

### 2020-08-18 inst/scripts/RunSims_NightReasc_WinErr_high.R

@@ -1,0 +1,260 @@
+# Author: Kevin See
+# Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model and TAC model. Night time passage and re-ascension rates are not equal
+# Created: 8/31/2016
+# Last Modified: 9/12/2016
+# Notes: 
+
+#-----------------------------------------------------------------
+# these functions contains the scripts to run the SCOBI estimator
+source('SCOBI/SCOBIv2.R')
+source('SCOBI/SCSrank.R')
+
+# load needed packages
+library(Hmisc) # needed for SCOBI model
+library(MCMCpack)
+library(FSA)
+library(Rcapture)
+library(boot)
+library(msm)
+library(lubridate)
+library(magrittr)
+library(plyr)
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(jagsUI)
+
+
+theme_set(theme_bw())
+
+# this contains the function to simulate data
+source('SimFnc.R')
+
+
+#-----------------------------------------------------------------
+# set mcmc parameters
+#-----------------------------------------------------------------
+# number of total samples
+mcmc.chainLength = 7000
+
+# number of burn-in samples
+mcmc.burn = 2000
+
+# thinning interval
+mcmc.thin = 10
+
+# number of MCMC chains
+mcmc.chains = 4
+
+# how many poseterior samples will we have?
+data.frame(Each.Chain = (mcmc.chainLength - mcmc.burn) / mcmc.thin, All.Chains = (mcmc.chainLength - mcmc.burn) / mcmc.thin * mcmc.chains)
+
+
+#-----------------------------------------------------------------
+# set up model location and which parameters to track with JAGS
+#-----------------------------------------------------------------
+model.loc = 'LGR_TotalEscape_AllOrigins.txt'
+
+jags.params = c('X.tot.all', 'X.tot.day', 'X.tot.night', 'X.tot.reasc', 'X.tot.new.wild', 'X.tot.new.hatch', 'X.tot.new.hnc', 'X.sigma', 'true.prop', 'win.prop.avg', 'win.prop.true', 'win.prop.sigma', 'reasc.avg', 'reasc.true', 'reasc.sigma', 'acf', 'org.prop', 'org.sigma', 'trap.rate.true', 'r', 'k')
+
+# set initial values
+jags.inits = function(){
+  return(list('X.log.all' = with(jags.data, log(Y.window + 1)),
[…]

### 2020-08-18 inst/scripts/RunSims_NightReasc_WinErr_low.R

@@ -1,0 +1,260 @@
+# Author: Kevin See
+# Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model and TAC model. Night time passage and re-ascension rates are not equal
+# Created: 8/31/2016
+# Last Modified: 9/12/2016
+# Notes: 
+
+#-----------------------------------------------------------------
+# these functions contains the scripts to run the SCOBI estimator
+source('SCOBI/SCOBIv2.R')
+source('SCOBI/SCSrank.R')
+
+# load needed packages
+library(Hmisc) # needed for SCOBI model
+library(MCMCpack)
+library(FSA)
+library(Rcapture)
+library(boot)
+library(msm)
+library(lubridate)
+library(magrittr)
+library(plyr)
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(jagsUI)
+
+
+theme_set(theme_bw())
+
+# this contains the function to simulate data
+source('SimFnc.R')
+
+
+#-----------------------------------------------------------------
+# set mcmc parameters
+#-----------------------------------------------------------------
+# number of total samples
+mcmc.chainLength = 7000
+
+# number of burn-in samples
+mcmc.burn = 2000
+
+# thinning interval
+mcmc.thin = 10
+
+# number of MCMC chains
+mcmc.chains = 4
+
+# how many poseterior samples will we have?
+data.frame(Each.Chain = (mcmc.chainLength - mcmc.burn) / mcmc.thin, All.Chains = (mcmc.chainLength - mcmc.burn) / mcmc.thin * mcmc.chains)
+
+
+#-----------------------------------------------------------------
+# set up model location and which parameters to track with JAGS
+#-----------------------------------------------------------------
+model.loc = 'LGR_TotalEscape_AllOrigins.txt'
+
+jags.params = c('X.tot.all', 'X.tot.day', 'X.tot.night', 'X.tot.reasc', 'X.tot.new.wild', 'X.tot.new.hatch', 'X.tot.new.hnc', 'X.sigma', 'true.prop', 'win.prop.avg', 'win.prop.true', 'win.prop.sigma', 'reasc.avg', 'reasc.true', 'reasc.sigma', 'acf', 'org.prop', 'org.sigma', 'trap.rate.true', 'r', 'k')
+
+# set initial values
+jags.inits = function(){
+  return(list('X.log.all' = with(jags.data, log(Y.window + 1)),
[…]

### 2020-08-18 inst/scripts/RunSims_TrapShutdown.R

@@ -1,0 +1,265 @@
+# Author: Kevin See
+# Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model
+# Created: 8/31/2016
+# Last Modified: 9/12/2016
+# Notes: 
+
+#-----------------------------------------------------------------
+# these functions contains the scripts to run the SCOBI estimator
+source('SCOBI/SCOBIv2.R')
+source('SCOBI/SCSrank.R')
+
+# load needed packages
+library(Hmisc) # needed for SCOBI model
+library(MCMCpack)
+library(FSA)
+library(Rcapture)
+library(boot)
+library(msm)
+library(lubridate)
+library(magrittr)
+library(plyr)
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(jagsUI)
+
+theme_set(theme_bw())
+
+#setwd('LGR_SimulationStudy')
+
+# this contains the function to simulate data
+source('SimFnc.R')
+
+
+#-----------------------------------------------------------------
+# set mcmc parameters
+#-----------------------------------------------------------------
+# number of total samples
+mcmc.chainLength = 7000
+
+# number of burn-in samples
+mcmc.burn = 2000
+
+# thinning interval
+mcmc.thin = 10
+
+# number of MCMC chains
+mcmc.chains = 4
+
+# how many poseterior samples will we have?
+data.frame(Each.Chain = (mcmc.chainLength - mcmc.burn) / mcmc.thin, All.Chains = (mcmc.chainLength - mcmc.burn) / mcmc.thin * mcmc.chains)
+
+
+#-----------------------------------------------------------------
+# set up model location and which parameters to track with JAGS
+#-----------------------------------------------------------------
+model.loc = 'LGR_TotalEscape_AllOrigins.txt'
+
+jags.params = c('X.tot.all', 'X.tot.day', 'X.tot.night', 'X.tot.reasc', 'X.tot.new.wild', 'X.tot.new.hatch', 'X.tot.new.hnc', 'X.sigma', 'true.prop', 'win.prop.avg', 'win.prop.true', 'win.prop.sigma', 'reasc.avg', 'reasc.true', 'reasc.sigma', 'acf', 'org.prop', 'org.sigma', 'trap.rate.true', 'r', 'k')
+
+# set initial values
+jags.inits = function(){
+  return(list('X.log.all' = with(jags.data, log(Y.window + 1)),
[…]

### 2020-08-18 inst/scripts/RunSims_TrapShutdown_WinErr_high.R

@@ -1,0 +1,265 @@
+# Author: Kevin See
+# Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model
+# Created: 8/31/2016
+# Last Modified: 9/12/2016
+# Notes: 
+
+#-----------------------------------------------------------------
+# these functions contains the scripts to run the SCOBI estimator
+source('SCOBI/SCOBIv2.R')
+source('SCOBI/SCSrank.R')
+
+# load needed packages
+library(Hmisc) # needed for SCOBI model
+library(MCMCpack)
+library(FSA)
+library(Rcapture)
+library(boot)
+library(msm)
+library(lubridate)
+library(magrittr)
+library(plyr)
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(jagsUI)
+
+theme_set(theme_bw())
+
+#setwd('LGR_SimulationStudy')
+
+# this contains the function to simulate data
+source('SimFnc.R')
+
+
+#-----------------------------------------------------------------
+# set mcmc parameters
+#-----------------------------------------------------------------
+# number of total samples
+mcmc.chainLength = 7000
+
+# number of burn-in samples
+mcmc.burn = 2000
+
+# thinning interval
+mcmc.thin = 10
+
+# number of MCMC chains
+mcmc.chains = 4
+
+# how many poseterior samples will we have?
+data.frame(Each.Chain = (mcmc.chainLength - mcmc.burn) / mcmc.thin, All.Chains = (mcmc.chainLength - mcmc.burn) / mcmc.thin * mcmc.chains)
+
+
+#-----------------------------------------------------------------
+# set up model location and which parameters to track with JAGS
+#-----------------------------------------------------------------
+model.loc = 'LGR_TotalEscape_AllOrigins.txt'
+
+jags.params = c('X.tot.all', 'X.tot.day', 'X.tot.night', 'X.tot.reasc', 'X.tot.new.wild', 'X.tot.new.hatch', 'X.tot.new.hnc', 'X.sigma', 'true.prop', 'win.prop.avg', 'win.prop.true', 'win.prop.sigma', 'reasc.avg', 'reasc.true', 'reasc.sigma', 'acf', 'org.prop', 'org.sigma', 'trap.rate.true', 'r', 'k')
+
+# set initial values
+jags.inits = function(){
+  return(list('X.log.all' = with(jags.data, log(Y.window + 1)),
[…]

### 2020-08-18 inst/scripts/RunSims_TrapShutdown_WinErr_low.R

@@ -1,0 +1,265 @@
+# Author: Kevin See
+# Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model
+# Created: 8/31/2016
+# Last Modified: 9/12/2016
+# Notes: 
+
+#-----------------------------------------------------------------
+# these functions contains the scripts to run the SCOBI estimator
+source('SCOBI/SCOBIv2.R')
+source('SCOBI/SCSrank.R')
+
+# load needed packages
+library(Hmisc) # needed for SCOBI model
+library(MCMCpack)
+library(FSA)
+library(Rcapture)
+library(boot)
+library(msm)
+library(lubridate)
+library(magrittr)
+library(plyr)
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(jagsUI)
+
+theme_set(theme_bw())
+
+#setwd('LGR_SimulationStudy')
+
+# this contains the function to simulate data
+source('SimFnc.R')
+
+
+#-----------------------------------------------------------------
+# set mcmc parameters
+#-----------------------------------------------------------------
+# number of total samples
+mcmc.chainLength = 7000
+
+# number of burn-in samples
+mcmc.burn = 2000
+
+# thinning interval
+mcmc.thin = 10
+
+# number of MCMC chains
+mcmc.chains = 4
+
+# how many poseterior samples will we have?
+data.frame(Each.Chain = (mcmc.chainLength - mcmc.burn) / mcmc.thin, All.Chains = (mcmc.chainLength - mcmc.burn) / mcmc.thin * mcmc.chains)
+
+
+#-----------------------------------------------------------------
+# set up model location and which parameters to track with JAGS
+#-----------------------------------------------------------------
+model.loc = 'LGR_TotalEscape_AllOrigins.txt'
+
+jags.params = c('X.tot.all', 'X.tot.day', 'X.tot.night', 'X.tot.reasc', 'X.tot.new.wild', 'X.tot.new.hatch', 'X.tot.new.hnc', 'X.sigma', 'true.prop', 'win.prop.avg', 'win.prop.true', 'win.prop.sigma', 'reasc.avg', 'reasc.true', 'reasc.sigma', 'acf', 'org.prop', 'org.sigma', 'trap.rate.true', 'r', 'k')
+
+# set initial values
+jags.inits = function(){
+  return(list('X.log.all' = with(jags.data, log(Y.window + 1)),
[…]

### 2020-08-18 inst/scripts/RunSims_Baseline.R
removed SCOBI model runs from sims

@@ -1,16 +1,11 @@
-# Author: Kevin See
+# Author: Ryan N. Kinzer
 # Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model
 # Created: 8/31/2016
 # Last Modified: 9/12/2016
-# Notes: 
-
-#-----------------------------------------------------------------
-# these functions contains the scripts to run the SCOBI estimator
-source('SCOBI/SCOBIv2.R')
-source('SCOBI/SCSrank.R')
+# Notes:
 
 # load needed packages
-library(Hmisc) # needed for SCOBI model
+library(Hmisc)
 library(MCMCpack)
 library(FSA)
 library(Rcapture)
@@ -82,43 +77,43 @@ tot_ptm = proc.time()
 i = 1
 while(i <= n_sim) {
   cat(paste('Starting simulation #', i, '\n'))
-  
+
   # simulate data
   my_sim = SimulateLGRdata(trap.rate.df = my_trap_rate)
-  
+
   # pull out data from simulation
   lgr_truth = my_sim$sim
-  
+
   # generate weekly observations
   # error rate is the CV of the daily window counts
   lgr_week = SimulateLGRobs(my_sim$parameters, lgr_truth, error_rate = 0.05, perfect.window = T)
-  
+
   # filter for spring/summer Chinook dates
   lgr_truth %<>%
     filter(Start_Date >= ymd('20150301'),
            Start_Date <= ymd('20150817'))
-  
+
   lgr_week$obs %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_week$by_origin %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, Origin, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_truth %<>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   #-------------#
   # ISEMP model #
   #-------------#
-  
+
   # pull data together for JAGS
   org_exist = lgr_week$obs %>% select(wild_fish, hatch_fish, HNC_fish) %>% colSums()
   org_exist = ifelse(org_exist > 0, 1, org_exist)
@@ -137,28 +132,28 @@ while(i <= n_sim) {
[…]

### 2020-08-18 inst/scripts/RunSims_Baseline_WinErr_high.R

@@ -1,16 +1,11 @@
-# Author: Kevin See
+# Author: Ryan N. Kinzer
 # Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model
 # Created: 8/31/2016
 # Last Modified: 9/12/2016
-# Notes: 
-
-#-----------------------------------------------------------------
-# these functions contains the scripts to run the SCOBI estimator
-source('SCOBI/SCOBIv2.R')
-source('SCOBI/SCSrank.R')
+# Notes:
 
 # load needed packages
-library(Hmisc) # needed for SCOBI model
+library(Hmisc)
 library(MCMCpack)
 library(FSA)
 library(Rcapture)
@@ -87,43 +82,43 @@ tot_ptm = proc.time()
 i = 1
 while(i <= n_sim) {
   cat(paste('Starting simulation #', i, '\n'))
-  
+
   # simulate data
   my_sim = SimulateLGRdata(trap.rate.df = my_trap_rate)
-  
+
   # pull out data from simulation
   lgr_truth = my_sim$sim
-  
+
   # generate weekly observations
   # error rate is the CV of the daily window counts
   lgr_week = SimulateLGRobs(my_sim$parameters, lgr_truth, error_rate = 0.1, perfect.window = F)
-  
+
   # filter for spring/summer Chinook dates
   lgr_truth %<>%
     filter(Start_Date >= ymd('20150301'),
            Start_Date <= ymd('20150817'))
-  
+
   lgr_week$obs %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_week$by_origin %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, Origin, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_truth %<>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   #-------------#
   # ISEMP model #
   #-------------#
-  
+
   # pull data together for JAGS
   org_exist = lgr_week$obs %>% select(wild_fish, hatch_fish, HNC_fish) %>% colSums()
   org_exist = ifelse(org_exist > 0, 1, org_exist)
@@ -142,39 +137,39 @@ while(i <= n_sim) {
[…]

### 2020-08-18 inst/scripts/RunSims_Baseline_WinErr_low.R

@@ -1,16 +1,11 @@
-# Author: Kevin See
+# Author: Ryan N. Kinzer
 # Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model
 # Created: 8/31/2016
 # Last Modified: 9/12/2016
-# Notes: 
-
-#-----------------------------------------------------------------
-# these functions contains the scripts to run the SCOBI estimator
-source('SCOBI/SCOBIv2.R')
-source('SCOBI/SCSrank.R')
+# Notes:
 
 # load needed packages
-library(Hmisc) # needed for SCOBI model
+library(Hmisc)
 library(MCMCpack)
 library(FSA)
 library(Rcapture)
@@ -87,43 +82,43 @@ tot_ptm = proc.time()
 i = 1
 while(i <= n_sim) {
   cat(paste('Starting simulation #', i, '\n'))
-  
+
   # simulate data
   my_sim = SimulateLGRdata(trap.rate.df = my_trap_rate)
-  
+
   # pull out data from simulation
   lgr_truth = my_sim$sim
-  
+
   # generate weekly observations
   # error rate is the CV of the daily window counts
   lgr_week = SimulateLGRobs(my_sim$parameters, lgr_truth, error_rate = 0.05, perfect.window = F)
-  
+
   # filter for spring/summer Chinook dates
   lgr_truth %<>%
     filter(Start_Date >= ymd('20150301'),
            Start_Date <= ymd('20150817'))
-  
+
   lgr_week$obs %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_week$by_origin %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, Origin, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_truth %<>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   #-------------#
   # ISEMP model #
   #-------------#
-  
+
   # pull data together for JAGS
   org_exist = lgr_week$obs %>% select(wild_fish, hatch_fish, HNC_fish) %>% colSums()
   org_exist = ifelse(org_exist > 0, 1, org_exist)
@@ -142,39 +137,39 @@ while(i <= n_sim) {
[…]

### 2020-08-18 inst/scripts/RunSims_NightReasc.R

@@ -1,16 +1,11 @@
-# Author: Kevin See
+# Author: Ryan N. Kinzer
 # Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model and TAC model. Night time passage and re-ascension rates are not equal
 # Created: 8/31/2016
 # Last Modified: 9/12/2016
-# Notes: 
-
-#-----------------------------------------------------------------
-# these functions contains the scripts to run the SCOBI estimator
-source('SCOBI/SCOBIv2.R')
-source('SCOBI/SCSrank.R')
+# Notes:
 
 # load needed packages
-library(Hmisc) # needed for SCOBI model
+library(Hmisc)
 library(MCMCpack)
 library(FSA)
 library(Rcapture)
@@ -87,45 +82,45 @@ tot_ptm = proc.time()
 i = 1
 while(i <= n_sim) {
   cat(paste('Starting simulation #', i, '\n'))
-  
+
   # simulate data
   my_sim = SimulateLGRdata(trap.rate.df = my_trap_rate,
                            fallback.rate = 0.1,
                            night.passage.rate = 0.05)
-  
+
   # pull out data from simulation
   lgr_truth = my_sim$sim
-  
+
   # generate weekly observations
   # error rate is the CV of the daily window counts
   lgr_week = SimulateLGRobs(my_sim$parameters, lgr_truth, error_rate = 0.05, perfect.window = T)
-  
+
   # filter for spring/summer Chinook dates
   lgr_truth %<>%
     filter(Start_Date >= ymd('20150301'),
            Start_Date <= ymd('20150817'))
-  
+
   lgr_week$obs %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_week$by_origin %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, Origin, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_truth %<>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   #-------------#
   # ISEMP model #
   #-------------#
-  
+
   # pull data together for JAGS
   org_exist = lgr_week$obs %>% select(wild_fish, hatch_fish, HNC_fish) %>% colSums()
[…]

### 2020-08-18 inst/scripts/RunSims_NightReasc_TrapShutdown.R

@@ -1,16 +1,11 @@
-# Author: Kevin See
+# Author: Ryan N. Kinzer
 # Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model and TAC model. Night time passage and re-ascension rates are not equal
 # Created: 8/31/2016
 # Last Modified: 9/12/2016
-# Notes: 
-
-#-----------------------------------------------------------------
-# these functions contains the scripts to run the SCOBI estimator
-source('SCOBI/SCOBIv2.R')
-source('SCOBI/SCSrank.R')
+# Notes:
 
 # load needed packages
-library(Hmisc) # needed for SCOBI model
+library(Hmisc)
 library(MCMCpack)
 library(FSA)
 library(Rcapture)
@@ -93,45 +88,45 @@ tot_ptm = proc.time()
 i = 1
 while(i <= n_sim) {
   cat(paste('Starting simulation #', i, '\n'))
-  
+
   # simulate data
   my_sim = SimulateLGRdata(trap.rate.df = my_trap_rate,
                            fallback.rate = 0.1,
                            night.passage.rate = 0.05)
-  
+
   # pull out data from simulation
   lgr_truth = my_sim$sim
-  
+
   # generate weekly observations
   # error rate is the CV of the daily window counts
   lgr_week = SimulateLGRobs(my_sim$parameters, lgr_truth, error_rate = 0.05, perfect.window = T)
-  
+
   # filter for spring/summer Chinook dates
   lgr_truth %<>%
     filter(Start_Date >= ymd('20150301'),
            Start_Date <= ymd('20150817'))
-  
+
   lgr_week$obs %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_week$by_origin %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, Origin, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_truth %<>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   #-------------#
   # ISEMP model #
   #-------------#
-  
+
   # pull data together for JAGS
   org_exist = lgr_week$obs %>% select(wild_fish, hatch_fish, HNC_fish) %>% colSums()
[…]

### 2020-08-18 inst/scripts/RunSims_NightReasc_TrapShutdown_WinErr_high.R

@@ -1,16 +1,11 @@
-# Author: Kevin See
+# Author: Ryan N. Kinzer
 # Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model and TAC model. Night time passage and re-ascension rates are not equal
 # Created: 8/31/2016
 # Last Modified: 9/12/2016
-# Notes: 
-
-#-----------------------------------------------------------------
-# these functions contains the scripts to run the SCOBI estimator
-source('SCOBI/SCOBIv2.R')
-source('SCOBI/SCSrank.R')
+# Notes:
 
 # load needed packages
-library(Hmisc) # needed for SCOBI model
+library(Hmisc)
 library(MCMCpack)
 library(FSA)
 library(Rcapture)
@@ -93,45 +88,45 @@ tot_ptm = proc.time()
 i = 1
 while(i <= n_sim) {
   cat(paste('Starting simulation #', i, '\n'))
-  
+
   # simulate data
   my_sim = SimulateLGRdata(trap.rate.df = my_trap_rate,
                            fallback.rate = 0.1,
                            night.passage.rate = 0.05)
-  
+
   # pull out data from simulation
   lgr_truth = my_sim$sim
-  
+
   # generate weekly observations
   # error rate is the CV of the daily window counts
   lgr_week = SimulateLGRobs(my_sim$parameters, lgr_truth, error_rate = 0.1, perfect.window = F)
-  
+
   # filter for spring/summer Chinook dates
   lgr_truth %<>%
     filter(Start_Date >= ymd('20150301'),
            Start_Date <= ymd('20150817'))
-  
+
   lgr_week$obs %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_week$by_origin %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, Origin, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_truth %<>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   #-------------#
   # ISEMP model #
   #-------------#
-  
+
   # pull data together for JAGS
   org_exist = lgr_week$obs %>% select(wild_fish, hatch_fish, HNC_fish) %>% colSums()
[…]

### 2020-08-18 inst/scripts/RunSims_NightReasc_TrapShutdown_WinErr_low.R

@@ -1,16 +1,11 @@
-# Author: Kevin See
+# Author: Ryan N. Kinzer
 # Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model and TAC model. Night time passage and re-ascension rates are not equal
 # Created: 8/31/2016
 # Last Modified: 9/12/2016
-# Notes: 
-
-#-----------------------------------------------------------------
-# these functions contains the scripts to run the SCOBI estimator
-source('SCOBI/SCOBIv2.R')
-source('SCOBI/SCSrank.R')
+# Notes:
 
 # load needed packages
-library(Hmisc) # needed for SCOBI model
+library(Hmisc)
 library(MCMCpack)
 library(FSA)
 library(Rcapture)
@@ -93,45 +88,45 @@ tot_ptm = proc.time()
 i = 1
 while(i <= n_sim) {
   cat(paste('Starting simulation #', i, '\n'))
-  
+
   # simulate data
   my_sim = SimulateLGRdata(trap.rate.df = my_trap_rate,
                            fallback.rate = 0.1,
                            night.passage.rate = 0.05)
-  
+
   # pull out data from simulation
   lgr_truth = my_sim$sim
-  
+
   # generate weekly observations
   # error rate is the CV of the daily window counts
   lgr_week = SimulateLGRobs(my_sim$parameters, lgr_truth, error_rate = 0.05, perfect.window = F)
-  
+
   # filter for spring/summer Chinook dates
   lgr_truth %<>%
     filter(Start_Date >= ymd('20150301'),
            Start_Date <= ymd('20150817'))
-  
+
   lgr_week$obs %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_week$by_origin %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, Origin, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_truth %<>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   #-------------#
   # ISEMP model #
   #-------------#
-  
+
   # pull data together for JAGS
   org_exist = lgr_week$obs %>% select(wild_fish, hatch_fish, HNC_fish) %>% colSums()
[…]

### 2020-08-18 inst/scripts/RunSims_NightReasc_WinErr_high.R

@@ -1,16 +1,11 @@
-# Author: Kevin See
+# Author: Ryan N. Kinzer
 # Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model and TAC model. Night time passage and re-ascension rates are not equal
 # Created: 8/31/2016
 # Last Modified: 9/12/2016
-# Notes: 
-
-#-----------------------------------------------------------------
-# these functions contains the scripts to run the SCOBI estimator
-source('SCOBI/SCOBIv2.R')
-source('SCOBI/SCSrank.R')
+# Notes:
 
 # load needed packages
-library(Hmisc) # needed for SCOBI model
+library(Hmisc)
 library(MCMCpack)
 library(FSA)
 library(Rcapture)
@@ -87,45 +82,45 @@ tot_ptm = proc.time()
 i = 1
 while(i <= n_sim) {
   cat(paste('Starting simulation #', i, '\n'))
-  
+
   # simulate data
   my_sim = SimulateLGRdata(trap.rate.df = my_trap_rate,
                            fallback.rate = 0.1,
                            night.passage.rate = 0.05)
-  
+
   # pull out data from simulation
   lgr_truth = my_sim$sim
-  
+
   # generate weekly observations
   # error rate is the CV of the daily window counts
   lgr_week = SimulateLGRobs(my_sim$parameters, lgr_truth, error_rate = 0.1, perfect.window = F)
-  
+
   # filter for spring/summer Chinook dates
   lgr_truth %<>%
     filter(Start_Date >= ymd('20150301'),
            Start_Date <= ymd('20150817'))
-  
+
   lgr_week$obs %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_week$by_origin %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, Origin, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_truth %<>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   #-------------#
   # ISEMP model #
   #-------------#
-  
+
   # pull data together for JAGS
   org_exist = lgr_week$obs %>% select(wild_fish, hatch_fish, HNC_fish) %>% colSums()
[…]

### 2020-08-18 inst/scripts/RunSims_NightReasc_WinErr_low.R

@@ -1,16 +1,11 @@
-# Author: Kevin See
+# Author: Ryan N. Kinzer
 # Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model and TAC model. Night time passage and re-ascension rates are not equal
 # Created: 8/31/2016
 # Last Modified: 9/12/2016
-# Notes: 
-
-#-----------------------------------------------------------------
-# these functions contains the scripts to run the SCOBI estimator
-source('SCOBI/SCOBIv2.R')
-source('SCOBI/SCSrank.R')
+# Notes:
 
 # load needed packages
-library(Hmisc) # needed for SCOBI model
+library(Hmisc)
 library(MCMCpack)
 library(FSA)
 library(Rcapture)
@@ -87,45 +82,45 @@ tot_ptm = proc.time()
 i = 1
 while(i <= n_sim) {
   cat(paste('Starting simulation #', i, '\n'))
-  
+
   # simulate data
   my_sim = SimulateLGRdata(trap.rate.df = my_trap_rate,
                            fallback.rate = 0.1,
                            night.passage.rate = 0.05)
-  
+
   # pull out data from simulation
   lgr_truth = my_sim$sim
-  
+
   # generate weekly observations
   # error rate is the CV of the daily window counts
   lgr_week = SimulateLGRobs(my_sim$parameters, lgr_truth, error_rate = 0.05, perfect.window = F)
-  
+
   # filter for spring/summer Chinook dates
   lgr_truth %<>%
     filter(Start_Date >= ymd('20150301'),
            Start_Date <= ymd('20150817'))
-  
+
   lgr_week$obs %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_week$by_origin %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, Origin, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_truth %<>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   #-------------#
   # ISEMP model #
   #-------------#
-  
+
   # pull data together for JAGS
   org_exist = lgr_week$obs %>% select(wild_fish, hatch_fish, HNC_fish) %>% colSums()
[…]

### 2020-08-18 inst/scripts/RunSims_TrapShutdown.R

@@ -1,16 +1,11 @@
-# Author: Kevin See
+# Author: Ryan N. Kinzer
 # Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model
 # Created: 8/31/2016
 # Last Modified: 9/12/2016
-# Notes: 
-
-#-----------------------------------------------------------------
-# these functions contains the scripts to run the SCOBI estimator
-source('SCOBI/SCOBIv2.R')
-source('SCOBI/SCSrank.R')
+# Notes:
 
 # load needed packages
-library(Hmisc) # needed for SCOBI model
+library(Hmisc)
 library(MCMCpack)
 library(FSA)
 library(Rcapture)
@@ -94,43 +89,43 @@ tot_ptm = proc.time()
 i = 1
 while(i <= n_sim) {
   cat(paste('Starting simulation #', i, '\n'))
-  
+
   # simulate data
   my_sim = SimulateLGRdata(trap.rate.df = my_trap_rate)
-  
+
   # pull out data from simulation
   lgr_truth = my_sim$sim
-  
+
   # generate weekly observations
   # error rate is the CV of the daily window counts
   lgr_week = SimulateLGRobs(my_sim$parameters, lgr_truth, error_rate = 0.05, perfect.window = T)
-  
+
   # filter for spring/summer Chinook dates
   lgr_truth %<>%
     filter(Start_Date >= ymd('20150301'),
            Start_Date <= ymd('20150817'))
-  
+
   lgr_week$obs %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_week$by_origin %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, Origin, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_truth %<>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   #-------------#
   # ISEMP model #
   #-------------#
-  
+
   # pull data together for JAGS
   org_exist = lgr_week$obs %>% select(wild_fish, hatch_fish, HNC_fish) %>% colSums()
   org_exist = ifelse(org_exist > 0, 1, org_exist)
@@ -149,39 +144,39 @@ while(i <= n_sim) {
[…]

### 2020-08-18 inst/scripts/RunSims_TrapShutdown_WinErr_high.R

@@ -1,16 +1,11 @@
-# Author: Kevin See
+# Author: Ryan N. Kinzer
 # Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model
 # Created: 8/31/2016
 # Last Modified: 9/12/2016
-# Notes: 
-
-#-----------------------------------------------------------------
-# these functions contains the scripts to run the SCOBI estimator
-source('SCOBI/SCOBIv2.R')
-source('SCOBI/SCSrank.R')
+# Notes:
 
 # load needed packages
-library(Hmisc) # needed for SCOBI model
+library(Hmisc)
 library(MCMCpack)
 library(FSA)
 library(Rcapture)
@@ -94,43 +89,43 @@ tot_ptm = proc.time()
 i = 1
 while(i <= n_sim) {
   cat(paste('Starting simulation #', i, '\n'))
-  
+
   # simulate data
   my_sim = SimulateLGRdata(trap.rate.df = my_trap_rate)
-  
+
   # pull out data from simulation
   lgr_truth = my_sim$sim
-  
+
   # generate weekly observations
   # error rate is the CV of the daily window counts
   lgr_week = SimulateLGRobs(my_sim$parameters, lgr_truth, error_rate = 0.1, perfect.window = F)
-  
+
   # filter for spring/summer Chinook dates
   lgr_truth %<>%
     filter(Start_Date >= ymd('20150301'),
            Start_Date <= ymd('20150817'))
-  
+
   lgr_week$obs %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_week$by_origin %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, Origin, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_truth %<>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   #-------------#
   # ISEMP model #
   #-------------#
-  
+
   # pull data together for JAGS
   org_exist = lgr_week$obs %>% select(wild_fish, hatch_fish, HNC_fish) %>% colSums()
   org_exist = ifelse(org_exist > 0, 1, org_exist)
@@ -149,39 +144,39 @@ while(i <= n_sim) {
[…]

### 2020-08-18 inst/scripts/RunSims_TrapShutdown_WinErr_low.R

@@ -1,16 +1,11 @@
-# Author: Kevin See
+# Author: Ryan N. Kinzer
 # Purpose: Simulate a number of LGR-like datasets, and estimate escapement using ISEMP model
 # Created: 8/31/2016
 # Last Modified: 9/12/2016
-# Notes: 
-
-#-----------------------------------------------------------------
-# these functions contains the scripts to run the SCOBI estimator
-source('SCOBI/SCOBIv2.R')
-source('SCOBI/SCSrank.R')
+# Notes:
 
 # load needed packages
-library(Hmisc) # needed for SCOBI model
+library(Hmisc)
 library(MCMCpack)
 library(FSA)
 library(Rcapture)
@@ -94,43 +89,43 @@ tot_ptm = proc.time()
 i = 1
 while(i <= n_sim) {
   cat(paste('Starting simulation #', i, '\n'))
-  
+
   # simulate data
   my_sim = SimulateLGRdata(trap.rate.df = my_trap_rate)
-  
+
   # pull out data from simulation
   lgr_truth = my_sim$sim
-  
+
   # generate weekly observations
   # error rate is the CV of the daily window counts
   lgr_week = SimulateLGRobs(my_sim$parameters, lgr_truth, error_rate = 0.05, perfect.window = F)
-  
+
   # filter for spring/summer Chinook dates
   lgr_truth %<>%
     filter(Start_Date >= ymd('20150301'),
            Start_Date <= ymd('20150817'))
-  
+
   lgr_week$obs %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_week$by_origin %<>%
     inner_join(lgr_truth %>%
                  select(Week, Start_Date) %>%
                  distinct()) %>%
     select(Start_Date, Week, Origin, everything()) %>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   lgr_truth %<>%
     mutate(Week = Week - min(Week) + 1)
-  
+
   #-------------#
   # ISEMP model #
   #-------------#
-  
+
   # pull data together for JAGS
   org_exist = lgr_week$obs %>% select(wild_fish, hatch_fish, HNC_fish) %>% colSums()
   org_exist = ifelse(org_exist > 0, 1, org_exist)
@@ -149,39 +144,39 @@ while(i <= n_sim) {
[…]

### 2020-08-18 analysis/paper/appendices/Appendix_STADEM_descrip.Rmd
compiled word and html docs

@@
 
-Escapement at LGD is estimated by combining two independent observations, trap catches and window counts, of the true number of fish crossing the dam in a state-space model with a weekly time-step.
+Escapement at Lower Granite Dam (LGD) is estimated by combining two independent observations, trap catches and window counts, of the true number of fish crossing the dam in a state-space model [@royle2008hierarchical] with a weekly time-step.
 Both are assumed to be corrupted observations of the true unknown number of fish crossing LGD each week.
@@
 In the formula below, $p_t$ is the proportion of fish observed at the window and $r$ is the shape parameter.
-If $r$ is estimated to be small provides evidence for over-dispersion, and as it grows very large, the negative binomial distribution behaves like a Poisson distribution.
+If $r$ is estimated to be small it provides evidence for over-dispersion, and as it grows very large, the negative binomial distribution behaves like a Poisson distribution.
 

### 2020-08-18 analysis/paper/appendices/Appendix_simulation_details.Rmd
compiled word and html docs

@@
 
-Simulation parameters for model evaluations were set to mimic typical ecapement of spring/summer Chinook to LGD with similar origin proportions, marking rates and run timing as those observed from return years 2010 - 2015.
+Simulation parameters for model evaluations were set to mimic typical ecapement of spring/summer Chinook Salmon to LGD with similar origin proportions, marking rates and run timing as those observed from return years 2010 - 2015.
 Escapement of each origin ($N_j$) was set at 25,000 wild, 70,000 hatchery and 5,000 hatchery no-clips spread randomly across 25 populations ($n$).

### 2020-08-18 analysis/paper/paper.Rmd
compiled word and html docs

@@
       - --lua-filter=../templates/pagebreak.lua
+    bookdown::html_document2:
+      fig_caption: yes
+      number_sections: FALSE
+      toc: no
+      fig_width: 7
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
     bookdown::pdf_document2:
@@
         in_header: ../templates/header_manuscript.tex
-    bookdown::html_document2:
-      fig_caption: yes
-      number_sections: FALSE
-      toc: no
-      fig_width: 7
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
 fontsize: 12pt
@@
 For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their spawning area.
-Escapement estimates facilitate effective fisheries management, and particularly, estimates of escapement for each component group (e.g., stock, population, age, origin [wild, hatchery]; @hess2014monitoring; @steinhorst2017abundance) provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations.
+Escapement estimates facilitate effective fisheries management, and particularly, estimates of escapement for each component group (e.g., stock, population, age, origin [wild, hatchery]; @hess2014monitoring; @Steinhorst2017) provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations.
 Accurate escapement estimates are increasingly important for depleted populations as they inform viable salmonid metrics [@mcelhany2000viable], and facilitate assessments of population viability and extinction risk [@nwfsc2015; @williams2016viability] and climate change vulerability [@crozier2019climate].

### 2020-08-18 analysis/paper/paper.tex
compiled word and html docs

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
  \setmainfont[]{Times New Roman}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources},
  pdfauthor={Kevin E. See1,, Ryan N. Kinzer2, and Mike W. Ackerman1},
  pdfkeywords={escapement; Chinook; steelhead; Lower Granite Dam; state-space model; Bayesian model},
  pdfborder={0 0 0},
  breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-2}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

% double spacing
\usepackage{setspace}
\doublespacing
% to double space figure captions
\usepackage{caption}
\captionsetup{font = doublespacing}

% line numbers
\usepackage{lineno}
\linenumbers
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\title{State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources}
\author{Kevin E. See\textsuperscript{1,*}, Ryan N. Kinzer\textsuperscript{2}, and Mike W. Ackerman\textsuperscript{1}}
\date{August 18, 2020}

\begin{document}
\maketitle
\begin{abstract}
This is the abstract
\end{abstract}

\textsuperscript{1} Biomark, Inc.~705 South 8th St., Boise, Idaho, 83702, USA\\
\textsuperscript{2} Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA

\textsuperscript{*} Correspondence: \href{mailto:Kevin.See@merck.com}{Kevin E. See \textless{}\href{mailto:Kevin.See@merck.com}{\nolinkurl{Kevin.See@merck.com}}\textgreater{}}

\textbf{Possible journals:}

First Choice:

\begin{itemize}
\tightlist
\item
  \textbf{\href{https://afspubs.onlinelibrary.wiley.com/journal/15488675}{North American Journal of Fisheries Management}}
\end{itemize}

Second Choice:

\begin{itemize}
\tightlist
\item
  \href{https://afspubs.onlinelibrary.wiley.com/journal/15488659}{Transactions of the American Fisheries Society}
\end{itemize}

Others:

\begin{itemize}
\item
  \href{http://www.nrcresearchpress.com/page/cjfas/editors}{Canadian Journal of Fisheries and Aquatic Sciences}
\item
  \href{https://www.journals.elsevier.com/fisheries-research}{Fisheries Research}
\item
  \href{http://fishbull.noaa.gov/}{Fishery Bulletin}
\item
  \href{http://onlinelibrary.wiley.com/journal/10.1111/(ISSN)1467-2979/homepage/ProductInformation.html}{Fish and Fisheries}
\end{itemize}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Fish escapement is the number of adults that survive juvenile and adult rearing, escape harvest and return to their natal habitat to potentially spawn (e.g.~Bue et al. (1998)). For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their spawning area. Escapement estimates facilitate effective fisheries management, and particularly, estimates of escapement for each component group (e.g., stock, population, age, origin {[}wild, hatchery{]}; Hess et al. (2014); Steinhorst et al. (2017)) provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations. Accurate escapement estimates are increasingly important for depleted populations as they inform viable salmonid metrics (McElhany et al. 2000), and facilitate assessments of population viability and extinction risk (Ford et al. 2015; Williams et al. 2016) and climate change vulerability (Crozier et al. 2019).

Populations of Chinook Salmon Oncorhynchus tshawytscha and steelhead trout O. mykiss in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric dams on the Snake and Columbia rivers) and tributary habitats (Nehlsen et al. 1991; McClure et al. 2003; Ford et al. 2015). As a result, Snake River spring-summer run Chinook Salmon (hereafter Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937). Snake River Chinook Salmon and steelhead have substantial recreational, commercial, cultural and subsistence value both within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries. The aggregate escapement of Snake River Chinook Salmon and steelhead populations, with the exception of Tucannon River, is measured at Lower Granite Dam located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations. Many fisheries management and conservation actions are made based on estimates of escapement at Lower Granite Dam parsed by species and origin (\textbf{Snake Basin Harvest Plans?}). Additionally, harvest openings and closures, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia River fisheries, are predicated on escapements at Lower Granite Dam (\textbf{could cite US v OR Management Agreement and Snake Basin Harvest Plans?}).

Chinook Salmon and steelhead returning to a majority of populations in the Snake River must ascend a fish ladder on Lower Granite Dam before migrating to their natal tributary spawning locations. Fish management agencies previously used counts of fish, by species, passing an observation window located on the side of the fish ladder as a census of fish escaping past the dam. For both species, total escapement was then parsed into groups (e.g., wild, hatchery) using observed marks and genetic data from a sample of fish captured at an adult trap located on the fish ladder upstream of the observation window (Camacho et al. 2017 p. @Steinhorst2017). Window counts as a census proved beneficial as being an easy, straight-forward method that was ascertained in near real-time. Moreover, downriver fisheries management arenas have used window counts at lower-river dams as escapement estimates for the past several years, and consistency in methods is often desirable for management decision making (\textbf{citation?}).

Using window counts, however, as a census of Chinook Salmon and steelhead passing Lower Granite Dam can be problematic as it avoids multiple sources of uncertainty and ignores known biological processes. First, live (i.e., in-person) window counts only occur from April through October each year. Additionally, live window counts only occur for 16-hours each day, and fish counters working at the observation window look directly into the fish ladders to identify and count all passing fish, by species, for only 50 minutes of each hour. Counts are then expanded to provide an estimate for the hour (USACE 2015). From November through March, the remainder of the year, video tape fish counting is used and only occurs for 10 hours each day; fish counters then read the video tapes and submit daily fish counts (Hatch et al. 1994). Typically, the observational error rates of live and video window counts are unknown, and sampling error rates are ignored. Additionally, two biological processes are unaccounted for: 1) fish that cross the dam during the 8-hours when the window is closed for counting (i.e., nighttime passage) which may result in an underestimate of escapement, and 2) fish that cross the dam multiple times (i.e., re-ascension), and therefore potentially double-counted. Finally, Chinook Salmon and steelhead that migrate through the ladder and past the dam may fallback over the dam (e.g., over spillway, through navigation locks), and later, may or may not re-ascend the fish ladder again (Boggs et al. 2004). Both fallback with no re-ascension and fallback with re-ascension potentially result in an overestimate of escapement (Dauble and Mueller 2000). Previously, it was assumed that nighttime passage rates and fallback/re-ascension rates cancelled each other out resulting in window counts providing an unbiased estimate of escapement (\textbf{citation?})

In this study, we describe a novel method to estimate aggregate and group escapement which incorporates all sources of known uncertainty, we demonstrate the estimation method is essentially unbiased, thus improving effective conservation and management decision making, and show night-time passage and fallback/re-ascension rates are unequal. Our method for estimating escapement, by species, past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of PIT tagged fish in the adult ladder to explicitly model nighttime passage, re-ascension, and error from both window and trap estimates using a state-space approach (Royle and Dorazio 2008). To meet desired management and conservation objectives, modeled escapement includes estimates of uncertainty and is parsed into weekly strata. Further, total and weekly estimates are parsed into three origin groups: wild, hatchery, and hatchery no-clip. Estimates of escapement account for fish that migrate through the ladder at night outside of observation hours and account for fish that may ascend the ladder multiple times due to fallback and re-ascension. Our model is implemented in the \textbf{ST}ate-space \textbf{A}dult \textbf{D}am \textbf{E}scapement \textbf{M}odel (STADEM) package for the statistical software R (R Core Team 2020), and is available for downloaded at \url{https://github.com/KevinSee/STADEM}. To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates. The STADEM model combines multiple imperfect sources of data to reduce bias in escapement estimates and provide more reasonable estimates of uncertainty. Additionally, the STADEM model framework can be applied elsewhere to estimate aggregate and group escapement, provided a passage barrier containing multiple counting and observational mechanisms exist, such as, a counting window, an adult trap sampling a portion of the run, and PIT tag antenna infrastructure.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{data-requirements}{%
\subsection{Data Requirements}\label{data-requirements}}

We used STADEM and three sources of data to estimate Chinook Salmon and steelhead escapement at Lower Granite Dam from 2010-2019. Data sources included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of previously PIT tagged fish detected in the adult fish ladder. Following, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; however, similar data could likely be obtained from other fish passage facilities.

\hypertarget{window-counts}{%
\subsubsection{Window Counts}\label{window-counts}}

Daily counts of adult Chinook Salmon and steelhead passing an observation window located on the Lower Granite fish ladder were the first source of data. Daily counts were made and provided by the US Army Corps of Engineers, and when summed, provide an estimate of the number of fish ascending and passing (i.e., escapement) Lower Granite Dam each season. Window counts were made for each species using video monitoring and direct in-person visual monitoring during daytime hours (Hatch et al. 1994). Video monitoring occured during the beginning and tail ends of the adult runs (March 1 -- March 31 and November 1 -- December) for 10 hours per day (0600 -- 1600 hours). Direct visual monitoring occured during peak run times (April 1 -- October 31) for 16 hours per day (0400 -- 2000 hours) (USACE 2015). During direct visual monitoring, observers recorded each adult (≥ 30cm), by species, passing the window for 50 minutes of each hour of operation. Salmonids under 30cm in length were not identified to species. The sum of the daily 50-minute counts were then multiplied by 1.2 to account for the 10 minutes when fish were not counted. Daytime window counts were not expanded for fish that may have ascended the ladder outside of operational hours (i.e., nighttime) (USACE 2015). Window counts were accessed through the Columbia Basin Research Data Access in Real Time (DART) website, \url{www.cbr.washington.edu/dart/query/adult_daily}, using their window count query. Counts were provided for each day the fish ladder was open to passage. Although window counts were assumed to be a census of every fish passing Lower Granite Dam, corrections were not applied for nighttime passage or re-ascending fish. Further, there was no estimation of daily or seasonal observation or sampling error.

\hypertarget{adult-fish-trap-data}{%
\subsubsection{Adult Fish Trap Data}\label{adult-fish-trap-data}}

The second source of data came from a sample of fish collected in the adult trap as they migrated past Lower Granite Dam (Ogden 2016a). The trap, also located within the adult fish ladder and upstream of the observation window, provided biological data (e.g., origin {[}wild, hatchery{]}, genetic stock, length, age, sex) for captured adults that allowed decomposition of the escapement into specific groups (e.g., Camacho et al. (2019), Steinhorst et al. (2017)). The trap was operational for 24 hours per day, and randomly sampled the daily run by opening four times per hour for a length of time determined by a set daily trapping rate. The trap rate was determined by a committee of collaborating management agencies with a goal of capturing a target number of wild fish, but also balancing fish handling concerns. Trap sample rates were typically 10-25\%, but fluctuated throughout the season due to high water temperatures, decreased flows, trap malfunctions and/or closures, fish handling logistics, etc. All captured fish were anesthetized, identified to species, examined for existing marks/tags, and measured for fork length. Additionally, each fish had scale and genetic tissue samples taken; scale samples were used to estimate age (Wright et al. 2015), tissue samples were used to determine sex and for genetic stock identification analysis (e.g., Hargrove et al. (2019)). Prior to release, all non-PIT tagged fish with an intact adipose fin (i.e., putatively wild) received a PIT tag. Final determination of wild, hatchery, and hatchery no-clip origins were assigned using a post-hoc analysis of marks, tags, and genetic information. Data from the adult trap were collected and managed by multiple agencies and were made available by the Idaho Department of Fish and Game (Camacho et al. 2019).

\hypertarget{pit-tag-data}{%
\subsubsection{PIT tag Data}\label{pit-tag-data}}

The last source of data was observations of PIT tagged adult Chinook Salmon and steelhead at detection sites located in the Lower Granite Dam fish ladder. These observations provide estimates of 1) a trapping rate, 2) the proportion of fish passing during nighttime hours (i.e., outside of window observation hours), and 3) the proportion of fish that ascend the fish ladder multiple times (i.e., re-ascension rate). Detections used in the model include all fish that were previously PIT tagged as juveniles or adults prior to reaching Lower Granite Dam (i.e., does not include newly tagged adults at the dam) and detected at adult detection sites in the dam passage system. PIT tag data was provided through DART and the adult ladder PIT tag query; \url{http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail}.

A trap rate estimate was derived using mark-recapture methods and PIT tag observations of both Chinook Salmon and steelhead at Lower Granite Dam adult detection sites. The ``mark'' group included all tags detected in the adult trap and the ``capture'' group included tags observed to cross the weir at the upstream end of the fish ladder as adults left the passage system. Using a mark-recapture model with differing capture probabilities, we estimated the trap rate on a weekly basis. Those estimates, with associated uncertainties, were then incorporated into the model as informed priors, while the model estimated the true trap rate based on all the data, including trap and window counts.

The nighttime passage rate was estimated as the count of tags that migrated through the fish ladder during non-window observation hours divided by the total number of tags passing the fish ladder, and was calculated on a weekly basis. The re-ascension rate was the count of tags observed passing the upstream most detection sites in the adult fish ladder (i.e., passing the dam) and later detected re-entering the downstream end of the fish ladder at a later time divided by the total number of tags leaving the fish ladder. Previously, we examined for differences in nighttime passage and re-ascension rates as estimated using wild fish only, versus combining hatchery and wild fish together, and found no difference. Therefore, we combine wild and hatchery PIT tagged fish observations to estimate common nighttime passage and re-ascension rates to increase sample sizes.

\hypertarget{model-framework}{%
\subsection{Model Framework}\label{model-framework}}

We estimated the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for nighttime passage, and fallback/re-ascension rates using a state-space modeling approach (Royle and Dorazio 2008) implemented in the STADEM package for the R statistical software (R Core Team 2020). We assumed that the window counts and the estimates from the trap (fish in the trap divided by trap rate that week) were generated by processes with observation error. In the case of the trap for example, we assumed there was sampling variation and uncertainty around our estimates of the true unknown trap rate. STADEM further accounted for the proportion of fish that ascended the ladder while the counting window was closed (i.e., night), as well as for fish potentially double-counted (or more) after falling back below the dam and later re-ascending the fish ladder. Last, adult sampling data from the trap (wild, hatchery, hatchery no-clip) were used to partition the total escapement estimate by origin (Figure \ref{fig:stadem-examp-fig}. Additional model details can be found in \protect\hyperlink{append1}{Appendix A}. The STADEM package is available from the primary author at a \url{https://github.com/KevinSee/STADEM}, and requires the use of the JAGS software (Plummer 2019) for Bayesian inference.

\hypertarget{simulations}{%
\subsection{Simulations}\label{simulations}}

We tested the STADEM model on a variety of simulated data sets. These simulated data sets contained a fixed number of unique adult fish of known origin crossing a dam, from a total of 25 fictional populations with differential run-timing (i.e., date of passage at Lower Granite Dam). Each simulated fish was given a date of ladder ascension, based on its population and the range of run timing for that population. Each fish was also simulated to cross the dam either while the window was open for counting, or not, and was given the chance to be ``caught'' in the simulated fish trap given the week when it ascended the dam, and the known trap rate that week. Fallback and re-ascension behavior was also simulated, with each fish having the possibility of falling back and re-ascending the ladder up to three times.

Our objective was to examine STADEM model estimates of origin-specific (wild, hatchery, hatchery no-clip) escapement from the combinations of two separate trapping rates, two fallback/re-ascension and nighttime passage combinations, and three window count error rates; resulting in twelve different scenarios (Table \ref{tab:lgr-sim-tab}). The simulation parameters such as proportion of origin, run-timing, nighttime passage rates, fallback and re-ascension rates and trap rates were based on observed values at Lower Granite Dam between 2010-2015. Further details about simulation procedures can be found in \protect\hyperlink{append2}{Appendix B}.

\hypertarget{lower-granite-application}{%
\subsection{Lower Granite Application}\label{lower-granite-application}}

Finally, we applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead returning to the Snake River to spawn during 2010 to 2019. Window counts for both species were accessed from DART via functions within STADEM. For Chinook Salmon, a spawn year refers to adults that migrate past the dam prior to August 17 each year and spawn that late summer and fall. For steelhead, the spawn year is defined as steelhead that migrate past Lower Granite Dam starting July 1 the previous year and prior to June 30 of the given year (e.g., spawn year 2017 steelhead migrate past Lower Granite between July 1, 2016 and June 30, 2017). Data from the adult trap was made available by Idaho Department of Fish and Game, and adult PIT tag detection data within the fish passage ladder at Lower Granite Dam was accessed from the PTAGIS regional database (\url{https://www.ptagis.org/}).

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{simulations-1}{%
\subsection{Simulations}\label{simulations-1}}

Simulation results of observed bias, sampling variation, precision, root mean squared deviation (RMSD), and coverage probabilities were qualitatively similar for hatchery (N = 70,000) and hatchery no-clip (N = 5,000) origins as observed in the wild origin (N = 25,000) comparisons. As such, only diagnostic measures of Lower Granite Dam model fits to a medium sized escapement level (e.g., wild origin escapement) are presented.

STADEM results were very similar across all scenarios (Figure \ref{fig:rel-bias-lgd}). Estimates were unbiased, with an average relative bias of 0.2--0.3\%. The CV of the estimates average 2.0--3.0\%, with coverage probabilities that always exceeded 95\%. We calculated RMSD as the square root of the sum of the variance of the estimate and the squared expected bias, which accounts for the size of the uncertainty in the estimator as well as its bias. The RMSD was near 500 for each scenario, representing an estimate within 2\% of the true value (Table \ref{tab:summ-tab}).

\hypertarget{lower-granite-application-1}{%
\subsection{Lower Granite Application}\label{lower-granite-application-1}}

We applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead for spawn years 2010--2019. Estimates of total escapement, as well as estimates of wild, hatchery, and hatchery no-clip estimates are presented in Table \ref{tab:lgr-tab}. Coefficients of variation ranged from 3.5-7.5\% for wild fish, 3.0-6.2\% for hatchery fish, 4.0-8.4\% for hatchery no-clip fish and 2.9-5.6\% for total unique fish past Lower Granite Dam.

Weekly estimates of total escapement over Lower Granite Dam tracked the window counts and trap estimates (Figure \ref{fig:jags-fit-fig}). STADEM point estimates were often between estimates based on window counts, and those based on the number of fish caught in the adult trap. However, when very few fish were caught in the trap, or there was more uncertainty about the trap rate that week, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \ref{fig:jags-fit-fig}. That year shows the utility of STADEM in dealing with missing data, as the trap was shut down for several weeks in July and August.

Examining the estimates of nighttime passage and re-ascension rates based on the observed PIT tags crossing over Lower Granite Dam, the two rates did not match up (Figure \ref{fig:night-reasc-diff-fig}). In particular, there are several weeks when the window counts are quite high, and the rates differ by as much as 10\%. Clearly, those two biological processes do not cancel each other out, and thus, employing a model that accounts for both will result in more accurate estimates of escapement.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

We have presented a novel method for estimating adult salmonid escapement past a large hydroelectric facility (e.g., Lower Granite Dam) that incorporates data from window counts, fish trap, and observations of PIT tagged fish in the adult passage ladder. Our model explicitly models nighttime passage, re-ascension, and potential error in both window and trap estimates. In doing so, we demonstrated that at Lower Granite Dam, nighttime passage and re-ascension rates do not always offset each other, and assuming they do, will lead to biased escapement estimates in some years. With minor adjustments this modelling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall Chinook Salmon, Pacific lamprey Lampetra tridentata), or elsewhere; provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run, and a tag observation or detection infrastructure (e.g., a PIT tag detection array or similar) exists. Our state-space model combined multiple imperfect sources of data to reduce bias in adult escapement estimates and provided more reasonable estimates of uncertainty. Accurate population or stock abundance estimates and uncertainty accounting for observation and process error can be particularly important when estimates are used or leveraged for management and conservation decisions such as population viability analyses.

Combining data from the adult fish trap with live and video window counts provides several benefits. First, it allows us to model observer error in the window counts, which is typically unknown. If estimates rely on window counts alone, quantifying observer error is impossible, and we believe it's prudent to capture and account for known sources of error to minimize management decision risk. Second, by incorporating both sources of information in a state-space framework, STADEM incorporates missing data at either the observation window or adult trap seamlessly. At Lower Granite Dam, the adult trap has been closed for brief or extended periods of time (i.e., days, weeks) intermittently over the past several years, often during peak run times. Trap closures are typically associated with elevated water temperatures resulting in potential fish handling stress and/or trap malfunctions. Given predicted Pacific Northwest climate change scenarios trap closures from high water temperatures may become more commonplace in the future; amplifying the need for a modeling framework that accounts for periods of missing data while still capturing estimate uncertainty. Additionally, having a framework in place that accounts for missing periods of data will allow for increased logistic flexibility if, for example, maintenance or construction is needed at the observational window or adult trap.

Although not currently set up for this, STADEM could be modified and run on a weekly basis or in near real-time to provide better in-season estimates for fisheries managers. Currently, the only roadblock to this, at Lower Granite Dam, is the identification of hatchery origin fish from phenotypically wild fish (i.e., hatchery no-clip) using genetic tissues samples (\textbf{\emph{Steele et al.~2013}}) collected at the adult trap, which currently is completed post hoc after the trapping season. The inclusion of genetic information typically results in a reduction in wild escapement estimates and an associated increase in hatchery no-clip escapement. However, if in-season management decisions do not require this correction or could accept the potential bias, origin calls at the trap could be used in-season as a first approximation to escapement. Final post-hoc estimates, parsed by origin, could then be finalized at season's end. All other data included in this model (e.g., window counts and PIT observations) are otherwise provided in almost real-time by DART. Provided the Lower Granite Dam adult fish trap database was updated and made available frequently, there are minimal obstacles for adapting the STADEM framework to provide in-season estimates of escapement.

Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary spawning areas(Kinzer et al. 2020a, 2020b). Estimates of escapement at Lower Granite Dam, by species and origin, include known uncertainty, and are available to further parse into sex- or age-structured escapement estimates (e.g., Camacho et al. (2017), Schrader et al. (2013)) that are important for fisheries management and productivity monitoring of wild populations. As an example, STADEM is being applied at Lower Granite to estimate the total unique wild fish migrating past the dam. Estimates of fish passing the dam are then combined with PIT tag observations at instream PIT tag detection systems throughout the Snake River basin and estimated movement or transition probabilities, simialar to Waterhouse et al. (2020), to estimate escapement to Snake River populations and locations throughout the basin (Orme et al. 2018). Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations that, joined with sex and age data collected at the adult fish trap (Hargrove et al. 2019), provides necessary information to evaluate productivity and population viability for select Snake River Chinook Salmon and steelhead groups (Kinzer et al. 2020b).

Although STADEM was developed with salmonid escapement at Lower Granite Dam in mind, it could be applied to any migratory fish species at locations with similar monitoring infrastructure. Justification and infrastructure exist for applying a modified STADEM framework for fish passing Bonneville Dam, the lowest dam on the Columbia River, or Priest Rapids Dam in the upper Columbia River. Both locations currently trap a sub-sample of passing Chinook Salmon and steelhead for biological information, and use window counts as a surrogate of true escapement, however, similar problems to those observed at Lower Granite Dam, such as, unaccounted observer and sampling error, night-time passage, and re-ascension exists at the sites. Certainly, estimating an unbiased total return to the entire Columbia River basin (i.e., Bonneville Dam) and Upper Columbia River with uncertainty would benefit managers and decision making.

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

Funding for this study and development of the STADEM model was partially provided by the Bonneville Power Administration under project 2003-017-00. Special thanks to Darren Ogden and staff at the Lower Granite Dam adult trap for their hard work and diligent data collection. Thank you to personnel at the Idaho Department of Fish and Game, particularly Paul Bunn, Tim Copeland, and Bill Schrader, for providing access to data from the adult fish trap and conceptualizing methods, and the staff at the Eagle Fish Lab for analyzing genetic samples. Special thanks to Columbia Basin Research staff and the Columbia River Data Access in Real Time (DART) application, and Susannah Iltis in particular. And finally, thank you to Rick Orme for his contributions to the development of STADEM, and all the other folks who have contributed through productive critique and conversations.

\newpage

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Boggs2004}{}%
Boggs, C., M. L. Keefer, C. Peery, T. C. Bjornn, and L. C. Stuehrenberg. 2004. Fallback, reascension, and adjusted fishway escapement estimates for adult chinook salmon and steelhead at columbia and snake river dams. Transactions of the American Fisheries Society 133(4):932--949. Taylor \& Francis.

\leavevmode\hypertarget{ref-Bue1998}{}%
Bue, B., S. Fried, S. Sharr, D. Sharp, J. Wilcock, and H. Geiger. 1998. Estimating salmon escapement using area-under-the-curve, aerial observer efficiency, and stream-life estimates: The prince william sound pink salmon example. North Pacific Anadromous Fish Commission Bulletin 1:240--250.

\leavevmode\hypertarget{ref-Camacho2019}{}%
Camacho, C. A., T. Delomas, M. Davison, M. E. Dobos, W. C. Schrader, M. R. Campbell, and T. Copeland. 2019. Wild juvenile steelhead and Chinook Salmon abundance and composition at Lower Granite Dam, migratory year 2018. Idaho Department of Fish and Game, 19-12.

\leavevmode\hypertarget{ref-Camacho2017}{}%
Camacho, C., K. Wright, J. Powell, W. C. Schrader, T. Copeland, M. Ackerman, M. Dobos, M. Corsi, and M. R. Campbell. 2017. Wild adult steelhead and Chinook salmon abundance and composition at Lower Granite Dam, spawn years 2009-2016. Idaho Department of Fish and Game, 17-06.

\leavevmode\hypertarget{ref-crozier2019climate}{}%
Crozier, L. G., M. M. McClure, T. Beechie, S. J. Bograd, D. A. Boughton, M. Carr, T. D. Cooney, J. B. Dunham, C. M. Greene, M. A. Haltuch, and others. 2019. Climate vulnerability assessment for pacific salmon and steelhead in the california current large marine ecosystem. PloS one 14(7):e0217711. Public Library of Science San Francisco, CA USA.

\leavevmode\hypertarget{ref-Dauble2000}{}%
Dauble, D. D., and R. P. Mueller. 2000. Upstream passage monitoring: Difficulties in estimating survival for adult chinook salmon in the columbia and snake rivers. Fisheries 25(8):24--34. Taylor \& Francis.

\leavevmode\hypertarget{ref-nwfsc2015}{}%
Ford, M., K. Barnas, T. Cooney, L. Crozier, M. Diaz, J.J. Hard, E. Holmes, D. Holzer, R. Kope, P. Lawson, M. Liermann, J. Myers, M. Rowse, D. Teel, D. V. Doornik, T. Wainwright, L. Weitkamp, and M. Williams. 2015. Status review update for pacific salmon and steelhead listed under the endangered species act: Pacific northwest. National Marine Fisheries Service, Northwest Fisheries Science Center.

\leavevmode\hypertarget{ref-Hargrove2019}{}%
Hargrove, J. S., T. A. Delomas, and M. Davison. 2019. Chinook and steelhead genotyping for genetic stock identification at Lower Granite Dam. Idaho Department of Fish and Game, 19-08.

\leavevmode\hypertarget{ref-Hatch1994}{}%
Hatch, D. R., D. R. Pederson, and J. Fryer. 1994. Feasibility of documenting and estimating adult fish passage at large hydroelectric facilities in the snake river using video technology; 1993 final report. Columbia River Inter-Tribal Fish Commission.

\leavevmode\hypertarget{ref-hess2014monitoring}{}%
Hess, J. E., J. M. Whiteaker, J. K. Fryer, and S. R. Narum. 2014. Monitoring stock-specific abundance, run timing, and straying of chinook salmon in the columbia river using genetic stock identification (gsi). North American Journal of Fisheries Management 34(1):184--201. Taylor \& Francis.

\leavevmode\hypertarget{ref-kinzer2020report}{}%
Kinzer, R. N., H. Arnsberg Bill, A. Maxwell, R. Orme, C. Rabe, and S. Vatland. 2020a. Snake river basin adult chinook salmon and steelhead monitoring 2019 annual report. Nez Perce Tribe, Department of Fisheries Resources Management, Research Division, Lapwai, ID.

\leavevmode\hypertarget{ref-iptds2020report}{}%
Kinzer, R. N., Orme Rick, M. Campbell, J. Hargrove, and K. See. 2020b. REPORT to noaa fisheries for 5-year esa status review: SNAKE river basin steelhead and chinook salmon population abundance, life history, and diversity metrics calculated from in-stream pit-tag observations (sy2010-sy2019).

\leavevmode\hypertarget{ref-mcclure2003large}{}%
McClure, M. M., E. E. Holmes, B. L. Sanderson, and C. E. Jordan. 2003. A large-scale, multispecies status assessment: Anadromous salmonids in the columbia river basin. Ecological Applications 13(4):964--989. Wiley Online Library.

\leavevmode\hypertarget{ref-mcelhany2000viable}{}%
McElhany, P., M. H. Rucklelshaus, M. J. Ford, T. C. Wainwright, and E. P. Bjorkstedt. 2000. Viable salmonid populations and the recovery of evolutionarily significant units.

\leavevmode\hypertarget{ref-nehlsen1991pacific}{}%
Nehlsen, W., J. E. Williams, and J. A. Lichatowich. 1991. Pacific salmon at the crossroads: Stocks at risk from california, oregon, idaho, and washington. Fisheries 16(2):4--21. Wiley Online Library.

\leavevmode\hypertarget{ref-Ogden2014}{}%
Ogden, D. A. 2014. Operation of the adult trap at lower granite dam 2013.

\leavevmode\hypertarget{ref-Ogden2016b}{}%
Ogden, D. A. 2016a. Operation of the adult trap at Lower Granite Dam 2015.

\leavevmode\hypertarget{ref-Ogden2016}{}%
Ogden, D. A. 2016b. Operation of the adult trap at Lower Granite Dam 2014.

\leavevmode\hypertarget{ref-orme2018population}{}%
Orme, R., N. Kinzer Ryan, and C. Albee. 2018. Population and tributary level escapement estimates of snake river natural-origin spring/summer chinook salmon and steelhead from in-stream pit tag detection systems-2019 annual report. Nez Perce Tribe, Department of Fisheries Resources Management, Research Division, Lapwai, ID.

\leavevmode\hypertarget{ref-R-rjags}{}%
Plummer, M. 2019. Rjags: Bayesian graphical models using mcmc.

\leavevmode\hypertarget{ref-R-Core-Team2020}{}%
R Core Team. 2020. R: A Language and Environment for Statistical Computing. R Foundation for Statistical Computing, Vienna, Austria.

\leavevmode\hypertarget{ref-royle2008hierarchical}{}%
Royle, J. A., and R. M. Dorazio. 2008. Hierarchical modeling and inference in ecology: The analysis of data from populations, metapopulations and communities. Elsevier.

\leavevmode\hypertarget{ref-Schrader2013}{}%
Schrader, W. C., M. P. Corsi, P. Kennedy, M. W. Ackerman, M. R. Campbell, K. K. Wright, and T. Copeland. 2013. Wild adult steelhead and Chinook salmon abundance and composition at Lower Granite dam, spawn year 2011. Idaho Department of Fish; Game, Report 13-15.

\leavevmode\hypertarget{ref-Shumway2010}{}%
Shumway, R. H., and D. S. Stoffer. 2010. Time series analysis and its applications: With R examples. Springer.

\leavevmode\hypertarget{ref-Steinhorst2017}{}%
Steinhorst, K., T. Copeland, M. W. Ackerman, W. C. Schrader, and E. C. Anderson. 2017. Abundance estimates and confidence intervals for the run composition of returning salmonids. Fishery Bulletin 115(1):1--12.

\leavevmode\hypertarget{ref-USACE2015}{}%
USACE. 2015. Annual fish passage report Columbia and Snake rivers. U.S. Army Corps of Engineers.

\leavevmode\hypertarget{ref-waterhouse2020bayesian}{}%
Waterhouse, L., J. White, K. See, A. Murdoch, and B. X. Semmens. 2020. A bayesian nested patch occupancy model to estimate steelhead movement and abundance. Ecological Applications:e02202. Wiley Online Library.

\leavevmode\hypertarget{ref-williams2016viability}{}%
Williams, T. H., B. C. Spence, D. A. Boughton, R. C. Johnson, E. G. R. Crozier, N. J. Mantua, M. R. O'Farrell, and S. T. Lindley. 2016. Viability assessment for pacific salmon and steelhead listed under the endangered species act: Southwest.

\leavevmode\hypertarget{ref-Wright2015}{}%
Wright, K. K., W. C. Schrader, L. Reinhardt, K. Hernandez, C. Hohman, and T. Copeland. 2015. Process and methods for assigning ages to anadromous salmonids from scale samples. Idaho Department of Fish; Game, 15-03.

\newpage

\hypertarget{tables}{%
\section{Tables}\label{tables}}

\begin{table}

\caption{\label{tab:lgr-sim-tab}Summary of simulation scenarios including varying adult trapping, fallback rates, re-ascension, nighttime passage, and window count error rates use to evaluate the performance of STADEM.}
\centering
\begin{tabular}[t]{llrrrl}
\toprule
Scenario & Trap rate & Fallback rate & Re-ascension rate & Nighttime passage rate & Window count error\\
\midrule
Baseline & 0.15 & 0.06 & 1 & 0.06 & No Error\\
Baseline Err H & 0.15 & 0.06 & 1 & 0.06 & 10\% Error\\
Baseline Err L & 0.15 & 0.06 & 1 & 0.06 & 5\% Error\\
N-R & 0.15 & 0.10 & 1 & 0.05 & No Error\\
N-R Err H & 0.15 & 0.10 & 1 & 0.05 & 10\% Error\\
\addlinespace
N-R Err L & 0.15 & 0.10 & 1 & 0.05 & 5\% Error\\
N-R trap down & 0.15 and 0.00 3 weeks & 0.06 & 1 & 0.06 & No Error\\
N-R trap down Err H & 0.15 and 0.00 3 weeks & 0.06 & 1 & 0.06 & 10\% Error\\
N-R trap down Err L & 0.15 and 0.00 3 weeks & 0.06 & 1 & 0.06 & 5\% Error\\
Trap down & 0.15 and 0.00 3 weeks & 0.10 & 1 & 0.05 & No Error\\
\addlinespace
Trap down Err H & 0.15 and 0.00 3 weeks & 0.10 & 1 & 0.05 & 10\% Error\\
Trap down Err L & 0.15 and 0.00 3 weeks & 0.10 & 1 & 0.05 & 5\% Error\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}

\caption{\label{tab:summ-tab}Summary statistics, including relative bias, mean coefficient of variation (CV), root mean squared deviation (RMSD) and coverage for results from each of the twelve simulation scenarios.}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
Scenario & Relative bias & Mean CV & RMSD & Coverage\\
\midrule
Baseline & 0.002 & 0.024 & 551.228 & 0.978\\
Baseline Err L & 0.002 & 0.024 & 525.404 & 0.984\\
Baseline Err H & 0.002 & 0.024 & 553.310 & 0.978\\
Trap Down & 0.003 & 0.030 & 534.096 & 0.996\\
Trap Down Err L & 0.002 & 0.030 & 546.582 & 0.998\\
\addlinespace
Trap Down Err H & 0.003 & 0.030 & 566.949 & 0.994\\
N-R & -0.001 & 0.023 & 503.934 & 0.988\\
N-R Err L & 0.002 & 0.024 & 478.226 & 0.990\\
N-R Err H & 0.002 & 0.024 & 508.481 & 0.984\\
N-R Trap Down & 0.002 & 0.030 & 523.817 & 0.994\\
\addlinespace
N-R Trap Down Err L & 0.003 & 0.030 & 568.434 & 0.996\\
N-R Trap Down Err H & 0.003 & 0.030 & 580.082 & 0.994\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}

\caption{\label{tab:lgr-tab}Window counts, and estimates of total, wild, hatchery and hatchery no-clip escapement, with coefficients of variation in parenthesis, for Chinook Salmon and steelhead from spawn years 2010 to 2019.}
\centering
\begin{tabular}[t]{lrlllll}
\toprule
Species & Year & Window counts & Total & Wild & Hatchery & Hatchery no-clip\\
\midrule
Chinook & 2010 & 134,684 & 133,706 (0.046) & 26,994 (0.053) & 99,609 (0.047) & 7,103 (0.079)\\
Chinook & 2011 & 134,594 & 123,610 (0.021) & 24,692 (0.027) & 93,845 (0.022) & 5,072 (0.046)\\
Chinook & 2012 & 84,771 & 83,875 (0.052) & 21,424 (0.048) & 57,929 (0.056) & 4,522 (0.069)\\
Chinook & 2013 & 70,966 & 69,500 (0.023) & 19,086 (0.033) & 44,159 (0.027) & 6,255 (0.048)\\
Chinook & 2014 & 114,673 & 106,741 (0.034) & 28,468 (0.037) & 68,895 (0.038) & 9,377 (0.048)\\
\addlinespace
Chinook & 2015 & 132,432 & 133,330 (0.03) & 23,887 (0.044) & 98,902 (0.033) & 10,540 (0.059)\\
Chinook & 2016 & 81,753 & 84,810 (0.031) & 17,300 (0.031) & 59,471 (0.034) & 8,039 (0.036)\\
Chinook & 2017 & 48,192 & 43,151 (0.039) & 5,162 (0.046) & 34,476 (0.041) & 3,513 (0.051)\\
Chinook & 2018 & 42,232 & 39,621 (0.037) & 7,004 (0.045) & 28,996 (0.039) & 3,621 (0.052)\\
Chinook & 2019 & 29,617 & 27,876 (0.104) & 4,758 (0.128) & 21,185 (0.105) & 1,932 (0.11)\\
\addlinespace
Steelhead & 2010 & 323,382 & 348,082 (0.035) & 45,320 (0.038) & 266,570 (0.036) & 36,193 (0.039)\\
Steelhead & 2011 & 208,296 & 217,640 (0.037) & 45,917 (0.037) & 148,564 (0.038) & 23,159 (0.041)\\
Steelhead & 2012 & 180,320 & 190,947 (0.023) & 40,476 (0.027) & 139,476 (0.024) & 10,995 (0.038)\\
Steelhead & 2013 & 109,186 & 121,265 (0.037) & 25,160 (0.042) & 85,707 (0.038) & 10,398 (0.058)\\
Steelhead & 2014 & 108,154 & 117,169 (0.037) & 28,222 (0.067) & 81,145 (0.042) & 7,803 (0.081)\\
\addlinespace
Steelhead & 2015 & 165,591 & 176,628 (0.029) & 47,959 (0.036) & 118,185 (0.03) & 10,483 (0.054)\\
Steelhead & 2016 & 136,126 & 144,310 (0.031) & 36,141 (0.037) & 101,923 (0.031) & 6,245 (0.056)\\
Steelhead & 2017 & 101,827 & 104,479 (0.035) & 15,472 (0.039) & 80,707 (0.035) & 8,299 (0.061)\\
Steelhead & 2018 & 74,097 & 69,584 (0.033) & 10,105 (0.038) & 56,865 (0.033) & 2,614 (0.055)\\
Steelhead & 2019 & 51,818 & 56,024 (0.104) & 11,073 (0.248) & 41,726 (0.079) & 3,224 (0.092)\\
\bottomrule
\end{tabular}
\end{table}

\newpage

\hypertarget{figures}{%
\section{Figures}\label{figures}}

\begin{figure}
\centering
\includegraphics{../figures/stadem-examp-fig-1.pdf}
\caption{\label{fig:stadem-examp-fig}Schematic of how the STADEM model works. Panel A shows the posterior of the estimate of fish crossing the dam while the window is open (dashed line shows observed window counts). That estimate is divided by the nighttime passage rate (B). The total fish is then discounted by the reascension rate to estimate unique fish (C). Those unique fish are then multiplied by the proportion of wild fish (D), to estimate unique wild fish.}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/rel-bias-lgd-1.pdf}
\caption{\label{fig:rel-bias-lgd}'Boxplots show relative bias of STADEM estimates for wild escapement across various scenarios. The boxes contain 50\% of the simulations, whiskers contain 95\% of the simulations, and points are outliers.}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/jags-fit-fig-1.pdf}
\caption{\label{fig:jags-fit-fig}Time-series plot showing estimates of total escapement for Chinook in 2014, including raw window counts, window counts adjusted for night-time passage, trap estimates and STADEM estimates. Gray ribbon represents the 95\% credible interval for STADEM estimates.}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/night-reasc-diff-fig-1.pdf}
\caption{\label{fig:night-reasc-diff-fig}Night-time passage rate plotted against re-ascension rate, calculated from observed PIT tags for each week of spawn years 2010-2019. Colors correspond to different spawn year, while the size of each point is proportional to the window count that week. The dashed line is the 1-1 line.}
\end{figure}

\newpage

\hypertarget{append1}{%
\section{Appendix A - STADEM Model Description}\label{append1}}

\hypertarget{total-and-weekly-escapement}{%
\subsection{Total and Weekly Escapement}\label{total-and-weekly-escapement}}

Escapement at LGD is estimated by combining two independent observations, trap catches and window counts, of the true number of fish crossing the dam in a state-space model with a weekly time-step. Both are assumed to be corrupted observations of the true unknown number of fish crossing LGD each week. The log of the true number of fish crossing (\(X_t\)), is modeled as a random walk process (Shumway and Stoffer 2010).

\[
\begin{aligned}
  \ln \left( X_{t} \right) &= \ln \left( X_{t-1} \right) + e_t \\
  e_t &\sim \mathcal{N}(0, \sigma^2_X)
\end{aligned}
\]

The number of fish caught in the trap, \(Y^T_t\), for week \(t\) is modeled as a binomial process based on the unknown true trap rate that week, \(\nu_t\), and the unknown true number of fish crossing the dam that week, \(X_t\). The true trap rate is estimated from a beta distribution with previously estimated parameters, \(\hat{\alpha_t}\) and \(\hat{\beta_t}\), informed by the mark-recapture estimate of the trap rate.

\[
\begin{aligned}
  \nu_t &\sim \text{Beta} \left(\hat{\alpha_t}, \hat{\beta_t} \right) \\
  Y^T_t &\sim \text{Bin} \left( \nu_t, X_t \right) \\
\end{aligned}
\]

The estimate of the weekly trap rate is derived based on previously PIT-tagged spring/summer Chinook and steelhead who are crossing LGD that week. The fish caught in the trap that week are considered the ``mark'' group, and all the PIT tagged fish who are detected at the upper end of the LGD fish ladder that week are considered the second capture group (which includes recaptures of the the ``marked'' fish). From this data, we can estimate the total number of previoiusly PIT tagged fish crossing LGD that week, and the proportion of those that are caught in the trap is the weekly trap rate. The uncertainty in estimates of total fish is translated into uncertainty in the trap rate, and this is then summarised with a beta distribution (\(\text{Beta} \left(\hat{\alpha_t}, \hat{\beta_t} \right)\)) that becomes an input into STADEM. Although the group of previously PIT tagged fish is not assumed to be representative of the overall run, the rate at which they are caught in the trap should be the same rate that the overall run experiences.

The number of fish counted at the window, \(Y^W_t\), is modeled as a (potentially) over-dispersed negative binomial process, with an expected value of \(X_t^{day}\), the number of fish crossing the dam while the window is open. This is simply the total number of fish crossing that week, \(X_t\), multiplied by the propoortion of fish crossing while the window is open for counting, \(\theta_t\), calculated on a weekly basis. In the formula below, \(p_t\) is the proportion of fish observed at the window and \(r\) is the shape parameter. If \(r\) is estimated to be small provides evidence for over-dispersion, and as it grows very large, the negative binomial distribution behaves like a Poisson distribution.

\[
\begin{aligned}
  X_t^{day} &= X_t * \theta_t \\
  p_t &= \frac{r}{\left( r + X_t^{day} \right)} \\
  Y^W_t &\sim \text{NegBin} \left( p_t, r \right) \\
\end{aligned}
\]

Thus, the unknown true number of fish crossing LGD each week, \(X_t\), is estimated from two different data source: window counts and fish sampled in the trap. The window counts provide an estimate (with some potential observer error) of the fish crossing during daytime hours, while the fish in the trap, when expanded by the estimated true trap rate, provide an estimate of the total fish crossing that week. For weeks when we have a more precise estimate of the trap rate (i.e.~weeks when lots of previously PIT tagged fish are crossing LGD), STADEM will tend to favor the estimate of total escapement based on the trap data, whereas when that trap rate is more uncertain (e.g.~fewer PIT tagged fish to use in estimating the trap rate), STADEM will rely more on the window counts to estimate total escapement. During peak run times, when lots of fish are crossing LGD, estimates based on trap data and trap rates will be more precise, while estimates from the window counts may have more observation error due to so many fish passing the window. For weeks when the trap is down, STADEM relies exclusively on the window counts and night-time passage data, but there will be more uncertainty in the estimates.

\hypertarget{day-time-passage-and-re-ascension-rates}{%
\subsection{Day-time Passage and Re-ascension Rates}\label{day-time-passage-and-re-ascension-rates}}

There are two other processes that must be accounted for, first, the proportion of fish that cross the dam while the window is closed for counting (night-time passage rate), and the second, the proportion of fish that are crossing the dam multiple times (re-ascension rate) and therefore potentially double-counted. Both rates can be estimated from previously PIT tagged fish that are crossing the dam each week.

The proportion of fish passing the window during non-operational hours, night-time passage rate, is just the complement of the rate of fish passing during the day when the window is operating. The daytime passage rate for week \(t\), \(\theta_t\), is modeled as a random walk process and estimated from a binomial distribution based on the number of PIT tags observed to cross the dam during operational hours, \(y^{day}_t\), and the total number of PIT tags observed to cross the dam at any point that week, \(N_t\) (Shumway and Stoffer 2010).

\[
\begin{aligned}
  y^{day}_t &\sim \text{Bin} \left(\theta_t, N_t \right) \\
  \text{logit} (\theta_{t}) &= \text{logit} \left( \theta_{t-1} \right) + g_t \\
  g_t &\sim \mathcal{N}(0, \sigma^2_\theta) \\
\end{aligned}
\]

The number of total fish crossing Lower Granite differs from the number of unique fish crossing Lower Granite because some fish fall back and re-ascend the dam. These fish are potentially double-counted at the window, and have the potential to be caught in the fish trap more than once. The number of tags known to be re-ascending the dam each week, \(y^{reasc}_t\), is modeled as a binomial process based on the estimated re-ascension rate, \(\eta_t\), and the total number of tags crossing the dam that week, \(N_t\). The logit of the re-ascension rate is modeled as a random walk process similar to day-time passage (Shumway and Stoffer 2010).

\[
\begin{aligned}
  y^{reasc}_t &\sim \text{Bin} \left(\eta_t, N_t \right) \\
  \text{logit} \left( \eta_{t} \right) &= \text{logit} \left(\eta_{t-1} \right) + f_t \\
  f_t &\sim \mathcal{N}(0, \sigma^2_\eta)
\end{aligned}
\]

\hypertarget{origin-proportions}{%
\subsection{Origin Proportions}\label{origin-proportions}}

After estimating the total number of fish to have crossed Lower Granite each week, \(X_t\), the total must be further refined into the number of wild fish, \(X_{w,t}\), hatchery fish, \(X_{h,t}\) and hatchery no-clip fish, \(X_{hnc, t}\). This is done by estimating a weekly origin proportion vector, \(\omega_t\) based on the random sample of fish caught in trap that week, \(Y^T_t\). The observed number of wild, \(Y^T_{w,t}\), hatchery, \(Y^T_{h,t}\), and hatchery no-clip, \(Y^T_{hnc,t}\), fish caught in the trap that week is assumed to come from a multinomial distribution with probability vector \(\omega_t\). The log-odds ratio of the proportions in \(\omega_t\), in relation to the proportion of hatchery fish, \(\omega_{h,t}\) is modeled as a random walk, so it can change through time. This allows the proportions of wild, hatchery and hatchery no-clip fish to shift throughout the season, based on the data available from the fish trap.

\[
\begin{aligned}
  \left( Y^T_{w,t}, Y^T_{h,t}, Y^T_{hnc,t} \right) &\sim \text{Multinom} \left( \omega_t, Y^T_t \right) \\
  \omega_t &= \frac{\exp(\phi_t)}{\sum{\exp(\phi_t})} \\
  \phi_{w,t} &= \ln \left( \frac{\omega_{w,t}}{\omega_{h,t}} \right) \\
  \phi_{hnc,t} &= \ln \left( \frac{\omega_{hnc,t}}{\omega_{h,t}} \right) \\
  \phi_{h,t} &= 0 \\
  \phi_{w,t} &= \phi_{w, t-1} + d_{w,t} \\
  \phi_{hnc,t} &= \phi_{hnc, t-1} + d_{hnc,t} \\
  d_t &\sim \mathcal{N}(0, \sigma^2_\omega) \\
\end{aligned}
\]

Finally, the number of unique fish crossing Lower Granite each week, \(X_{w,t}\), is the product of the total fish crossing that week, \(X_t\) multiplied by one minus the re-ascension rate, \((1 - \eta_t)\), and the origin proportion vector, \(\omega_t\).

\[
\begin{aligned}
  \begin{bmatrix} X_{w,t} \\ X_{h,t} \\ X_{hnc,t} \end{bmatrix}
  &= X_t * (1 - \eta_t) * \begin{bmatrix} \omega_{w,t} \\ \omega_{h,t} \\ \omega_{hnc,t} \end{bmatrix}
\end{aligned}
\]

\hypertarget{model-fitting}{%
\subsection{Model Fitting}\label{model-fitting}}

The model was fit using the JAGS program (Plummer 2019), run with R software (R Core Team 2020). Uninformative priors were used for \(\sigma_X, \sigma_\eta, \sigma_\theta, \sigma_\omega\) and \(\ln(X_1)\) (Uniform(0,10)), as well as logit(\(\eta_1\)) and logit(\(\theta_1\)) (\(\mathcal{N}(0, 1000)\)), and finally \(\phi_{w,1}\) and \(\phi_{hnc,1}\) (\(\mathcal{N}(0, 100)\)).

\newpage

\hypertarget{append2}{%
\section{Appendix B - Simulation Details}\label{append2}}

To simulate fish passing a dam, an \(\mathbb{R}\) software function was developed (R Core Team 2020). The function randomly samples observations from assumed probability distribution functions (\(\mathcal{PDF}\)) with known parameters. Total unique fish, \(N\), and a vector, \(\omega\), containing the proportions of wild (\(w\)), hatchery (\(h\)) and hatchery no-clip (\(hnc\)) fish passing the dam is set to establish known ``truths'' of escapement by origin.

\[
\begin{aligned}
\left[N_{w}, N_{h}, N_{hnc} \right] &= N * \left[\omega_{w}, \omega_{h}, \omega_{hnc} \right]
\end{aligned}
\]

Escapement is then randomly divided across a set number of populations, \(n\), by randomly drawing proportions, \(\phi_{j,p}\), of origin group \(j\) in each poplulation \(p\) using a dirichlet \(\mathcal{PDF}\). The dirichlet function is parameterized from a vector, \(\eta_j\), containing 1's and 0's designating populations with origin \(j\) fish returning. \(\eta_j\) originates from a bernoulli \(\mathcal{PDF}\) and the proportion of populations with each origin, \(\tau_j\). Wild fish are assumed to be in all populations; \(\tau_w = 1.0\). The product of sampled population proportions \(\phi_{j,p}\) and fixed \(N_j\) yields a random variable of abundance for each origin in each population, \(N_{j,p}\). Summing across origin abundances then gives a random total population abundance, \(N_p\), crossing the dam.

\[
\begin{aligned}
  \left[\eta_{j,p=1},  ... , \eta_{j,p=n}\right] &\sim \text{Bern}(\tau_j) \\
  \left[\phi_{j,p=1},  ... , \phi_{j,p=n}\right] &\sim \text{Dir}(\eta_{j,p=1},  ... , \eta_{j,p=n}) \\
  N_{j,p} &= N_j * \phi_{j,p} \\
  N_p &= \sum_{j=w, h, hnc} N_{j,p}
\end{aligned}
\]

Mean arrival date, \(\bar{a}_p\), for each population returning to the dam is drawn from a normal \(\mathcal{PDF}\) with hyper-parameters \(\mu_a\) and \(\sigma^2_a\). Similarly, the variance or spread in run-timing within populations is the absolute value of random variables drawn from a normal \(\mathcal{PDF}\) with hyper-parameters \(\mu_s\) and \(\sigma^2_s\).

\[
\begin{aligned}
 \left[\bar{a}_p, ..., \bar{a}_n\right] &\sim \text{Norm}(\mu_a, \sigma^2_a) \\
 \left[s_p, ..., s_n\right] &\sim \text{Norm}(\mu_s, \sigma^2_s)
\end{aligned}
\]

After sampling the mean date of arrival and variances for each population, the date of arrival, \(a_{i,p}\), for individual fish, \(i\), within each population are drawn from a normal \(\mathcal{PDF}\) with population parameters \(\bar{a}_p\) and \(s^2_p\). This simulates a random arrival day that is similar for all fish returning to the same population, regardless of origin.

\[
\begin{aligned}
 date_{i,p} &\sim \text{Norm}(\bar{a}_p, s^2_p) \\
\end{aligned}
\]

To examine the sensitivites of models to different fish behavior and dam operational scenarios, seven addtional attributes are randomly assigned to each individual fish. Each attribute is randomly assigned a TRUE/FALSE using a bernoulli \(\mathcal{PDF}\) and a fixed probability parameter. Fish passage during the day-time (i.e., during periods of window operation) is modeled using one minus the night-time passage rate (\(1 - \nu\)). Window observations are conditioned on fish passing during the day and being observed at a set rate, \(\gamma\). Whether fish \(i\) is sampled by the adult trap is modeled on the weekly set trap rate, \(\delta_t\). The rate of previously PIT-tagged fish is determined by \(\lambda\), and their subsequent detection at the ladder PIT antenna is governed by \(\kappa\). Fallback behavior is modeled with a common rate across all populations, \(\psi\). Re-ascension occurs with probability \(\rho\), conditioned on fish \(i\) falling back. If fish \(i\) falls back and re-ascends, the entire process described above is repeated, with some time-lag between initial ascension and re-ascension that is governed by a Poisson \(\mathcal{PDF}\) with mean = 2 days. Fish may fall-back and re-ascend up to 3 times, allowing for the possibility of the same fish being counted or trapped multiple times.

\[
\begin{aligned}
  day_{i} &\sim \text{Bern}(1-\nu) \\
  window_{i|d = TRUE} &\sim \text{Bern}(\gamma) \\
  trapped_{i} &\sim \text{Bern}(\delta_t) \\
  tagged_{i} &\sim \text{Bern}(\lambda) \\
  ladder_{i|m = TRUE} &\sim \text{Bern}(\kappa) \\  
  fallback_{i} &\sim \text{Bern}(\psi) \\
  re-ascend_{i|f = TRUE} &\sim \text{Bern}(\rho)
\end{aligned}
\]

Simulation parameters for model evaluations were set to mimic typical ecapement of spring/summer Chinook to LGD with similar origin proportions, marking rates and run timing as those observed from return years 2010 - 2015. Escapement of each origin (\(N_j\)) was set at 25,000 wild, 70,000 hatchery and 5,000 hatchery no-clips spread randomly across 25 populations (\(n\)). Of the 25 populations, each had a 1.0 probability of containing wild fish, 0.50 probability of having hatchery fish and 0.15 probability of receiving hatchery no-clip (\(\tau_j\)); resulting in an expected 25 wild, 12.5 hatchery and 3.75 hatchery no-clip populations. Mean arrival dates and variability were estimated from PIT-tag detection data queried from the Columbia Basin Research Data Access in Real Time (\href{http://www.cbr.washington.edu/dart/query/pit_adult_window}{DART}) website and organized by release subbasin. Mean arrival date across all subbasins and 2010 - 2015 return years was June \(19^{th}\) (\(\mu_a = 171\)) with a standard deviation of 13 days (\(\sigma_a\)). While the observed spread (i.e., variance) of arrival dates within subbasins was determined to have a mean (\(\mu_s\)) of 22 days and a standard deviation of 7 days (\(\sigma_s\)).

For the specific simulated scenarios, we were interested in STADEM model estimates of origin specific escapment from the combinations of two seperate trapping rates, two fallback, re-ascension and night-passage combinations and three window count error rates; resulting in twelve different scenarios. First, trapping rates were set static at 0.15 across all weeks for six scenarios to mimic an optimium trap operation for an expected return of 25,000 wild fish (i.e., trap \(\approx\) 4,000 wild fish). For the remaining six scenarios, trapping rates for weeks 30, 31 and 32 (i.e., July \(22^{nd}\) to August \(11^{th}\)) were changed to 0.00 to test STADEM sensitivities to potential trap shut downs similar to those observed in 2013, 2014 and 2015 (Ogden 2014, 2016b, 2016a). To simulate and control the number of re-ascending and night-time passing fish to model response, we altered fallback and night-time passage rates while holding the re-ascension rate constant at \(\rho = 1.0\). Altering fallback rates and holding re-ascension constant allowed for a more simple control of the number of fish re-ascending; because the number of re-ascending fish is a function of the number of fallbacks and the re-ascension rate. Six scenarios had equal rates of fallback and night-time passage set at \(\psi = \nu = 0.06\) (Boggs et al. 2004) which means other estimator assumptions (Schrader et al. 2013). The other six scenarios set fallback at \(\psi = 0.10\) and night-time passage at \(\nu = 0.05\) to create a 5.0\% positive bias of unique fish at the window. A potential 5.0\% weekly bias was determined from PIT-tag data and within the range of observed weekly difference for return years 2010 - 2015 (Figure \ref{fig:night-reasc-diff-fig}). Finally, we desired to test the sensitivites of STADEM to potential rates of window count error; 0\%, 5\% and 10\% (Hatch et al. 1994). Error at the window was simulated as unbiased (i.e., expect varying high and low counts to cancel each other) and is introduced to the sum of daily window counts using a normal \(\mathcal{PDF}\).

\end{document}

### 2020-08-18 analysis/paper/paper.Rmd
re-knit pdf, word and html versions after Ryan's edits.

@@
 output:
-    bookdown::word_document2:
-      fig_caption: yes
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
-      number_sections: FALSE
-      toc: no
-      always_allow_html: true
-      fig_width: 7
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
-    bookdown::html_document2:
-      fig_caption: yes
-      number_sections: FALSE
-      toc: no
-      fig_width: 7
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
     bookdown::pdf_document2:
@@
         in_header: ../templates/header_manuscript.tex
+    bookdown::html_document2:
+      fig_caption: yes
+      number_sections: FALSE
+      toc: no
+      fig_width: 7
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
+    bookdown::word_document2:
+      fig_caption: yes
+      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+      number_sections: FALSE
+      toc: no
+      always_allow_html: true
+      fig_width: 7
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
 fontsize: 12pt

### 2020-08-20 analysis/paper/paper.Rmd
Corrected a few spelling errors and other minor edits. Re-knit to all 3 formats

@@
         in_header: ../templates/header_manuscript.tex
-    bookdown::html_document2:
-      fig_caption: yes
-      number_sections: FALSE
-      toc: no
-      fig_width: 7
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
     bookdown::word_document2:
@@
       always_allow_html: true
+      fig_width: 7
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
+    bookdown::html_document2:
+      fig_caption: yes
+      number_sections: FALSE
+      toc: no
       fig_width: 7
@@
 Escapement estimates facilitate effective fisheries management, and particularly, estimates of escapement for each component group (e.g., stock, population, age, origin [wild, hatchery]; @hess2014monitoring; @Steinhorst2017) provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations.
-Accurate escapement estimates are increasingly important for depleted populations as they inform viable salmonid metrics [@mcelhany2000viable], and facilitate assessments of population viability and extinction risk [@nwfsc2015; @williams2016viability] and climate change vulerability [@crozier2019climate].
+Accurate escapement estimates are increasingly important for depleted populations as they inform viable salmonid metrics [@mcelhany2000viable], and facilitate assessments of population viability and extinction risk [@nwfsc2015; @williams2016viability] and climate change vulnerability [@crozier2019climate].
 
-Populations of Chinook Salmon Oncorhynchus tshawytscha and steelhead trout O.
-mykiss in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric dams on the Snake and Columbia rivers) and tributary habitats [@nehlsen1991pacific; @mcclure2003large; @nwfsc2015].
+Populations of Chinook Salmon, Oncorhynchus tshawytscha, and steelhead trout, O.
+mykiss, in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric dams on the Snake and Columbia rivers) and tributary habitats [@nehlsen1991pacific; @mcclure2003large; @nwfsc2015].
 As a result, Snake River spring-summer run Chinook Salmon (hereafter Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937).
@@
 Fish management agencies previously used counts of fish, by species, passing an observation window located on the side of the fish ladder as a census of fish escaping past the dam.
-For both species, total escapement was then parsed into groups (e.g., wild, hatchery) using observed marks and genetic data from a sample of fish captured at an adult trap located on the fish ladder upstream of the observation window [@Camacho2017, @Steinhorst2017].
+For both species, total escapement was then parsed into groups (e.g., wild, hatchery) using observed marks and genetic data from a sample of fish captured at an adult trap located on the fish ladder upstream of the observation window [@Camacho2017; @Steinhorst2017].
 Window counts as a census proved beneficial as being an easy, straight-forward method that was ascertained in near real-time.
@@
 Data sources included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of previously PIT tagged fish detected in the adult fish ladder.
-Following, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; however, similar data could likely be obtained from other fish passage facilities.
+Below, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; however, similar data could likely be obtained from other fish passage facilities.
 
@@
 Window counts were made for each species using video monitoring and direct in-person visual monitoring during daytime hours [@Hatch1994].
-Video monitoring occured during the beginning and tail ends of the adult runs (March 1 – March 31 and November 1 – December) for 10 hours per day (0600 – 1600 hours).
-Direct visual monitoring occured during peak run times (April 1 – October 31) for 16 hours per day (0400 – 2000 hours) [@USACE2015].
+Video monitoring occurred during the beginning and tail ends of the adult runs (March 1 – March 31 and November 1 – December) for 10 hours per day (0600 – 1600 hours).
+Direct visual monitoring occurred during peak run times (April 1 – October 31) for 16 hours per day (0400 – 2000 hours) [@USACE2015].
 During direct visual monitoring, observers recorded each adult (≥ 30cm), by species, passing the window for 50 minutes of each hour of operation.
@@
 
+<!-- Don't you think it's important to explain why we need to estimate the trap rate?
+Because otherwise after reading the description of the adult fish trap data, I would assume the trap rate is known -->
+
 The nighttime passage rate was estimated as the count of tags that migrated through the fish ladder during non-window observation hours divided by the total number of tags passing the fish ladder, and was calculated on a weekly basis.
 The re-ascension rate was the count of tags observed passing the upstream most detection sites in the adult fish ladder (i.e., passing the dam) and later detected re-entering the downstream end of the fish ladder at a later time divided by the total number of tags leaving the fish ladder.
-Previously, we examined for differences in nighttime passage and re-ascension rates as estimated using wild fish only, versus combining hatchery and wild fish together, and found no difference.
+Previously, we looked for differences in nighttime passage and re-ascension rates estimated using wild fish only, versus combining hatchery and wild fish together, and found no difference.
 Therefore, we combine wild and hatchery PIT tagged fish observations to estimate common nighttime passage and re-ascension rates to increase sample sizes.
@@
 STADEM further accounted for the proportion of fish that ascended the ladder while the counting window was closed (i.e., night), as well as for fish potentially double-counted (or more) after falling back below the dam and later re-ascending the fish ladder.
-Last, adult sampling data from the trap (wild, hatchery, hatchery no-clip) were used to partition the total escapement estimate by origin (Figure \@ref(fig:stadem-examp-fig).
+Finally, adult sampling data from the trap (wild, hatchery, hatchery no-clip) were used to partition the total escapement estimate by origin (Figure \@ref(fig:stadem-examp-fig)).
 Additional model details can be found in [Appendix A](#append1).
-The STADEM package is available from the primary author at a [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM), and requires the use of the JAGS software [@R-rjags] for Bayesian inference.
+The STADEM package is available from the primary author at [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM), and requires the use of the JAGS software [@R-rjags] for Bayesian inference.
 
@@
 
-Finally, we applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead returning to the Snake River to spawn during 2010 to 2019.
+Finally, we applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead returning to the Snake River to spawn for the spawn years 2010 to 2019.
 Window counts for both species were accessed from DART via functions within STADEM.
@@
 
-Simulation results of observed bias, sampling variation, precision, root mean squared deviation (RMSD), and coverage probabilities were qualitatively similar for hatchery (N = 70,000) and hatchery no-clip (N = 5,000) origins as observed in the wild origin (N = 25,000) comparisons.
+Simulation results of observed bias, sampling variation, precision, root mean squared deviation (RMSD), and coverage probabilities were qualitatively similar for the hatchery (N = 70,000) and hatchery no-clip (N = 5,000) origin groups as observed in the wild origin group (N = 25,000) comparisons.
 As such, only diagnostic measures of Lower Granite Dam model fits to a medium sized escapement level (e.g., wild origin escapement) are presented.
@@
 Estimates were unbiased, with an average relative bias of 0.2–0.3%.
-The CV of the estimates average 2.0–3.0%, with coverage probabilities that always exceeded 95%.
+The CV of the estimates averaged 2.0–3.0%, with coverage probabilities that always exceeded 95%.
 We calculated RMSD as the square root of the sum of the variance of the estimate and the squared expected bias, which accounts for the size of the uncertainty in the estimator as well as its bias.
@@
 Estimates of total escapement, as well as estimates of wild, hatchery, and hatchery no-clip estimates are presented in Table \@ref(tab:lgr-tab).
-Coefficients of variation ranged from 3.5-7.5% for wild fish, 3.0-6.2% for hatchery fish, 4.0-8.4% for hatchery no-clip fish and 2.9-5.6% for total unique fish past Lower Granite Dam.
+Coefficients of variation ranged from 2.7-24.8% for wild fish, 2.2-10.5% for hatchery fish, 3.6-11% for hatchery no-clip fish and 2.1-10.4% for total unique fish past Lower Granite Dam.
+
+<!-- All of the high values were from 2019, and the ranges drop quite a bit if we exclude that year. Max is more like 6%. Worth trying to mention that?  -->
 
@@
 However, when very few fish were caught in the trap, or there was more uncertainty about the trap rate that week, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \@ref(fig:jags-fit-fig).
-That year shows the utility of STADEM in dealing with missing data, as the trap was shut down for several weeks in July and August.
+That year also shows the utility of STADEM in dealing with missing data, as the trap was shut down for several weeks in July and August.
 
@@
 Our model explicitly models nighttime passage, re-ascension, and potential error in both window and trap estimates.
-In doing so, we demonstrated that at Lower Granite Dam, nighttime passage and re-ascension rates do not always offset each other, and assuming they do, will lead to biased escapement estimates in some years.
-With minor adjustments this modelling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall Chinook Salmon, Pacific lamprey Lampetra tridentata), or elsewhere; provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run, and a tag observation or detection infrastructure (e.g., a PIT tag detection array or similar) exists.
+In doing so, we demonstrated that at Lower Granite Dam, nighttime passage and re-ascension rates do not always offset each other, and assuming they do will lead to biased escapement estimates in some years.
+With minor adjustments this modelling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall Chinook Salmon, Pacific lamprey, Lampetra tridentata), or elsewhere; provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run, and a tag observation or detection infrastructure (e.g., a PIT tag detection array or similar) exists.
 Our state-space model combined multiple imperfect sources of data to reduce bias in adult escapement estimates and provided more reasonable estimates of uncertainty.
@@
 Currently, the only roadblock to this, at Lower Granite Dam, is the identification of hatchery origin fish from phenotypically wild fish (i.e., hatchery no-clip) using genetic tissues samples (***Steele et al.
-2013***) collected at the adult trap, which currently is completed post hoc after the trapping season.
+2013***) collected at the adult trap, which currently is completed post-hoc after the trapping season.
 The inclusion of genetic information typically results in a reduction in wild escapement estimates and an associated increase in hatchery no-clip escapement.
@@
 
-Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary spawning areas[@iptds2020report; @kinzer2020report].
-Estimates of escapement at Lower Granite Dam, by species and origin, include known uncertainty, and are available to further parse into sex- or age-structured escapement estimates (e.g., @Camacho2017, @Schrader2013) that are important for fisheries management and productivity monitoring of wild populations.
+Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary spawning areas [@iptds2020report; @kinzer2020report].
+Estimates of escapement at Lower Granite Dam, by species and origin, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates (e.g., @Camacho2017, @Schrader2013) that are important for fisheries management and productivity monitoring of wild populations.
 As an example, STADEM is being applied at Lower Granite to estimate the total unique wild fish migrating past the dam.
-Estimates of fish passing the dam are then combined with PIT tag observations at instream PIT tag detection systems throughout the Snake River basin and estimated movement or transition probabilities, simialar to @waterhouse2020bayesian, to estimate escapement to Snake River populations and locations throughout the basin [@orme2018population].
+Estimates of fish passing the dam are then combined with estimated movement or transition probabilities based on PIT tag observations at instream PIT tag detection systems throughout the Snake River basin, similar to @waterhouse2020bayesian, to estimate escapement to Snake River populations and locations throughout the basin [@orme2018population].
 Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations that, joined with sex and age data collected at the adult fish trap [@Hargrove2019], provides necessary information to evaluate productivity and population viability for select Snake River Chinook Salmon and steelhead groups [@iptds2020report].
@@
 Justification and infrastructure exist for applying a modified STADEM framework for fish passing Bonneville Dam, the lowest dam on the Columbia River, or Priest Rapids Dam in the upper Columbia River.
-Both locations currently trap a sub-sample of passing Chinook Salmon and steelhead for biological information, and use window counts as a surrogate of true escapement, however, similar problems to those observed at Lower Granite Dam, such as, unaccounted observer and sampling error, night-time passage, and re-ascension exists at the sites.
+Both locations currently trap a sub-sample of passing Chinook Salmon and/or steelhead for biological information, and use window counts as a surrogate of true escapement.
+However, each has at least some similar problems to those observed at Lower Granite Dam, such as unaccounted observer and sampling error, night-time passage, and/or re-ascension.
 Certainly, estimating an unbiased total return to the entire Columbia River basin (i.e., Bonneville Dam) and Upper Columbia River with uncertainty would benefit managers and decision making.

### 2020-08-26 analysis/paper/paper.Rmd
a few last minute edits from Jay

@@
 output:
+    bookdown::html_document2:
+      fig_caption: yes
+      number_sections: FALSE
+      toc: no
+      fig_width: 7
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
     bookdown::pdf_document2:
@@
       always_allow_html: true
-      fig_width: 7
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
-    bookdown::html_document2:
-      fig_caption: yes
-      number_sections: FALSE
-      toc: no
       fig_width: 7
@@
 Escapement estimates facilitate effective fisheries management, and particularly, estimates of escapement for each component group (e.g., stock, population, age, origin [wild, hatchery]; @hess2014monitoring; @Steinhorst2017) provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations.
-Accurate escapement estimates are increasingly important for depleted populations as they inform viable salmonid metrics [@mcelhany2000viable], and facilitate assessments of population viability and extinction risk [@nwfsc2015; @williams2016viability] and climate change vulnerability [@crozier2019climate].
+Accurate escapement estimates are increasingly important for depleted populations as they inform status metrics [@mcelhany2000viable], and facilitate assessments of population viability and extinction risk [@nwfsc2015; @williams2016viability] and climate change vulnerability [@crozier2019climate].
 
@@
 As a result, Snake River spring-summer run Chinook Salmon (hereafter Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937).
-Snake River Chinook Salmon and steelhead have substantial recreational, commercial, cultural and subsistence value both within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries.
-The aggregate escapement of Snake River Chinook Salmon and steelhead populations, with the exception of Tucannon River, is measured at Lower Granite Dam located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations.
-Many fisheries management and conservation actions are made based on estimates of escapement at Lower Granite Dam parsed by species and origin (**Snake Basin Harvest Plans?**).
-Additionally, harvest openings and closures, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia River fisheries, are predicated on escapements at Lower Granite Dam (**could cite US v OR Management Agreement and Snake Basin Harvest Plans?**).
+Snake River Chinook Salmon and steelhead have substantial cultural, recreational, commercial and subsistence value both within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries.
+The aggregate escapement of Snake River Chinook Salmon and steelhead populations, with the exception of Tucannon River, is monitored at Lower Granite Dam located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations.
+Many fisheries management and conservation actions are made based on estimates of escapement at Lower Granite Dam parsed by species and origin (**cite BiOp, NPCC FWP, and status assessment**).
+Additionally, harvest openings and closures, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia River fisheries, are predicated on escapements at Lower Granite Dam.
 
 Chinook Salmon and steelhead returning to a majority of populations in the Snake River must ascend a fish ladder on Lower Granite Dam before migrating to their natal tributary spawning locations.
-Fish management agencies previously used counts of fish, by species, passing an observation window located on the side of the fish ladder as a census of fish escaping past the dam.
+Fish management agencies previously used counts of fish, by species, passing an observation window within the fish ladder as a census of fish escaping past the dam.
 For both species, total escapement was then parsed into groups (e.g., wild, hatchery) using observed marks and genetic data from a sample of fish captured at an adult trap located on the fish ladder upstream of the observation window [@Camacho2017; @Steinhorst2017].
@@
 Typically, the observational error rates of live and video window counts are unknown, and sampling error rates are ignored.
-Additionally, two biological processes are unaccounted for: 1) fish that cross the dam during the 8-hours when the window is closed for counting (i.e., nighttime passage) which may result in an underestimate of escapement, and 2) fish that cross the dam multiple times (i.e., re-ascension), and therefore potentially double-counted.
-Finally, Chinook Salmon and steelhead that migrate through the ladder and past the dam may fallback over the dam (e.g., over spillway, through navigation locks), and later, may or may not re-ascend the fish ladder again [@Boggs2004].
+Additionally, two biological processes are unaccounted for: 1) fish that cross the dam during the 8-hours when the window is closed for counting (i.e., nighttime passage) which may result in an underestimate of escapement, and 2) fish that migrate through the ladder and past the dam may fallback over the dam (e.g., over spillway, through navigation locks), and later, may or may not re-ascend the fish ladder again [@Boggs2004].
 Both fallback with no re-ascension and fallback with re-ascension potentially result in an overestimate of escapement [@Dauble2000].
@@
 
-In this study, we describe a novel method to estimate aggregate and group escapement which incorporates all sources of known uncertainty, we demonstrate the estimation method is essentially unbiased, thus improving effective conservation and management decision making, and show night-time passage and fallback/re-ascension rates are unequal.
+In this study, we describe a novel method to estimate aggregate and group escapement which incorporates all sources of known uncertainty, we demonstrate the estimation method is essentially unbiased, thus better informing conservation and management decision making, and show night-time passage and fallback/re-ascension rates are unequal.
 Our method for estimating escapement, by species, past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of PIT tagged fish in the adult ladder to explicitly model nighttime passage, re-ascension, and error from both window and trap estimates using a state-space approach [@royle2008hierarchical].
@@
 To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates.
-The STADEM model combines multiple imperfect sources of data to reduce bias in escapement estimates and provide more reasonable estimates of uncertainty.
-Additionally, the STADEM model framework can be applied elsewhere to estimate aggregate and group escapement, provided a passage barrier containing multiple counting and observational mechanisms exist, such as, a counting window, an adult trap sampling a portion of the run, and PIT tag antenna infrastructure.
+The STADEM model combines multiple imperfect sources of data to reduce bias in escapement estimates and provide more reasonable estimates of uncertainty.<!-- RK comment: the following sentence is better left in the discussion and can be removed from this paragraph.
+-- Additionally, the STADEM model framework can be applied elsewhere to estimate aggregate and group escapement, provided a passage barrier containing multiple counting and observational mechanisms exist, such as, a counting window, an adult trap sampling a portion of the run, and PIT tag antenna infrastructure.
+-->
 
@@
 Data sources included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of previously PIT tagged fish detected in the adult fish ladder.
-Below, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; however, similar data could likely be obtained from other fish passage facilities.
+Below, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; knowing, similar data could likely be obtained from other fish passage facilities.
 
@@
 
-Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary spawning areas [@iptds2020report; @kinzer2020report].
+Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary or population specific spawning areas [@iptds2020report; @kinzer2020report].
 Estimates of escapement at Lower Granite Dam, by species and origin, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates (e.g., @Camacho2017, @Schrader2013) that are important for fisheries management and productivity monitoring of wild populations.

### 2020-08-26 analysis/paper/paper.Rmd
MA edits to manuscript

@@
       correspondence: false
-  - Mike W. Ackerman:
+  - Michael W. Ackerman:
       email: mike.ackerman@merck.com
@@
       toc: no
+      fig_width: 7
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
+    bookdown::word_document2:
+      fig_caption: yes
+      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+      number_sections: FALSE
+      toc: no
+      always_allow_html: true
       fig_width: 7
@@
         in_header: ../templates/header_manuscript.tex
-    bookdown::word_document2:
-      fig_caption: yes
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
-      number_sections: FALSE
-      toc: no
-      always_allow_html: true
-      fig_width: 7
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
 fontsize: 12pt
@@
 - packages.bib
+- AckermanLibrary.bib
 csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
@@
 
-Fish escapement is the number of adults that survive juvenile and adult rearing, escape harvest and return to their natal habitat to potentially spawn (e.g.
+Fish escapement often refers to the number of adults that survive juvenile and subadult rearing, escape harvest and return to their natal habitat to potentially spawn (e.g.
 @Bue1998).
 For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their spawning area.
-Escapement estimates facilitate effective fisheries management, and particularly, estimates of escapement for each component group (e.g., stock, population, age, origin [wild, hatchery]; @hess2014monitoring; @Steinhorst2017) provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations.
+Escapement estimates facilitate effective fisheries management, and particularly, estimates of escapement for component groups (e.g., by stock, population, age, origin [wild, hatchery]; @hess2014monitoring; @Steinhorst2017; @Camacho2017) provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations.
 Accurate escapement estimates are increasingly important for depleted populations as they inform status metrics [@mcelhany2000viable], and facilitate assessments of population viability and extinction risk [@nwfsc2015; @williams2016viability] and climate change vulnerability [@crozier2019climate].
 
-Populations of Chinook Salmon, Oncorhynchus tshawytscha, and steelhead trout, O.
-mykiss, in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric dams on the Snake and Columbia rivers) and tributary habitats [@nehlsen1991pacific; @mcclure2003large; @nwfsc2015].
-As a result, Snake River spring-summer run Chinook Salmon (hereafter Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937).
-Snake River Chinook Salmon and steelhead have substantial cultural, recreational, commercial and subsistence value both within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries.
-The aggregate escapement of Snake River Chinook Salmon and steelhead populations, with the exception of Tucannon River, is monitored at Lower Granite Dam located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations.
-Many fisheries management and conservation actions are made based on estimates of escapement at Lower Granite Dam parsed by species and origin (**cite BiOp, NPCC FWP, and status assessment**).
-Additionally, harvest openings and closures, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia River fisheries, are predicated on escapements at Lower Granite Dam.
+Populations of Chinook salmon *Oncorhynchus tshawytscha* and steelhead trout *O.
+mykiss* in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric dams on the Snake and Columbia rivers) and tributary habitats [@nehlsen1991pacific; @mcclure2003large; @nwfsc2015].
+As a result, Snake River spring-summer run Chinook salmon (hereafter Chinook salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937).
+Snake River Chinook salmon and steelhead have substantial cultural, recreational, commercial and subsistence value both within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries.
+The aggregate escapement of Snake River Chinook salmon and steelhead populations, with the exception of Tucannon River, is monitored at Lower Granite Dam located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations.
+Many fisheries management and conservation actions are made based on estimates of escapement at Lower Granite Dam parsed by species and origin [@NMFS2019BiOp; @NFSC2015; @NPCC2014].
+Additionally, harvest openings and closures, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia rivers fisheries, are predicated on escapements at Lower Granite Dam.
 
-Chinook Salmon and steelhead returning to a majority of populations in the Snake River must ascend a fish ladder on Lower Granite Dam before migrating to their natal tributary spawning locations.
+Chinook salmon and steelhead returning to a majority of populations in the Snake River must ascend a fish ladder on Lower Granite Dam before migrating to their natal tributary spawning locations.
 Fish management agencies previously used counts of fish, by species, passing an observation window within the fish ladder as a census of fish escaping past the dam.
-For both species, total escapement was then parsed into groups (e.g., wild, hatchery) using observed marks and genetic data from a sample of fish captured at an adult trap located on the fish ladder upstream of the observation window [@Camacho2017; @Steinhorst2017].
+For both species, total escapement was then parsed into groups (e.g., wild, hatchery) using observed marks and genetic data from a sample of fish captured at an adult trap located on the fish ladder upstream of the observation window [@Camacho2017; @Steinhorst2017; @Steele2019].
 Window counts as a census proved beneficial as being an easy, straight-forward method that was ascertained in near real-time.
-Moreover, downriver fisheries management arenas have used window counts at lower-river dams as escapement estimates for the past several years, and consistency in methods is often desirable for management decision making (**citation?**).
+Moreover, downriver fisheries management arenas have used window counts at lower-river dams as escapement estimates for the past several years, and consistency in methods is often desirable for management decision making [@WDFW2019].
 
-Using window counts, however, as a census of Chinook Salmon and steelhead passing Lower Granite Dam can be problematic as it avoids multiple sources of uncertainty and ignores known biological processes.
+Using window counts, however, as a census of Chinook salmon and steelhead passing Lower Granite Dam can be problematic as it avoids multiple sources of uncertainty and disregards known biological processes.
 First, live (i.e., in-person) window counts only occur from April through October each year.
-Additionally, live window counts only occur for 16-hours each day, and fish counters working at the observation window look directly into the fish ladders to identify and count all passing fish, by species, for only 50 minutes of each hour.
-Counts are then expanded to provide an estimate for the hour [@USACE2015].
+Additionally, live window counts only occur for 16-hours each day, and fish counters working at the observation window look directly into the fish ladders to identify and count all passing fish, by species, for 50 minutes of each hour.
+Counts are then expanded to provide an estimate for the entire hour [@USACE2015].
 From November through March, the remainder of the year, video tape fish counting is used and only occurs for 10 hours each day; fish counters then read the video tapes and submit daily fish counts [@Hatch1994].
@@
 Both fallback with no re-ascension and fallback with re-ascension potentially result in an overestimate of escapement [@Dauble2000].
-Previously, it was assumed that nighttime passage rates and fallback/re-ascension rates cancelled each other out resulting in window counts providing an unbiased estimate of escapement (**citation?**)
+Previously, it was assumed that nighttime passage rates and fallback/re-ascension rates canceled each other out resulting in window counts providing an unbiased estimate of escapement [@Camacho2017].
 
-In this study, we describe a novel method to estimate aggregate and group escapement which incorporates all sources of known uncertainty, we demonstrate the estimation method is essentially unbiased, thus better informing conservation and management decision making, and show night-time passage and fallback/re-ascension rates are unequal.
+In this study, we describe a novel method to estimate aggregate and group escapements which incorporate all sources of known uncertainty and demonstrate that the estimation method is essentially unbiased, thus better informing conservation and management decision making.
+We also show that observed nighttime passage and fallback/re-ascension rates are typically unequal.
 Our method for estimating escapement, by species, past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of PIT tagged fish in the adult ladder to explicitly model nighttime passage, re-ascension, and error from both window and trap estimates using a state-space approach [@royle2008hierarchical].
@@
 To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates.
-The STADEM model combines multiple imperfect sources of data to reduce bias in escapement estimates and provide more reasonable estimates of uncertainty.<!-- RK comment: the following sentence is better left in the discussion and can be removed from this paragraph.
+The STADEM model combines multiple imperfect sources of data to reduce bias in escapement estimates and provides improved estimates of uncertainty.
+
+<!-- RK comment: the following sentence is better left in the discussion and can be removed from this paragraph.
 -- Additionally, the STADEM model framework can be applied elsewhere to estimate aggregate and group escapement, provided a passage barrier containing multiple counting and observational mechanisms exist, such as, a counting window, an adult trap sampling a portion of the run, and PIT tag antenna infrastructure.
@@
 
-We used STADEM and three sources of data to estimate Chinook Salmon and steelhead escapement at Lower Granite Dam from 2010-2019.
+We used STADEM and three sources of data to estimate Chinook salmon and steelhead escapement at Lower Granite Dam from 2010-2019.
 Data sources included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of previously PIT tagged fish detected in the adult fish ladder.
@@
 
-Daily counts of adult Chinook Salmon and steelhead passing an observation window located on the Lower Granite fish ladder were the first source of data.
+Daily counts of adult Chinook salmon and steelhead passing an observation window located on the Lower Granite Dam fish ladder were the first source of data.
 Daily counts were made and provided by the US Army Corps of Engineers, and when summed, provide an estimate of the number of fish ascending and passing (i.e., escapement) Lower Granite Dam each season.
@@
 The second source of data came from a sample of fish collected in the adult trap as they migrated past Lower Granite Dam [@Ogden2016b].
-The trap, also located within the adult fish ladder and upstream of the observation window, provided biological data (e.g., origin [wild, hatchery], genetic stock, length, age, sex) for captured adults that allowed decomposition of the escapement into specific groups (e.g., @Camacho2019, @Steinhorst2017).
+The trap, also located within the adult fish ladder and upstream of the observation window, provided biological data (e.g., origin [wild, hatchery], genetic stock, length, age, sex) for captured adults that allowed decomposition of the escapement into specific groups (e.g., @Camacho2017, @Steinhorst2017).
 The trap was operational for 24 hours per day, and randomly sampled the daily run by opening four times per hour for a length of time determined by a set daily trapping rate.
 The trap rate was determined by a committee of collaborating management agencies with a goal of capturing a target number of wild fish, but also balancing fish handling concerns.
-Trap sample rates were typically 10-25%, but fluctuated throughout the season due to high water temperatures, decreased flows, trap malfunctions and/or closures, fish handling logistics, etc.
+Trap sample rates were typically 10-25%, but fluctuated throughout the season due to high water temperatures, decreased flows, trap malfunctions and/or closures, fish handling logistics, in-season forecast adjustments, etc.
+
 All captured fish were anesthetized, identified to species, examined for existing marks/tags, and measured for fork length.
-Additionally, each fish had scale and genetic tissue samples taken; scale samples were used to estimate age [@Wright2015], tissue samples were used to determine sex and for genetic stock identification analysis (e.g., @Hargrove2019).
+For adipose-intact (unclipped) adults, which includes wild and hatchery-no clip individuals, some portion or all of fish trapped had scale and genetic tissue samples taken.
+Scale samples were used to estimate age [@Wright2015] and genetic tissue samples were used to determine sex [@Campbell2012a] and estimate the origin of wild fish via genetic stock identification (e.g., @Hargrove2019).
+Starting with adults spawning in 2013, every unclipped Chinook salmon and steelhead trapped at LGR was genotyped to simplify collaborative logistics and better estimate the proportion of phenotypically wild fish of truly hatchery origin.
+Previous to that, only fish determined to be wild in origin at the trap were sampled for scale and genetics.
+@Camacho2017 provide further details on trap sample rates and valid sample selection.
 Prior to release, all non-PIT tagged fish with an intact adipose fin (i.e., putatively wild) received a PIT tag.
-Final determination of wild, hatchery, and hatchery no-clip origins were assigned using a post-hoc analysis of marks, tags, and genetic information.
-Data from the adult trap were collected and managed by multiple agencies and were made available by the Idaho Department of Fish and Game [@Camacho2019].
+Final determination of wild, hatchery, and hatchery no-clip origins were assigned using a post-hoc analysis of marks and tags, including genetic parentage-based tags [@Steele2013; @Steele2019].
+Data from the adult trap were collected and managed by multiple agencies and were made available by the Idaho Department of Fish and Game [@Camacho2017].
 
@@
 
-The last source of data was observations of PIT tagged adult Chinook Salmon and steelhead at detection sites located in the Lower Granite Dam fish ladder.
-These observations provide estimates of 1) a trapping rate, 2) the proportion of fish passing during nighttime hours (i.e., outside of window observation hours), and 3) the proportion of fish that ascend the fish ladder multiple times (i.e., re-ascension rate).
+The last source of data was observations of PIT tagged adult Chinook salmon and steelhead at detection sites located in the Lower Granite Dam fish ladder.
+These observations provide estimates of 1) a trapping rate, 2) the proportion of fish passing during nighttime hours, and 3) the proportion of fish that ascend the fish ladder multiple times (i.e., the re-ascension rate).
 Detections used in the model include all fish that were previously PIT tagged as juveniles or adults prior to reaching Lower Granite Dam (i.e., does not include newly tagged adults at the dam) and detected at adult detection sites in the dam passage system.
@@
 
-A trap rate estimate was derived using mark-recapture methods and PIT tag observations of both Chinook Salmon and steelhead at Lower Granite Dam adult detection sites.
+A trap rate estimate was derived using mark-recapture methods and PIT tag observations of both Chinook salmon and steelhead at Lower Granite Dam adult detection sites.
 The "mark" group included all tags detected in the adult trap and the "capture" group included tags observed to cross the weir at the upstream end of the fish ladder as adults left the passage system.
@@
 Those estimates, with associated uncertainties, were then incorporated into the model as informed priors, while the model estimated the true trap rate based on all the data, including trap and window counts.
+We want to estimate the "true" trapping rate as the set trapping rate, the recorded time that the trap is open to trap adults, does not always reflect the true proportion of fish that are captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error, among others.
+Therefore, we use the mark-recapture estimate of the trap rate to estimate a true trapping rate.
+
 <!--The true trapping rate was estimated as the recorded time of the trap being open does not always reflect the true proportion of fish that are captured in the trap, due to trap malfunctions, sort-by-code fish opening the trap more frequently than expected, or process error (Rick Orme, *per.
@@
 We assumed that the window counts and the estimates from the trap (fish in the trap divided by trap rate that week) were generated by processes with observation error.
-In the case of the trap for example, we assumed there was sampling variation and uncertainty around our estimates of the true unknown trap rate.
+In the case of the trap, for example, we assumed there was sampling variation and uncertainty around our estimates of the true unknown trap rate.
 STADEM further accounted for the proportion of fish that ascended the ladder while the counting window was closed (i.e., night), as well as for fish potentially double-counted (or more) after falling back below the dam and later re-ascending the fish ladder.
@@
 These simulated data sets contained a fixed number of unique adult fish of known origin crossing a dam, from a total of 25 fictional populations with differential run-timing (i.e., date of passage at Lower Granite Dam).
-Each simulated fish was given a date of ladder ascension, based on its population and the range of run timing for that population.
+Each simulated fish was given a date of ladder ascension, based on its population and the range of observed run timing for that population.
 Each fish was also simulated to cross the dam either while the window was open for counting, or not, and was given the chance to be "caught" in the simulated fish trap given the week when it ascended the dam, and the known trap rate that week.
@@
 
-Finally, we applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead returning to the Snake River to spawn for the spawn years 2010 to 2019.
+Finally, we applied STADEM to data from Lower Granite Dam for Chinook salmon and steelhead returning to the Snake River during spawn years 2010 to 2019.
 Window counts for both species were accessed from DART via functions within STADEM.
-For Chinook Salmon, a spawn year refers to adults that migrate past the dam prior to August 17 each year and spawn that late summer and fall.
+For Chinook salmon, a spawn year refers to adults that migrate past the dam prior to August 17 each year and spawn that late summer and fall.
 For steelhead, the spawn year is defined as steelhead that migrate past Lower Granite Dam starting July 1 the previous year and prior to June 30 of the given year (e.g., spawn year 2017 steelhead migrate past Lower Granite between July 1, 2016 and June 30, 2017).
@@
 
-We applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead for spawn years 2010–2019.
+We applied STADEM to data from Lower Granite Dam for Chinook salmon and steelhead for spawn years 2010–2019.
 Estimates of total escapement, as well as estimates of wild, hatchery, and hatchery no-clip estimates are presented in Table \@ref(tab:lgr-tab).
@@
 
-<!-- All of the high values were from 2019, and the ranges drop quite a bit if we exclude that year. Max is more like 6%. Worth trying to mention that?  -->
+<!-- 
+All of the high values were from 2019, and the ranges drop quite a bit if we exclude that year. Max is more like 6%. Worth trying to mention that?
+
+MA: Maybe worth including in the Discussion?
+-->
 
@@
 
-Examining the estimates of nighttime passage and re-ascension rates based on the observed PIT tags crossing over Lower Granite Dam, the two rates did not match up (Figure \@ref(fig:night-reasc-diff-fig)).
+Examining the estimates of nighttime passage and re-ascension rates based on the observed PIT tags crossing over Lower Granite Dam (Figure \@ref(fig:night-reasc-diff-fig)), the two rates did not match up in most cases.
 In particular, there are several weeks when the window counts are quite high, and the rates differ by as much as 10%.
-Clearly, those two biological processes do not cancel each other out, and thus, employing a model that accounts for both will result in more accurate estimates of escapement.
+Clearly, those two biological processes do not cancel each other out in nearly all cases, and thus, employing a model that accounts for both will result in more accurate estimates of escapement.
 
@@
 
-We have presented a novel method for estimating adult salmonid escapement past a large hydroelectric facility (e.g., Lower Granite Dam) that incorporates data from window counts, fish trap, and observations of PIT tagged fish in the adult passage ladder.
+We have presented a novel method for estimating adult salmonid escapement past a large hydroelectric facility (e.g., Lower Granite Dam) that incorporates data from window counts, a fish trap, and observations of PIT tagged fish in the adult passage ladder.
 Our model explicitly models nighttime passage, re-ascension, and potential error in both window and trap estimates.
 In doing so, we demonstrated that at Lower Granite Dam, nighttime passage and re-ascension rates do not always offset each other, and assuming they do will lead to biased escapement estimates in some years.
-With minor adjustments this modelling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall Chinook Salmon, Pacific lamprey, Lampetra tridentata), or elsewhere; provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run, and a tag observation or detection infrastructure (e.g., a PIT tag detection array or similar) exists.
+With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall Chinook salmon, Pacific lamprey *Lampetra tridentata*), or elsewhere; provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run, and tag observation or detection infrastructure (e.g., a PIT tag detection array or similar) exists.
 Our state-space model combined multiple imperfect sources of data to reduce bias in adult escapement estimates and provided more reasonable estimates of uncertainty.
@@
 
-<!-- I can't find the Steele et al. 2013 reference -->
-
 Although not currently set up for this, STADEM could be modified and run on a weekly basis or in near real-time to provide better in-season estimates for fisheries managers.
-Currently, the only roadblock to this, at Lower Granite Dam, is the identification of hatchery origin fish from phenotypically wild fish (i.e., hatchery no-clip) using genetic tissues samples (***Steele et al.
-2013***) collected at the adult trap, which currently is completed post-hoc after the trapping season.
+Currently, the only roadblock to this, at Lower Granite Dam, is the identification of hatchery origin fish from phenotypically wild fish (i.e., hatchery no-clip) using genetic tissue samples [@Steele2013; @Steele2019] collected at the adult trap, which currently is completed post-hoc after the trapping season.
 The inclusion of genetic information typically results in a reduction in wild escapement estimates and an associated increase in hatchery no-clip escapement.
@@
 All other data included in this model (e.g., window counts and PIT observations) are otherwise provided in almost real-time by DART.
-Provided the Lower Granite Dam adult fish trap database was updated and made available frequently, there are minimal obstacles for adapting the STADEM framework to provide in-season estimates of escapement.
+Provided the Lower Granite Dam adult fish trap database was updated and available in near real-time, there are minimal obstacles for adapting the STADEM framework to provide in-season estimates of escapement.
 
-Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary or population specific spawning areas [@iptds2020report; @kinzer2020report].
+Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook salmon and steelhead past Lower Granite Dam, and returning to tributary or population specific spawning areas [@iptds2020report; @kinzer2020report].
 Estimates of escapement at Lower Granite Dam, by species and origin, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates (e.g., @Camacho2017, @Schrader2013) that are important for fisheries management and productivity monitoring of wild populations.
-As an example, STADEM is being applied at Lower Granite to estimate the total unique wild fish migrating past the dam.
+As an example, STADEM is being applied at Lower Granite Dam to estimate the total unique wild fish migrating past the dam.
 Estimates of fish passing the dam are then combined with estimated movement or transition probabilities based on PIT tag observations at instream PIT tag detection systems throughout the Snake River basin, similar to @waterhouse2020bayesian, to estimate escapement to Snake River populations and locations throughout the basin [@orme2018population].
-Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations that, joined with sex and age data collected at the adult fish trap [@Hargrove2019], provides necessary information to evaluate productivity and population viability for select Snake River Chinook Salmon and steelhead groups [@iptds2020report].
+Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations that, joined with sex and age data collected at the adult fish trap [@Hargrove2019], provides necessary information to evaluate productivity and population viability for select Snake River Chinook salmon and steelhead groups [@iptds2020report].
 
@@
 Justification and infrastructure exist for applying a modified STADEM framework for fish passing Bonneville Dam, the lowest dam on the Columbia River, or Priest Rapids Dam in the upper Columbia River.
-Both locations currently trap a sub-sample of passing Chinook Salmon and/or steelhead for biological information, and use window counts as a surrogate of true escapement.
-However, each has at least some similar problems to those observed at Lower Granite Dam, such as unaccounted observer and sampling error, night-time passage, and/or re-ascension.
+Both locations currently trap a sub-sample of passing Chinook salmon and/or steelhead for biological information, and use window counts as a surrogate of true escapement.
+However, each has at least some similar problems to those observed at Lower Granite Dam, such as unaccounted observer and sampling error, nighttime passage, and/or re-ascension.
 Certainly, estimating an unbiased total return to the entire Columbia River basin (i.e., Bonneville Dam) and Upper Columbia River with uncertainty would benefit managers and decision making.
@@
 Special thanks to Darren Ogden and staff at the Lower Granite Dam adult trap for their hard work and diligent data collection.
-Thank you to personnel at the Idaho Department of Fish and Game, particularly Paul Bunn, Tim Copeland, and Bill Schrader, for providing access to data from the adult fish trap and conceptualizing methods, and the staff at the Eagle Fish Lab for analyzing genetic samples.
+Thank you to personnel at the Idaho Department of Fish and Game, particularly Paul Bunn, Tim Copeland, and Bill Schrader, for providing access to data from the adult fish trap and conceptualizing methods, and to Matthew Campbell and staff at the Eagle Fish Genetics Laboratory for analyzing genetic samples.
 Special thanks to Columbia Basin Research staff and the Columbia River Data Access in Real Time (DART) application, and Susannah Iltis in particular.
@@
   kable(booktabs = T,
-        caption = "Window counts, and estimates of total, wild, hatchery and hatchery no-clip escapement, with coefficients of variation in parenthesis, for Chinook Salmon and steelhead from spawn years 2010 to 2019.") %>%
+        caption = "Window counts, and estimates of total, wild, hatchery and hatchery no-clip escapement, with coefficients of variation in parenthesis, for Chinook salmon and steelhead from spawn years 2010 to 2019.") %>%
   kable_styling()
@@
 
-```{r jags-fit-fig, fig.cap = "Time-series plot showing estimates of total escapement for Chinook in 2014, including raw window counts, window counts adjusted for night-time passage, trap estimates and STADEM estimates.
+```{r jags-fit-fig, fig.cap = "Time-series plot showing estimates of total escapement for Chinook in 2014, including raw window counts, window counts adjusted for nighttime passage, trap estimates and STADEM estimates.
 Gray ribbon represents the 95% credible interval for STADEM estimates."}
@@
 
-```{r night-reasc-diff-fig, fig.cap = "Night-time passage rate plotted against re-ascension rate, calculated from observed PIT tags for each week of spawn years 2010-2019.
+```{r night-reasc-diff-fig, fig.cap = "Nighttime passage rate plotted against re-ascension rate, calculated from observed PIT tags for each week of spawn years 2010-2019.
 Colors correspond to different spawn year, while the size of each point is proportional to the window count that week.

### 2020-08-27 README.Rmd
added some things to the gitignore and gitattributes, updated the README file
removed Katie Lawry from README
Moved list of journals from paper.Rmd to README. changed "Chinook salmon" to "Chinook Salmon" to match AFS guidelines.

@@
 # Please put your title here to include it in the file below.
-Title <- "Estimating total escapment past Lower Granite Dam using all possible data"
+Title <- "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
 ```
+
+**Possible journals:**
+
+First Choice:
+
+* **[North American Journal of Fisheries Management](https://afspubs.onlinelibrary.wiley.com/journal/15488675)**
+
+Second Choice:
+
+* [Transactions of the American Fisheries Society](https://afspubs.onlinelibrary.wiley.com/journal/15488659)
+
+Others:
+
+* [Canadian Journal of Fisheries and Aquatic Sciences](http://www.nrcresearchpress.com/page/cjfas/editors)
+
+* [Fisheries Research](https://www.journals.elsevier.com/fisheries-research)
+
+* [Fishery Bulletin](http://fishbull.noaa.gov/)
+
+* [Fish and Fisheries](http://onlinelibrary.wiley.com/journal/10.1111/(ISSN)1467-2979/homepage/ProductInformation.html)
 
@@
 
-> See, K.E., R.N. Kinzer and M.W. Ackerman, (2020). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
+> See, K.E., R.N. Kinzer and M.W. Ackerman (2020). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
 
@@
 
-> See, K.E., R.N. Kinzer and M.W. Ackerman, (2020). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> See, K.E., R.N. Kinzer and M.W. Ackerman (2020). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
 
@@
 Kinzer and M.W.
-Ackerman, (`r format(Sys.Date(), "%Y")`).
+Ackerman (`r format(Sys.Date(), "%Y")`).
 _Compendium of R code and data for `r Title`_.

### 2020-08-27 analysis/paper/paper.Rmd
Moved list of journals from paper.Rmd to README. changed "Chinook salmon" to "Chinook Salmon" to match AFS guidelines.

@@
 
-**Possible journals:**
-
-First Choice:
-
-* **[North American Journal of Fisheries Management](https://afspubs.onlinelibrary.wiley.com/journal/15488675)**
-
-Second Choice:
-
-* [Transactions of the American Fisheries Society](https://afspubs.onlinelibrary.wiley.com/journal/15488659)
-
-Others:
-
-* [Canadian Journal of Fisheries and Aquatic Sciences](http://www.nrcresearchpress.com/page/cjfas/editors)
-
-* [Fisheries Research](https://www.journals.elsevier.com/fisheries-research)
-
-* [Fishery Bulletin](http://fishbull.noaa.gov/)
-
-* [Fish and Fisheries](http://onlinelibrary.wiley.com/journal/10.1111/(ISSN)1467-2979/homepage/ProductInformation.html)
-
 # Introduction
@@
 
-Populations of Chinook salmon *Oncorhynchus tshawytscha* and steelhead trout *O.
+Populations of Chinook Salmon *Oncorhynchus tshawytscha* and steelhead trout *O.
 mykiss* in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric dams on the Snake and Columbia rivers) and tributary habitats [@nehlsen1991pacific; @mcclure2003large; @nwfsc2015].
-As a result, Snake River spring-summer run Chinook salmon (hereafter Chinook salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937).
-Snake River Chinook salmon and steelhead have substantial cultural, recreational, commercial and subsistence value both within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries.
-The aggregate escapement of Snake River Chinook salmon and steelhead populations, with the exception of Tucannon River, is monitored at Lower Granite Dam located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations.
+As a result, Snake River spring-summer run Chinook Salmon (hereafter Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937).
+Snake River Chinook Salmon and steelhead have substantial cultural, recreational, commercial and subsistence value both within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries.
+The aggregate escapement of Snake River Chinook Salmon and steelhead populations, with the exception of Tucannon River, is monitored at Lower Granite Dam located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations.
 Many fisheries management and conservation actions are made based on estimates of escapement at Lower Granite Dam parsed by species and origin [@NMFS2019BiOp; @NFSC2015; @NPCC2014].
@@
 
-Chinook salmon and steelhead returning to a majority of populations in the Snake River must ascend a fish ladder on Lower Granite Dam before migrating to their natal tributary spawning locations.
+Chinook Salmon and steelhead returning to a majority of populations in the Snake River must ascend a fish ladder on Lower Granite Dam before migrating to their natal tributary spawning locations.
 Fish management agencies previously used counts of fish, by species, passing an observation window within the fish ladder as a census of fish escaping past the dam.
@@
 
-Using window counts, however, as a census of Chinook salmon and steelhead passing Lower Granite Dam can be problematic as it avoids multiple sources of uncertainty and disregards known biological processes.
+Using window counts, however, as a census of Chinook Salmon and steelhead passing Lower Granite Dam can be problematic as it avoids multiple sources of uncertainty and disregards known biological processes.
 First, live (i.e., in-person) window counts only occur from April through October each year.
@@
 
-We used STADEM and three sources of data to estimate Chinook salmon and steelhead escapement at Lower Granite Dam from 2010-2019.
+We used STADEM and three sources of data to estimate Chinook Salmon and steelhead escapement at Lower Granite Dam from 2010-2019.
 Data sources included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of previously PIT tagged fish detected in the adult fish ladder.
@@
 
-Daily counts of adult Chinook salmon and steelhead passing an observation window located on the Lower Granite Dam fish ladder were the first source of data.
+Daily counts of adult Chinook Salmon and steelhead passing an observation window located on the Lower Granite Dam fish ladder were the first source of data.
 Daily counts were made and provided by the US Army Corps of Engineers, and when summed, provide an estimate of the number of fish ascending and passing (i.e., escapement) Lower Granite Dam each season.
@@
 Scale samples were used to estimate age [@Wright2015] and genetic tissue samples were used to determine sex [@Campbell2012a] and estimate the origin of wild fish via genetic stock identification (e.g., @Hargrove2019).
-Starting with adults spawning in 2013, every unclipped Chinook salmon and steelhead trapped at LGR was genotyped to simplify collaborative logistics and better estimate the proportion of phenotypically wild fish of truly hatchery origin.
+Starting with adults spawning in 2013, every unclipped Chinook Salmon and steelhead trapped at LGR was genotyped to simplify collaborative logistics and better estimate the proportion of phenotypically wild fish of truly hatchery origin.
 Previous to that, only fish determined to be wild in origin at the trap were sampled for scale and genetics.
@@
 
-The last source of data was observations of PIT tagged adult Chinook salmon and steelhead at detection sites located in the Lower Granite Dam fish ladder.
+The last source of data was observations of PIT tagged adult Chinook Salmon and steelhead at detection sites located in the Lower Granite Dam fish ladder.
 These observations provide estimates of 1) a trapping rate, 2) the proportion of fish passing during nighttime hours, and 3) the proportion of fish that ascend the fish ladder multiple times (i.e., the re-ascension rate).
@@
 
-A trap rate estimate was derived using mark-recapture methods and PIT tag observations of both Chinook salmon and steelhead at Lower Granite Dam adult detection sites.
+A trap rate estimate was derived using mark-recapture methods and PIT tag observations of both Chinook Salmon and steelhead at Lower Granite Dam adult detection sites.
 The "mark" group included all tags detected in the adult trap and the "capture" group included tags observed to cross the weir at the upstream end of the fish ladder as adults left the passage system.
@@
 
-Finally, we applied STADEM to data from Lower Granite Dam for Chinook salmon and steelhead returning to the Snake River during spawn years 2010 to 2019.
+Finally, we applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead returning to the Snake River during spawn years 2010 to 2019.
 Window counts for both species were accessed from DART via functions within STADEM.
-For Chinook salmon, a spawn year refers to adults that migrate past the dam prior to August 17 each year and spawn that late summer and fall.
+For Chinook Salmon, a spawn year refers to adults that migrate past the dam prior to August 17 each year and spawn that late summer and fall.
 For steelhead, the spawn year is defined as steelhead that migrate past Lower Granite Dam starting July 1 the previous year and prior to June 30 of the given year (e.g., spawn year 2017 steelhead migrate past Lower Granite between July 1, 2016 and June 30, 2017).
@@
 
-We applied STADEM to data from Lower Granite Dam for Chinook salmon and steelhead for spawn years 2010–2019.
+We applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead for spawn years 2010–2019.
 Estimates of total escapement, as well as estimates of wild, hatchery, and hatchery no-clip estimates are presented in Table \@ref(tab:lgr-tab).
@@
 In doing so, we demonstrated that at Lower Granite Dam, nighttime passage and re-ascension rates do not always offset each other, and assuming they do will lead to biased escapement estimates in some years.
-With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall Chinook salmon, Pacific lamprey *Lampetra tridentata*), or elsewhere; provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run, and tag observation or detection infrastructure (e.g., a PIT tag detection array or similar) exists.
+With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall Chinook Salmon, Pacific lamprey *Lampetra tridentata*), or elsewhere; provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run, and tag observation or detection infrastructure (e.g., a PIT tag detection array or similar) exists.
 Our state-space model combined multiple imperfect sources of data to reduce bias in adult escapement estimates and provided more reasonable estimates of uncertainty.
@@
 
-Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook salmon and steelhead past Lower Granite Dam, and returning to tributary or population specific spawning areas [@iptds2020report; @kinzer2020report].
+Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary or population specific spawning areas [@iptds2020report; @kinzer2020report].
 Estimates of escapement at Lower Granite Dam, by species and origin, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates (e.g., @Camacho2017, @Schrader2013) that are important for fisheries management and productivity monitoring of wild populations.
@@
 Estimates of fish passing the dam are then combined with estimated movement or transition probabilities based on PIT tag observations at instream PIT tag detection systems throughout the Snake River basin, similar to @waterhouse2020bayesian, to estimate escapement to Snake River populations and locations throughout the basin [@orme2018population].
-Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations that, joined with sex and age data collected at the adult fish trap [@Hargrove2019], provides necessary information to evaluate productivity and population viability for select Snake River Chinook salmon and steelhead groups [@iptds2020report].
+Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations that, joined with sex and age data collected at the adult fish trap [@Hargrove2019], provides necessary information to evaluate productivity and population viability for select Snake River Chinook Salmon and steelhead groups [@iptds2020report].
 
@@
 Justification and infrastructure exist for applying a modified STADEM framework for fish passing Bonneville Dam, the lowest dam on the Columbia River, or Priest Rapids Dam in the upper Columbia River.
-Both locations currently trap a sub-sample of passing Chinook salmon and/or steelhead for biological information, and use window counts as a surrogate of true escapement.
+Both locations currently trap a sub-sample of passing Chinook Salmon and/or steelhead for biological information, and use window counts as a surrogate of true escapement.
 However, each has at least some similar problems to those observed at Lower Granite Dam, such as unaccounted observer and sampling error, nighttime passage, and/or re-ascension.

### 2020-08-27 analysis/paper/paper.Rmd
added abstract.
deleted a few comments in the text.

Slight revision to the methods description of which fish had genetic samples taken.

@@
 abstract: |
-  This is the abstract
+  Accurate estimates of the number of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successfully management of fisheries in that area.
+Here we describe a state-space model that estimates such escapement past a dam by using window counts, PIT tag observations and data from an adult fish trap, accounting for issues such as night-time passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate.
+We tested this approach using a simulation framework that mimicked several levels of observation error, differences between night-time passage and re-ascension rates and the possibility of the adult trap being closed for some period of time, and found it to produce unbiased estimates across all tested scenarios.
+We also applied this model to data from the Lower Granite dam to produces estimates of wild, clipped hatchery and unclipped hatchery spring-summer run Chinook Salmon and steelhead from spawn years 2010-2019.
 keywords: |
-  escapement; Chinook; steelhead; Lower Granite Dam; state-space model; Bayesian model
+  escapement; Chinook; steelhead; salmon; Lower Granite Dam; state-space model; Bayesian model
 highlights: |
@@
 To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates.
+We then applied this model to Chinook Salmon and steelhead returns at Lower Granite dam for spawn years 2010 - 2019.
 The STADEM model combines multiple imperfect sources of data to reduce bias in escapement estimates and provides improved estimates of uncertainty.
-
-<!-- RK comment: the following sentence is better left in the discussion and can be removed from this paragraph.
--- Additionally, the STADEM model framework can be applied elsewhere to estimate aggregate and group escapement, provided a passage barrier containing multiple counting and observational mechanisms exist, such as, a counting window, an adult trap sampling a portion of the run, and PIT tag antenna infrastructure.
--->
 
@@
 Scale samples were used to estimate age [@Wright2015] and genetic tissue samples were used to determine sex [@Campbell2012a] and estimate the origin of wild fish via genetic stock identification (e.g., @Hargrove2019).
-Starting with adults spawning in 2013, every unclipped Chinook Salmon and steelhead trapped at LGR was genotyped to simplify collaborative logistics and better estimate the proportion of phenotypically wild fish of truly hatchery origin.
-Previous to that, only fish determined to be wild in origin at the trap were sampled for scale and genetics.
+Prior to 2013, only fish determined to be wild in origin at the trap were sampled for scale and genetics.
+Starting in 2013, every unclipped Chinook Salmon and steelhead trapped at LGR was genotyped to simplify collaborative logistics and better estimate the proportion of phenotypically wild fish of truly hatchery origin.
 @Camacho2017 provide further details on trap sample rates and valid sample selection.
@@
 Therefore, we use the mark-recapture estimate of the trap rate to estimate a true trapping rate.
-
-<!--The true trapping rate was estimated as the recorded time of the trap being open does not always reflect the true proportion of fish that are captured in the trap, due to trap malfunctions, sort-by-code fish opening the trap more frequently than expected, or process error (Rick Orme, *per.
-comm.*).
-Therefore, we use the mark-recapture estimate of trap rate.
--->
-
-<!-- Don't you think it's important to explain why we need to estimate the trap rate?
-Because otherwise after reading the description of the adult fish trap data, I would assume the trap rate is known -->
 
@@
 Coefficients of variation ranged from 2.7-24.8% for wild fish, 2.2-10.5% for hatchery fish, 3.6-11% for hatchery no-clip fish and 2.1-10.4% for total unique fish past Lower Granite Dam.
-
-<!-- 
-All of the high values were from 2019, and the ranges drop quite a bit if we exclude that year. Max is more like 6%. Worth trying to mention that?
-
-MA: Maybe worth including in the Discussion?
--->
+Excluding 2019, which had the lowest escapement estimates for both species, the maximum CVs were 6.6% for wild fish, 5.6% for clipped hatchery fish, 8% for unclipped hatchery fish and 5.2% for total unque fish.
 
@@
   scale_color_brewer(palette = 'Set1') +
+  # scale_color_grey(start = 0.9,
+  #                  end = 0.3) +
   geom_vline(xintercept = 4,

### 2020-08-27 analysis/paper/paper.Rmd
re-knit all 3 versions of output

@@
 output:
+    bookdown::pdf_document2:
+      fig_caption: yes
+      number_sections: FALSE
+      toc: no
+      geometry: margin=1in
+      latex_engine: "xelatex"
+      fig_width: 7
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks2.lua
+      - --lua-filter=../templates/pagebreak.lua
+      includes:
+        in_header: ../templates/header_manuscript.tex
     bookdown::html_document2:
@@
       - --lua-filter=../templates/pagebreak.lua
-    bookdown::pdf_document2:
-      fig_caption: yes
-      number_sections: FALSE
-      toc: no
-      geometry: margin=1in
-      latex_engine: "xelatex"
-      fig_width: 7
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks2.lua
-      - --lua-filter=../templates/pagebreak.lua
-      includes:
-        in_header: ../templates/header_manuscript.tex
 fontsize: 12pt

### 2020-09-01 inst/scripts/STADEM_model_runs.R
added some code to drop intermediate objects from STADEM loop

@@ -92,5 +92,12 @@ for(i in 1:length(species)){
     save(stadem_mod, stadem_list,
          file = paste0(stademFolder,'/LGR_STADEM_', spp, '_', yr, '.rda'))
 
+    rm(incl_jacks,
+       start_date,
+       end_date,
+       stadem_list,
+       jags_data_list,
+       stadem_mod)
+
   } # close j loop
 } # close i loop

### 2020-09-01 analysis/paper/paper.Rmd
Better described how the trap rate was estimated and fixed a typo in the caption of figure showing relative bias in simulations.

@@
 Those estimates, with associated uncertainties, were then incorporated into the model as informed priors, while the model estimated the true trap rate based on all the data, including trap and window counts.
-We want to estimate the "true" trapping rate as the set trapping rate, the recorded time that the trap is open to trap adults, does not always reflect the true proportion of fish that are captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error, among others.
+We want to estimate the "true" trapping rate because the set trapping rate (i.e.
+the recorded time that the trap is open to trap adults) does not always reflect the true proportion of fish that are captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error, among others.
 Therefore, we use the mark-recapture estimate of the trap rate to estimate a true trapping rate.
+
+<!-- Fix the first two sentences below to make it clear the rates are estimated as a binomial with a rate that has a random walk in logit space -->
 
@@
 
-We estimated the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for nighttime passage, and fallback/re-ascension rates using a state-space modeling approach [@royle2008hierarchical] implemented in the STADEM package for the R statistical software [@R-Core-Team2020].
+We estimated the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for nighttime passage and fallback/re-ascension rates using a state-space modeling approach [@royle2008hierarchical] implemented in the STADEM package for the R statistical software [@R-Core-Team2020].
 We assumed that the window counts and the estimates from the trap (fish in the trap divided by trap rate that week) were generated by processes with observation error.
@@
 
-```{r rel-bias-lgd, fig.cap = "'Boxplots show relative bias of STADEM estimates for wild escapement across various scenarios.
+```{r rel-bias-lgd, fig.cap = "Boxplots show relative bias of STADEM estimates for wild escapement across various scenarios.
 The boxes contain 50% of the simulations, whiskers contain 95% of the simulations, and points are outliers."}

### 2020-09-01 analysis/paper/appendices/Appendix_STADEM_descrip.Rmd
changed how the trap rate is derived (now done within the STADEM framework, rather than outside and brought in with priors).
Improved some of the subscripts to make them more clear.

Changed the priors to make them more uniformative

@@
 The number of fish caught in the trap, $Y^T_t$, for week $t$ is modeled as a binomial process based on the unknown true trap rate that week, $\nu_t$, and the unknown true number of fish crossing the dam that week, $X_t$.
-The true trap rate is estimated from a beta distribution with previously estimated parameters, $\hat{\alpha_t}$ and $\hat{\beta_t}$, informed by the mark-recapture estimate of the trap rate.
+The estimate of the true weekly trap rate is derived based on previously PIT-tagged spring/summer Chinook and steelhead who are crossing LGD that week, using a Lincoln-Peterson mark-recapture model [@Seber2002].
+The fish, from both species, caught in the trap that week are considered the "mark" group ($m_t$), and all the previously PIT tagged fish who are detected at the upper end of the LGD fish ladder that week are considered the second capture group, $M_t$ (which includes recaptures of the the "marked" fish).
+The proportion of previously marked fish that are caught in the adult trap, $m/M$ can be modeled with a binomial distribution using the same trap rate parameter, $\nu_t$.
+Although the group of previously PIT tagged fish is not assumed to be representative of the overall run, the rate at which they are caught in the trap should be the same rate that the overall run experiences.
+The more tagged fish crossing the dam in a particular week, the more certain we can be of the true trap rate.
 
@@
 \begin{aligned}
-  \nu_t &\sim \text{Beta} \left(\hat{\alpha_t}, \hat{\beta_t} \right) \\
+  m_t &\sim \text{Bin} \left( \nu_t, M_t \right) \\
   Y^T_t &\sim \text{Bin} \left( \nu_t, X_t \right) \\
@@
 $$
-
-The estimate of the weekly trap rate is derived based on previously PIT-tagged spring/summer Chinook and steelhead who are crossing LGD that week.
-The fish caught in the trap that week are considered the "mark" group, and all the PIT tagged fish who are detected at the upper end of the LGD fish ladder that week are considered the second capture group (which includes recaptures of the the "marked" fish).
-From this data, we can estimate the total number of previoiusly PIT tagged fish crossing LGD that week, and the proportion of those that are caught in the trap is the weekly trap rate.
-The uncertainty in estimates of total fish is translated into uncertainty in the trap rate, and this is then summarised with a beta distribution ($\text{Beta} \left(\hat{\alpha_t}, \hat{\beta_t} \right)$) that becomes an input into STADEM.
-Although the group of previously PIT tagged fish is not assumed to be representative of the overall run, the rate at which they are caught in the trap should be the same rate that the overall run experiences.
 
@@
 
-After estimating the total number of fish to have crossed Lower Granite each week, $X_t$, the total must be further refined into the number of wild fish, $X_{w,t}$, hatchery fish, $X_{h,t}$ and hatchery no-clip fish, $X_{hnc, t}$.
+After estimating the total number of fish to have crossed Lower Granite each week, $X_t$, the total must be further refined into the number of wild fish, $X_{w,t}$, hatchery clipped fish, $X_{hc,t}$ and hatchery no-clip fish, $X_{hnc, t}$.
 This is done by estimating a weekly origin proportion vector, $\omega_t$ based on the random sample of fish caught in trap that week, $Y^T_t$.
-The observed number of wild, $Y^T_{w,t}$, hatchery, $Y^T_{h,t}$, and hatchery no-clip, $Y^T_{hnc,t}$, fish caught in the trap that week is assumed to come from a multinomial distribution with probability vector $\omega_t$.
-The log-odds ratio of the proportions in $\omega_t$, in relation to the proportion of hatchery fish, $\omega_{h,t}$ is modeled as a random walk, so it can change through time.
-This allows the proportions of wild, hatchery and hatchery no-clip fish to shift throughout the season, based on the data available from the fish trap.
+The observed number of wild, $Y^T_{w,t}$, hatchery clipped, $Y^T_{hc,t}$, and hatchery no-clip, $Y^T_{hnc,t}$, fish caught in the trap that week is assumed to come from a multinomial distribution with probability vector $\omega_t$.
+The log-odds ratio of the proportions in $\omega_t$, in relation to the proportion of clipped hatchery fish, $\omega_{hc,t}$ is modeled as a random walk, so it can change through time.
+This allows the proportions of wild, hatchery clipped and hatchery no-clip fish to shift throughout the season, based on the data available from the fish trap.
 
@@
 \begin{aligned}
-  \left( Y^T_{w,t}, Y^T_{h,t}, Y^T_{hnc,t} \right) &\sim \text{Multinom} \left( \omega_t, Y^T_t \right) \\
+  \left( Y^T_{w,t}, Y^T_{hc,t}, Y^T_{hnc,t} \right) &\sim \text{Multinom} \left( \omega_t, Y^T_t \right) \\
   \omega_t &= \frac{\exp(\phi_t)}{\sum{\exp(\phi_t})} \\
-  \phi_{w,t} &= \ln \left( \frac{\omega_{w,t}}{\omega_{h,t}} \right) \\
-  \phi_{hnc,t} &= \ln \left( \frac{\omega_{hnc,t}}{\omega_{h,t}} \right) \\
-  \phi_{h,t} &= 0 \\
+  \phi_{hc,t} &= 0 \\
+  \phi_{w,t} &= \ln \left( \frac{\omega_{w,t}}{\omega_{hc,t}} \right) \\
+  \phi_{hnc,t} &= \ln \left( \frac{\omega_{hnc,t}}{\omega_{hc,t}} \right) \\
   \phi_{w,t} &= \phi_{w, t-1} + d_{w,t} \\
   \phi_{hnc,t} &= \phi_{hnc, t-1} + d_{hnc,t} \\
-  d_t &\sim \mathcal{N}(0, \sigma^2_\omega) \\
+  d_{w,t} &\sim \mathcal{N}(0, \sigma^2_\omega) \\
+  d_{hnc,t} &\sim \mathcal{N}(0, \sigma^2_\omega) \\
 \end{aligned}
@@
 \begin{aligned}
-  \begin{bmatrix} X_{w,t} \\ X_{h,t} \\ X_{hnc,t} \end{bmatrix}
-  &= X_t * (1 - \eta_t) * \begin{bmatrix} \omega_{w,t} \\ \omega_{h,t} \\ \omega_{hnc,t} \end{bmatrix}
+  \begin{bmatrix} X_{w,t} \\ X_{hc,t} \\ X_{hnc,t} \end{bmatrix}
+  &= X_t * (1 - \eta_t) * \begin{bmatrix} \omega_{w,t} \\ \omega_{hc,t} \\ \omega_{hnc,t} \end{bmatrix}
 \end{aligned}
@@
 The model was fit using the JAGS program [@R-rjags], run with R software [@R-Core-Team2020].
-Uninformative priors were used for $\sigma_X, \sigma_\eta, \sigma_\theta, \sigma_\omega$ and $\ln(X_1)$ (Uniform(0,10)), as well as logit($\eta_1$) and logit($\theta_1$) ($\mathcal{N}(0, 1000)$), and finally $\phi_{w,1}$ and $\phi_{hnc,1}$ ($\mathcal{N}(0, 100)$).
+Variance parameters $\sigma_X, \sigma_\eta, \sigma_\theta$, and $\sigma_\omega$, as well as the initial abundance, $X_1$, and the overdispersion parameter of the negative binomial, $r$, were given half-Cauchy priors with mean of 0 and scale of 100.
+The initial day-time passage and re-ascension rates, $\theta_1$ and $\eta_1$ were given Uniform(0,1) priors.
+Finally, $\phi_{w,1}$ and $\phi_{hnc,1}$ were given priors of Uniform(-3,3), in an effort to make $\omega_1$ as uniformative as possible.

### 2020-09-01 analysis/paper/paper.Rmd
improved the description of how the trap rate is estimated.

@@
 
-A trap rate estimate was derived using mark-recapture methods and PIT tag observations of both Chinook Salmon and steelhead at Lower Granite Dam adult detection sites.
-The "mark" group included all tags detected in the adult trap and the "capture" group included tags observed to cross the weir at the upstream end of the fish ladder as adults left the passage system.
-Using a mark-recapture model with differing capture probabilities, we estimated the trap rate on a weekly basis.
-Those estimates, with associated uncertainties, were then incorporated into the model as informed priors, while the model estimated the true trap rate based on all the data, including trap and window counts.
+A trap rate estimate was derived using Lincoln-Peterson mark-recapture methods and PIT tag observations of both Chinook Salmon and steelhead at Lower Granite Dam adult detection sites.
+The "mark" group included all tags (of both species) detected in the adult trap and the "capture" group included tags observed to cross the weir at the upstream end of the fish ladder as adults left the passage system.
+The proportion of the tags detected at the weir that were also caught in the trap each week was assumed to reflect the same trap rate that all adults of the target species experienced as they crossed the ladder.
+The sample size of previously tagged fish detected at Lower Granite Dam influences the uncertainty in that trap rate, which is also informed by the number of adults caught in the trap and the window counts.
 We want to estimate the "true" trapping rate because the set trapping rate (i.e.
@@
 
-<!-- Fix the first two sentences below to make it clear the rates are estimated as a binomial with a rate that has a random walk in logit space -->
-
-The nighttime passage rate was estimated as the count of tags that migrated through the fish ladder during non-window observation hours divided by the total number of tags passing the fish ladder, and was calculated on a weekly basis.
-The re-ascension rate was the count of tags observed passing the upstream most detection sites in the adult fish ladder (i.e., passing the dam) and later detected re-entering the downstream end of the fish ladder at a later time divided by the total number of tags leaving the fish ladder.
+The nighttime passage rate was based on the count of tags that migrated through the fish ladder during non-window observation hours and the total number of tags passing the fish ladder, and was estimated on a weekly basis.
+The re-ascension rate incorporated the count of tags observed passing the upstream most detection sites in the adult fish ladder (i.e., passing the dam) and later detected re-entering the downstream end of the fish ladder at a later time andthe total number of tags leaving the fish ladder.
 Previously, we looked for differences in nighttime passage and re-ascension rates estimated using wild fish only, versus combining hatchery and wild fish together, and found no difference.

### 2020-09-01 analysis/paper/appendices/Appendix_simulation_details.Rmd
re-ran all the STADEM year/species combinations with most up-to-date version of STADEM

--- 2020-08-07 analysis/paper/appendices/Appendix_simulation_details.Rmd
@@
 
-Escapement is then randomly divided across a set number of populations, $n$, by randomly drawing proportions, $\phi_{j,p}$, of origin group $j$ in each poplulation $p$ using a dirichlet $\mathcal{PDF}$.
-The dirichlet function is parameterized from a vector, $\eta_j$, containing 1's and 0's designating populations with origin $j$ fish returning.
-$\eta_j$ originates from a bernoulli $\mathcal{PDF}$ and the proportion of populations with each origin, $\tau_j$.
+Escapement of each origin is then randomly divided across a set number of populations, $n$, by randomly drawing proportions, $\phi_{j,p}$, of origin group $j$ in each population $p$ using a Dirichlet $\mathcal{PDF}$.
+The Dirichlet function is parameterized from a vector, $\eta_j$, containing 1's and 0's designating populations with origin $j$ fish returning.
+$\eta_j$ originates from a Bernoulli $\mathcal{PDF}$ and the proportion of populations with each origin, $\tau_j$.
 Wild fish are assumed to be in all populations; $\tau_w = 1.0$.
@@
 
-To examine the sensitivites of models to different fish behavior and dam operational scenarios, seven addtional attributes are randomly assigned to each individual fish.
-Each attribute is randomly assigned a TRUE/FALSE using a bernoulli $\mathcal{PDF}$ and a fixed probability parameter.
+To examine the sensitivities of models to different fish behavior and dam operational scenarios, seven additional attributes are randomly assigned to each individual fish.
+Each attribute is randomly assigned a TRUE/FALSE using a Bernoulli $\mathcal{PDF}$ and a fixed probability parameter.
 Fish passage during the day-time (i.e., during periods of window operation) is modeled using one minus the night-time passage rate ($1 - \nu$).
@@
   day_{i} &\sim \text{Bern}(1-\nu) \\
-  window_{i|d = TRUE} &\sim \text{Bern}(\gamma) \\
+  window_{i} &\sim \text{Bern}(\gamma \times day_i) \\
   trapped_{i} &\sim \text{Bern}(\delta_t) \\
   tagged_{i} &\sim \text{Bern}(\lambda) \\
-  ladder_{i|m = TRUE} &\sim \text{Bern}(\kappa) \\  
+  ladder_{i} &\sim \text{Bern}(\kappa \times tagged_i) \\  
   fallback_{i} &\sim \text{Bern}(\psi) \\
-  re-ascend_{i|f = TRUE} &\sim \text{Bern}(\rho)
+  re-ascend_{i} &\sim \text{Bern}(\rho \times fallback_i)
 \end{aligned}
@@
 
-Simulation parameters for model evaluations were set to mimic typical ecapement of spring/summer Chinook to LGD with similar origin proportions, marking rates and run timing as those observed from return years 2010 - 2015.
+Simulation parameters for model evaluations were set to mimic typical escapement of spring/summer Chinook Salmon to LGD with similar origin proportions, marking rates and run timing as those observed from return years 2010 - 2015.
 Escapement of each origin ($N_j$) was set at 25,000 wild, 70,000 hatchery and 5,000 hatchery no-clips spread randomly across 25 populations ($n$).
@@
 
-For the specific simulated scenarios, we were interested in STADEM model estimates of origin specific escapment from the combinations of two seperate trapping rates, two fallback, re-ascension and night-passage combinations and three window count error rates; resulting in twelve different scenarios.
-First, trapping rates were set static at 0.15 across all weeks for six scenarios to mimic an optimium trap operation for an expected return of 25,000 wild fish (i.e., trap $\approx$ 4,000 wild fish).
+For the specific simulated scenarios, we were interested in STADEM model estimates of origin specific escapement from the combinations of two separate trapping rates, two fallback, re-ascension and night-passage combinations and three window count error rates; resulting in twelve different scenarios.
+First, trapping rates were set static at 0.15 across all weeks for six scenarios to mimic an optimum trap operation for an expected return of 25,000 wild fish (i.e., trap $\approx$ 4,000 wild fish).
 For the remaining six scenarios, trapping rates for weeks 30, 31 and 32 (i.e., July $22^{nd}$ to August $11^{th}$) were changed to 0.00 to test STADEM sensitivities to potential trap shut downs similar to those observed in 2013, 2014 and 2015 [@Ogden2014; @Ogden2016; @Ogden2016b].
@@
 A potential 5.0% weekly bias was determined from PIT-tag data and within the range of observed weekly difference for return years 2010 - 2015 (Figure \@ref(fig:night-reasc-diff-fig)).
-Finally, we desired to test the sensitivites of STADEM to potential rates of window count error; 0%, 5% and 10% [@Hatch1994].
+Finally, we desired to test the sensitivities of STADEM to potential rates of window count error; 0%, 5% and 10% [@Hatch1994].
 Error at the window was simulated as unbiased (i.e., expect varying high and low counts to cancel each other) and is introduced to the sum of daily window counts using a normal $\mathcal{PDF}$.

### 2020-09-01 analysis/paper/paper.Rmd
corrected a typo in the abstract
made the legend for the time-series plot have 2 rows

@@
 output:
+    bookdown::html_document2:
+      fig_caption: yes
+      number_sections: FALSE
+      toc: no
+      fig_width: 7
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
+    bookdown::word_document2:
+      fig_caption: yes
+      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+      number_sections: FALSE
+      toc: no
+      always_allow_html: true
+      fig_width: 7
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
     bookdown::pdf_document2:
@@
         in_header: ../templates/header_manuscript.tex
-    bookdown::word_document2:
-      fig_caption: yes
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
-      number_sections: FALSE
-      toc: no
-      always_allow_html: true
-      fig_width: 7
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
-    bookdown::html_document2:
-      fig_caption: yes
-      number_sections: FALSE
-      toc: no
-      fig_width: 7
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
 fontsize: 12pt
@@
 We tested this approach using a simulation framework that mimicked several levels of observation error, differences between night-time passage and re-ascension rates and the possibility of the adult trap being closed for some period of time, and found it to produce unbiased estimates across all tested scenarios.
-We also applied this model to data from the Lower Granite dam to produces estimates of wild, clipped hatchery and unclipped hatchery spring-summer run Chinook Salmon and steelhead from spawn years 2010-2019.
+We also applied this model to data from the Lower Granite dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring-summer run Chinook Salmon and steelhead from spawn years 2010-2019.
 keywords: |
@@
 Estimates of total escapement, as well as estimates of wild, hatchery, and hatchery no-clip estimates are presented in Table \@ref(tab:lgr-tab).
-Coefficients of variation ranged from 2.7-24.8% for wild fish, 2.2-10.5% for hatchery fish, 3.6-11% for hatchery no-clip fish and 2.1-10.4% for total unique fish past Lower Granite Dam.
-Excluding 2019, which had the lowest escapement estimates for both species, the maximum CVs were 6.6% for wild fish, 5.6% for clipped hatchery fish, 8% for unclipped hatchery fish and 5.2% for total unque fish.
+Coefficients of variation ranged from 2.5-16% for wild fish, 2.3-9.4% for hatchery fish, 3.4-24.2% for hatchery no-clip fish and 2.2-9.8% for total unique fish past Lower Granite Dam.
+Excluding 2019, which had the lowest escapement estimates for both species, the maximum CVs were 7.1% for wild fish, 4.8% for clipped hatchery fish, 7.8% for unclipped hatchery fish and 4.7% for total unque fish.
 
@@
 plotTimeSeries(stadem_mod,
-               weeklyData = stadem_list$weeklyData)
+               weeklyData = stadem_list$weeklyData) +
+  guides(color = guide_legend(nrow = 2),
+         shape = guide_legend(nrow = 2))
 ```

### 2020-09-04 analysis/paper/appendices/Appendix_simulation_details.Rmd
RK edits to appendix after EB's comments

@@
 Escapement of each origin is then randomly divided across a set number of populations, $n$, by randomly drawing proportions, $\phi_{j,p}$, of origin group $j$ in each population $p$ using a Dirichlet $\mathcal{PDF}$.
-The Dirichlet function is parameterized from a vector, $\eta_j$, containing 1's and 0's designating populations with origin $j$ fish returning.
-$\eta_j$ originates from a Bernoulli $\mathcal{PDF}$ and the proportion of populations with each origin, $\tau_j$.
+The Dirichlet function is parameterized from a vector, $\zeta_j$, containing 1's and 0's designating populations with origin $j$ fish returning.
+$\zeta_j$ originates from a Binomial $\mathcal{PDF}$ with $n$ populations, and the proportion of populations with each origin, $\tau_j$.
 Wild fish are assumed to be in all populations; $\tau_w = 1.0$.
@@
 \begin{aligned}
-  \left[\eta_{j,p=1},  ... , \eta_{j,p=n}\right] &\sim \text{Bern}(\tau_j) \\
-  \left[\phi_{j,p=1},  ... , \phi_{j,p=n}\right] &\sim \text{Dir}(\eta_{j,p=1},  ... , \eta_{j,p=n}) \\
+  \left[\zeta_{j,p=1},  ... , \zeta_{j,p=n}\right] &\sim \text{Binomial}(n, \tau_j) \\
+  \left[\phi_{j,p=1},  ... , \phi_{j,p=n}\right] &\sim \text{Dir}(\zeta_{j,p=1},  ... , \zeta_{j,p=n}) \\
   N_{j,p} &= N_j * \phi_{j,p} \\
@@
 Finally, we desired to test the sensitivities of STADEM to potential rates of window count error; 0%, 5% and 10% [@Hatch1994].
-Error at the window was simulated as unbiased (i.e., expect varying high and low counts to cancel each other) and is introduced to the sum of daily window counts using a normal $\mathcal{PDF}$.
+To simulate window count error, we assumed the observed daily count was a random variable from a normal distribution with a mean equal to the true daily count, and a standard deviation equal to the applied error rate (i.e., 0%, 5%, 10%) multiplied by the true daily count.
+This method simulated observed counts as unbiased, and allowed for possible under and overcounts at the window.

### 2020-09-04 analysis/paper/paper.Rmd
RK edits to appendix after EB's comments

@@
       - --lua-filter=../templates/pagebreak.lua
-    bookdown::word_document2:
-      fig_caption: yes
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
-      number_sections: FALSE
-      toc: no
-      always_allow_html: true
-      fig_width: 7
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
     bookdown::pdf_document2:
@@
         in_header: ../templates/header_manuscript.tex
+    bookdown::word_document2:
+      fig_caption: yes
+      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+      number_sections: FALSE
+      toc: no
+      always_allow_html: true
+      fig_width: 7
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
 fontsize: 12pt

### 2020-09-05 analysis/paper/paper.Rmd
updated with new STADEM runs for 2019. Changed the results about max CV of escapement estimates.

@@
 output:
-    bookdown::html_document2:
-      fig_caption: yes
-      number_sections: FALSE
-      toc: no
-      fig_width: 7
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
     bookdown::pdf_document2:
@@
         in_header: ../templates/header_manuscript.tex
+    bookdown::html_document2:
+      fig_caption: yes
+      number_sections: FALSE
+      toc: no
+      fig_width: 7
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
     bookdown::word_document2:
@@
 Estimates of total escapement, as well as estimates of wild, hatchery, and hatchery no-clip estimates are presented in Table \@ref(tab:lgr-tab).
-Coefficients of variation ranged from 2.5-16% for wild fish, 2.3-9.4% for hatchery fish, 3.4-24.2% for hatchery no-clip fish and 2.2-9.8% for total unique fish past Lower Granite Dam.
-Excluding 2019, which had the lowest escapement estimates for both species, the maximum CVs were 7.1% for wild fish, 4.8% for clipped hatchery fish, 7.8% for unclipped hatchery fish and 4.7% for total unque fish.
+Coefficients of variation ranged from 2.5-7.1% for wild fish, 2.3-5.1% for clipped hatchery fish, 3.4-8.8% for hatchery no-clip fish and 2.2-4.7% for total unique fish past Lower Granite Dam.
 

### 2020-09-05 analysis/paper/paper.Rmd
incorporated new STADEM results and changed results on CV of escapement estimates in manuscript.

@@
       - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
+    bookdown::html_document2:
+      fig_caption: yes
+      number_sections: FALSE
+      toc: no
+      fig_width: 7
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks2.lua
       - --lua-filter=../templates/pagebreak.lua

### 2020-09-11 analysis/paper/appendices/Appendix_STADEM_descrip.Rmd
incorporated suggestions from John Hargrove into manuscript

@@
 The estimate of the true weekly trap rate is derived based on previously PIT-tagged spring/summer Chinook and steelhead who are crossing LGD that week, using a Lincoln-Peterson mark-recapture model [@Seber2002].
-The fish, from both species, caught in the trap that week are considered the "mark" group ($m_t$), and all the previously PIT tagged fish who are detected at the upper end of the LGD fish ladder that week are considered the second capture group, $M_t$ (which includes recaptures of the the "marked" fish).
+The fish, from both species, caught in the trap that week are considered the "mark" group ($m_t$), and all the previously PIT tagged fish who are detected at the upper end of the LGD fish ladder that week are considered the second "capture" group, $M_t$ (which includes recaptures of the the "marked" fish).
 The proportion of previously marked fish that are caught in the adult trap, $m/M$ can be modeled with a binomial distribution using the same trap rate parameter, $\nu_t$.
@@
 The number of fish counted at the window, $Y^W_t$, is modeled as a (potentially) over-dispersed negative binomial process, with an expected value of $X_t^{day}$, the number of fish crossing the dam while the window is open.
-This is simply the total number of fish crossing that week, $X_t$, multiplied by the propoortion of fish crossing while the window is open for counting, $\theta_t$, calculated on a weekly basis.
+This is simply the total number of fish crossing that week, $X_t$, multiplied by the proportion of fish crossing while the window is open for counting, $\theta_t$, calculated on a weekly basis.
 In the formula below, $p_t$ is the proportion of fish observed at the window and $r$ is the shape parameter.
@@
 During peak run times, when lots of fish are crossing LGD, estimates based on trap data and trap rates will be more precise, while estimates from the window counts may have more observation error due to so many fish passing the window.
-For weeks when the trap is down, STADEM relies exclusively on the window counts and night-time passage data, but there will be more uncertainty in the estimates.
+For weeks when the trap is down, STADEM relies exclusively on the window counts and nighttime passage data, but there will be more uncertainty in the estimates.
 
@@
 
-There are two other processes that must be accounted for, first, the proportion of fish that cross the dam while the window is closed for counting (night-time passage rate), and the second, the proportion of fish that are crossing the dam multiple times (re-ascension rate) and therefore potentially double-counted.
+There are two other processes that must be accounted for, first, the proportion of fish that cross the dam while the window is closed for counting (nighttime passage rate), and the second, the proportion of fish that are crossing the dam multiple times (re-ascension rate) and therefore potentially double-counted.
 Both rates can be estimated from previously PIT tagged fish that are crossing the dam each week.
 
-The proportion of fish passing the window during non-operational hours, night-time passage rate, is just the complement of the rate of fish passing during the day when the window is operating.
+The proportion of fish passing the window during non-operational hours, nighttime passage rate, is just the complement of the rate of fish passing during the day when the window is operating.
 The daytime passage rate for week $t$, $\theta_t$, is modeled as a random walk process and estimated from a binomial distribution based on the number of PIT tags observed to cross the dam during operational hours, $y^{day}_t$, and the total number of PIT tags observed to cross the dam at any point that week, $N_t$ [@Shumway2010].
@@
 
-The number of total fish crossing Lower Granite differs from the number of unique fish crossing Lower Granite because some fish fall back and re-ascend the dam.
+The number of total fish crossing Lower Granite differs from the number of unique fish crossing Lower Granite because some fish fallback and re-ascend the dam.
 These fish are potentially double-counted at the window, and have the potential to be caught in the fish trap more than once.

### 2020-09-11 analysis/paper/paper.Rmd
incorporated suggestions from John Hargrove into manuscript

--- 2020-09-05 analysis/paper/paper.Rmd
@@
       - --lua-filter=../templates/pagebreak.lua
+    bookdown::html_document2:
+      fig_caption: yes
+      number_sections: FALSE
+      toc: no
+      fig_width: 7
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks2.lua
+      - --lua-filter=../templates/pagebreak.lua
 fontsize: 12pt
@@
 abstract: |
-  Accurate estimates of the number of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successfully management of fisheries in that area.
-Here we describe a state-space model that estimates such escapement past a dam by using window counts, PIT tag observations and data from an adult fish trap, accounting for issues such as night-time passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate.
-We tested this approach using a simulation framework that mimicked several levels of observation error, differences between night-time passage and re-ascension rates and the possibility of the adult trap being closed for some period of time, and found it to produce unbiased estimates across all tested scenarios.
+  Accurate estimates of the number of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successful management of fisheries in the Snake River basin.
+Here we describe a state-space model that estimates such escapement past a dam by using window counts, passive integrated transponder (PIT) tag observations and data from an adult fish trap, accounting for issues such as nighttime passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate.
+We tested this approach using a simulation framework that mimicked several levels of observation error, differences between nighttime passage and re-ascension rates and the possibility of the adult trap being closed for some period of time, and found it to produce unbiased estimates across all tested scenarios.
 We also applied this model to data from the Lower Granite dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring-summer run Chinook Salmon and steelhead from spawn years 2010-2019.
@@
                         "ISEMP" = "STADEM")) %>%
+  rowwise() %>%
   mutate(se = (uppCI - lowCI) / (2 * qnorm(0.975)),
@@
          rel_bias = bias / Truth,
-         MSE = se^2 + bias^2,
+         MSE = mean((bias)^2),
          inCI = ifelse(inCI == 1, T, ifelse(inCI == 0, F, NA)),
@@
 For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their spawning area.
-Escapement estimates facilitate effective fisheries management, and particularly, estimates of escapement for component groups (e.g., by stock, population, age, origin [wild, hatchery]; @hess2014monitoring; @Steinhorst2017; @Camacho2017) provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations.
-Accurate escapement estimates are increasingly important for depleted populations as they inform status metrics [@mcelhany2000viable], and facilitate assessments of population viability and extinction risk [@nwfsc2015; @williams2016viability] and climate change vulnerability [@crozier2019climate].
+Escapement estimates facilitate effective fisheries management, and particularly, estimates of escapement for component groups (e.g., by stock, population, age, origin [wild, hatchery]; @hess2014monitoring; @Steinhorst2017; @Camacho2018) provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations.
+Accurate escapement estimates are increasingly important for depleted populations as they inform status metrics [@mcelhany2000viable] and facilitate assessments of population viability, extinction risk [@nwfsc2015; @williams2016viability] and climate change vulnerability [@crozier2019climate].
 
@@
 
-Chinook Salmon and steelhead returning to a majority of populations in the Snake River must ascend a fish ladder on Lower Granite Dam before migrating to their natal tributary spawning locations.
+The majority of Chinook Salmon and steelhead returning to the Snake River must ascend a fish ladder on Lower Granite Dam before migrating to their natal tributary spawning locations.
 Fish management agencies previously used counts of fish, by species, passing an observation window within the fish ladder as a census of fish escaping past the dam.
-For both species, total escapement was then parsed into groups (e.g., wild, hatchery) using observed marks and genetic data from a sample of fish captured at an adult trap located on the fish ladder upstream of the observation window [@Camacho2017; @Steinhorst2017; @Steele2019].
-Window counts as a census proved beneficial as being an easy, straight-forward method that was ascertained in near real-time.
+For both species, total escapement was then parsed into groups (e.g., wild, hatchery) using observed marks and genetic data from a sample of fish captured at an adult trap located on the fish ladder upstream of the observation window [@Camacho2018; @Steinhorst2017; @Steele2019].
+Treating window counts as a census proved beneficial as being an easy, straight-forward method that was ascertained in near real-time.
 Moreover, downriver fisheries management arenas have used window counts at lower-river dams as escapement estimates for the past several years, and consistency in methods is often desirable for management decision making [@WDFW2019].
 
-Using window counts, however, as a census of Chinook Salmon and steelhead passing Lower Granite Dam can be problematic as it avoids multiple sources of uncertainty and disregards known biological processes.
-First, live (i.e., in-person) window counts only occur from April through October each year.
-Additionally, live window counts only occur for 16-hours each day, and fish counters working at the observation window look directly into the fish ladders to identify and count all passing fish, by species, for 50 minutes of each hour.
+However, using window counts as a census of Chinook Salmon and steelhead passing Lower Granite Dam can be problematic as it fails to account for multiple sources of uncertainty and disregards known biological processes.
+In-person window counts where observers look directly into the fish ladders to identify and count all passing fish, by species, occur for 50 minutes per hour, 16 hours a day from April through October (which corresponds with peak run timing for several species).
 Counts are then expanded to provide an estimate for the entire hour [@USACE2015].
-From November through March, the remainder of the year, video tape fish counting is used and only occurs for 10 hours each day; fish counters then read the video tapes and submit daily fish counts [@Hatch1994].
+For the remainder of the year (November through March), fish passage is video-taped for 10 hours each day; fish counters then read the video tapes and submit daily fish counts.
 Typically, the observational error rates of live and video window counts are unknown, and sampling error rates are ignored.
-Additionally, two biological processes are unaccounted for: 1) fish that cross the dam during the 8-hours when the window is closed for counting (i.e., nighttime passage) which may result in an underestimate of escapement, and 2) fish that migrate through the ladder and past the dam may fallback over the dam (e.g., over spillway, through navigation locks), and later, may or may not re-ascend the fish ladder again [@Boggs2004].
-Both fallback with no re-ascension and fallback with re-ascension potentially result in an overestimate of escapement [@Dauble2000].
-Previously, it was assumed that nighttime passage rates and fallback/re-ascension rates canceled each other out resulting in window counts providing an unbiased estimate of escapement [@Camacho2017].
+@Hatch1994 conducted a study of the counting system at Lower Granite dam in 1992, and found that Chinook Salmon were undercounted at the window by in-person window counters compared to video counts.
+There was a significant difference between daily standard counts (16 or 10 hours per day) and daily 24-hour video counts for both Chinook Salmon and steelhead, accounted for by estimates of passage rates while the counting window is closed of 3.5% (Chinook Salmon) and 6.6% (steelhead).
+They also found some species misidentification issues, which could result in either under- or over-counting of a particular species.
+While that study identified potential issues with the current window counting procedure, it was only conducted in a single year, nearly 30 years ago, at a time when Chinook Salmon and steelhead total returns were lower than the past decade.
+
+Besides potential errors in the window counts, two biological processes are unaccounted for: 1) fish that cross the dam during the 8+ hours when the window is closed for counting (i.e., nighttime passage) which may result in an underestimate of escapement, and 2) fish that migrate through the ladder and past the dam may fallback over the dam (e.g., over spillway, through navigation locks), and later, may or may not re-ascend the fish ladder again [@Boggs2004].
+Both fallback with and without re-ascension potentially result in an overestimate of escapement [@Dauble2000].
+Previously, it was assumed that nighttime passage rates and fallback/re-ascension rates canceled each other out resulting in window counts providing an unbiased estimate of escapement [@Camacho2018].
+
+<!-- John Hargrove says: You mention in the abstract uncertainty in the trap rate as a source of error.
+This may be a good place to identify what error is and how it likely affects estimates of escapement.
+
+I don't know that it's appropriate to bring that up here, since the trap rate has not been previously used to estimate escapement at all, and has been irrelevant for previous analyses.-->
 
@@
 We also show that observed nighttime passage and fallback/re-ascension rates are typically unequal.
-Our method for estimating escapement, by species, past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of PIT tagged fish in the adult ladder to explicitly model nighttime passage, re-ascension, and error from both window and trap estimates using a state-space approach [@royle2008hierarchical].
+Our method for estimating escapement, by species, past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of fish previously tagged with passive integrated transponder (PIT) tags in the adult ladder to explicitly model nighttime passage, re-ascension, and error from both window and trap estimates using a state-space approach [@royle2008hierarchical].
 To meet desired management and conservation objectives, modeled escapement includes estimates of uncertainty and is parsed into weekly strata.
-Further, total and weekly estimates are parsed into three origin groups: wild, hatchery, and hatchery no-clip.
-Estimates of escapement account for fish that migrate through the ladder at night outside of observation hours and account for fish that may ascend the ladder multiple times due to fallback and re-ascension.
+Further, total and weekly estimates are parsed into three origin groups: wild fish, hatchery fish with a clipped adipose fin, and unclipped hatchery fish.
+Estimates of escapement account for fish that migrate through the ladder at night outside of observation hours (nighttime passage) and account for fish that may ascend the ladder multiple times due to fallback and re-ascension.
 Our model is implemented in the **ST**ate-space **A**dult **D**am **E**scapement **M**odel (STADEM) package for the statistical software R [@R-Core-Team2020], and is available for downloaded at [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM).
@@
 We used STADEM and three sources of data to estimate Chinook Salmon and steelhead escapement at Lower Granite Dam from 2010-2019.
-Data sources included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of previously PIT tagged fish detected in the adult fish ladder.
+Data sources included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of fish PIT tagged before they reached Lower Granite Dam detected in the adult fish ladder.
 Below, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; knowing, similar data could likely be obtained from other fish passage facilities.
@@
 Daily counts of adult Chinook Salmon and steelhead passing an observation window located on the Lower Granite Dam fish ladder were the first source of data.
-Daily counts were made and provided by the US Army Corps of Engineers, and when summed, provide an estimate of the number of fish ascending and passing (i.e., escapement) Lower Granite Dam each season.
+Daily counts were made and provided by the US Army Corps of Engineers, and when summed, provide an estimate of the number of fish ascending and passing Lower Granite Dam each season.
 Window counts were made for each species using video monitoring and direct in-person visual monitoring during daytime hours [@Hatch1994].
@@
 The sum of the daily 50-minute counts were then multiplied by 1.2 to account for the 10 minutes when fish were not counted.
-Daytime window counts were not expanded for fish that may have ascended the ladder outside of operational hours (i.e., nighttime) [@USACE2015].
+Daytime window counts were not expanded for fish that may have ascended the ladder outside of operational hours (i.e., nighttime passage) [@USACE2015].
 Window counts were accessed through the Columbia Basin Research Data Access in Real Time (DART) website, [www.cbr.washington.edu/dart/query/adult_daily](www.cbr.washington.edu/dart/query/adult_daily), using their window count query.
@@
 The second source of data came from a sample of fish collected in the adult trap as they migrated past Lower Granite Dam [@Ogden2016b].
-The trap, also located within the adult fish ladder and upstream of the observation window, provided biological data (e.g., origin [wild, hatchery], genetic stock, length, age, sex) for captured adults that allowed decomposition of the escapement into specific groups (e.g., @Camacho2017, @Steinhorst2017).
+The trap, also located within the adult fish ladder and upstream of the observation window, provided biological data (e.g., origin [wild, hatchery], genetic stock, length, age, sex) for captured adults that allowed decomposition of the escapement into specific groups (e.g., @Camacho2018, @Steinhorst2017).
 The trap was operational for 24 hours per day, and randomly sampled the daily run by opening four times per hour for a length of time determined by a set daily trapping rate.
@@
 
-All captured fish were anesthetized, identified to species, examined for existing marks/tags, and measured for fork length.
-For adipose-intact (unclipped) adults, which includes wild and hatchery-no clip individuals, some portion or all of fish trapped had scale and genetic tissue samples taken.
+<!-- John Hargrove: Are there many more items to list – if not, then consider writing them out rather than use etc. -->
+
+All captured fish were anesthetized, identified to species, examined for existing marks/tags, measured for fork length and visually identified as wild or hatchery.
+The most widely usee marking of hatchery fish is the adipose fin clip, although coded wire tags are used in less than 10% of the hatchery releases.
+Some subset of hatchery fish are either intentionally or unintentionally released without an adipose fin clip, and these are referred to as unclipped hatchery fish, or hatchery no-clip (HNC).
+For adipose-intact (unclipped) adults, which includes wild and hatchery no-clip individuals, either a portion or all of fish trapped (depending on the year) had scale and genetic tissue samples taken.
 Scale samples were used to estimate age [@Wright2015] and genetic tissue samples were used to determine sex [@Campbell2012a] and estimate the origin of wild fish via genetic stock identification (e.g., @Hargrove2019).
@@
 Starting in 2013, every unclipped Chinook Salmon and steelhead trapped at LGR was genotyped to simplify collaborative logistics and better estimate the proportion of phenotypically wild fish of truly hatchery origin.
-@Camacho2017 provide further details on trap sample rates and valid sample selection.
+@Camacho2018 provide further details on trap sample rates and valid sample selection.
 Prior to release, all non-PIT tagged fish with an intact adipose fin (i.e., putatively wild) received a PIT tag.
-Final determination of wild, hatchery, and hatchery no-clip origins were assigned using a post-hoc analysis of marks and tags, including genetic parentage-based tags [@Steele2013; @Steele2019].
-Data from the adult trap were collected and managed by multiple agencies and were made available by the Idaho Department of Fish and Game [@Camacho2017].
+Final determination of wild, clipped hatchery, or unclipped hatchery origins were assigned using a post-hoc analysis of marks and tags, including parentage-based tqagging [@Steele2013; @Steele2019].
+Data from the adult trap were collected and managed by multiple agencies and were made available by the Idaho Department of Fish and Game [@Camacho2018].
 
@@
 The last source of data was observations of PIT tagged adult Chinook Salmon and steelhead at detection sites located in the Lower Granite Dam fish ladder.
-These observations provide estimates of 1) a trapping rate, 2) the proportion of fish passing during nighttime hours, and 3) the proportion of fish that ascend the fish ladder multiple times (i.e., the re-ascension rate).
+These observations provide estimates of 1) a trapping rate, 2) the nighttime passage rate, and 3) the re-ascension rate.
 Detections used in the model include all fish that were previously PIT tagged as juveniles or adults prior to reaching Lower Granite Dam (i.e., does not include newly tagged adults at the dam) and detected at adult detection sites in the dam passage system.
@@
 The sample size of previously tagged fish detected at Lower Granite Dam influences the uncertainty in that trap rate, which is also informed by the number of adults caught in the trap and the window counts.
-We want to estimate the "true" trapping rate because the set trapping rate (i.e.
+The set trapping rate (i.e.
 the recorded time that the trap is open to trap adults) does not always reflect the true proportion of fish that are captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error, among others.
-Therefore, we use the mark-recapture estimate of the trap rate to estimate a true trapping rate.
+Therefore, we use the mark-recapture approach to estimate a "true" trapping rate.
 
-The nighttime passage rate was based on the count of tags that migrated through the fish ladder during non-window observation hours and the total number of tags passing the fish ladder, and was estimated on a weekly basis.
-The re-ascension rate incorporated the count of tags observed passing the upstream most detection sites in the adult fish ladder (i.e., passing the dam) and later detected re-entering the downstream end of the fish ladder at a later time andthe total number of tags leaving the fish ladder.
+We used PIT tag data to estimate nighttime passage and re-ascension rates, using the total number of tags passing the fish ladder for both, and estimating both on a weekly basis.
+The nighttime passage rate was based on the count of PIT tags that migrated through the fish ladder during non-window observation hours and the total number of tags passing the fish ladder.
+The re-ascension rate incorporated the count of tags observed passing the upstream most detection sites in the adult fish ladder (i.e., passing the dam) and later detected re-entering the downstream end of the fish ladder at a later time and the total number of tags leaving the fish ladder.
 Previously, we looked for differences in nighttime passage and re-ascension rates estimated using wild fish only, versus combining hatchery and wild fish together, and found no difference.
@@
 In the case of the trap, for example, we assumed there was sampling variation and uncertainty around our estimates of the true unknown trap rate.
-STADEM further accounted for the proportion of fish that ascended the ladder while the counting window was closed (i.e., night), as well as for fish potentially double-counted (or more) after falling back below the dam and later re-ascending the fish ladder.
+STADEM adjusted the window counts for the nighttime passage rate, and both window and trap estimates for the re-ascension rate to estimate the number of unique fish that crossed teh dam.
 Finally, adult sampling data from the trap (wild, hatchery, hatchery no-clip) were used to partition the total escapement estimate by origin (Figure \@ref(fig:stadem-examp-fig)).
@@
 
-Our objective was to examine STADEM model estimates of origin-specific (wild, hatchery, hatchery no-clip) escapement from the combinations of two separate trapping rates, two fallback/re-ascension and nighttime passage combinations, and three window count error rates; resulting in twelve different scenarios (Table \@ref(tab:lgr-sim-tab)).
+Our objective was to examine STADEM model estimates of origin-specific (wild, hatchery, hatchery no-clip) escapement from the combinations of two trap rate scenarios (constant and shut down for 3 weeks), two fallback/re-ascension and nighttime passage combinations, and three window count error rates; resulting in twelve different scenarios (Table \@ref(tab:lgr-sim-tab)).
 The simulation parameters such as proportion of origin, run-timing, nighttime passage rates, fallback and re-ascension rates and trap rates were based on observed values at Lower Granite Dam between 2010-2015.
+We generated 99 simulations for each scenario, and ran STADEM on each one.
 Further details about simulation procedures can be found in [Appendix B](#append2).
 
+We grouped the simulation results by origin (wild, hatchery clipped and hatchery no-clip) and evaluated them by several measures.
+Relative bias is the difference between the simulated "truth" and the STADEM estimate, divided by the simulated "truth".
+We estimated the precision by examining the average coefficient of variation (CV) of the estimates.
+We calculated the root mean squared error (RMSE) as the square root of the mean of the squared bias in the estimate.
+Finally, we evaluated coverage probabilities by determining what proportion of the model results generated an estimated 95% credible interval that contained the simulated true value.
+
 ## Lower Granite Application
+
+<!-- John Hargrove: While I think this is a noteworthy application of STADEM, as written there is no justification of why you did this or how these values compare to other estimates made at LGR.
+In other words, were you interested in trends in abundance?
+Estimated coefficients of variation?
+Performance relative to another estimation procedure?
+
+We could do a comparison with the total window counts, or a comparison with SCOBI results, but I don't know that either is terribly relevant.
+SCOBI has been replaced by STADEM, I think.
+Comparing window counts with STADEM estimates of total unique fish, we could summarise the RMSE (~ 5000 for Chinook, ~10,000 for steelhead) or the normalized RMSE (~ 6% for Chinook, !
+7% for steelhead).
+Seem worth doing?
+Either that, or just explicitly state that we're doing a case study of STADEM at Lower Granite?
+-->
 
@@
 
-Simulation results of observed bias, sampling variation, precision, root mean squared deviation (RMSD), and coverage probabilities were qualitatively similar for the hatchery (N = 70,000) and hatchery no-clip (N = 5,000) origin groups as observed in the wild origin group (N = 25,000) comparisons.
-As such, only diagnostic measures of Lower Granite Dam model fits to a medium sized escapement level (e.g., wild origin escapement) are presented.
+The STADEM performance evaluation statistics were very similar for hatchery clipped (N = 70,000), hatchery no-clip (N = 5,000) and wild fish (N = 25,000).
+In the interest of brevity, we only present the results from wild fish, which corresponds to a medium sized escapement level.
 
@@
 Estimates were unbiased, with an average relative bias of 0.2–0.3%.
-The CV of the estimates averaged 2.0–3.0%, with coverage probabilities that always exceeded 95%.
-We calculated RMSD as the square root of the sum of the variance of the estimate and the squared expected bias, which accounts for the size of the uncertainty in the estimator as well as its bias.
-The RMSD was near 500 for each scenario, representing an estimate within 2% of the true value (Table \@ref(tab:summ-tab)).
+The CV of the estimates averaged 2.0–3.0%, with higher CV's being associated with scenarios when the trap was closed for 3 weeks.
+The coverage probabilities always exceeded 95% across all scenarios.
+The RMSE was near 500 for each scenario, representing an estimate within 2% of the true value, demonstrating the accuracy of STADEM (Table \@ref(tab:summ-tab)).
 
@@
 We applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead for spawn years 2010–2019.
-Estimates of total escapement, as well as estimates of wild, hatchery, and hatchery no-clip estimates are presented in Table \@ref(tab:lgr-tab).
+Estimates of total escapement, as well as estimates of wild, hatchery clipped, and hatchery no-clip estimates are presented in Table \@ref(tab:lgr-tab).
+Estimates of total unique fish escaping past Lower Granite Dam were sometimes higher and sometimes lower than the raw window counts, indicating that the relative strength of nighttime passage and re-ascension rates differed across years.
 Coefficients of variation ranged from 2.5-7.1% for wild fish, 2.3-5.1% for clipped hatchery fish, 3.4-8.8% for hatchery no-clip fish and 2.2-4.7% for total unique fish past Lower Granite Dam.
 
-Weekly estimates of total escapement over Lower Granite Dam tracked the window counts and trap estimates (Figure \@ref(fig:jags-fit-fig)).
+Weekly estimates of total escapement over Lower Granite Dam tracked the window counts and trap estimates (One example: Figure \@ref(fig:jags-fit-fig)).
 STADEM point estimates were often between estimates based on window counts, and those based on the number of fish caught in the adult trap.
-However, when very few fish were caught in the trap, or there was more uncertainty about the trap rate that week, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \@ref(fig:jags-fit-fig).
+However, for weeks when very few fish were caught in the trap, or there was more uncertainty about the trap rate, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \@ref(fig:jags-fit-fig).
 That year also shows the utility of STADEM in dealing with missing data, as the trap was shut down for several weeks in July and August.
 
-Examining the estimates of nighttime passage and re-ascension rates based on the observed PIT tags crossing over Lower Granite Dam (Figure \@ref(fig:night-reasc-diff-fig)), the two rates did not match up in most cases.
+<!-- John Hargrove: Is there a reason why you only present a singular year  in Figure 3?
+I acknowledge that having 10 years of data may make it unreadable – but perhaps justify why 2014 was selected or include all 10 years as separate plots using facet_wrap in ggplot.
+
+I think showing 10 time series plots is unnecessary (Table 3 shows how estimates match up with window counts).
+I think 2014 is a nice illustration of general agreement, with some examples of how STADEM doesn't always track the trap estimate, and how it deals with missing data.
+Do we need to make that more clear, or leave it as is?-->
+
+Examining the weekly estimates of nighttime passage and re-ascension rates at Lower Granite Dam (Figure \@ref(fig:night-reasc-diff-fig)), the two rates did not match up in most weeks.
 In particular, there are several weeks when the window counts are quite high, and the rates differ by as much as 10%.
-Clearly, those two biological processes do not cancel each other out in nearly all cases, and thus, employing a model that accounts for both will result in more accurate estimates of escapement.
+When nighttime passage is larger than re-ascension, the window counts will be biased low, and vice versa.
 
@@
 Trap closures are typically associated with elevated water temperatures resulting in potential fish handling stress and/or trap malfunctions.
-Given predicted Pacific Northwest climate change scenarios trap closures from high water temperatures may become more commonplace in the future; amplifying the need for a modeling framework that accounts for periods of missing data while still capturing estimate uncertainty.
+Given predicted Pacific Northwest climate change scenarios [@Zhang2019] trap closures from high water temperatures may become more commonplace in the future; amplifying the need for a modeling framework that accounts for periods of missing data while still capturing estimate uncertainty.
 Additionally, having a framework in place that accounts for missing periods of data will allow for increased logistic flexibility if, for example, maintenance or construction is needed at the observational window or adult trap.
 
+<!-- John Hargrove on the above paragraph, specifically the sentence about the LGD adult trap having been closed over the last decade: Do you have this data for all spawn years you presented?
+If so, quantitatively state how common this happens.
+I believe these data may be presented in Ogden’s reports.
+Alternatively, Camacho’s reports typically list out open and close dates as well as trap rates for different spawn years.
+-->
+
 Although not currently set up for this, STADEM could be modified and run on a weekly basis or in near real-time to provide better in-season estimates for fisheries managers.
-Currently, the only roadblock to this, at Lower Granite Dam, is the identification of hatchery origin fish from phenotypically wild fish (i.e., hatchery no-clip) using genetic tissue samples [@Steele2013; @Steele2019] collected at the adult trap, which currently is completed post-hoc after the trapping season.
-The inclusion of genetic information typically results in a reduction in wild escapement estimates and an associated increase in hatchery no-clip escapement.
+Currently, the only roadblock to this at Lower Granite Dam is the identification of hatchery no-clip fish using genetic tissue samples [@Steele2013; @Steele2019] collected at the adult trap, which currently is completed post-hoc after the trapping season.
+The inclusion of genetic information typically results in a reduction in wild escapement estimates and an associated increase in hatchery no-clip escapement [@Hargrove2020].
 However, if in-season management decisions do not require this correction or could accept the potential bias, origin calls at the trap could be used in-season as a first approximation to escapement.
@@
 
-Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary or population specific spawning areas [@iptds2020report; @kinzer2020report].
-Estimates of escapement at Lower Granite Dam, by species and origin, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates (e.g., @Camacho2017, @Schrader2013) that are important for fisheries management and productivity monitoring of wild populations.
+Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary or population specific spawning areas [@IPTDSW2020; @kinzer2020report].
+Estimates of escapement at Lower Granite Dam, by species and origin, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates (e.g., @Camacho2018, @Schrader2013) that are important for fisheries management and productivity monitoring of wild populations.
 As an example, STADEM is being applied at Lower Granite Dam to estimate the total unique wild fish migrating past the dam.
 Estimates of fish passing the dam are then combined with estimated movement or transition probabilities based on PIT tag observations at instream PIT tag detection systems throughout the Snake River basin, similar to @waterhouse2020bayesian, to estimate escapement to Snake River populations and locations throughout the basin [@orme2018population].
-Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations that, joined with sex and age data collected at the adult fish trap [@Hargrove2019], provides necessary information to evaluate productivity and population viability for select Snake River Chinook Salmon and steelhead groups [@iptds2020report].
+Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations that, joined with sex and age data collected at the adult fish trap [@Hargrove2019], provides necessary information to evaluate productivity and population viability for select Snake River Chinook Salmon and steelhead groups [@IPTDSW2020].
+
+<!-- John Hargrove: Have you considered comparing estimate that you produced  to those that failed to account for fallback/reascension to see how much difference there is?
+In other words, you could draw estimates from Camacho and Schrader docs and readily compare – may serve to better illustrate your point of the superiority of accounting for these errors relative to less complicated models
+
+I quickly pulled together the estimates from IDFG's latest report, 2010-2018 wild escapement.
+Their estimates of wild Chinook are about 4% higher than STADEM, and about 3% lower for wild steelhead.
+So far we've avoided explicitly comparing our results with existing methods in an effort to avoid confrontation.
+Maybe that's moot if IDFG has started adopting STADEM results, but looks like as of March 2019 they had not.
+Thoughts?-->
 
@@
                          "0.15 and 0.00 3 weeks"),
-         `Fallback rate` = c(0.06,
+         `Fallback & Re-ascension rate` = c(0.06,
                              0.10),
-         `Re-ascension rate` = 1,
+         # `Re-ascension rate` = 1,
          `Window count error` = factor(c("No Error",
-                                         "10% Error",
-                                         "5% Error"),
+                                         "5% Error",
+                                         "10% Error"),
                                        levels = c("No Error",
-                                                  "10% Error",
-                                                  "5% Error"))) %>%
-  mutate(`Nighttime passage rate` = if_else(`Fallback rate` == 0.06,
+                                                  "5% Error",
+                                                  "10% Error"))) %>%
+  mutate(`Nighttime passage rate` = if_else(`Fallback & Re-ascension rate` == 0.06,
                                             0.06,
@@
                                 "Trap down"),
-                              each = 3),
-                          c("",
-                            "Err H", 
-                            "Err L"))) %>%
+                              each = 3)),
+         Scenario = if_else(`Window count error` == "5% Error",
+                            paste(Scenario, "Err L"),
+                            if_else(`Window count error` == "10% Error",
+                                    paste(Scenario, "Err H"),
+                                    Scenario))) %>%
   select(Scenario, 
-         `Trap rate`:`Re-ascension rate`,
+         `Trap rate`:`Fallback & Re-ascension rate`,
          `Nighttime passage rate`,
@@
         booktabs = T,
-        caption = "Summary of simulation scenarios including varying adult trapping, fallback rates, re-ascension, nighttime passage, and window count error rates use to evaluate the performance of STADEM.") %>%
+        caption = "Summary of simulation scenarios including varying adult trapping, fallback and re-ascension rates, nighttime passage, and window count error rates use to evaluate the performance of STADEM.") %>%
   kable_styling()
@@
             `Mean CV` = mean(cv),
-            RMSD = sqrt(var(est) + mean(bias)^2),
+            # RMSD = sqrt(var(est) + mean(bias)^2),
+            RMSE = sqrt(mean(bias^2)),
+            MAE = mean(abs(bias)),
             Coverage = inCI/nsims) %>%
@@
          `Mean CV`,
-         RMSD,
+         # RMSD,
+         RMSE,
+         # MAE,
          Coverage)
@@
         booktabs = T,
-        caption = "Summary statistics, including relative bias, mean coefficient of variation (CV), root mean squared deviation (RMSD) and coverage for results from each of the twelve simulation scenarios.") %>%
+        caption = "Summary statistics, including relative bias, mean coefficient of variation (CV), root mean squared error (RMSE) and 95% credible interval coverage for results from each of the twelve simulation scenarios.") %>%
   kable_styling()
@@
 That estimate is divided by the nighttime passage rate (B).
-The total fish is then discounted by the reascension rate to estimate unique fish (C).
+The total fish is then discounted by the re-ascension rate to estimate unique fish (C).
 Those unique fish are then multiplied by the proportion of wild fish (D), to estimate unique wild fish."}
@@
 arrow_height = 1.05
-my_xlim = c(-3, 7.5)
+my_xlim = c(-4, 7.5)
 my_ylim = c(0.06, 1.1)
@@
                 geom = 'blank') +
-  scale_color_brewer(palette = 'Set1') +
+  stat_function(fun = dnorm,
+                args = list(mean = -1, sd = 0.75),
+                aes(color = 'Unique Hatchery Fish'),
+                geom = 'blank') +
+  stat_function(fun = dnorm,
+                args = list(mean = -1, sd = 0.75),
+                aes(color = 'Unique HNC Fish'),
+                geom = 'blank') +
+  # scale_color_brewer(palette = 'Set1') +
+  scale_color_brewer(palette = "Spectral") +
+  # scale_color_viridis_d() +
   # scale_color_grey(start = 0.9,
@@
                 lwd = 2) +
-  labs(title = 'Including Night Passage')
+  labs(title = 'Nighttime Passage')
 
@@
                 lwd = 2) +
-  labs(title = 'Account for Reascension')
+  labs(title = 'Re-ascension')
 
@@
   stat_function(fun = dnorm,
-                args = list(mean = -1, sd = 0.75),
+                args = list(mean = -1, sd = 0.85),
                 aes(color = 'Unique Wild Fish'),
+                lwd = 2) +
+  stat_function(fun = dnorm,
+                args = list(mean = 1, sd = 0.75),
+                aes(color = 'Unique Hatchery Fish'),
+                lwd = 2) +
+  stat_function(fun = dnorm,
+                args = list(mean = -2, sd = 1.25),
+                aes(color = 'Unique HNC Fish'),
                 lwd = 2) +
@@
                arrow = arrow(length = arrow_length)) +
-  labs(title = 'Proportion Wild')
+  labs(title = 'Origin Proportion')
 
@@
 
-```{r rel-bias-lgd, fig.cap = "Boxplots show relative bias of STADEM estimates for wild escapement across various scenarios.
-The boxes contain 50% of the simulations, whiskers contain 95% of the simulations, and points are outliers."}
-hdi_box = function(x) {
-  r = c(hdi(x, 0.95), hdi(x, 0.5), median(x))
-  # r = quantile(x, probs = c(0.025, 0.975, 0.25, 0.75, 0.5))
-  names(r) = c('ymin', 'ymax', 'lower', 'upper', 'middle')
-  r
-}
-
-hdi_out = function(x) {
-  subset(x, x < hdi(x, 0.95)[1] | x > hdi(x, 0.95)[2])
-  # subset(x, x < quantile(x, probs = c(0.025)) | x > quantile(x, probs = c(0.975)))
-}
+```{r rel-bias-lgd, fig.cap = "Boxplots of the relative bias of STADEM estimates for wild escapement across various scenarios (See Table 1)."}
+# hdi_box = function(x) {
+#   r = c(hdi(x, 0.95), hdi(x, 0.5), median(x))
+#   # r = quantile(x, probs = c(0.025, 0.975, 0.25, 0.75, 0.5))
+#   names(r) = c('ymin', 'ymax', 'lower', 'upper', 'middle')
+#   r
+# }
+# 
+# hdi_out = function(x) {
+#   subset(x, x < hdi(x, 0.95)[1] | x > hdi(x, 0.95)[2])
+#   # subset(x, x < quantile(x, probs = c(0.025)) | x > quantile(x, probs = c(0.975)))
+# }
 
@@
   mutate(grp = paste(Variable, Model, sep = ' / ')) %>%
-  # mutate(Scenario = factor(Scenario, 
-  #                          levels = sort(levels(res_long_df$Scenario), decreasing = F))) %>%
   filter(Variable == "Wild") %>%
@@
              fill = grp)) +
-  stat_summary(fun.data = hdi_box, 
-               geom = 'boxplot') +
-  stat_summary(fun = hdi_out,
-               geom = 'point',
-               color = 'gray') +
+  # stat_summary(fun.data = hdi_box,
+  #              geom = 'boxplot') +
+  # stat_summary(fun = hdi_out,
+  #              geom = 'point',
+  #              color = 'gray') +
+  geom_boxplot() +
   geom_hline(yintercept = 0,

### 2020-09-12 analysis/paper/appendices/Appendix_STADEM_descrip.Rmd
incorported comments and revisions from Sarah Hoffman's review

@@
 
-Escapement at Lower Granite Dam (LGD) is estimated by combining two independent observations, trap catches and window counts, of the true number of fish crossing the dam in a state-space model [@royle2008hierarchical] with a weekly time-step.
+Escapement at Lower Granite Dam (LGD) is estimated by combining two independent observations (trap catches and window counts) of the true number of fish crossing the dam in a state-space model [@royle2008hierarchical] with a weekly time-step.
 Both are assumed to be corrupted observations of the true unknown number of fish crossing LGD each week.
@@
 For weeks when we have a more precise estimate of the trap rate (i.e.
-weeks when lots of previously PIT tagged fish are crossing LGD), STADEM will tend to favor the estimate of total escapement based on the trap data, whereas when that trap rate is more uncertain (e.g.
+weeks when many previously PIT tagged fish are crossing LGD), STADEM will tend to favor the estimate of total escapement based on the trap data, whereas when that trap rate is more uncertain (e.g.
 fewer PIT tagged fish to use in estimating the trap rate), STADEM will rely more on the window counts to estimate total escapement.
-During peak run times, when lots of fish are crossing LGD, estimates based on trap data and trap rates will be more precise, while estimates from the window counts may have more observation error due to so many fish passing the window.
+During peak run times, when many fish are crossing LGD, estimates based on trap data and trap rates will be more precise, while estimates from the window counts may have more observation error due to densely crowded (and visually obstructed) fish passing the window.
 For weeks when the trap is down, STADEM relies exclusively on the window counts and nighttime passage data, but there will be more uncertainty in the estimates.
@@
 
-There are two other processes that must be accounted for, first, the proportion of fish that cross the dam while the window is closed for counting (nighttime passage rate), and the second, the proportion of fish that are crossing the dam multiple times (re-ascension rate) and therefore potentially double-counted.
+There are two other processes that must be accounted for: first, the proportion of fish that cross the dam while the window is closed for counting (nighttime passage rate), and second, the proportion of fish that are crossing the dam multiple times (re-ascension rate) and therefore potentially double-counted.
 Both rates can be estimated from previously PIT tagged fish that are crossing the dam each week.

### 2020-09-12 analysis/paper/appendices/Appendix_simulation_details.Rmd
incorporated suggestions from John Hargrove into manuscript
incorported comments and revisions from Sarah Hoffman's review

@@
 
-To examine the sensitivities of models to different fish behavior and dam operational scenarios, seven additional attributes are randomly assigned to each individual fish.
+To model different fish behavior and dam operational scenarios, seven additional attributes are randomly assigned to each individual fish.
 Each attribute is randomly assigned a TRUE/FALSE using a Bernoulli $\mathcal{PDF}$ and a fixed probability parameter.
@@
 If fish $i$ falls back and re-ascends, the entire process described above is repeated, with some time-lag between initial ascension and re-ascension that is governed by a Poisson $\mathcal{PDF}$ with mean = 2 days.
-Fish may fall-back and re-ascend up to 3 times, allowing for the possibility of the same fish being counted or trapped multiple times.
+Fish may fallback and re-ascend up to 3 times, allowing for the possibility of the same fish being counted or trapped multiple times.
 

### 2020-09-12 analysis/paper/paper.Rmd
incorported comments and revisions from Sarah Hoffman's review

@@
 abstract: |
-  Accurate estimates of the number of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successful management of fisheries in the Snake River basin.
+  Accurate estimates of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successful management of fisheries in the Snake River basin.
 Here we describe a state-space model that estimates such escapement past a dam by using window counts, passive integrated transponder (PIT) tag observations and data from an adult fish trap, accounting for issues such as nighttime passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate.
-We tested this approach using a simulation framework that mimicked several levels of observation error, differences between nighttime passage and re-ascension rates and the possibility of the adult trap being closed for some period of time, and found it to produce unbiased estimates across all tested scenarios.
+We tested this approach using a simulation framework that mimicked several levels of observation error, differences between nighttime passage and re-ascension rates and the possibility of the adult trap being closed for some period of time.
+Our results demonstrate that the model produced unbiased estimates across all tested scenarios.
 We also applied this model to data from the Lower Granite dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring-summer run Chinook Salmon and steelhead from spawn years 2010-2019.
@@
 For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their spawning area.
-Escapement estimates facilitate effective fisheries management, and particularly, estimates of escapement for component groups (e.g., by stock, population, age, origin [wild, hatchery]; @hess2014monitoring; @Steinhorst2017; @Camacho2018) provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations.
+Escapement estimates facilitate effective fisheries management, particularly estimates of escapement for component groups (e.g., by stock, population, age, origin [wild, hatchery]; @hess2014monitoring; @Steinhorst2017; @Camacho2018) which provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations.
 Accurate escapement estimates are increasingly important for depleted populations as they inform status metrics [@mcelhany2000viable] and facilitate assessments of population viability, extinction risk [@nwfsc2015; @williams2016viability] and climate change vulnerability [@crozier2019climate].
@@
 mykiss* in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric dams on the Snake and Columbia rivers) and tributary habitats [@nehlsen1991pacific; @mcclure2003large; @nwfsc2015].
-As a result, Snake River spring-summer run Chinook Salmon (hereafter Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937).
-Snake River Chinook Salmon and steelhead have substantial cultural, recreational, commercial and subsistence value both within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries.
+As a result, Snake River spring-summer run (arriving at Lower Granite between March 1 and August 17) Chinook Salmon (hereafter Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937).
+Snake River Chinook Salmon and steelhead have substantial cultural, recreational, commercial and subsistence value within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries.
 The aggregate escapement of Snake River Chinook Salmon and steelhead populations, with the exception of Tucannon River, is monitored at Lower Granite Dam located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations.
@@
 For both species, total escapement was then parsed into groups (e.g., wild, hatchery) using observed marks and genetic data from a sample of fish captured at an adult trap located on the fish ladder upstream of the observation window [@Camacho2018; @Steinhorst2017; @Steele2019].
-Treating window counts as a census proved beneficial as being an easy, straight-forward method that was ascertained in near real-time.
+Treating window counts as a census proved beneficial as being an easy, straight-forward method ascertained in near real-time.
 Moreover, downriver fisheries management arenas have used window counts at lower-river dams as escapement estimates for the past several years, and consistency in methods is often desirable for management decision making [@WDFW2019].
@@
 
-Besides potential errors in the window counts, two biological processes are unaccounted for: 1) fish that cross the dam during the 8+ hours when the window is closed for counting (i.e., nighttime passage) which may result in an underestimate of escapement, and 2) fish that migrate through the ladder and past the dam may fallback over the dam (e.g., over spillway, through navigation locks), and later, may or may not re-ascend the fish ladder again [@Boggs2004].
+Besides potential errors in the window counts, two biological processes are unaccounted for: 1) fish that cross the dam during the 8+ hours when the window is unmonitored (i.e., nighttime passage) which may result in an underestimate of escapement, and 2) fish that migrate through the ladder and past the dam may fallback over the dam (e.g., over spillway, through navigation locks) and later may re-ascend the fish ladder again [@Boggs2004] and be double-counted at the window (re-ascension).
 Both fallback with and without re-ascension potentially result in an overestimate of escapement [@Dauble2000].
@@
 In this study, we describe a novel method to estimate aggregate and group escapements which incorporate all sources of known uncertainty and demonstrate that the estimation method is essentially unbiased, thus better informing conservation and management decision making.
-We also show that observed nighttime passage and fallback/re-ascension rates are typically unequal.
-Our method for estimating escapement, by species, past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of fish previously tagged with passive integrated transponder (PIT) tags in the adult ladder to explicitly model nighttime passage, re-ascension, and error from both window and trap estimates using a state-space approach [@royle2008hierarchical].
+We also test whether observed nighttime passage and fallback/re-ascension rates are typically equal.
+Our method for estimating species specific escapement past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of fish previously tagged with passive integrated transponder (PIT) tags in the adult ladder to explicitly model nighttime passage, re-ascension, and observation error using a state-space approach [@royle2008hierarchical] which separates process variance (e.g.
+week to week variance in true escapment) from observation error variance (e.g.
+observation error at the window, or sampling variance at the trap).
 To meet desired management and conservation objectives, modeled escapement includes estimates of uncertainty and is parsed into weekly strata.
 Further, total and weekly estimates are parsed into three origin groups: wild fish, hatchery fish with a clipped adipose fin, and unclipped hatchery fish.
-Estimates of escapement account for fish that migrate through the ladder at night outside of observation hours (nighttime passage) and account for fish that may ascend the ladder multiple times due to fallback and re-ascension.
+Estimates of escapement account for fish that migrate through the ladder outside of observation hours (nighttime passage) and those that ascend the ladder multiple times (re-ascension).
 Our model is implemented in the **ST**ate-space **A**dult **D**am **E**scapement **M**odel (STADEM) package for the statistical software R [@R-Core-Team2020], and is available for downloaded at [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM).
@@
 Data sources included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of fish PIT tagged before they reached Lower Granite Dam detected in the adult fish ladder.
-Below, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; knowing, similar data could likely be obtained from other fish passage facilities.
+Below, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; similar data could likely be obtained from other fish passage facilities.
 
@@
 
-Daily counts of adult Chinook Salmon and steelhead passing an observation window located on the Lower Granite Dam fish ladder were the first source of data.
-Daily counts were made and provided by the US Army Corps of Engineers, and when summed, provide an estimate of the number of fish ascending and passing Lower Granite Dam each season.
+Daily counts of adult Chinook Salmon and steelhead passing an observation window located on the Lower Granite Dam fish ladder were made and provided by the US Army Corps of Engineers.
+When summed, they provide an estimate of the number of fish ascending and passing Lower Granite Dam each season.
 Window counts were made for each species using video monitoring and direct in-person visual monitoring during daytime hours [@Hatch1994].
@@
 Counts were provided for each day the fish ladder was open to passage.
-Although window counts were assumed to be a census of every fish passing Lower Granite Dam, corrections were not applied for nighttime passage or re-ascending fish.
-Further, there was no estimation of daily or seasonal observation or sampling error.
 
@@
 The trap, also located within the adult fish ladder and upstream of the observation window, provided biological data (e.g., origin [wild, hatchery], genetic stock, length, age, sex) for captured adults that allowed decomposition of the escapement into specific groups (e.g., @Camacho2018, @Steinhorst2017).
-The trap was operational for 24 hours per day, and randomly sampled the daily run by opening four times per hour for a length of time determined by a set daily trapping rate.
-The trap rate was determined by a committee of collaborating management agencies with a goal of capturing a target number of wild fish, but also balancing fish handling concerns.
+The trap was operational for 24 hours per day and randomly sampled the run by opening four times per hour for a length of time determined by a set daily trapping rate.
+The trap rate was determined by a committee of collaborating management agencies with a goal of capturing a target number of wild fish while also balancing fish handling concerns.
 Trap sample rates were typically 10-25%, but fluctuated throughout the season due to high water temperatures, decreased flows, trap malfunctions and/or closures, fish handling logistics, in-season forecast adjustments, etc.
@@
 
-All captured fish were anesthetized, identified to species, examined for existing marks/tags, measured for fork length and visually identified as wild or hatchery.
+All captured fish were anesthetized, speciated, examined for existing marks/tags, measured for fork length and visually identified as wild or hatchery.
 The most widely usee marking of hatchery fish is the adipose fin clip, although coded wire tags are used in less than 10% of the hatchery releases.
@@
 Prior to 2013, only fish determined to be wild in origin at the trap were sampled for scale and genetics.
-Starting in 2013, every unclipped Chinook Salmon and steelhead trapped at LGR was genotyped to simplify collaborative logistics and better estimate the proportion of phenotypically wild fish of truly hatchery origin.
+Starting in 2013, every unclipped Chinook Salmon and steelhead trapped at Lower Granite Dam was genotyped to simplify collaborative logistics and better estimate the proportion of unclipped hatchery fish that appear phenotypically wild.
 @Camacho2018 provide further details on trap sample rates and valid sample selection.
@@
 
-### PIT tag Data
+### PIT Tag Data
 
@@
 
-We used PIT tag data to estimate nighttime passage and re-ascension rates, using the total number of tags passing the fish ladder for both, and estimating both on a weekly basis.
+We used PIT tag data to estimate nighttime passage and re-ascension rates, using the total number of tags passing the fish ladder for both and estimating both on a weekly basis.
 The nighttime passage rate was based on the count of PIT tags that migrated through the fish ladder during non-window observation hours and the total number of tags passing the fish ladder.
@@
 Estimates of total unique fish escaping past Lower Granite Dam were sometimes higher and sometimes lower than the raw window counts, indicating that the relative strength of nighttime passage and re-ascension rates differed across years.
-Coefficients of variation ranged from 2.5-7.1% for wild fish, 2.3-5.1% for clipped hatchery fish, 3.4-8.8% for hatchery no-clip fish and 2.2-4.7% for total unique fish past Lower Granite Dam.
+CVs ranged from 2.5-7.1% for wild fish, 2.3-5.1% for clipped hatchery fish, 3.4-8.8% for hatchery no-clip fish and 2.2-4.7% for total unique fish past Lower Granite Dam.
 
 Weekly estimates of total escapement over Lower Granite Dam tracked the window counts and trap estimates (One example: Figure \@ref(fig:jags-fit-fig)).
-STADEM point estimates were often between estimates based on window counts, and those based on the number of fish caught in the adult trap.
-However, for weeks when very few fish were caught in the trap, or there was more uncertainty about the trap rate, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \@ref(fig:jags-fit-fig).
+STADEM point estimates were often between estimates based on window counts and those based on the number of fish caught in the adult trap.
+However, for weeks when very few fish were caught in the trap or there was more uncertainty about the trap rate, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \@ref(fig:jags-fit-fig).
 That year also shows the utility of STADEM in dealing with missing data, as the trap was shut down for several weeks in July and August.
@@
 
-Examining the weekly estimates of nighttime passage and re-ascension rates at Lower Granite Dam (Figure \@ref(fig:night-reasc-diff-fig)), the two rates did not match up in most weeks.
-In particular, there are several weeks when the window counts are quite high, and the rates differ by as much as 10%.
+Estimates of weekly nighttime passage and re-ascenstion rates did not match in most cases (Figure \@ref(fig:night-reasc-diff-fig)).
+In particular, there are several weeks when the window counts are quite large and the rates differ by as much as 10%.
 When nighttime passage is larger than re-ascension, the window counts will be biased low, and vice versa.
@@
 We have presented a novel method for estimating adult salmonid escapement past a large hydroelectric facility (e.g., Lower Granite Dam) that incorporates data from window counts, a fish trap, and observations of PIT tagged fish in the adult passage ladder.
-Our model explicitly models nighttime passage, re-ascension, and potential error in both window and trap estimates.
+Our model explicitly incorporates nighttime passage, re-ascension, and potential error in both window and trap estimates.
 In doing so, we demonstrated that at Lower Granite Dam, nighttime passage and re-ascension rates do not always offset each other, and assuming they do will lead to biased escapement estimates in some years.
-With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall Chinook Salmon, Pacific lamprey *Lampetra tridentata*), or elsewhere; provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run, and tag observation or detection infrastructure (e.g., a PIT tag detection array or similar) exists.
-Our state-space model combined multiple imperfect sources of data to reduce bias in adult escapement estimates and provided more reasonable estimates of uncertainty.
+With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall Chinook Salmon, Pacific lamprey *Lampetra tridentata*), or elsewhere, provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run and tag observation or detection infrastructure (e.g., a PIT tag detection array or similar) exists.
+Our state-space model combined multiple imperfect sources of data to reduce bias in adult escapement estimates and provided quantitative estimates of uncertainty.
 Accurate population or stock abundance estimates and uncertainty accounting for observation and process error can be particularly important when estimates are used or leveraged for management and conservation decisions such as population viability analyses.
@@
 First, it allows us to model observer error in the window counts, which is typically unknown.
-If estimates rely on window counts alone, quantifying observer error is impossible, and we believe it's prudent to capture and account for known sources of error to minimize management decision risk.
+If estimates rely on window counts alone, quantifying observer error is impossible.
+Capturing and accounting for known sources of error is prudent to minimize management decision risk.
 Second, by incorporating both sources of information in a state-space framework, STADEM incorporates missing data at either the observation window or adult trap seamlessly.
@@
 Trap closures are typically associated with elevated water temperatures resulting in potential fish handling stress and/or trap malfunctions.
-Given predicted Pacific Northwest climate change scenarios [@Zhang2019] trap closures from high water temperatures may become more commonplace in the future; amplifying the need for a modeling framework that accounts for periods of missing data while still capturing estimate uncertainty.
+Given predicted Pacific Northwest climate change scenarios [@Zhang2019] trap closures from high water temperatures may become more commonplace in the future, amplifying the need for a modeling framework that accounts for periods of missing data while still capturing estimate uncertainty.
 Additionally, having a framework in place that accounts for missing periods of data will allow for increased logistic flexibility if, for example, maintenance or construction is needed at the observational window or adult trap.
@@
 
-Although not currently set up for this, STADEM could be modified and run on a weekly basis or in near real-time to provide better in-season estimates for fisheries managers.
+STADEM could be modified and run on a weekly basis or in near real-time to provide better in-season estimates for fisheries managers.
 Currently, the only roadblock to this at Lower Granite Dam is the identification of hatchery no-clip fish using genetic tissue samples [@Steele2013; @Steele2019] collected at the adult trap, which currently is completed post-hoc after the trapping season.
@@
 However, if in-season management decisions do not require this correction or could accept the potential bias, origin calls at the trap could be used in-season as a first approximation to escapement.
-Final post-hoc estimates, parsed by origin, could then be finalized at season’s end.
-All other data included in this model (e.g., window counts and PIT observations) are otherwise provided in almost real-time by DART.
+Final post-hoc estimates parsed by origin could then be finalized at season’s end.
+All other data included in this model (e.g., window counts and PIT observations) are otherwise provided in near real-time by DART.
 Provided the Lower Granite Dam adult fish trap database was updated and available in near real-time, there are minimal obstacles for adapting the STADEM framework to provide in-season estimates of escapement.
@@
 Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary or population specific spawning areas [@IPTDSW2020; @kinzer2020report].
-Estimates of escapement at Lower Granite Dam, by species and origin, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates (e.g., @Camacho2018, @Schrader2013) that are important for fisheries management and productivity monitoring of wild populations.
+Estimates of species and origin specific escapement at Lower Granite Dam, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates (e.g., @Camacho2018, @Schrader2013) that are important for fisheries management and productivity monitoring of wild populations.
 As an example, STADEM is being applied at Lower Granite Dam to estimate the total unique wild fish migrating past the dam.
 Estimates of fish passing the dam are then combined with estimated movement or transition probabilities based on PIT tag observations at instream PIT tag detection systems throughout the Snake River basin, similar to @waterhouse2020bayesian, to estimate escapement to Snake River populations and locations throughout the basin [@orme2018population].
-Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations that, joined with sex and age data collected at the adult fish trap [@Hargrove2019], provides necessary information to evaluate productivity and population viability for select Snake River Chinook Salmon and steelhead groups [@IPTDSW2020].
+Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations.
+With sex and age data collected at the adult fish trap [@Hargrove2019], this approach provides necessary information to evaluate productivity and population viability for select Snake River Chinook Salmon and steelhead groups [@IPTDSW2020].
 
@@
 Justification and infrastructure exist for applying a modified STADEM framework for fish passing Bonneville Dam, the lowest dam on the Columbia River, or Priest Rapids Dam in the upper Columbia River.
-Both locations currently trap a sub-sample of passing Chinook Salmon and/or steelhead for biological information, and use window counts as a surrogate of true escapement.
-However, each has at least some similar problems to those observed at Lower Granite Dam, such as unaccounted observer and sampling error, nighttime passage, and/or re-ascension.
+Both locations currently trap a sub-sample of passing Chinook Salmon and/or steelhead for biological information and use window counts as a surrogate of true escapement.
+However, each has at least some similar problems to those observed at Lower Granite Dam such as unaccounted observer and sampling error, nighttime passage, and/or re-ascension.
 Certainly, estimating an unbiased total return to the entire Columbia River basin (i.e., Bonneville Dam) and Upper Columbia River with uncertainty would benefit managers and decision making.
@@
 ```{r stadem-examp-fig, fig.cap = "Schematic of how the STADEM model works.
-Panel A shows the posterior of the estimate of fish crossing the dam while the window is open (dashed line shows observed window counts).
+(A) shows the posterior of the estimate of fish crossing the dam while the window is open (dashed line shows observed window counts).
 That estimate is divided by the nighttime passage rate (B).
 The total fish is then discounted by the re-ascension rate to estimate unique fish (C).
-Those unique fish are then multiplied by the proportion of wild fish (D), to estimate unique wild fish."}
+Those unique fish are then multiplied by the proportions by origin (D), to estimate unique fish by origin."}
 
@@
 ```{r jags-fit-fig, fig.cap = "Time-series plot showing estimates of total escapement for Chinook in 2014, including raw window counts, window counts adjusted for nighttime passage, trap estimates and STADEM estimates.
-Gray ribbon represents the 95% credible interval for STADEM estimates."}
+The gray ribbon represents the 95% credible interval for STADEM estimates."}
 

### 2020-09-12 analysis/paper/paper.Rmd
re-knit all 3 versions

@@
         booktabs = T,
+        linesep = "",
         caption = "Summary of simulation scenarios including varying adult trapping, fallback and re-ascension rates, nighttime passage, and window count error rates use to evaluate the performance of STADEM.") %>%
-  kable_styling()
+  kable_styling(latex_options = c('scale_down',
+                                  "hold_position"))
 
@@
 
+\newpage
+
 ```{r summ-tab}
+
 sum_tab <- res_long_df %>%
@@
             `Mean CV` = mean(cv),
-            # RMSD = sqrt(var(est) + mean(bias)^2),
             RMSE = sqrt(mean(bias^2)),
-            MAE = mean(abs(bias)),
             Coverage = inCI/nsims) %>%
@@
          `Mean CV`,
-         # RMSD,
          RMSE,
-         # MAE,
          Coverage)
 
-# mu_rmse <- sum_tab %>%
-#   group_by(Model) %>%
-#   summarise(mu = mean(RMSE))
+sum_tab %>%
+  kable(digits = c(0, 3, 3, 0, 3),
+        booktabs = T,
+        linesep = "",
+        caption = "Summary statistics, including relative bias, mean coefficient of variation (CV), root mean squared error (RMSE) and 95 credible interval coverage for results from each of the twelve simulation scenarios.") %>%
+  kable_styling(latex_options = c('scale_down',
+                                  "hold_position"))
 
-sum_tab %>%
-  kable(digits = 3,
-        booktabs = T,
-        caption = "Summary statistics, including relative bias, mean coefficient of variation (CV), root mean squared error (RMSE) and 95% credible interval coverage for results from each of the twelve simulation scenarios.") %>%
-  kable_styling()
 ```
+
+\newpage
 
@@
   kable(booktabs = T,
+        linesep = "",
         caption = "Window counts, and estimates of total, wild, hatchery and hatchery no-clip escapement, with coefficients of variation in parenthesis, for Chinook Salmon and steelhead from spawn years 2010 to 2019.") %>%
-  kable_styling()
+  kable_styling(latex_options = c('scale_down',
+                                  "hold_position"))
+
 ```

### 2020-09-16 analysis/paper/appendices/Appendix_simulation_details.Rmd
clarified how vector of 0s and 1s is generated describing which populations have fish of a particular origin type.
changed the formatting of PDF in the simulation appendix

@@
 To simulate fish passing a dam, an $\mathbb{R}$ software function was developed [@R-Core-Team2020].
-The function randomly samples observations from assumed probability distribution functions ($\mathcal{PDF}$) with known parameters.
+The function randomly samples observations from assumed probability distribution functions (pdf) with known parameters.
 Total unique fish, $N$, and a vector, $\omega$, containing the proportions of wild ($w$), hatchery ($h$) and hatchery no-clip ($hnc$) fish passing the dam is set to establish known "truths" of escapement by origin.
@@
 
-Escapement of each origin is then randomly divided across a set number of populations, $n$, by randomly drawing proportions, $\phi_{j,p}$, of origin group $j$ in each population $p$ using a Dirichlet $\mathcal{PDF}$.
+Escapement of each origin is then randomly divided across a set number of populations, $n$, by randomly drawing proportions, $\phi_{j,p}$, of origin group $j$ in each population $p$ using a Dirichlet pdf.
 The Dirichlet function is parameterized from a vector, $\zeta_j$, containing 1's and 0's designating populations with origin $j$ fish returning.
-$\zeta_j$ originates from a Binomial $\mathcal{PDF}$ with $n$ populations, and the proportion of populations with each origin, $\tau_j$.
+For each population $p$, $\zeta_{j,p}$ is drawn from a Bernoulli pdf using the proportion of populations that contain each origin, $\tau_j$.
 Wild fish are assumed to be in all populations; $\tau_w = 1.0$.
@@
 \begin{aligned}
-  \left[\zeta_{j,p=1},  ... , \zeta_{j,p=n}\right] &\sim \text{Binomial}(n, \tau_j) \\
+  \zeta_{j,p} &\sim \text{Bernoulli}(\tau_j) \\
   \left[\phi_{j,p=1},  ... , \phi_{j,p=n}\right] &\sim \text{Dir}(\zeta_{j,p=1},  ... , \zeta_{j,p=n}) \\
@@
 
-Mean arrival date, $\bar{a}_p$, for each population returning to the dam is drawn from a normal $\mathcal{PDF}$ with hyper-parameters $\mu_a$ and $\sigma^2_a$.
-Similarly, the variance or spread in run-timing within populations is the absolute value of random variables drawn from a normal $\mathcal{PDF}$ with hyper-parameters $\mu_s$ and $\sigma^2_s$.
+Mean arrival date, $\bar{a}_p$, for each population returning to the dam is drawn from a normal pdf with hyper-parameters $\mu_a$ and $\sigma^2_a$.
+Similarly, the variance or spread in run-timing within populations is the absolute value of random variables drawn from a normal pdf with hyper-parameters $\mu_s$ and $\sigma^2_s$.
 
@@
  \left[\bar{a}_p, ..., \bar{a}_n\right] &\sim \text{Norm}(\mu_a, \sigma^2_a) \\
- \left[s_p, ..., s_n\right] &\sim \text{Norm}(\mu_s, \sigma^2_s)
+ \left[s_p, ..., s_n\right] &\sim \left | \text{Norm}(\mu_s, \sigma^2_s) \right |
 \end{aligned}
@@
 
-After sampling the mean date of arrival and variances for each population, the date of arrival, $a_{i,p}$, for individual fish, $i$, within each population are drawn from a normal $\mathcal{PDF}$ with population parameters $\bar{a}_p$ and $s^2_p$.
+After sampling the mean date of arrival and variances for each population, the date of arrival, $a_{i,p}$, for individual fish, $i$, within each population are drawn from a normal pdf with population parameters $\bar{a}_p$ and $s^2_p$.
 This simulates a random arrival day that is similar for all fish returning to the same population, regardless of origin.
@@
 To model different fish behavior and dam operational scenarios, seven additional attributes are randomly assigned to each individual fish.
-Each attribute is randomly assigned a TRUE/FALSE using a Bernoulli $\mathcal{PDF}$ and a fixed probability parameter.
+Each attribute is randomly assigned a TRUE/FALSE using a Bernoulli pdf and a fixed probability parameter.
 Fish passage during the day-time (i.e., during periods of window operation) is modeled using one minus the night-time passage rate ($1 - \nu$).
@@
 Re-ascension occurs with probability $\rho$, conditioned on fish $i$ falling back.
-If fish $i$ falls back and re-ascends, the entire process described above is repeated, with some time-lag between initial ascension and re-ascension that is governed by a Poisson $\mathcal{PDF}$ with mean = 2 days.
+If fish $i$ falls back and re-ascends, the entire process described above is repeated, with some time-lag between initial ascension and re-ascension that is governed by a Poisson pdf with mean = 2 days.
 Fish may fallback and re-ascend up to 3 times, allowing for the possibility of the same fish being counted or trapped multiple times.

### 2020-09-16 inst/scripts/plot_time_series.R
added script to plot time series for all years for each species, and save the faceted plot.

@@ -1,0 +1,132 @@
+#------------------------------------------------------------------------------
+# Script loads STADEM model runs, and creates time-series plots of model estimates,
+# window counts and trap estimates for each species / year combination
+#
+# Author: Kevin See
+#------------------------------------------------------------------------------
+
+# load needed packages
+library(tidyverse)
+library(magrittr)
+
+#------------------------------------------------------------------------------
+# get plotting data from multiple years
+plot_df = crossing(Species = c("Chinook",
+                               "Steelhead"),
+                   Year = 2010:2019) %>%
+  mutate(plot_title = paste(Species, Year)) %>%
+  rowwise() %>%
+  mutate(plot_data = map2(Species,
+                          Year,
+                          .f = function(x, y) {
+                            load(paste0('analysis/data/derived_data/STADEM_results/LGR_STADEM_', Species, '_', Year, '.rda'))
+                            plot_df = stadem_mod$summary[grep('^X.all', rownames(stadem_mod$summary)),] %>%
+                              as_tibble(rownames = 'var') %>%
+                              mutate(week = as.integer(str_extract(var, "[0-9]+")),
+                                     param = str_extract_all(var, "[:alpha:]+", simplify = T)[,3],
+                                     param = ifelse(param == '', 'all', param)) %>%
+                              select(var, param, week, everything()) %>%
+                              left_join(stadem_list$weeklyData %>%
+                                          rename(week = week_num) %>%
+                                          rowwise() %>%
+                                          mutate(trap_lowCI = max(0, trap_est + trap_est_se * qnorm(0.025)),
+                                                 trap_uppCI = trap_est + trap_est_se * qnorm(0.975)) %>%
+                                          ungroup())%>%
+                              mutate_at(vars(trap_est),
+                                        list(~ if_else(. == Inf,
+                                                       as.numeric(NA),
+                                                       .))) %>%
+                              select(-Species)
+                            return(plot_df)
+                          })) %>%
+  unnest(cols = plot_data) %>%
+  select(Species, Year, plot_title, everything()) %>%
+  filter(param == "all")
+
+# drop initial weeks when window counts == 0
+plot_df %<>%
+  group_by(Species, Year) %>%
[…]

### 2020-09-17 analysis/paper/paper.Rmd
Implemented some suggestions from John Hargrove, after discussing with Ryan Kinzer

@@
 output:
-    bookdown::pdf_document2:
+    bookdown::html_document2:
       fig_caption: yes
@@
       toc: no
-      geometry: margin=1in
-      latex_engine: "xelatex"
       fig_width: 7
@@
       - --lua-filter=../templates/pagebreak.lua
-      includes:
-        in_header: ../templates/header_manuscript.tex
     bookdown::word_document2:
@@
       - --lua-filter=../templates/pagebreak.lua
-    bookdown::html_document2:
+    bookdown::pdf_document2:
       fig_caption: yes
@@
       toc: no
+      geometry: margin=1in
+      latex_engine: "xelatex"
       fig_width: 7
+      # fig_height: 6
       pandoc_args:
@@
       - --lua-filter=../templates/pagebreak.lua
+      includes:
+        in_header: ../templates/header_manuscript.tex
 fontsize: 12pt
@@
 library(lubridate)
-library(msm)
-library(HDInterval)
+# library(msm)
+# library(HDInterval)
+library(magrittr)
 library(ggpubr)
@@
 # set default theme for ggplot2
-theme_set(theme_bw())
+# theme_set(theme_pubr(base_family = "Times New Roman",
+#                      base_size = 12))
+theme_set(theme_pubr())
 
@@
 
-Fish escapement often refers to the number of adults that survive juvenile and subadult rearing, escape harvest and return to their natal habitat to potentially spawn (e.g.
-@Bue1998).
+Fish escapement often refers to the number of adults that survive juvenile and subadult rearing, escape harvest and return to their natal habitat to potentially spawn [e.g.
+@Bue1998].
 For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their spawning area.
-Escapement estimates facilitate effective fisheries management, particularly estimates of escapement for component groups (e.g., by stock, population, age, origin [wild, hatchery]; @hess2014monitoring; @Steinhorst2017; @Camacho2018) which provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations.
+Escapement estimates facilitate effective fisheries management, particularly estimates of escapement for component groups [e.g., by stock, population, age, origin [wild, hatchery], @hess2014monitoring; @Steinhorst2017; @Camacho2018] which provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations.
 Accurate escapement estimates are increasingly important for depleted populations as they inform status metrics [@mcelhany2000viable] and facilitate assessments of population viability, extinction risk [@nwfsc2015; @williams2016viability] and climate change vulnerability [@crozier2019climate].
@@
 Previously, it was assumed that nighttime passage rates and fallback/re-ascension rates canceled each other out resulting in window counts providing an unbiased estimate of escapement [@Camacho2018].
-
-<!-- John Hargrove says: You mention in the abstract uncertainty in the trap rate as a source of error.
-This may be a good place to identify what error is and how it likely affects estimates of escapement.
-
-I don't know that it's appropriate to bring that up here, since the trap rate has not been previously used to estimate escapement at all, and has been irrelevant for previous analyses.-->
 
@@
 The second source of data came from a sample of fish collected in the adult trap as they migrated past Lower Granite Dam [@Ogden2016b].
-The trap, also located within the adult fish ladder and upstream of the observation window, provided biological data (e.g., origin [wild, hatchery], genetic stock, length, age, sex) for captured adults that allowed decomposition of the escapement into specific groups (e.g., @Camacho2018, @Steinhorst2017).
+The trap, also located within the adult fish ladder and upstream of the observation window, provided biological data (e.g., origin [wild, hatchery], genetic stock, length, age, sex) for captured adults that allowed decomposition of the escapement into specific groups [e.g., @Camacho2018; @Steinhorst2017].
 The trap was operational for 24 hours per day and randomly sampled the run by opening four times per hour for a length of time determined by a set daily trapping rate.
 The trap rate was determined by a committee of collaborating management agencies with a goal of capturing a target number of wild fish while also balancing fish handling concerns.
-Trap sample rates were typically 10-25%, but fluctuated throughout the season due to high water temperatures, decreased flows, trap malfunctions and/or closures, fish handling logistics, in-season forecast adjustments, etc.
-
-<!-- John Hargrove: Are there many more items to list – if not, then consider writing them out rather than use etc. -->
+Trap sample rates were typically 10-25%, but fluctuated throughout the season due to, for example, high water temperatures, decreased flows, trap malfunctions and/or closures, fish handling logistics, in-season forecast adjustments, or adjustments due to shifting species composition throughout the year.
 
@@
 
-<!-- John Hargrove: While I think this is a noteworthy application of STADEM, as written there is no justification of why you did this or how these values compare to other estimates made at LGR.
-In other words, were you interested in trends in abundance?
-Estimated coefficients of variation?
-Performance relative to another estimation procedure?
-
-We could do a comparison with the total window counts, or a comparison with SCOBI results, but I don't know that either is terribly relevant.
-SCOBI has been replaced by STADEM, I think.
-Comparing window counts with STADEM estimates of total unique fish, we could summarise the RMSE (~ 5000 for Chinook, ~10,000 for steelhead) or the normalized RMSE (~ 6% for Chinook, !
-7% for steelhead).
-Seem worth doing?
-Either that, or just explicitly state that we're doing a case study of STADEM at Lower Granite?
--->
-
 Finally, we applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead returning to the Snake River during spawn years 2010 to 2019.
@@
 Data from the adult trap was made available by Idaho Department of Fish and Game, and adult PIT tag detection data within the fish passage ladder at Lower Granite Dam was accessed from the PTAGIS regional database ([https://www.ptagis.org/](https://www.ptagis.org/)).
+
+All data used in this manuscript, as well as associated code for both the simulations and the Lower Granite appllication, can be found at [https://www.github.com/KevinSee/ManuscriptSTADEM]([https://www.github.com/KevinSee/ManuscriptSTADEM]).
 
@@
 
-Weekly estimates of total escapement over Lower Granite Dam tracked the window counts and trap estimates (One example: Figure \@ref(fig:jags-fit-fig)).
+Weekly estimates of total escapement over Lower Granite Dam tracked the window counts and trap estimates (One example: Figure \@ref(fig:jags-fit-fig).
+Similar plots for all model runs are available on the manuscript GitHub page: [https://www.github.com/KevinSee/ManuscriptSTADEM]([https://www.github.com/KevinSee/ManuscriptSTADEM])).
 STADEM point estimates were often between estimates based on window counts and those based on the number of fish caught in the adult trap.
@@
 That year also shows the utility of STADEM in dealing with missing data, as the trap was shut down for several weeks in July and August.
+The model's uncertainty is always smaller than the uncertainty from the trap estimates alone, whereas the window counts alone provide no estimate of uncertainty.
 
@@
 I think 2014 is a nice illustration of general agreement, with some examples of how STADEM doesn't always track the trap estimate, and how it deals with missing data.
-Do we need to make that more clear, or leave it as is?-->
+Do we need to make that more clear, or leave it as is?
+
+After experimenting with 4 years, Ryan and Kevin decided even 4 was too busy. Let's stick with one, and note that the rest are on the github repo-->
 
@@
 Second, by incorporating both sources of information in a state-space framework, STADEM incorporates missing data at either the observation window or adult trap seamlessly.
-At Lower Granite Dam, the adult trap has been closed for brief or extended periods of time (i.e., days, weeks) intermittently over the past several years, often during peak run times.
-Trap closures are typically associated with elevated water temperatures resulting in potential fish handling stress and/or trap malfunctions.
+At Lower Granite Dam, the adult trap has been closed for brief or extended periods of time (i.e., days, weeks) intermittently in five of the last ten years, often during peak run times [@USACE2010; @USACE2011; @USACE2012; @USACE2013; @Ogden2016; @USACE2015; @USACE2016; @USACE2017; @USACE2018; @USACE2019].
+Trap closures are typically associated with elevated water temperatures resulting in potential fish handling stress and/or trap malfunctions [@Ogden2016b].
 Given predicted Pacific Northwest climate change scenarios [@Zhang2019] trap closures from high water temperatures may become more commonplace in the future, amplifying the need for a modeling framework that accounts for periods of missing data while still capturing estimate uncertainty.
 Additionally, having a framework in place that accounts for missing periods of data will allow for increased logistic flexibility if, for example, maintenance or construction is needed at the observational window or adult trap.
-
-<!-- John Hargrove on the above paragraph, specifically the sentence about the LGD adult trap having been closed over the last decade: Do you have this data for all spawn years you presented?
-If so, quantitatively state how common this happens.
-I believe these data may be presented in Ogden’s reports.
-Alternatively, Camacho’s reports typically list out open and close dates as well as trap rates for different spawn years.
--->
 
@@
 Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary or population specific spawning areas [@IPTDSW2020; @kinzer2020report].
-Estimates of species and origin specific escapement at Lower Granite Dam, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates (e.g., @Camacho2018, @Schrader2013) that are important for fisheries management and productivity monitoring of wild populations.
+Estimates of species and origin specific escapement at Lower Granite Dam, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates [e.g., @Camacho2018; @Schrader2013] that are important for fisheries management and productivity monitoring of wild populations.
 As an example, STADEM is being applied at Lower Granite Dam to estimate the total unique wild fish migrating past the dam.
@@
 Maybe that's moot if IDFG has started adopting STADEM results, but looks like as of March 2019 they had not.
-Thoughts?-->
+
+Ryan and Kevin decided to not go down this path. One quick comparison is total escapement vs. window counts; which shows up in one of our tables, and we discuss it briefly in the results. -->
 
@@
   kable_styling(latex_options = c('scale_down',
-                                  "hold_position"))
+                                  "hold_position")) %>%
+  column_spec(column = 2,
+              width = "1in") %>%
+  column_spec(column = c(3, 4, 5),
+              width = "1.2in")
 
@@
         linesep = "",
-        caption = "Summary statistics, including relative bias, mean coefficient of variation (CV), root mean squared error (RMSE) and 95 credible interval coverage for results from each of the twelve simulation scenarios.") %>%
+        caption = "Summary statistics, including relative bias, mean coefficient of variation (CV), root mean squared error (RMSE) and 95\\% credible interval coverage for results from each of the twelve simulation scenarios.") %>%
   kable_styling(latex_options = c('scale_down',
@@
 
-```{r stadem-examp-fig, fig.cap = "Schematic of how the STADEM model works.
+```{r stadem-examp-fig, fig.width=3.5, fig.height=7, fig.cap = "Schematic of how the STADEM model works.
 (A) shows the posterior of the estimate of fish crossing the dam while the window is open (dashed line shows observed window counts).
@@
 my_xlim = c(-4, 7.5)
-my_ylim = c(0.06, 1.1)
+my_ylim = c(0.08, 1.2)
 
@@
   # scale_color_brewer(palette = 'Set1') +
+  # scale_color_viridis_d() +
   scale_color_brewer(palette = "Spectral") +
-  # scale_color_viridis_d() +
   # scale_color_grey(start = 0.9,
@@
                   ylim = my_ylim) +
-  theme_classic() +
+  # theme_classic() +
+  theme_pubr(legend = "bottom") +
+  guides(color = guide_legend(nrow = 3,
+                              title.position = 'top')) +
   theme(axis.text = element_blank(),
         axis.ticks = element_blank(),
-        legend.position = 'bottom') +
+        axis.title.y = element_blank()) +
   labs(x = 'Escapement',
-       y = 'Density',
        color = 'Fish Type',
-       title = 'Daytime Fish')
+       title = '   Daytime Fish')
 
@@
                 lwd = 2) +
-  labs(title = 'Nighttime Passage')
+  labs(title = '   Nighttime Passage')
 
@@
                 lwd = 2) +
-  labs(title = 'Re-ascension')
+  labs(title = '   Re-ascension')
 
@@
                arrow = arrow(length = arrow_length)) +
-  labs(title = 'Origin Proportion')
+  labs(title = '   Origin Proportion')
 
@@
                     labels = 'AUTO',
-                    ncol = 2,
-                    nrow = 2,
+                    ncol = 1,
+                    nrow = 4,
                     common.legend = T,
@@
 ```
+
+\newpage
 
@@
   filter(Variable == "Wild") %>%
-  ggplot(aes(x = Scenario,
-             y = rel_bias,
+  ggplot(aes(x = rel_bias,
+             y = fct_rev(Scenario),
              fill = grp)) +
@@
   #              color = 'gray') +
-  geom_boxplot() +
-  geom_hline(yintercept = 0,
+  geom_boxplot(fill = 'gray80') +
+  geom_vline(xintercept = 0,
              linetype = 2) +
-  labs(y = 'Relative Bias',
+  # facet_wrap(~ Variable) +
+  theme_pubr(legend = "none") +
+    labs(x = 'Relative Bias',
+       y = "Scenario",
        fill = 'Fish / Model') +
-  facet_wrap(~ Variable) +
-  theme(legend.position = 'none',
-        axis.text.x = element_text(angle=45, hjust = 1))
+  coord_fixed(0.03)
 
@@
 
-```{r jags-fit-fig, fig.cap = "Time-series plot showing estimates of total escapement for Chinook in 2014, including raw window counts, window counts adjusted for nighttime passage, trap estimates and STADEM estimates.
-The gray ribbon represents the 95% credible interval for STADEM estimates."}
+\newpage
 
-load('../data/derived_data/STADEM_results/LGR_STADEM_Chinook_2014.rda')
-plotTimeSeries(stadem_mod,
-               weeklyData = stadem_list$weeklyData) +
-  guides(color = guide_legend(nrow = 2),
-         shape = guide_legend(nrow = 2))
+```{r jags-fit-fig, fig.cap = "Time-series plot showing estimates of total escapement for Chinook in 2014, including window counts, trap estimates and STADEM estimates.
+The gray ribbon represents the 95% credible intervals for STADEM estimates, while the red ribbon represents the 95% confidence intervals for the trap estimates."}
+
+# get plotting data from particular year
+plot_df = tibble(Species = "Chinook",
+                 Year = "2014") %>%
+  mutate(plot_title = paste(Species, Year)) %>%
+  rowwise() %>%
+  mutate(plot_data = map2(Species, 
+                          Year,
+                          .f = function(x, y) {
+                            load(paste0('../data/derived_data/STADEM_results/LGR_STADEM_', Species, '_', Year, '.rda'))
+                            plot_df = stadem_mod$summary[grep('^X.all', rownames(stadem_mod$summary)),] %>%
+                              as_tibble(rownames = 'var') %>%
+                              mutate(week = as.integer(str_extract(var, "[0-9]+")),
+                                     param = str_extract_all(var, "[:alpha:]+", simplify = T)[,3],
+                                     param = ifelse(param == '', 'all', param)) %>%
+                              select(var, param, week, everything()) %>%
+                              left_join(stadem_list$weeklyData %>%
+                                          rename(week = week_num) %>%
+                                          rowwise() %>%
+                                          mutate(trap_lowCI = max(0, trap_est + trap_est_se * qnorm(0.025)),
+                                                 trap_uppCI = trap_est + trap_est_se * qnorm(0.975)) %>%
+                                          ungroup())%>%
+                              mutate_at(vars(trap_est),
+                                        list(~ if_else(. == Inf,
+                                                       as.numeric(NA),
+                                                       .))) %>%
+                              select(-Species)
+                            return(plot_df)
+                          })) %>%
+  unnest(cols = plot_data) %>%
+  select(Species, Year, plot_title, everything()) %>%
+  filter(param == "all")
+
+# drop initial weeks when window counts == 0
+plot_df %<>%
+  group_by(Species, Year) %>%
+  filter(Start_Date >= min(Start_Date[win_cnt > 0]))
+
+# scale some variables for plotting
+plot_df %<>%
+    mutate_at(vars(trap_lowCI,
+                 trap_uppCI,
+                 `2.5%`,
+                 `97.5%`,
+                 win_cnt,
+                 trap_est,
+                 `50%`),
+            list(~ . / 10000))
+
+# set size of points
+pt_size = 4
+
+# color version
+ts_col = plot_df %>%
+  ggplot(aes(x = Start_Date)) +
+  geom_ribbon(aes(ymin = trap_lowCI,
+                  ymax = trap_uppCI,
+                  fill = "Trap Est."),
+              alpha = 0.4) +
+  geom_ribbon(aes(ymin = `2.5%`,
+                  ymax = `97.5%`,
+                  fill = "Model"),
+              alpha = 0.4) +
+  geom_point(aes(y = win_cnt,
+                 shape = 'Window Count',
+                 color = 'Window Count'),
+             size = pt_size) +
+  geom_point(aes(y = trap_est,
+                 shape = 'Trap Est.',
+                 color = 'Trap Est.'),
+             size = pt_size) +
+  geom_line(aes(y = `50%`,
+                color = 'Model')) +
+  geom_point(aes(y = `50%`,
+                 shape = 'Model',
+                 color = 'Model'),
+             size = pt_size) +
+  scale_color_manual(values = c('Model' = 'black',
+                                'Window Count' = 'blue',
+                                'Trap Est.' = 'red'),
+                     name = "Source") +
+  scale_fill_manual(values = c('Model' = 'black',
+                                'Window Count' = 'blue',
+                                'Trap Est.' = 'red'),
+                     guide = "none") +
+  scale_shape_manual(values = c("Model" = 18,
+                                "Window Count" = 1,
+                                "Trap Est." = 8),
+                     name = "Source") +
+  theme_pubr() +
+  theme(legend.position = 'bottom') +
+  facet_wrap(~ plot_title,
+             scales = "free",
+             ncol = 2) +
+  labs(x = 'Date',
+       y = 'Total Escapement (x 10,000)')
+
+# black and white version
+ts_bw = ts_col +
+  scale_color_grey(start = 0,
+                   end = 0,
+                   name = "Source") +
+  scale_fill_manual(values = c('Model' = 'gray20',
+                                'Trap Est.' = 'gray70'),
+                     guide = "none")
+
+ts_col
+# ts_bw
+
 ```
 
-```{r night-reasc-diff-fig, fig.cap = "Nighttime passage rate plotted against re-ascension rate, calculated from observed PIT tags for each week of spawn years 2010-2019.
+\newpage
+
+```{r night-reasc-diff-fig, fig.height=5, fig.cap = "Nighttime passage rate plotted against re-ascension rate on the logit scale, calculated from observed PIT tags for each week of spawn years 2010-2019.
 Colors correspond to different spawn year, while the size of each point is proportional to the window count that week.
@@
 
-rate_comp %>%
+rate_p_base = rate_comp %>%
   filter(win_cnt > 0) %>%
@@
          !reasc_rate %in% c(1, 0)) %>%
-  # arrange(desc(night_rate))
   ggplot(aes(x = night_rate,
-             y = reasc_rate,
+             y = reasc_rate)) +
+  geom_abline(linetype = 2) +
+  facet_wrap(~ Species,
+             ncol = 1) +
+  theme_pubr(legend = 'none') +
+  coord_fixed(ratio = 1) +
+  scale_x_continuous(trans = "logit",
+                     breaks = scales::pretty_breaks(3)) +
+  scale_y_continuous(trans = "logit",
+                     breaks = scales::pretty_breaks()) +
+  labs(x = "Nighttime Passage Rate",
+       y = "Re-ascension Rate")
+
+# color version
+rate_p_col = rate_p_base +
+  scale_color_brewer(palette = "Paired") +
+  geom_point(aes(size = win_cnt,
                  color = as.factor(Year))) +
-  # scale_color_viridis_d() +
-  scale_color_brewer(palette = "Paired") +
-  geom_abline(linetype = 2) +
-  geom_point(aes(size = win_cnt)) +
-  facet_wrap(~ Species,
-             scales = 'free') +
-  theme(legend.position = 'bottom') +
-  # scale_x_continuous(trans = "log",
-  #                    breaks = scales::pretty_breaks()) +
-  # scale_y_continuous(trans = "log",
-  #                    breaks = scales::pretty_breaks()) +
-  guides(size = guide_legend(nrow = 2)) +
-  labs(x = "Nighttime Passage Rate",
-       y = "Re-ascension Rate",
-       color = "Spawn\nYear",
-       size = "Window\nCount")
+  labs(color = "Spawn Year",
+       size = "Window Count") +
+  theme_pubr(legend = "left")
+
+# black and white version
+rate_p_bw = rate_p_base +
+  geom_point(aes(size = win_cnt),
+             shape = 21,
+             fill = 'gray40',
+             color = 'black',
+             alpha = 0.3)
+
+rate_p_col
+# rate_p_bw
 ```

### 2020-09-17 analysis/paper/paper.Rmd
cleaned up some extraneous lines

@@
 
-<!-- The following line inserts a page break  -->
 \newpage
@@
 
+# choose appropriate color scale for color vs. black & white
 p1 = ggplot(data.frame(x = c(-5, 8)),
@@
                 geom = 'blank') +
-  # scale_color_brewer(palette = 'Set1') +
-  # scale_color_viridis_d() +
   scale_color_brewer(palette = "Spectral") +
@@
                   ylim = my_ylim) +
-  # theme_classic() +
   theme_pubr(legend = "bottom") +
@@
 examp_p
+
 ```

### 2020-09-17 analysis/paper/paper.Rmd
re-knit all 3 versions

@@
 
-Ryan and Kevin decided to not go down this path. One quick comparison is total escapement vs. window counts; which shows up in one of our tables, and we discuss it briefly in the results. -->
+Ryan and Kevin decided to not go down this path.
+One quick comparison is total escapement vs.
+window counts; which shows up in one of our tables, and we discuss it briefly in the results.
+The other is figure 4, showing nighttime passage and re-ascension rates, and how they don't match.
+-->
 
@@
 
-```{r stadem-examp-fig, fig.width=3.5, fig.height=7, fig.cap = "Schematic of how the STADEM model works.
+```{r stadem-examp-fig, fig.width=3.5, fig.asp = 2, fig.cap = "Schematic of how the STADEM model works.
 (A) shows the posterior of the estimate of fish crossing the dam while the window is open (dashed line shows observed window counts).
@@
              linetype = 2) +
-  coord_cartesian(xlim = my_xlim,
-                  ylim = my_ylim) +
   theme_pubr(legend = "bottom") +
@@
        color = 'Fish Type',
-       title = '   Daytime Fish')
+       title = '   Daytime Fish') +
+  coord_cartesian(xlim = my_xlim,
+                  ylim = my_ylim)
+# coord_fixed(ratio = 4,
+  #             xlim = my_xlim,
+  #             ylim = my_ylim)
 
@@
                     nrow = 4,
+                    align = 'v',
                     common.legend = T,

### 2020-09-17 analysis/paper/appendices/Appendix_STADEM_descrip.Rmd
corrected some references in the appendices, after updating the reference.bib file

@@
 
-Escapement at Lower Granite Dam (LGD) is estimated by combining two independent observations (trap catches and window counts) of the true number of fish crossing the dam in a state-space model [@royle2008hierarchical] with a weekly time-step.
+Escapement at Lower Granite Dam (LGD) is estimated by combining two independent observations (trap catches and window counts) of the true number of fish crossing the dam in a state-space model [@Royle2008] with a weekly time-step.
 Both are assumed to be corrupted observations of the true unknown number of fish crossing LGD each week.

### 2020-09-17 analysis/paper/appendices/Appendix_simulation_details.Rmd
made the equations using normal distributation match the notation in the other appendix
added some \left( and \right) to some latex code.
added a sentence at the end pointing to the github repo for all the code.
corrected some references in the appendices, after updating the reference.bib file

@@
   \zeta_{j,p} &\sim \text{Bernoulli}(\tau_j) \\
-  \left[\phi_{j,p=1},  ... , \phi_{j,p=n}\right] &\sim \text{Dir}(\zeta_{j,p=1},  ... , \zeta_{j,p=n}) \\
+  \left[\phi_{j,p=1},  ... , \phi_{j,p=n}\right] &\sim \text{Dir} \left(\zeta_{j,p=1},  ... , \zeta_{j,p=n} \right) \\
   N_{j,p} &= N_j * \phi_{j,p} \\
-  N_p &= \sum_{j=w, h, hnc} N_{j,p}
+  N_p &= \sum_{j \in w, h, hnc} N_{j,p}
 \end{aligned}
@@
 \begin{aligned}
- \left[\bar{a}_p, ..., \bar{a}_n\right] &\sim \text{Norm}(\mu_a, \sigma^2_a) \\
- \left[s_p, ..., s_n\right] &\sim \left | \text{Norm}(\mu_s, \sigma^2_s) \right |
+ \left[\bar{a}_p, ..., \bar{a}_n\right] &\sim \mathcal{N}(\mu_a, \sigma^2_a) \\
+ \left[s_p, ..., s_n\right] &\sim \left | \mathcal{N}(\mu_s, \sigma^2_s) \right |
 \end{aligned}
@@
 \begin{aligned}
- date_{i,p} &\sim \text{Norm}(\bar{a}_p, s^2_p) \\
+ date_{i,p} &\sim \mathcal{N}(\bar{a}_p, s^2_p) \\
 \end{aligned}
@@
 First, trapping rates were set static at 0.15 across all weeks for six scenarios to mimic an optimum trap operation for an expected return of 25,000 wild fish (i.e., trap $\approx$ 4,000 wild fish).
-For the remaining six scenarios, trapping rates for weeks 30, 31 and 32 (i.e., July $22^{nd}$ to August $11^{th}$) were changed to 0.00 to test STADEM sensitivities to potential trap shut downs similar to those observed in 2013, 2014 and 2015 [@Ogden2014; @Ogden2016; @Ogden2016b].
+For the remaining six scenarios, trapping rates for weeks 30, 31 and 32 (i.e., July $22^{nd}$ to August $11^{th}$) were changed to 0.00 to test STADEM sensitivities to potential trap shut downs similar to those observed in 2013, 2014 and 2015 [@Ogden2014; @Ogden2016a; @Ogden2016b].
 To simulate and control the number of re-ascending and night-time passing fish to model response, we altered fallback and night-time passage rates while holding the re-ascension rate constant at $\rho = 1.0$.
@@
 This method simulated observed counts as unbiased, and allowed for possible under and overcounts at the window.
+
+All code for simulating data and fitting STADEM to that data can be found at [https://www.github.com/KevinSee/ManuscriptSTADEM]([https://www.github.com/KevinSee/ManuscriptSTADEM]).

### 2020-09-17 analysis/paper/paper.Rmd
updated all citation calls, corrected a few instances of Lower Granite dam (should be Lower Granite Dam), and re-knit all versions

@@
 output:
-    bookdown::html_document2:
-      fig_caption: yes
-      number_sections: FALSE
-      toc: no
-      fig_width: 7
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks2.lua
-      - --lua-filter=../templates/pagebreak.lua
     bookdown::pdf_document2:
@@
       always_allow_html: true
+      fig_width: 6
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
+    bookdown::html_document2:
+      fig_caption: yes
+      number_sections: FALSE
+      toc: no
       fig_width: 7
@@
       - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/author-info-blocks2.lua
       - --lua-filter=../templates/pagebreak.lua
@@
 - packages.bib
-- AckermanLibrary.bib
 csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
@@
 Our results demonstrate that the model produced unbiased estimates across all tested scenarios.
-We also applied this model to data from the Lower Granite dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring-summer run Chinook Salmon and steelhead from spawn years 2010-2019.
+We also applied this model to data from the Lower Granite Dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring-summer run Chinook Salmon and steelhead from spawn years 2010-2019.
 keywords: |
@@
 # set default theme for ggplot2
-# theme_set(theme_pubr(base_family = "Times New Roman",
-#                      base_size = 12))
 theme_set(theme_pubr())
@@
   options(knitr.table.format = "pandoc")
+}
+
+if(doc.type %in% c("docx", "html")) {
+  theme_set(theme_pubr(base_family = "Times New Roman"))
 }
@@
 For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their spawning area.
-Escapement estimates facilitate effective fisheries management, particularly estimates of escapement for component groups [e.g., by stock, population, age, origin [wild, hatchery], @hess2014monitoring; @Steinhorst2017; @Camacho2018] which provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations.
-Accurate escapement estimates are increasingly important for depleted populations as they inform status metrics [@mcelhany2000viable] and facilitate assessments of population viability, extinction risk [@nwfsc2015; @williams2016viability] and climate change vulnerability [@crozier2019climate].
+Escapement estimates facilitate effective fisheries management, particularly estimates of escapement for component groups [e.g., by stock, population, age, origin [wild, hatchery], @Hess2014; @Steinhorst2017; @Camacho2018] which provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations.
+Accurate escapement estimates are increasingly important for depleted populations as they inform status metrics [@McElhany2000] and facilitate assessments of population viability, extinction risk [@NWFSC2015; @Williams2016] and climate change vulnerability [@Crozier2019].
 
 Populations of Chinook Salmon *Oncorhynchus tshawytscha* and steelhead trout *O.
-mykiss* in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric dams on the Snake and Columbia rivers) and tributary habitats [@nehlsen1991pacific; @mcclure2003large; @nwfsc2015].
+mykiss* in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric dams on the Snake and Columbia rivers) and tributary habitats [@Nehlsen1991; @McClure2003; @NWFSC2015].
 As a result, Snake River spring-summer run (arriving at Lower Granite between March 1 and August 17) Chinook Salmon (hereafter Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937).
@@
 The aggregate escapement of Snake River Chinook Salmon and steelhead populations, with the exception of Tucannon River, is monitored at Lower Granite Dam located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations.
-Many fisheries management and conservation actions are made based on estimates of escapement at Lower Granite Dam parsed by species and origin [@NMFS2019BiOp; @NFSC2015; @NPCC2014].
+Many fisheries management and conservation actions are made based on estimates of escapement at Lower Granite Dam parsed by species and origin [@NMFS2019; @NWFSC2015; @NPCC2014].
 Additionally, harvest openings and closures, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia rivers fisheries, are predicated on escapements at Lower Granite Dam.
@@
 Typically, the observational error rates of live and video window counts are unknown, and sampling error rates are ignored.
-@Hatch1994 conducted a study of the counting system at Lower Granite dam in 1992, and found that Chinook Salmon were undercounted at the window by in-person window counters compared to video counts.
+@Hatch1994 conducted a study of the counting system at Lower Granite Dam in 1992, and found that Chinook Salmon were undercounted at the window by in-person window counters compared to video counts.
 There was a significant difference between daily standard counts (16 or 10 hours per day) and daily 24-hour video counts for both Chinook Salmon and steelhead, accounted for by estimates of passage rates while the counting window is closed of 3.5% (Chinook Salmon) and 6.6% (steelhead).
@@
 We also test whether observed nighttime passage and fallback/re-ascension rates are typically equal.
-Our method for estimating species specific escapement past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of fish previously tagged with passive integrated transponder (PIT) tags in the adult ladder to explicitly model nighttime passage, re-ascension, and observation error using a state-space approach [@royle2008hierarchical] which separates process variance (e.g.
+Our method for estimating species specific escapement past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of fish previously tagged with passive integrated transponder (PIT) tags in the adult ladder to explicitly model nighttime passage, re-ascension, and observation error using a state-space approach [@Royle2008] which separates process variance (e.g.
 week to week variance in true escapment) from observation error variance (e.g.
@@
 To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates.
-We then applied this model to Chinook Salmon and steelhead returns at Lower Granite dam for spawn years 2010 - 2019.
+We then applied this model to Chinook Salmon and steelhead returns at Lower Granite Dam for spawn years 2010 - 2019.
 The STADEM model combines multiple imperfect sources of data to reduce bias in escapement estimates and provides improved estimates of uncertainty.
@@
 For adipose-intact (unclipped) adults, which includes wild and hatchery no-clip individuals, either a portion or all of fish trapped (depending on the year) had scale and genetic tissue samples taken.
-Scale samples were used to estimate age [@Wright2015] and genetic tissue samples were used to determine sex [@Campbell2012a] and estimate the origin of wild fish via genetic stock identification (e.g., @Hargrove2019).
+Scale samples were used to estimate age [@Wright2015] and genetic tissue samples were used to determine sex [@Campbell2012] and estimate the origin of wild fish via genetic stock identification (e.g., @Hargrove2019).
 Prior to 2013, only fish determined to be wild in origin at the trap were sampled for scale and genetics.
@@
 
-We estimated the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for nighttime passage and fallback/re-ascension rates using a state-space modeling approach [@royle2008hierarchical] implemented in the STADEM package for the R statistical software [@R-Core-Team2020].
+We estimated the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for nighttime passage and fallback/re-ascension rates using a state-space modeling approach [@Royle2008] implemented in the STADEM package for the R statistical software [@R-Core-Team2020].
 We assumed that the window counts and the estimates from the trap (fish in the trap divided by trap rate that week) were generated by processes with observation error.
@@
 Second, by incorporating both sources of information in a state-space framework, STADEM incorporates missing data at either the observation window or adult trap seamlessly.
-At Lower Granite Dam, the adult trap has been closed for brief or extended periods of time (i.e., days, weeks) intermittently in five of the last ten years, often during peak run times [@USACE2010; @USACE2011; @USACE2012; @USACE2013; @Ogden2016; @USACE2015; @USACE2016; @USACE2017; @USACE2018; @USACE2019].
+At Lower Granite Dam, the adult trap has been closed for brief or extended periods of time (i.e., days, weeks) intermittently in five of the last ten years, often during peak run times [@USACE2010; @USACE2011; @USACE2012; @USACE2013; @Ogden2016a; @USACE2015; @USACE2016; @USACE2017; @USACE2018; @USACE2019].
 Trap closures are typically associated with elevated water temperatures resulting in potential fish handling stress and/or trap malfunctions [@Ogden2016b].
@@
 
-Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary or population specific spawning areas [@IPTDSW2020; @kinzer2020report].
+Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary or population specific spawning areas [@IPTDSW2020; @Kinzer2020].
 Estimates of species and origin specific escapement at Lower Granite Dam, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates [e.g., @Camacho2018; @Schrader2013] that are important for fisheries management and productivity monitoring of wild populations.
 As an example, STADEM is being applied at Lower Granite Dam to estimate the total unique wild fish migrating past the dam.
-Estimates of fish passing the dam are then combined with estimated movement or transition probabilities based on PIT tag observations at instream PIT tag detection systems throughout the Snake River basin, similar to @waterhouse2020bayesian, to estimate escapement to Snake River populations and locations throughout the basin [@orme2018population].
+Estimates of fish passing the dam are then combined with estimated movement or transition probabilities based on PIT tag observations at instream PIT tag detection systems throughout the Snake River basin, similar to @Waterhouse2020, to estimate escapement to Snake River populations and locations throughout the basin [@Orme2019].
 Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations.
@@
                      name = "Source") +
-  theme_pubr() +
-  theme(legend.position = 'bottom') +
+  theme_pubr(legend = "bottom") +
   facet_wrap(~ plot_title,

### 2020-09-18 analysis/paper/paper.Rmd
edits to manuscript Intro

@@
 institute:
-  - BM: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
+  - BM: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
   - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
@@
 Here we describe a state-space model that estimates such escapement past a dam by using window counts, passive integrated transponder (PIT) tag observations and data from an adult fish trap, accounting for issues such as nighttime passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate.
-We tested this approach using a simulation framework that mimicked several levels of observation error, differences between nighttime passage and re-ascension rates and the possibility of the adult trap being closed for some period of time.
+We tested the approach using a simulation framework that mimicked several levels of observation error, differences between nighttime passage and re-ascension rates and the possibility of the adult trap being closed for some period of time.
 Our results demonstrate that the model produced unbiased estimates across all tested scenarios.
@@
 
-Fish escapement often refers to the number of adults that survive juvenile and subadult rearing, escape harvest and return to their natal habitat to potentially spawn [e.g.
-@Bue1998].
+Fish escapement often refers to the number of adults that survive juvenile and subadult rearing, escape harvest and return to their natal habitat to potentially spawn [e.g., @Bue1998].
 For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their spawning area.
@@
 Populations of Chinook Salmon *Oncorhynchus tshawytscha* and steelhead trout *O.
-mykiss* in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric dams on the Snake and Columbia rivers) and tributary habitats [@Nehlsen1991; @McClure2003; @NWFSC2015].
-As a result, Snake River spring-summer run (arriving at Lower Granite between March 1 and August 17) Chinook Salmon (hereafter Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937).
+mykiss* in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric projects on the Snake and Columbia rivers) and tributary habitats [@Nehlsen1991; @McClure2003; @NWFSC2015].
+As a result, Snake River spring-summer run Chinook Salmon (hereafter Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937).
 Snake River Chinook Salmon and steelhead have substantial cultural, recreational, commercial and subsistence value within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries.
@@
 
-The majority of Chinook Salmon and steelhead returning to the Snake River must ascend a fish ladder on Lower Granite Dam before migrating to their natal tributary spawning locations.
+The majority of Chinook Salmon and steelhead returning to the Snake River must ascend a fish ladder on Lower Granite Dam before migrating upriver.
 Fish management agencies previously used counts of fish, by species, passing an observation window within the fish ladder as a census of fish escaping past the dam.
@@
 However, using window counts as a census of Chinook Salmon and steelhead passing Lower Granite Dam can be problematic as it fails to account for multiple sources of uncertainty and disregards known biological processes.
-In-person window counts where observers look directly into the fish ladders to identify and count all passing fish, by species, occur for 50 minutes per hour, 16 hours a day from April through October (which corresponds with peak run timing for several species).
+In-person window counts, where observers look directly into the fish ladders to identify and count all passing fish, by species, occur for 50 minutes per hour, 16 hours a day from April through October (which corresponds with peak run timing for several species).
 Counts are then expanded to provide an estimate for the entire hour [@USACE2015].
@@
 @Hatch1994 conducted a study of the counting system at Lower Granite Dam in 1992, and found that Chinook Salmon were undercounted at the window by in-person window counters compared to video counts.
-There was a significant difference between daily standard counts (16 or 10 hours per day) and daily 24-hour video counts for both Chinook Salmon and steelhead, accounted for by estimates of passage rates while the counting window is closed of 3.5% (Chinook Salmon) and 6.6% (steelhead).
+There was a significant difference between daily standard counts (16 or 10 hours per day) and daily 24-hour video counts; they found a significant nighttime passage rate (crossing while the counting window was closed) with 3.5% of Chinook Salmon and 6.6% of steelhead adults passing during "nighttime" hours.
 They also found some species misidentification issues, which could result in either under- or over-counting of a particular species.
-While that study identified potential issues with the current window counting procedure, it was only conducted in a single year, nearly 30 years ago, at a time when Chinook Salmon and steelhead total returns were lower than the past decade.
+While that study identified potential issues with the current window counting procedure, it was only conducted in a single year, nearly 30 years ago, at a time when Chinook Salmon and steelhead total returns were lower than in the past decade.
 
-Besides potential errors in the window counts, two biological processes are unaccounted for: 1) fish that cross the dam during the 8+ hours when the window is unmonitored (i.e., nighttime passage) which may result in an underestimate of escapement, and 2) fish that migrate through the ladder and past the dam may fallback over the dam (e.g., over spillway, through navigation locks) and later may re-ascend the fish ladder again [@Boggs2004] and be double-counted at the window (re-ascension).
-Both fallback with and without re-ascension potentially result in an overestimate of escapement [@Dauble2000].
-Previously, it was assumed that nighttime passage rates and fallback/re-ascension rates canceled each other out resulting in window counts providing an unbiased estimate of escapement [@Camacho2018].
+Besides potential errors in the window counts, two biological processes are unaccounted for: 1) fish that cross the dam during the 8+ hours when the window is unmonitored (i.e., nighttime passage) which may result in an underestimate of escapement, and 2) fish that migrate through the ladder and past the dam may fallback over the dam (e.g., over a  spillway, through navigation locks) and later may re-ascend the fish ladder again [@Boggs2004] and be double-counted at the window (re-ascension).
+Both fallback without and with re-ascension potentially result in an overestimate of escapement [@Dauble2000].
+Previously, it was often assumed that nighttime passage rates and fallback/re-ascension rates canceled each other out resulting in window counts providing an unbiased estimate of escapement [@Camacho2018].
 
@@
 We also test whether observed nighttime passage and fallback/re-ascension rates are typically equal.
-Our method for estimating species specific escapement past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of fish previously tagged with passive integrated transponder (PIT) tags in the adult ladder to explicitly model nighttime passage, re-ascension, and observation error using a state-space approach [@Royle2008] which separates process variance (e.g.
-week to week variance in true escapment) from observation error variance (e.g.
-observation error at the window, or sampling variance at the trap).
+Our method for estimating species-specific escapement past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of fish previously tagged with passive integrated transponder (PIT) tags in the adult ladder to explicitly model nighttime passage, re-ascension, and observation error using a state-space approach [@Royle2008] which separates process variance (e.g., week to week variance in true escapment) from observation error variance (e.g., observation error at the window, or sampling variance at the trap).
 To meet desired management and conservation objectives, modeled escapement includes estimates of uncertainty and is parsed into weekly strata.
@@
 Estimates of escapement account for fish that migrate through the ladder outside of observation hours (nighttime passage) and those that ascend the ladder multiple times (re-ascension).
-Our model is implemented in the **ST**ate-space **A**dult **D**am **E**scapement **M**odel (STADEM) package for the statistical software R [@R-Core-Team2020], and is available for downloaded at [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM).
+Our model is implemented in the **ST**ate-space **A**dult **D**am **E**scapement **M**odel (STADEM) package for the statistical software R [@R-Core-Team2020], and is available for download or installation from [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM).
 To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates.
@@
 We used STADEM and three sources of data to estimate Chinook Salmon and steelhead escapement at Lower Granite Dam from 2010-2019.
-Data sources included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of fish PIT tagged before they reached Lower Granite Dam detected in the adult fish ladder.
+Data sources included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of fish PIT tagged prior to arriving at Lower Granite Dam and detected in the adult fish ladder.
 Below, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; similar data could likely be obtained from other fish passage facilities.

### 2020-09-21 analysis/paper/paper.Rmd
finished edits to body of ms

@@
 output:
+    bookdown::html_document2:
+      fig_caption: yes
+      number_sections: FALSE
+      toc: no
+      fig_width: 7
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks2.lua
+      - --lua-filter=../templates/pagebreak.lua
+    bookdown::word_document2:
+      fig_caption: yes
+      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+      number_sections: FALSE
+      toc: no
+      always_allow_html: true
+      fig_width: 6
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
     bookdown::pdf_document2:
@@
         in_header: ../templates/header_manuscript.tex
-    bookdown::word_document2:
-      fig_caption: yes
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
-      number_sections: FALSE
-      toc: no
-      always_allow_html: true
-      fig_width: 6
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
-    bookdown::html_document2:
-      fig_caption: yes
-      number_sections: FALSE
-      toc: no
-      fig_width: 7
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks2.lua
-      - --lua-filter=../templates/pagebreak.lua
 fontsize: 12pt
@@
 
-Daily counts of adult Chinook Salmon and steelhead passing an observation window located on the Lower Granite Dam fish ladder were made and provided by the US Army Corps of Engineers.
+Daily counts of adult Chinook Salmon and steelhead passing the observation window located on the Lower Granite Dam fish ladder were estimated and provided by the US Army Corps of Engineers.
 When summed, they provide an estimate of the number of fish ascending and passing Lower Granite Dam each season.
-Window counts were made for each species using video monitoring and direct in-person visual monitoring during daytime hours [@Hatch1994].
-Video monitoring occurred during the beginning and tail ends of the adult runs (March 1 – March 31 and November 1 – December) for 10 hours per day (0600 – 1600 hours).
-Direct visual monitoring occurred during peak run times (April 1 – October 31) for 16 hours per day (0400 – 2000 hours) [@USACE2015].
+Window counts were made for each species using direct in-person visual monitoring and video monitoring during daytime hours [@Hatch1994].
+Direct visual monitoring occurred during peak run times (April 1 – October 31) for 16 hours per day (0400 – 2000 hours).
+Video monitoring occurred during the beginning and tail ends of the adult runs (March 1 – March 31 and November 1 – December) for 10 hours per day (0600 – 1600 hours) [@USACE2015].
 During direct visual monitoring, observers recorded each adult (≥ 30cm), by species, passing the window for 50 minutes of each hour of operation.
-Salmonids under 30cm in length were not identified to species.
 The sum of the daily 50-minute counts were then multiplied by 1.2 to account for the 10 minutes when fish were not counted.
@@
 The second source of data came from a sample of fish collected in the adult trap as they migrated past Lower Granite Dam [@Ogden2016b].
-The trap, also located within the adult fish ladder and upstream of the observation window, provided biological data (e.g., origin [wild, hatchery], genetic stock, length, age, sex) for captured adults that allowed decomposition of the escapement into specific groups [e.g., @Camacho2018; @Steinhorst2017].
+The trap, also located within the adult fish ladder and upstream of the observation window, provides an opportunity to collect biological data (e.g., origin [wild, hatchery], genetic stock, length, age, sex) from captured adults which allows for decomposition of the escapement into specific groups [e.g., @Camacho2018; @Steinhorst2017].
 The trap was operational for 24 hours per day and randomly sampled the run by opening four times per hour for a length of time determined by a set daily trapping rate.
@@
 
-All captured fish were anesthetized, speciated, examined for existing marks/tags, measured for fork length and visually identified as wild or hatchery.
-The most widely usee marking of hatchery fish is the adipose fin clip, although coded wire tags are used in less than 10% of the hatchery releases.
-Some subset of hatchery fish are either intentionally or unintentionally released without an adipose fin clip, and these are referred to as unclipped hatchery fish, or hatchery no-clip (HNC).
-For adipose-intact (unclipped) adults, which includes wild and hatchery no-clip individuals, either a portion or all of fish trapped (depending on the year) had scale and genetic tissue samples taken.
-Scale samples were used to estimate age [@Wright2015] and genetic tissue samples were used to determine sex [@Campbell2012] and estimate the origin of wild fish via genetic stock identification (e.g., @Hargrove2019).
+All captured fish were anesthetized, speciated, examined for existing marks or tags, measured for fork length and visually identified as wild or hatchery.
+The most widely used marking of hatchery fish is a clipped (removed) adipose fin, although coded wire tags are used in less than 10% of the hatchery releases.
+Some subset of hatchery fish are either intentionally or unintentionally released without a clipped adipose fin, and these are referred to as unclipped hatchery fish, or hatchery no-clips (HNC).
+For adults with intact adipose fins (unclipped), which includes wild and hatchery no-clip individuals, either a portion or all of fish trapped (depending on the year) had scale and genetic tissue samples taken.
+Scale samples were used to estimate age [@Wright2015] and genetic tissue samples were used to determine sex [@Campbell2012] and estimate the location of origin of wild fish using genetic stock identification (e.g., @Hargrove2019).
 Prior to 2013, only fish determined to be wild in origin at the trap were sampled for scale and genetics.
@@
 Prior to release, all non-PIT tagged fish with an intact adipose fin (i.e., putatively wild) received a PIT tag.
-Final determination of wild, clipped hatchery, or unclipped hatchery origins were assigned using a post-hoc analysis of marks and tags, including parentage-based tqagging [@Steele2013; @Steele2019].
+Final determination of wild, clipped hatchery, or unclipped hatchery origins were assigned using a post-hoc analysis of marks and tags, including parentage-based tagging [@Steele2013; @Steele2019].
 Data from the adult trap were collected and managed by multiple agencies and were made available by the Idaho Department of Fish and Game [@Camacho2018].
@@
 The last source of data was observations of PIT tagged adult Chinook Salmon and steelhead at detection sites located in the Lower Granite Dam fish ladder.
-These observations provide estimates of 1) a trapping rate, 2) the nighttime passage rate, and 3) the re-ascension rate.
-Detections used in the model include all fish that were previously PIT tagged as juveniles or adults prior to reaching Lower Granite Dam (i.e., does not include newly tagged adults at the dam) and detected at adult detection sites in the dam passage system.
+These observations provide daily estimates of 1) a trapping rate, 2) a nighttime passage rate, and 3) a re-ascension rate.
+Detections used in the model include all fish that were previously PIT tagged as juveniles or adults prior to arriving at Lower Granite Dam (i.e., does not include newly tagged adults at the dam) and detected at adult detection sites in the dam passage system.
 PIT tag data was provided through DART and the adult ladder PIT tag query; [http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail](http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail).
 
-A trap rate estimate was derived using Lincoln-Peterson mark-recapture methods and PIT tag observations of both Chinook Salmon and steelhead at Lower Granite Dam adult detection sites.
-The "mark" group included all tags (of both species) detected in the adult trap and the "capture" group included tags observed to cross the weir at the upstream end of the fish ladder as adults left the passage system.
-The proportion of the tags detected at the weir that were also caught in the trap each week was assumed to reflect the same trap rate that all adults of the target species experienced as they crossed the ladder.
+A trap rate estimate was derived using Lincoln-Peterson mark-recapture methods [@Seber2002] and PIT tag observations of both Chinook Salmon and steelhead at Lower Granite Dam adult detection sites.
+The "mark" group included all tags (both species) detected in the adult trap and the "capture" group included tags observed to cross the weir at the upstream end of the fish ladder as adults left the passage system.
+The proportion of tags detected leaving the passage system that were also caught in the trap each week was assumed to reflect the trap rate experienced by target species as they migrated through the ladder.
 The sample size of previously tagged fish detected at Lower Granite Dam influences the uncertainty in that trap rate, which is also informed by the number of adults caught in the trap and the window counts.
-The set trapping rate (i.e.
-the recorded time that the trap is open to trap adults) does not always reflect the true proportion of fish that are captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error, among others.
+The set trapping rate (i.e., the recorded time that the trap is open to trap adults) does not always reflect the true trapping rate or proportion of fish that are actually captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error, among others.
 Therefore, we use the mark-recapture approach to estimate a "true" trapping rate.
 
-We used PIT tag data to estimate nighttime passage and re-ascension rates, using the total number of tags passing the fish ladder for both and estimating both on a weekly basis.
+We used PIT tag data to estimate nighttime passage and re-ascension rates, using the total number of tags passing the fish ladder for both and estimating each on a weekly basis.
 The nighttime passage rate was based on the count of PIT tags that migrated through the fish ladder during non-window observation hours and the total number of tags passing the fish ladder.
@@
 
-We estimated the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for nighttime passage and fallback/re-ascension rates using a state-space modeling approach [@Royle2008] implemented in the STADEM package for the R statistical software [@R-Core-Team2020].
+We estimated the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for estimated nighttime passage and fallback/re-ascension rates using a state-space modeling approach [@Royle2008].
+Our model is implemented in the STADEM package for the R statistical software [@R-Core-Team2020].
 We assumed that the window counts and the estimates from the trap (fish in the trap divided by trap rate that week) were generated by processes with observation error.
 In the case of the trap, for example, we assumed there was sampling variation and uncertainty around our estimates of the true unknown trap rate.
-STADEM adjusted the window counts for the nighttime passage rate, and both window and trap estimates for the re-ascension rate to estimate the number of unique fish that crossed teh dam.
+STADEM adjusted the window counts for the nighttime passage rate, and both window and trap estimates for the re-ascension rate to estimate the number of unique fish that crossed the dam.
 Finally, adult sampling data from the trap (wild, hatchery, hatchery no-clip) were used to partition the total escapement estimate by origin (Figure \@ref(fig:stadem-examp-fig)).
@@
 The simulation parameters such as proportion of origin, run-timing, nighttime passage rates, fallback and re-ascension rates and trap rates were based on observed values at Lower Granite Dam between 2010-2015.
-We generated 99 simulations for each scenario, and ran STADEM on each one.
+We generated 99 simulations for each scenario, and ran STADEM on each one, resulting in a total of 1,188 model runs.
 Further details about simulation procedures can be found in [Appendix B](#append2).
 
-We grouped the simulation results by origin (wild, hatchery clipped and hatchery no-clip) and evaluated them by several measures.
+We grouped the simulation results by fish origin and evaluated them by several measures.
 Relative bias is the difference between the simulated "truth" and the STADEM estimate, divided by the simulated "truth".
@@
 
-Finally, we applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead returning to the Snake River during spawn years 2010 to 2019.
+We applied STADEM to empirical data from Lower Granite Dam for Chinook Salmon and steelhead returning to the Snake River during spawn years 2010 to 2019.
 Window counts for both species were accessed from DART via functions within STADEM.
@@
 STADEM results were very similar across all scenarios (Figure \@ref(fig:rel-bias-lgd)).
-Estimates were unbiased, with an average relative bias of 0.2–0.3%.
+Estimates of wild escapement were unbiased, with an average relative bias of 0.2–0.3%.
 The CV of the estimates averaged 2.0–3.0%, with higher CV's being associated with scenarios when the trap was closed for 3 weeks.
@@
 Weekly estimates of total escapement over Lower Granite Dam tracked the window counts and trap estimates (One example: Figure \@ref(fig:jags-fit-fig).
-Similar plots for all model runs are available on the manuscript GitHub page: [https://www.github.com/KevinSee/ManuscriptSTADEM]([https://www.github.com/KevinSee/ManuscriptSTADEM])).
+Similar plots for all model runs [each species and year combination] are available on the manuscript GitHub page: [https://www.github.com/KevinSee/ManuscriptSTADEM]([https://www.github.com/KevinSee/ManuscriptSTADEM])).
 STADEM point estimates were often between estimates based on window counts and those based on the number of fish caught in the adult trap.
@@
 We have presented a novel method for estimating adult salmonid escapement past a large hydroelectric facility (e.g., Lower Granite Dam) that incorporates data from window counts, a fish trap, and observations of PIT tagged fish in the adult passage ladder.
-Our model explicitly incorporates nighttime passage, re-ascension, and potential error in both window and trap estimates.
+Our model explicitly accounts for nighttime passage, re-ascension, and potential error in both window and trap estimates.
 In doing so, we demonstrated that at Lower Granite Dam, nighttime passage and re-ascension rates do not always offset each other, and assuming they do will lead to biased escapement estimates in some years.
-With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall Chinook Salmon, Pacific lamprey *Lampetra tridentata*), or elsewhere, provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run and tag observation or detection infrastructure (e.g., a PIT tag detection array or similar) exists.
+With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall Chinook Salmon, Pacific lamprey *Lampetra tridentata*), or elsewhere, provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run and tag observation or detection infrastructure (e.g., a PIT tag detection array or similar).
 Our state-space model combined multiple imperfect sources of data to reduce bias in adult escapement estimates and provided quantitative estimates of uncertainty.
-Accurate population or stock abundance estimates and uncertainty accounting for observation and process error can be particularly important when estimates are used or leveraged for management and conservation decisions such as population viability analyses.
+Accurate population or group abundance estimates and uncertainty accounting for observation and process error can be particularly important when estimates are used for management and conservation decisions such as population viability analyses.
 
@@
 
-Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary or population specific spawning areas [@IPTDSW2020; @Kinzer2020].
-Estimates of species and origin specific escapement at Lower Granite Dam, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates [e.g., @Camacho2018; @Schrader2013] that are important for fisheries management and productivity monitoring of wild populations.
+Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate escapement of spring-summer run Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary or population specific spawning areas [@IPTDSW2020; @Kinzer2020].
+Estimates of species and origin-specific escapement at Lower Granite Dam, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates [e.g., @Camacho2018; @Schrader2013] that are important for fisheries management and productivity monitoring of wild populations.
 As an example, STADEM is being applied at Lower Granite Dam to estimate the total unique wild fish migrating past the dam.

### 2020-09-21 analysis/paper/paper.Rmd
edits to captions and re-knit .html

@@
         linesep = "",
-        caption = "Summary of simulation scenarios including varying adult trapping, fallback and re-ascension rates, nighttime passage, and window count error rates use to evaluate the performance of STADEM.") %>%
+        caption = "Summary of simulation scenarios including varying adult trapping, fallback and re-ascension, nighttime passage, and window count error rates used to evaluate the performance of STADEM.") %>%
   kable_styling(latex_options = c('scale_down',
@@
 
-```{r stadem-examp-fig, fig.width=3.5, fig.asp = 2, fig.cap = "Schematic of how the STADEM model works.
+```{r stadem-examp-fig, fig.width=3.5, fig.asp = 2, fig.cap = "Schematic showing the STADEM model framework.
 (A) shows the posterior of the estimate of fish crossing the dam while the window is open (dashed line shows observed window counts).
@@
 
-```{r jags-fit-fig, fig.cap = "Time-series plot showing estimates of total escapement for Chinook in 2014, including window counts, trap estimates and STADEM estimates.
+```{r jags-fit-fig, fig.cap = "Time-series plot showing estimates of total escapement for Chinook Salmon in 2014, including window counts, trap estimates and STADEM estimates.
 The gray ribbon represents the 95% credible intervals for STADEM estimates, while the red ribbon represents the 95% confidence intervals for the trap estimates."}
@@
 ```{r night-reasc-diff-fig, fig.height=5, fig.cap = "Nighttime passage rate plotted against re-ascension rate on the logit scale, calculated from observed PIT tags for each week of spawn years 2010-2019.
-Colors correspond to different spawn year, while the size of each point is proportional to the window count that week.
+Colors correspond to different spawn years, while the size of each point is proportional to the window count that week.
 The dashed line is the 1-1 line."}

### 2020-09-23 analysis/paper/paper.Rmd
added code to refine the figures, and save black and white and color versions of them as .tiff files, as required by NAFMJ. Added all .tiff files to Git LFS.

@@
 output:
-    bookdown::html_document2:
-      fig_caption: yes
-      number_sections: FALSE
-      toc: no
-      fig_width: 7
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks2.lua
-      - --lua-filter=../templates/pagebreak.lua
     bookdown::word_document2:
@@
   fig.path = "../figures/",
-  dpi = 600
+  dpi = 300
 )
+
+# set default values for saving alternative figures (color versions)
+my_dpi = knitr::opts_chunk$get()$dpi
+my_ext = "tiff"
 ```
@@
 Special thanks to Columbia Basin Research staff and the Columbia River Data Access in Real Time (DART) application, and Susannah Iltis in particular.
-And finally, thank you to Rick Orme for his contributions to the development of STADEM, and all the other folks who have contributed through productive critique and conversations.
+And finally, thank you to Rick Orme for his contributions to the development of STADEM, and all the other people who have contributed through productive critique and conversations.
 
@@
 
+\listoffigures
+
+\newpage
+
 # Figures
 
-```{r stadem-examp-fig, fig.width=3.5, fig.asp = 2, fig.cap = "Schematic showing the STADEM model framework.
+```{r stadem-examp-fig, fig.height = 5, fig.cap = "Schematic showing the STADEM model framework.
 (A) shows the posterior of the estimate of fish crossing the dam while the window is open (dashed line shows observed window counts).
@@
   scale_color_brewer(palette = "Spectral") +
-  # scale_color_grey(start = 0.9,
-  #                  end = 0.3) +
   geom_vline(xintercept = 4,
@@
 
-examp_p = ggarrange(p1, p2a, p3a, p4,
+examp_col = ggarrange(p1, p2a, p3a, p4,
                     labels = 'AUTO',
-                    ncol = 1,
-                    nrow = 4,
+                    ncol = 2,
+                    nrow = 2,
                     align = 'v',
@@
 
-examp_p
+bw_scale = c(0.9, 0.1)
+examp_bw = ggarrange(p1 +
+                       scale_color_grey(start = bw_scale[1],
+                                        end = bw_scale[2]), 
+                   p2a +
+                       scale_color_grey(start = bw_scale[1],
+                                        end = bw_scale[2]), 
+                   p3a +
+                       scale_color_grey(start = bw_scale[1],
+                                        end = bw_scale[2]), 
+                   p4 +
+                       scale_color_grey(start = bw_scale[1],
+                                        end = bw_scale[2]),
+                    labels = 'AUTO',
+                    ncol = 2,
+                    nrow = 2,
+                    align = 'v',
+                    common.legend = T,
+                    legend = 'bottom')
+
+examp_bw
+
+# save black and white version
+ggsave(paste0("../figures/stadem-examp-fig-bw.", my_ext),
+       examp_bw,
+       height = 5,
+       width = 7,
+       dpi = my_dpi)
+
+# save color version
+ggsave(paste0("../figures/stadem-examp-fig-col.", my_ext),
+       examp_col,
+       height = 5,
+       width = 7,
+       dpi = my_dpi)
 
@@
 rel_bias_p
+
+# save black and white version
+ggsave(paste0("../figures/rel-bias-lgd-bw.", my_ext),
+       rel_bias_p,
+       height = 5,
+       width = 3.5,
+       dpi = my_dpi)
+
 ```
@@
 
-ts_col
-# ts_bw
+# ts_col
+ts_bw
 
+# save black and white version
+ggsave(paste0("../figures/jags-fit-fig-bw.", my_ext),
+       ts_bw,
+       height = 5,
+       width = 7,
+       dpi = my_dpi)
+
+# save color version
+ggsave(paste0("../figures/jags-fit-fig-col.", my_ext),
+       ts_col,
+       height = 5,
+       width = 7,
+       dpi = my_dpi)
 ```
@@
 ```{r night-reasc-diff-fig, fig.height=5, fig.cap = "Nighttime passage rate plotted against re-ascension rate on the logit scale, calculated from observed PIT tags for each week of spawn years 2010-2019.
-Colors correspond to different spawn years, while the size of each point is proportional to the window count that week.
+The size of each point is proportional to the window count that week.
 The dashed line is the 1-1 line."}
@@
        size = "Window Count") +
-  theme_pubr(legend = "left")
+  theme_pubr(legend = "left") +
+  theme(axis.text = element_text(size = 10))
 
@@
              color = 'black',
-             alpha = 0.3)
+             alpha = 0.3) +
+  labs(size = "Window Count") +
+  theme_pubr(legend = "bottom") +
+  theme(axis.text = element_text(size = 10)) +
+  guides(size = guide_legend(nrow = 2))
 
-rate_p_col
-# rate_p_bw
+# rate_p_col
+rate_p_bw
+
+# save black and white version
+ggsave(paste0("../figures/night-reasc-diff-fig-bw.", my_ext),
+       rate_p_bw,
+       width = 3.5,
+       height = 7,
+       dpi = my_dpi)
+
+# save color version
+ggsave(paste0("../figures/night-reasc-diff-fig-col.", my_ext),
+       rate_p_col,
+       width = 5,
+       height = 7,
+       dpi = my_dpi)
 ```

### 2020-09-23 analysis/paper/paper.Rmd
added Brice Semmens to the Acknowlegements section. Changed caption of time-series plot to reflect black and white version. Divided window count size by 1000 in rate comparison figure.

@@
 output:
+    bookdown::html_document2:
+      fig_caption: yes
+      number_sections: FALSE
+      toc: no
+      fig_width: 7
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks2.lua
+      - --lua-filter=../templates/pagebreak.lua
     bookdown::word_document2:
@@
       - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
-    bookdown::html_document2:
-      fig_caption: yes
-      number_sections: FALSE
-      toc: no
-      fig_width: 7
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks2.lua
       - --lua-filter=../templates/pagebreak.lua
@@
 Special thanks to Columbia Basin Research staff and the Columbia River Data Access in Real Time (DART) application, and Susannah Iltis in particular.
-And finally, thank you to Rick Orme for his contributions to the development of STADEM, and all the other people who have contributed through productive critique and conversations.
+And finally, thank you to Rick Orme and Brice Semmens for their contributions to the development of STADEM, and all the other people who have contributed through productive critique and conversations.
 
@@
 ```{r jags-fit-fig, fig.cap = "Time-series plot showing estimates of total escapement for Chinook Salmon in 2014, including window counts, trap estimates and STADEM estimates.
-The gray ribbon represents the 95% credible intervals for STADEM estimates, while the red ribbon represents the 95% confidence intervals for the trap estimates."}
+The dark gray ribbon represents the 95% credible intervals for STADEM estimates, while the light gray ribbon represents the 95% confidence intervals for the trap estimates."}
 
@@
   scale_color_brewer(palette = "Paired") +
-  geom_point(aes(size = win_cnt,
+  geom_point(aes(size = win_cnt / 1000,
                  color = as.factor(Year))) +
   labs(color = "Spawn Year",
-       size = "Window Count") +
+       size = "Window Count\n (x 1,000)") +
   theme_pubr(legend = "left") +
@@
 rate_p_bw = rate_p_base +
-  geom_point(aes(size = win_cnt),
+  geom_point(aes(size = win_cnt / 1000),
              shape = 21,
@@
              alpha = 0.3) +
-  labs(size = "Window Count") +
+  labs(size = "Window Count\n (x 1,000)") +
   theme_pubr(legend = "bottom") +

### 2020-09-23 analysis/paper/paper.Rmd
shrunk the axis text on figure 3. Added AFS style guide to "templates" folder

@@
 output:
-    bookdown::html_document2:
-      fig_caption: yes
-      number_sections: FALSE
-      toc: no
-      fig_width: 7
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks2.lua
-      - --lua-filter=../templates/pagebreak.lua
-    bookdown::word_document2:
-      fig_caption: yes
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
-      number_sections: FALSE
-      toc: no
-      always_allow_html: true
-      fig_width: 6
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
     bookdown::pdf_document2:
@@
         in_header: ../templates/header_manuscript.tex
+    bookdown::word_document2:
+      fig_caption: yes
+      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+      number_sections: FALSE
+      toc: no
+      always_allow_html: true
+      fig_width: 6
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
+    bookdown::html_document2:
+      fig_caption: yes
+      number_sections: FALSE
+      toc: no
+      fig_width: 7
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks2.lua
+      - --lua-filter=../templates/pagebreak.lua
 fontsize: 12pt
@@
   theme_pubr(legend = "none") +
+  theme(axis.text = element_text(size = 10)) +
   labs(x = 'Relative Bias',

### 2020-09-24 analysis/paper/paper.Rmd
removed comments that have been resolved.
Deleted a reference to a manuscript in review. re-knit

@@
 output:
+    bookdown::word_document2:
+      fig_caption: yes
+      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+      number_sections: FALSE
+      toc: no
+      always_allow_html: true
+      fig_width: 6
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
     bookdown::pdf_document2:
@@
         in_header: ../templates/header_manuscript.tex
-    bookdown::word_document2:
-      fig_caption: yes
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
-      number_sections: FALSE
-      toc: no
-      always_allow_html: true
-      fig_width: 6
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
     bookdown::html_document2:
@@
 For adults with intact adipose fins (unclipped), which includes wild and hatchery no-clip individuals, either a portion or all of fish trapped (depending on the year) had scale and genetic tissue samples taken.
-Scale samples were used to estimate age [@Wright2015] and genetic tissue samples were used to determine sex [@Campbell2012] and estimate the location of origin of wild fish using genetic stock identification (e.g., @Hargrove2019).
+Scale samples were used to estimate age [@Wright2015] and genetic tissue samples were used to determine sex [@Campbell2012] and estimate the location of origin of wild fish using genetic stock identification [e.g., @Hargrove2019].
 Prior to 2013, only fish determined to be wild in origin at the trap were sampled for scale and genetics.
@@
 The model's uncertainty is always smaller than the uncertainty from the trap estimates alone, whereas the window counts alone provide no estimate of uncertainty.
-
-<!-- John Hargrove: Is there a reason why you only present a singular year in Figure 3?
-I acknowledge that having 10 years of data may make it unreadable – but perhaps justify why 2014 was selected or include all 10 years as separate plots using facet_wrap in ggplot.
-
-I think showing 10 time series plots is unnecessary (Table 3 shows how estimates match up with window counts).
-I think 2014 is a nice illustration of general agreement, with some examples of how STADEM doesn't always track the trap estimate, and how it deals with missing data.
-Do we need to make that more clear, or leave it as is?
-
-After experimenting with 4 years, Ryan and Kevin decided even 4 was too busy. Let's stick with one, and note that the rest are on the github repo-->
 
@@
 Currently, the only roadblock to this at Lower Granite Dam is the identification of hatchery no-clip fish using genetic tissue samples [@Steele2013; @Steele2019] collected at the adult trap, which currently is completed post-hoc after the trapping season.
-The inclusion of genetic information typically results in a reduction in wild escapement estimates and an associated increase in hatchery no-clip escapement [@Hargrove2020].
+The inclusion of genetic information typically results in a reduction in wild escapement estimates and an associated increase in hatchery no-clip escapement.
 However, if in-season management decisions do not require this correction or could accept the potential bias, origin calls at the trap could be used in-season as a first approximation to escapement.
@@
 With sex and age data collected at the adult fish trap [@Hargrove2019], this approach provides necessary information to evaluate productivity and population viability for select Snake River Chinook Salmon and steelhead groups [@IPTDSW2020].
-
-<!-- John Hargrove: Have you considered comparing estimate that you produced  to those that failed to account for fallback/reascension to see how much difference there is?
-In other words, you could draw estimates from Camacho and Schrader docs and readily compare – may serve to better illustrate your point of the superiority of accounting for these errors relative to less complicated models
-
-I quickly pulled together the estimates from IDFG's latest report, 2010-2018 wild escapement.
-Their estimates of wild Chinook are about 4% higher than STADEM, and about 3% lower for wild steelhead.
-So far we've avoided explicitly comparing our results with existing methods in an effort to avoid confrontation.
-Maybe that's moot if IDFG has started adopting STADEM results, but looks like as of March 2019 they had not.
-
-Ryan and Kevin decided to not go down this path.
-One quick comparison is total escapement vs.
-window counts; which shows up in one of our tables, and we discuss it briefly in the results.
-The other is figure 4, showing nighttime passage and re-ascension rates, and how they don't match.
--->
 

### 2020-09-24 analysis/paper/paper.Rmd
added a few people to the Acknowledgements

@@
       - --lua-filter=../templates/pagebreak.lua
+    bookdown::html_document2:
+      fig_caption: yes
+      number_sections: FALSE
+      toc: no
+      fig_width: 7
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks2.lua
+      - --lua-filter=../templates/pagebreak.lua
     bookdown::pdf_document2:
@@
         in_header: ../templates/header_manuscript.tex
-    bookdown::html_document2:
-      fig_caption: yes
-      number_sections: FALSE
-      toc: no
-      fig_width: 7
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks2.lua
-      - --lua-filter=../templates/pagebreak.lua
 fontsize: 12pt
@@
 Special thanks to Columbia Basin Research staff and the Columbia River Data Access in Real Time (DART) application, and Susannah Iltis in particular.
-And finally, thank you to Rick Orme and Brice Semmens for their contributions to the development of STADEM, and all the other people who have contributed through productive critique and conversations.
+And finally, thank you to Rick Orme and Brice Semmens for their contributions to the development of STADEM, as well as Eric Buhle, Sarah Hoffman, John Hargrove and all the other people who have contributed through productive critique and conversations.
 

### 2021-02-23 analysis/paper/paper.Rmd
added some transparent elements to one plot

@@
   theme_pubr(legend = "bottom") +
+  theme(panel.background = element_rect(color = "transparent"),
+        plot.background = element_rect(color = "transparent"),
+        legend.background = element_rect(color = "transparent")) +
   facet_wrap(~ plot_title,

### 2021-02-23 inst/scripts/SimFnc.R
updated all simulation functions to use purrr rather than plyr

@@ -9,7 +9,7 @@ SimulateLGRdata = function(N.lgr = 100000,
                            hnc.prob = .05,
                            n.pops = 25,
                            hatch.pop.prob = 0.5,
-                           hnc.pop.prob = 0.15,                           
+                           hnc.pop.prob = 0.15,
                            run.mu.mu = 171,
                            run.mu.sd = 13,
                            run.sd.mu = 22,
@@ -38,32 +38,32 @@ SimulateLGRdata = function(N.lgr = 100000,
   # ladder.rate: detection probability of PIT tag antenna in the ladder
   # start.date: date from which simulation begins
   # trap.rate.df: data.frame with Week and trap.rate columns
-  
+
   # required packages to run this function
   # library(MCMCpack)
   # library(lubridate)
-  # library(plyr)
+  # library(purrr)
   # library(dplyr)
   # library(tidyr)
   # library(magrittr)
-  
+
   # quick clean-up of column names
   trap.rate.df %<>%
-    rename(trap_open = trap.open)
-  
+    dplyr::rename(trap_open = trap.open)
+
   # R.V. of population mean run-timing given assigned hyper-parameters
   # R.V. of population mean run-timing given assigned hyper-parameters
   pop.mu <- rnorm(n.pops, run.mu.mu[1], run.mu.sd[1])
-  
+
   # R.V. of population standard deviation of run-timing given assigned hyper-parameters
   pop.sd <- abs(rnorm(n.pops, run.sd.mu[1], run.sd.sd[1]))
-  
+
   # LGR escapement by origin
   if(h.prob > 0){
     N.hat <- N.lgr * h.prob
     hat.pop.p <- rdirichlet(1, alpha = rbinom(n.pops,1,hatch.pop.prob))
       if(!any(is.na(hat.pop.p))){
-        N.hat.pop <- rmultinom(1, N.hat, hat.pop.p)  
+        N.hat.pop <- rmultinom(1, N.hat, hat.pop.p)
       } else {
         N.hat <- 0
         N.hat.pop <- rep(0,n.pops)
@@ -72,12 +72,12 @@ SimulateLGRdata = function(N.lgr = 100000,
     N.hat <- 0
     N.hat.pop <- rep(0,n.pops)
   }
-  
-  if(hnc.prob >0){  
+
+  if(hnc.prob >0){
     N.hnc <- N.lgr * hnc.prob
[…]

### 2021-02-23 inst/scripts/RunSims_Baseline.R
tried to update one round of simulations to use purrr rather than plyr

@@ -7,19 +7,20 @@
 # load needed packages
 library(Hmisc)
 library(MCMCpack)
+library(gtools)
 library(FSA)
 library(Rcapture)
 library(boot)
 library(msm)
 library(lubridate)
 library(magrittr)
-library(plyr)
+library(purrr)
 library(dplyr)
 library(tidyr)
 library(jagsUI)
 
 # this contains the function to simulate data
-source('SimFnc.R')
+source('inst/scripts/SimFnc.R')
 
 #-----------------------------------------------------------------
 # set mcmc parameters
@@ -61,7 +62,8 @@ n_sim = 500
 # make it consistent and constant throughout the season
 my_trap_rate = data.frame(Week = 1:52,
                           trap.rate = 0.15,
-                          trap.open = T)
+                          trap.open = T) %>%
+  as_tibble()
 
 #-----------------------------------------------------------------
 # run simulations

### 2021-03-09 analysis/paper/appendices/Appendix_STADEM_descrip.Rmd
started to address reviewer comments

@@
 
-Thus, the unknown true number of fish crossing LGD each week, $X_t$, is estimated from two different data source: window counts and fish sampled in the trap.
+Thus, the unknown true number of fish crossing LGD each week, $X_t$, is estimated from two different data sources: window counts and fish sampled in the trap.
 The window counts provide an estimate (with some potential observer error) of the fish crossing during daytime hours, while the fish in the trap, when expanded by the estimated true trap rate, provide an estimate of the total fish crossing that week.

### 2021-03-09 analysis/paper/paper.Rmd
started to address reviewer comments

@@
 Our results demonstrate that the model produced unbiased estimates across all tested scenarios.
-We also applied this model to data from the Lower Granite Dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring-summer run Chinook Salmon and steelhead from spawn years 2010-2019.
+We also applied this model to data from the Lower Granite Dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring/summer-run Chinook Salmon and steelhead from spawn years 2010-2019.
 keywords: |
@@
 
-Fish escapement often refers to the number of adults that survive juvenile and subadult rearing, escape harvest and return to their natal habitat to potentially spawn [e.g., @Bue1998].
-For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their spawning area.
+Fish escapement refers to the number of adults that survive juvenile and subadult rearing, escape harvest, and achieve a size and age to potentially spawn [e.g., @Bue1998].
+For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their natal habitat to spawn.
 Escapement estimates facilitate effective fisheries management, particularly estimates of escapement for component groups [e.g., by stock, population, age, origin [wild, hatchery], @Hess2014; @Steinhorst2017; @Camacho2018] which provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations.
@@
 mykiss* in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric projects on the Snake and Columbia rivers) and tributary habitats [@Nehlsen1991; @McClure2003; @NWFSC2015].
-As a result, Snake River spring-summer run Chinook Salmon (hereafter Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937).
-Snake River Chinook Salmon and steelhead have substantial cultural, recreational, commercial and subsistence value within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries.
-The aggregate escapement of Snake River Chinook Salmon and steelhead populations, with the exception of Tucannon River, is monitored at Lower Granite Dam located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations.
+As a result, Snake River spring/summer-run Chinook Salmon (hereafter sp/sum Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937).
+Snake River sp/sum Chinook Salmon and steelhead have substantial cultural, recreational, commercial and subsistence value within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries.
+The aggregate escapement of Snake River sp/sum Chinook Salmon and steelhead populations, with the exception of Tucannon River (a tributary of the Snake River whose confluence is downstream of Lower Granite Dam), is monitored at Lower Granite Dam located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations.
 Many fisheries management and conservation actions are made based on estimates of escapement at Lower Granite Dam parsed by species and origin [@NMFS2019; @NWFSC2015; @NPCC2014].
@@
 
-The majority of Chinook Salmon and steelhead returning to the Snake River must ascend a fish ladder on Lower Granite Dam before migrating upriver.
+The majority of sp/sum Chinook Salmon and steelhead returning to the Snake River must ascend a fish ladder on Lower Granite Dam before migrating upriver.
 Fish management agencies previously used counts of fish, by species, passing an observation window within the fish ladder as a census of fish escaping past the dam.
@@
 
-However, using window counts as a census of Chinook Salmon and steelhead passing Lower Granite Dam can be problematic as it fails to account for multiple sources of uncertainty and disregards known biological processes.
+However, using window counts as a census of sp/sum Chinook Salmon and steelhead passing Lower Granite Dam can be problematic as it fails to account for multiple sources of uncertainty and disregards known biological processes.
 In-person window counts, where observers look directly into the fish ladders to identify and count all passing fish, by species, occur for 50 minutes per hour, 16 hours a day from April through October (which corresponds with peak run timing for several species).
@@
 Typically, the observational error rates of live and video window counts are unknown, and sampling error rates are ignored.
-@Hatch1994 conducted a study of the counting system at Lower Granite Dam in 1992, and found that Chinook Salmon were undercounted at the window by in-person window counters compared to video counts.
-There was a significant difference between daily standard counts (16 or 10 hours per day) and daily 24-hour video counts; they found a significant nighttime passage rate (crossing while the counting window was closed) with 3.5% of Chinook Salmon and 6.6% of steelhead adults passing during "nighttime" hours.
+@Hatch1994 conducted a study of the counting system at Lower Granite Dam in 1992, and found that sp/sum Chinook Salmon were undercounted at the window by in-person window counters compared to video counts.
+There was a significant difference between daily standard counts (16 or 10 hours per day) and daily 24-hour video counts; they found a significant nighttime passage rate (crossing while the counting window was closed) with 3.5% of sp/sum Chinook Salmon and 6.6% of steelhead adults passing during "nighttime" hours.
 They also found some species misidentification issues, which could result in either under- or over-counting of a particular species.
-While that study identified potential issues with the current window counting procedure, it was only conducted in a single year, nearly 30 years ago, at a time when Chinook Salmon and steelhead total returns were lower than in the past decade.
-
-Besides potential errors in the window counts, two biological processes are unaccounted for: 1) fish that cross the dam during the 8+ hours when the window is unmonitored (i.e., nighttime passage) which may result in an underestimate of escapement, and 2) fish that migrate through the ladder and past the dam may fallback over the dam (e.g., over a  spillway, through navigation locks) and later may re-ascend the fish ladder again [@Boggs2004] and be double-counted at the window (re-ascension).
-Both fallback without and with re-ascension potentially result in an overestimate of escapement [@Dauble2000].
-Previously, it was often assumed that nighttime passage rates and fallback/re-ascension rates canceled each other out resulting in window counts providing an unbiased estimate of escapement [@Camacho2018].
+While that study identified potential issues with the current window counting procedure, it was only conducted in a single year, nearly 30 years ago, at a time when sp/sum Chinook Salmon and steelhead total returns were lower than in the past decade.
+
+One additional biological process, the fallback and re-ascension of adult migrants at the dam, should also be accounted for to estimate the true escapement at Lower Granite Dam.
+Adult fish that migrate through the ladder and past the dam may fallback over the dam (e.g., over a spillway, through navigation locks) and later may re-ascend the fish ladder again [@Boogs2004] and be double-counted at the window (re-ascension).
+Whereas unaccounted nighttime passage can lead to an underestimate in escapement, both fallback without and with re-ascension instead potentially lead to an overestimate of escapement [@Dauble2000].
+Previously, it was often assumed that these two biological processes canceled each other out, resulting in window counts providing an unbiased estimate of escapement [@Camacho2018].
 
@@
 To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates.
-We then applied this model to Chinook Salmon and steelhead returns at Lower Granite Dam for spawn years 2010 - 2019.
+We then applied this model to sp/sum Chinook Salmon and steelhead returns at Lower Granite Dam for spawn years 2010 - 2019.
 The STADEM model combines multiple imperfect sources of data to reduce bias in escapement estimates and provides improved estimates of uncertainty.
@@
 
-We used STADEM and three sources of data to estimate Chinook Salmon and steelhead escapement at Lower Granite Dam from 2010-2019.
+We used STADEM and three sources of data to estimate sp/sum Chinook Salmon and steelhead escapement at Lower Granite Dam from 2010-2019.
 Data sources included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of fish PIT tagged prior to arriving at Lower Granite Dam and detected in the adult fish ladder.
@@
 
-Daily counts of adult Chinook Salmon and steelhead passing the observation window located on the Lower Granite Dam fish ladder were estimated and provided by the US Army Corps of Engineers.
+Daily counts of adult sp/sum Chinook Salmon and steelhead passing the observation window located on the Lower Granite Dam fish ladder were estimated and provided by the US Army Corps of Engineers.
 When summed, they provide an estimate of the number of fish ascending and passing Lower Granite Dam each season.
@@
 
-The second source of data came from a sample of fish collected in the adult trap as they migrated past Lower Granite Dam [@Ogden2016b].
+The second source of data are from a sample of fish collected in the adult trap as they migrated past Lower Granite Dam [@Ogden2016b].
 The trap, also located within the adult fish ladder and upstream of the observation window, provides an opportunity to collect biological data (e.g., origin [wild, hatchery], genetic stock, length, age, sex) from captured adults which allows for decomposition of the escapement into specific groups [e.g., @Camacho2018; @Steinhorst2017].
@@
 
-All captured fish were anesthetized, speciated, examined for existing marks or tags, measured for fork length and visually identified as wild or hatchery.
+All captured fish were anesthetized, identified to species, examined for existing marks or tags, measured for fork length and visually identified as wild or hatchery.
 The most widely used marking of hatchery fish is a clipped (removed) adipose fin, although coded wire tags are used in less than 10% of the hatchery releases.
@@
 Prior to 2013, only fish determined to be wild in origin at the trap were sampled for scale and genetics.
-Starting in 2013, every unclipped Chinook Salmon and steelhead trapped at Lower Granite Dam was genotyped to simplify collaborative logistics and better estimate the proportion of unclipped hatchery fish that appear phenotypically wild.
+Starting in 2013, every unclipped sp/sum Chinook Salmon and steelhead trapped at Lower Granite Dam was genotyped to simplify collaborative logistics and better estimate the proportion of unclipped hatchery fish that appear phenotypically wild.
 @Camacho2018 provide further details on trap sample rates and valid sample selection.
@@
 
-The last source of data was observations of PIT tagged adult Chinook Salmon and steelhead at detection sites located in the Lower Granite Dam fish ladder.
+The last source of data were observations of PIT tagged adult sp/sum Chinook Salmon and steelhead at detection sites located in the Lower Granite Dam fish ladder.
 These observations provide daily estimates of 1) a trapping rate, 2) a nighttime passage rate, and 3) a re-ascension rate.
-Detections used in the model include all fish that were previously PIT tagged as juveniles or adults prior to arriving at Lower Granite Dam (i.e., does not include newly tagged adults at the dam) and detected at adult detection sites in the dam passage system.
-PIT tag data was provided through DART and the adult ladder PIT tag query; [http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail](http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail).
-
-A trap rate estimate was derived using Lincoln-Peterson mark-recapture methods [@Seber2002] and PIT tag observations of both Chinook Salmon and steelhead at Lower Granite Dam adult detection sites.
+Detections used in the model included all fish that were previously PIT tagged as juveniles or adults prior to arriving at Lower Granite Dam (i.e., does not include newly tagged adults at the dam) and detected at adult detection sites in the dam passage system.
+PIT tag data were provided through DART and the adult ladder PIT tag query; [http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail](http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail).
+
+A trap rate estimate was derived using Lincoln-Peterson mark-recapture methods [@Seber2002] and PIT tag observations of both sp/sum Chinook Salmon and steelhead at Lower Granite Dam adult detection sites.
 The "mark" group included all tags (both species) detected in the adult trap and the "capture" group included tags observed to cross the weir at the upstream end of the fish ladder as adults left the passage system.
@@
 
-We used PIT tag data to estimate nighttime passage and re-ascension rates, using the total number of tags passing the fish ladder for both and estimating each on a weekly basis.
+We used PIT tag observation data from the adult fish ladder to estimate both nighttime passage and re-ascension rates; each estimated on a weekly basis.
 The nighttime passage rate was based on the count of PIT tags that migrated through the fish ladder during non-window observation hours and the total number of tags passing the fish ladder.
@@
 
-We applied STADEM to empirical data from Lower Granite Dam for Chinook Salmon and steelhead returning to the Snake River during spawn years 2010 to 2019.
+We applied STADEM to empirical data from Lower Granite Dam for sp/sum Chinook Salmon and steelhead returning to the Snake River during spawn years 2010 to 2019.
 Window counts for both species were accessed from DART via functions within STADEM.
-For Chinook Salmon, a spawn year refers to adults that migrate past the dam prior to August 17 each year and spawn that late summer and fall.
+For sp/sum Chinook Salmon, a spawn year refers to adults that migrate past the dam prior to August 17 each year and spawn that late summer and fall.
 For steelhead, the spawn year is defined as steelhead that migrate past Lower Granite Dam starting July 1 the previous year and prior to June 30 of the given year (e.g., spawn year 2017 steelhead migrate past Lower Granite between July 1, 2016 and June 30, 2017).
-Data from the adult trap was made available by Idaho Department of Fish and Game, and adult PIT tag detection data within the fish passage ladder at Lower Granite Dam was accessed from the PTAGIS regional database ([https://www.ptagis.org/](https://www.ptagis.org/)).
+Data from the adult trap were made available by Idaho Department of Fish and Game, and adult PIT tag detection data within the fish passage ladder at Lower Granite Dam were accessed from the PTAGIS regional database ([https://www.ptagis.org/](https://www.ptagis.org/)).
 
@@
 
-We applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead for spawn years 2010–2019.
+We applied STADEM to data from Lower Granite Dam for sp/sum Chinook Salmon and steelhead for spawn years 2010–2019.
 Estimates of total escapement, as well as estimates of wild, hatchery clipped, and hatchery no-clip estimates are presented in Table \@ref(tab:lgr-tab).
@@
 
-Estimates of weekly nighttime passage and re-ascenstion rates did not match in most cases (Figure \@ref(fig:night-reasc-diff-fig)).
-In particular, there are several weeks when the window counts are quite large and the rates differ by as much as 10%.
+Estimates of weekly nighttime passage and re-ascension rates were not equal in most cases (Figure \@ref(fig:night-reasc-diff-fig)).
+In particular, there were several weeks when the window counts were quite large, but the nighttime passage and re-ascension rates differed by as much as 10%, in which case assuming them equal would lead to a biased estimate of escapement.
 When nighttime passage is larger than re-ascension, the window counts will be biased low, and vice versa.
@@
 In doing so, we demonstrated that at Lower Granite Dam, nighttime passage and re-ascension rates do not always offset each other, and assuming they do will lead to biased escapement estimates in some years.
-With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall Chinook Salmon, Pacific lamprey *Lampetra tridentata*), or elsewhere, provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run and tag observation or detection infrastructure (e.g., a PIT tag detection array or similar).
+With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall-run Chinook Salmon, Pacific lamprey *Lampetra tridentata*), or elsewhere, provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run and tag observation or detection infrastructure (e.g., a PIT tag detection array or similar).
 Our state-space model combined multiple imperfect sources of data to reduce bias in adult escapement estimates and provided quantitative estimates of uncertainty.
@@
 
-Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate escapement of spring-summer run Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary or population specific spawning areas [@IPTDSW2020; @Kinzer2020].
+Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate escapement of sp/sum Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary or population specific spawning areas [@IPTDSW2020; @Kinzer2020].
 Estimates of species and origin-specific escapement at Lower Granite Dam, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates [e.g., @Camacho2018; @Schrader2013] that are important for fisheries management and productivity monitoring of wild populations.
@@
 Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations.
-With sex and age data collected at the adult fish trap [@Hargrove2019], this approach provides necessary information to evaluate productivity and population viability for select Snake River Chinook Salmon and steelhead groups [@IPTDSW2020].
+With sex and age data collected at the adult fish trap [@Hargrove2019], this approach provides necessary information to evaluate productivity and population viability for select Snake River sp/sum Chinook Salmon and steelhead groups [@IPTDSW2020].
 
@@
         linesep = "",
-        caption = "Window counts, and estimates of total, wild, hatchery and hatchery no-clip escapement, with coefficients of variation in parenthesis, for Chinook Salmon and steelhead from spawn years 2010 to 2019.") %>%
+        caption = "Window counts, and estimates of total, wild, hatchery and hatchery no-clip escapement, with coefficients of variation in parenthesis, for spring/summer-run Chinook Salmon and steelhead from spawn years 2010 to 2019.") %>%
   kable_styling(latex_options = c('scale_down',
@@
 
-```{r jags-fit-fig, fig.cap = "Time-series plot showing estimates of total escapement for Chinook Salmon in 2014, including window counts, trap estimates and STADEM estimates.
+```{r jags-fit-fig, fig.cap = "Time-series plot showing estimates of total escapement for spring/summer-run Chinook Salmon in 2014, including window counts, trap estimates and STADEM estimates.
 The dark gray ribbon represents the 95% credible intervals for STADEM estimates, while the light gray ribbon represents the 95% confidence intervals for the trap estimates."}

### 2021-03-13 analysis/paper/paper.Rmd
made a few changes to some figures and tables.

@@
          cv = se / est,
-         bias = est - Truth,
+         # bias = est - Truth,
+         bias = Truth - est,
          rel_bias = bias / Truth,
@@
                                 "N-R",
-                                "N-R trap down",
-                                "Trap down"),
+                                "Trap down",
+                                "N-R trap down"),
                               each = 3)),
@@
             Sample_Mean = mean(est),
+            `Mean bias` = mean(bias),
             `Relative bias` = mean(rel_bias),
             SE = sqrt(var(est)),
+            `Mean SE` = mean(se),
             `Mean CV` = mean(cv),
@@
          `Relative bias`,
+         `Mean bias`,
+         `Mean SE`,
          `Mean CV`,
@@
 sum_tab %>%
-  kable(digits = c(0, 3, 3, 0, 3),
+  kable(digits = c(0, 3, 1, 0, 3, 0, 3),
         booktabs = T,
@@
               values_from = "prnt_val") %>%
+  left_join(lgr_est %>%
+              group_by(Species, Year) %>%
+              summarise(win_bias = mean[variable == "win_cnt"] - mean[variable == "X.tot.new.all"],
+                        win_rel_bias = win_bias / mean[variable == "win_cnt"])) %>%
+  mutate(across(win_rel_bias,
+                ~ paste(round(. * 100, 1), "%"))) %>%
   select(Species, 
@@
          `Window counts` = win_cnt,
+         `Window bias` = win_rel_bias,
          Total = X.tot.new.all,
@@
         linesep = "",
-        caption = "Window counts, and estimates of total, wild, hatchery and hatchery no-clip escapement, with coefficients of variation in parenthesis, for spring/summer-run Chinook Salmon and steelhead from spawn years 2010 to 2019.") %>%
+        caption = "Window counts, the relative bias of window counts compared to estimates of total escapement, and estimates of total, wild, hatchery and hatchery no-clip escapement, with coefficients of variation in parenthesis, for spring/summer-run Chinook Salmon and steelhead from spawn years 2010 to 2019.") %>%
   kable_styling(latex_options = c('scale_down',
@@
 
-```{r stadem-examp-fig, fig.height = 5, fig.cap = "Schematic showing the STADEM model framework.
+```{r dag-fig, fig.height = 5, fig.cap = "Directed acyclic graph showing the STADEM model framework."}
+include_graphics("../figures/ModelSchematic.png")
+```
+
+```{r stadem-examp-fig, eval = F, fig.height = 5, fig.cap = "Schematic showing the STADEM model framework.
 (A) shows the posterior of the estimate of fish crossing the dam while the window is open (dashed line shows observed window counts).
@@
 
-```{r rel-bias-lgd, fig.cap = "Boxplots of the relative bias of STADEM estimates for wild escapement across various scenarios (See Table 1)."}
+```{r rel-bias-lgd-old, eval = F, fig.cap = "Boxplots of the relative bias of STADEM estimates for wild escapement across various scenarios (See Table 1)."}
 # hdi_box = function(x) {
@@
 
+```{r rel-bias-lgd, fig.cap = "Boxplots of the relative bias of window counts and STADEM estimates for total unique fish and STADEM estimates of unique wild fish across various scenarios (See Table 1)."}
+rel_bias_p = res_df %>%
+  mutate(across(sim,
+                as.integer)) %>%
+  filter(grepl('Unique', Variable)) %>%
+  group_by(Scenario,
+           sim) %>%
+  summarise(Truth = sum(Truth),
+            win_cnts = sum(SCOBI_est),
+            model_est = sum(ISEMP_est),
+            .groups = "drop") %>%
+  tibble::add_column(Variable = "Total Unique Fish",
+                     .after = "sim") %>%
+  bind_rows(res_df %>%
+              mutate(across(sim,
+                            as.integer)) %>%
+              filter(Variable == "Unique.Wild.Fish") %>%
+              mutate(Variable = recode(Variable,
+                                       "Unique.Wild.Fish" = "Unique Wild Fish")) %>%
+              select(Scenario, sim,
+                     Variable,
+                     Truth,
+                     model_est = ISEMP_est)) %>%
+  mutate(model_bias = model_est - Truth,
+         win_bias = win_cnts - Truth) %>%
+  mutate(across(ends_with("bias"),
+                ~ . / Truth)) %>%
+  select(Scenario, sim,
+         Variable,
+         ends_with("bias")) %>%
+  pivot_longer(cols = ends_with("bias"),
+               names_to = "source",
+               values_to = "bias") %>%
+  mutate(source = recode(source,
+                         "model_bias" = "Model",
+                         "win_bias" = "Window Count")) %>%
+  ggplot(aes(x = bias,
+             y = fct_rev(Scenario),
+             fill = source)) +
+  scale_fill_grey(start = 0.5,
+                  end = 0.9) +
+  geom_boxplot() +
+  geom_vline(xintercept = 0,
+             linetype = 2) +
+  theme(axis.text = element_text(size = 10)) +
+  facet_wrap(~ Variable) +
+  labs(x = "Relative Bias",
+       y = "Scenario",
+       fill = "Source") +
+  coord_fixed(0.03)
+
+rel_bias_p
+
+# save black and white version
+ggsave(paste0("../figures/rel-bias-lgd-bw.", my_ext),
+       rel_bias_p,
+       height = 5,
+       width = 3.5,
+       dpi = my_dpi)
+
+```
+
 \newpage
@@
                           .f = function(x, y) {
-                            load(paste0('../data/derived_data/STADEM_results/LGR_STADEM_', Species, '_', Year, '.rda'))
-                            plot_df = stadem_mod$summary[grep('^X.all', rownames(stadem_mod$summary)),] %>%
+                            load(paste0('../data/derived_data/STADEM_results/LGR_STADEM_', x, '_', y, '.rda'))
+                            # plot_df = stadem_mod$summary[grep('^X.all', rownames(stadem_mod$summary)),] %>%
+                            plot_df = stadem_mod$summary[grep('^X.new', rownames(stadem_mod$summary)),] %>%  
                               as_tibble(rownames = 'var') %>%
@@
                                      param = str_extract_all(var, "[:alpha:]+", simplify = T)[,3],
-                                     param = ifelse(param == '', 'all', param)) %>%
+                                     param = recode(param, "tot" = "all")) %>%
+                                     # param = ifelse(param == '', 'all', param)) %>%
                               select(var, param, week, everything()) %>%

### 2021-03-16 inst/scripts/plot_time_series.R
updated code to make all time series plots to depict model estimates of unique escapement

@@ -20,11 +20,14 @@ plot_df = crossing(Species = c("Chinook",
                           Year,
                           .f = function(x, y) {
                             load(paste0('analysis/data/derived_data/STADEM_results/LGR_STADEM_', Species, '_', Year, '.rda'))
-                            plot_df = stadem_mod$summary[grep('^X.all', rownames(stadem_mod$summary)),] %>%
+                            # plot_df = stadem_mod$summary[grep('^X.all', rownames(stadem_mod$summary)),] %>%
+                            plot_df = stadem_mod$summary[grep('^X.new', rownames(stadem_mod$summary)),] %>%
                               as_tibble(rownames = 'var') %>%
                               mutate(week = as.integer(str_extract(var, "[0-9]+")),
                                      param = str_extract_all(var, "[:alpha:]+", simplify = T)[,3],
-                                     param = ifelse(param == '', 'all', param)) %>%
+                                     param = ifelse(param == '', 'all', param),
+                                     param = recode(param,
+                                                    "tot" = "all")) %>%
                               select(var, param, week, everything()) %>%
                               left_join(stadem_list$weeklyData %>%
                                           rename(week = week_num) %>%
@@ -111,7 +114,7 @@ for(spp in unique(plot_df$Species)) {
                scales = "free",
                ncol = 2) +
     labs(x = 'Date',
-         y = 'Total Escapement (x 10,000)')
+         y = 'Escapement (x 10,000)')
 
   # black and white version
   ts_bw = ts_col +

### 2021-03-16 analysis/paper/paper.Rmd
made some preliminary revisions including beefing up the methods and others

@@
          cv = se / est,
-         # bias = est - Truth,
-         bias = Truth - est,
+         bias = est - Truth,
          rel_bias = bias / Truth,
@@
 
-A trap rate estimate was derived using Lincoln-Peterson mark-recapture methods [@Seber2002] and PIT tag observations of both sp/sum Chinook Salmon and steelhead at Lower Granite Dam adult detection sites.
+<!-- A trap rate estimate was derived using Lincoln-Peterson mark-recapture methods [@Seber2002] and PIT tag observations of both sp/sum Chinook Salmon and steelhead at Lower Granite Dam adult detection sites.
 The "mark" group included all tags (both species) detected in the adult trap and the "capture" group included tags observed to cross the weir at the upstream end of the fish ladder as adults left the passage system.
@@
 Therefore, we use the mark-recapture approach to estimate a "true" trapping rate.
-
-We used PIT tag observation data from the adult fish ladder to estimate both nighttime passage and re-ascension rates; each estimated on a weekly basis.
+-->
+
+<!-- We used PIT tag observation data from the adult fish ladder to estimate both nighttime passage and re-ascension rates; each estimated on a weekly basis.
 The nighttime passage rate was based on the count of PIT tags that migrated through the fish ladder during non-window observation hours and the total number of tags passing the fish ladder.
@@
 Therefore, we combine wild and hatchery PIT tagged fish observations to estimate common nighttime passage and re-ascension rates to increase sample sizes.
+-->
 
@@
 
-We estimated the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for estimated nighttime passage and fallback/re-ascension rates using a state-space modeling approach [@Royle2008].
-Our model is implemented in the STADEM package for the R statistical software [@R-Core-Team2020].
-We assumed that the window counts and the estimates from the trap (fish in the trap divided by trap rate that week) were generated by processes with observation error.
-In the case of the trap, for example, we assumed there was sampling variation and uncertainty around our estimates of the true unknown trap rate.
-STADEM adjusted the window counts for the nighttime passage rate, and both window and trap estimates for the re-ascension rate to estimate the number of unique fish that crossed the dam.
-Finally, adult sampling data from the trap (wild, hatchery, hatchery no-clip) were used to partition the total escapement estimate by origin (Figure \@ref(fig:stadem-examp-fig)).
-Additional model details can be found in [Appendix A](#append1).
+### Day-time Passage and Re-ascension Rates
+
+There are two processes that must be accounted for to correct any bias in the window counts: first, the proportion of fish that cross the dam while the window is closed for counting (nighttime passage rate), and second, the proportion of fish that are crossing the dam multiple times (re-ascension rate) and therefore potentially double-counted.
+Both rates can be estimated from previously PIT tagged fish that are crossing the dam each week.
+
+The proportion of fish passing the window during non-operational hours, nighttime passage rate, is just the complement of the rate of fish passing during the day when the window is operating.
+The daytime passage rate for week $t$, $\theta_t$, is modeled as a random walk process and estimated from a binomial distribution based on the number of PIT tags observed to cross the dam during operational hours, $y^{day}_t$, and the total number of PIT tags observed to cross the dam at any point that week, $N_t$ [@Shumway2010].
+
+$$
+\begin{aligned}
+  y^{day}_t &\sim \text{Bin} \left(\theta_t, N_t \right) \\
+  \text{logit} (\theta_{t}) &= \text{logit} \left( \theta_{t-1} \right) + g_t \\
+  g_t &\sim \mathcal{N}(0, \sigma^2_\theta) \\
+\end{aligned}
+$$
+
+The number of total fish crossing Lower Granite differs from the number of unique fish crossing Lower Granite because some fish fallback and re-ascend the dam.
+These fish are potentially double-counted at the window, and have the potential to be caught in the fish trap more than once.
+The number of tags known to be re-ascending the dam each week, $y^{reasc}_t$, is modeled as a binomial process based on the estimated re-ascension rate, $\eta_t$, and the total number of tags crossing the dam that week, $N_t$.
+The logit of the re-ascension rate is modeled as a random walk process similar to day-time passage [@Shumway2010].
+
+$$
+\begin{aligned}
+  y^{reasc}_t &\sim \text{Bin} \left(\eta_t, N_t \right) \\
+  \text{logit} \left( \eta_{t} \right) &= \text{logit} \left(\eta_{t-1} \right) + f_t \\
+  f_t &\sim \mathcal{N}(0, \sigma^2_\eta)
+\end{aligned}
+$$
+
+We looked for differences in nighttime passage and re-ascension rates estimated using wild fish only, versus combining hatchery and wild fish together, and found no difference.
+Therefore, we combine wild and hatchery PIT tagged fish observations to estimate common nighttime passage and re-ascension rates to increase sample sizes.
+
+### Total and Weekly Escapement
+
+Escapement at Lower Granite Dam (LGD) is estimated by combining two independent observations (trap catches and window counts) of the true number of fish crossing the dam in a state-space model [@Royle2008] with a weekly time-step.
+Both are assumed to be corrupted observations of the true unknown number of fish crossing LGD each week.
+We chose to use a weekly time-step to ensure sufficient sample sizes of PIT tag detections within each time strata, but this choice could be easily modified.
+The log of the true number of fish crossing ($X_t$), is modeled as a random walk process [@Shumway2010].
+
+$$
+\begin{aligned}
+  \ln \left( X_{t} \right) &= \ln \left( X_{t-1} \right) + e_t \\
+  e_t &\sim \mathcal{N}(0, \sigma^2_X)
+\end{aligned}
+$$
+
+The number of fish caught in the trap, $Y^T_t$, for week $t$ is modeled as a binomial process based on the unknown true trap rate that week, $\nu_t$, and the unknown true number of fish crossing the dam that week, $X_t$.
+The set trapping rate (i.e., the recorded time that the trap is open to trap adults) does not always reflect the true trapping rate or proportion of fish that are actually captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error, among others.
+Therefore, we use a mark-recapture approach to estimate a "true" trapping rate.
+The estimate of the true weekly trap rate is derived based on previously PIT-tagged sp/sum Chinook Salmon and steelhead that are crossing LGD that week, using a Lincoln-Peterson mark-recapture model [@Seber2002].
+The fish, from both species, caught in the trap that week are considered the "mark" group ($m_t$), and all the previously PIT tagged fish who are detected at the upper end of the LGD fish ladder that week are considered the second "capture" group, $M_t$ (which includes recaptures of the the "marked" fish).
+The proportion of previously marked fish that are caught in the adult trap, $m/M$ can be modeled with a binomial distribution using the same trap rate parameter, $\nu_t$.
+Although the group of previously PIT tagged fish is not assumed to be representative of the overall run, the rate at which they are caught in the trap should be the same rate that the overall run experiences.
+The more tagged fish crossing the dam in a particular week, the more certain we can be of the true trap rate.
+
+$$
+\begin{aligned}
+  m_t &\sim \text{Bin} \left( \nu_t, M_t \right) \\
+  Y^T_t &\sim \text{Bin} \left( \nu_t, X_t \right) \\
+\end{aligned}
+$$
+
+The number of fish counted at the window, $Y^W_t$, is modeled as a (potentially) over-dispersed negative binomial process, with an expected value of $X_t^{day}$, the number of fish crossing the dam while the window is open.
+This is simply the total number of fish crossing that week, $X_t$, multiplied by the proportion of fish crossing while the window is open for counting, $\theta_t$.
+In the formula below, $p_t$ is the proportion of fish observed at the window and $r$ is the shape parameter.
+If $r$ is estimated to be small it provides evidence for over-dispersion, and as it grows very large, the negative binomial distribution behaves like a Poisson distribution.
+
+$$
+\begin{aligned}
+  X_t^{day} &= X_t * \theta_t \\
+  p_t &= \frac{r}{\left( r + X_t^{day} \right)} \\
+  Y^W_t &\sim \text{NegBin} \left( p_t, r \right) \\
+\end{aligned}
+$$
+
+Thus, the unknown true number of fish crossing LGD each week, $X_t$, is estimated from two different data sources: window counts and fish sampled in the trap.
+The window counts provide an estimate (with some potential observer error) of the fish crossing during daytime hours, while the fish in the trap, when expanded by the estimated true trap rate, provide an estimate of the total fish crossing that week.
+The use of both estimates allows for an estimation of observer error in the window counts which has previously been inestimable.
+For weeks when the trap is down, STADEM relies exclusively on the window counts and nighttime passage data, but there will be more uncertainty in the estimates.
+
+### Origin Proportions
+
+After estimating the total number of fish to have crossed Lower Granite each week, $X_t$, the total must be further refined into the number of wild fish, $X_{w,t}$, hatchery clipped fish, $X_{hc,t}$ and hatchery no-clip fish, $X_{hnc, t}$.
+This is done by estimating a weekly origin proportion vector, $\omega_t$ based on the random sample of fish caught in trap that week, $Y^T_t$.
+The observed number of wild, $Y^T_{w,t}$, hatchery clipped, $Y^T_{hc,t}$, and hatchery no-clip, $Y^T_{hnc,t}$, fish caught in the trap that week is assumed to come from a multinomial distribution with probability vector $\omega_t$.
+The log-odds ratio of the proportions in $\omega_t$, in relation to the proportion of clipped hatchery fish, $\omega_{hc,t}$ is modeled as a random walk, so it can change through time, while smoothing over large weekly flucuations in the proportions due to small sample size.
+This allows the proportions of wild, hatchery clipped and hatchery no-clip fish to shift throughout the season, based on the data available from the fish trap.
+
+$$
+\begin{aligned}
+  \left( Y^T_{w,t}, Y^T_{hc,t}, Y^T_{hnc,t} \right) &\sim \text{Multinom} \left( \omega_t, Y^T_t \right) \\
+  \omega_t &= \frac{\exp(\phi_t)}{\sum{\exp(\phi_t})} \\
+  \phi_{hc,t} &= 0 \\
+  \phi_{w,t} &= \ln \left( \frac{\omega_{w,t}}{\omega_{hc,t}} \right) \\
+  \phi_{hnc,t} &= \ln \left( \frac{\omega_{hnc,t}}{\omega_{hc,t}} \right) \\
+  \phi_{w,t} &= \phi_{w, t-1} + d_{w,t} \\
+  \phi_{hnc,t} &= \phi_{hnc, t-1} + d_{hnc,t} \\
+  d_{w,t} &\sim \mathcal{N}(0, \sigma^2_\omega) \\
+  d_{hnc,t} &\sim \mathcal{N}(0, \sigma^2_\omega) \\
+\end{aligned}
+$$
+
+Finally, the number of unique fish by origin crossing Lower Granite each week ($X_{w,t}$, $X_{hc,t}$, and $X_{hnc,t}$) is the product of the total fish crossing that week, $X_t$, multiplied by one minus the re-ascension rate, $(1 - \eta_t)$, and the origin proportion vector, $\omega_t$.
+
+$$
+\begin{aligned}
+  \begin{bmatrix} X_{w,t} \\ X_{hc,t} \\ X_{hnc,t} \end{bmatrix}
+  &= X_t * (1 - \eta_t) * \begin{bmatrix} \omega_{w,t} \\ \omega_{hc,t} \\ \omega_{hnc,t} \end{bmatrix}
+\end{aligned}
+$$
+
+### Model Fitting
+
+The model was fit using the JAGS program [@R-rjags], run with R software [@R-Core-Team2020].
 The STADEM package is available from the primary author at [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM), and requires the use of the JAGS software [@R-rjags] for Bayesian inference.
+Variance parameters $\sigma_X, \sigma_\eta, \sigma_\theta$, and $\sigma_\omega$, as well as the initial abundance, $X_1$, and the overdispersion parameter of the negative binomial, $r$, were given half-Cauchy priors with mean of 0 and scale of 100.
+The initial day-time passage and re-ascension rates, $\theta_1$ and $\eta_1$ were given Uniform(0,1) priors.
+Finally, $\phi_{w,1}$ and $\phi_{hnc,1}$ were given priors of Uniform(-3,3), in an effort to make $\omega_1$ as uniformative as possible.
 
@@
 
-We tested the STADEM model on a variety of simulated data sets.
+We tested the STADEM model on a variety of simulated data sets, constructed to mimic the dynamics at Lower Granite Dam.
 These simulated data sets contained a fixed number of unique adult fish of known origin crossing a dam, from a total of 25 fictional populations with differential run-timing (i.e., date of passage at Lower Granite Dam).
@@
 
-Our objective was to examine STADEM model estimates of origin-specific (wild, hatchery, hatchery no-clip) escapement from the combinations of two trap rate scenarios (constant and shut down for 3 weeks), two fallback/re-ascension and nighttime passage combinations, and three window count error rates; resulting in twelve different scenarios (Table \@ref(tab:lgr-sim-tab)).
+Our objective was to examine STADEM model estimates of origin-specific (wild, hatchery, hatchery no-clip) escapement from the combinations of two trap rate scenarios (constant and shut down for 3 weeks), two fallback/re-ascension and nighttime passage combinations (equal and offsetting, and unequal), and three window count error rates (none, low and high); resulting in twelve different scenarios (Table \@ref(tab:lgr-sim-tab)).
 The simulation parameters such as proportion of origin, run-timing, nighttime passage rates, fallback and re-ascension rates and trap rates were based on observed values at Lower Granite Dam between 2010-2015.
 We generated 99 simulations for each scenario, and ran STADEM on each one, resulting in a total of 1,188 model runs.
-Further details about simulation procedures can be found in [Appendix B](#append2).
+Further details about simulation procedures can be found in [Appendix A](#append1).
 
@@
 We estimated the precision by examining the average coefficient of variation (CV) of the estimates.
-We calculated the root mean squared error (RMSE) as the square root of the mean of the squared bias in the estimate.
+We calculated the root mean squared error (RMSE) as the square root of the mean of the squared bias in the estimate, which is equivalent to the square root of the variance of the estimator plus the square of the expected bias.
 Finally, we evaluated coverage probabilities by determining what proportion of the model results generated an estimated 95% credible interval that contained the simulated true value.
+We also compared the model estimates of total unique fish with the simulated truth and the window counts, as an evaluation of how well window counts captured this metric.
 
@@
 
-STADEM results were very similar across all scenarios (Figure \@ref(fig:rel-bias-lgd)).
+STADEM results showed similar patterns for estimates of unique total fish and unique wild fish, in that both were unbiased with similar variance in the estimator across all scenarios (Figure \@ref(fig:rel-bias-lgd)).
+For unique total fish, window counts were unbiased, except in scenarios when night passage and re-ascension rates were not equal (N-R).
+In those scenarios, they were biased by nearly 5%, the difference between night passage and re-ascension rates (Table \@ref(tab:lgr-sim-tab)).
+The variance of the window counts also grew as the simulated observer error grew (no error vs.
+Err_L vs.
+Err_H), whereas the variance in the STADEM estimates remained fairly constant, albeit often larger than the window count variance.
+
 Estimates of wild escapement were unbiased, with an average relative bias of 0.2–0.3%.
@@
 The coverage probabilities always exceeded 95% across all scenarios.
-The RMSE was near 500 for each scenario, representing an estimate within 2% of the true value, demonstrating the accuracy of STADEM (Table \@ref(tab:summ-tab)).
+The RMSE was near 500 for each scenario, representing an estimate within 2% of the true value, demonstrating the accuracy and precision of STADEM (Table \@ref(tab:summ-tab)).
 
@@
 Estimates of total unique fish escaping past Lower Granite Dam were sometimes higher and sometimes lower than the raw window counts, indicating that the relative strength of nighttime passage and re-ascension rates differed across years.
-CVs ranged from 2.5-7.1% for wild fish, 2.3-5.1% for clipped hatchery fish, 3.4-8.8% for hatchery no-clip fish and 2.2-4.7% for total unique fish past Lower Granite Dam.
+Window count bias ranged from -3% to 11% for sp/sum Chinook Salmon and -9.9% to 6.6% for steelhead.
+Coefficients of variation ranged from 2.5 to 7.1% for wild fish, 2.3 to 5.1% for clipped hatchery fish, 3.4 to 8.8% for hatchery no-clip fish and 2.2 to 4.7% for total unique fish past Lower Granite Dam.
 
@@
 Similar plots for all model runs [each species and year combination] are available on the manuscript GitHub page: [https://www.github.com/KevinSee/ManuscriptSTADEM]([https://www.github.com/KevinSee/ManuscriptSTADEM])).
-STADEM point estimates were often between estimates based on window counts and those based on the number of fish caught in the adult trap.
-However, for weeks when very few fish were caught in the trap or there was more uncertainty about the trap rate, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \@ref(fig:jags-fit-fig).
-That year also shows the utility of STADEM in dealing with missing data, as the trap was shut down for several weeks in July and August.
-The model's uncertainty is always smaller than the uncertainty from the trap estimates alone, whereas the window counts alone provide no estimate of uncertainty.
 
@@
 In doing so, we demonstrated that at Lower Granite Dam, nighttime passage and re-ascension rates do not always offset each other, and assuming they do will lead to biased escapement estimates in some years.
-With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall-run Chinook Salmon, Pacific lamprey *Lampetra tridentata*), or elsewhere, provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run and tag observation or detection infrastructure (e.g., a PIT tag detection array or similar).
 Our state-space model combined multiple imperfect sources of data to reduce bias in adult escapement estimates and provided quantitative estimates of uncertainty.
 Accurate population or group abundance estimates and uncertainty accounting for observation and process error can be particularly important when estimates are used for management and conservation decisions such as population viability analyses.
+
+STADEM estimates were often slightly lower than those based on the number of fish caught in the adult trap because the trap estimates do not account for fallback and re-ascension.
+STADEM estimates were higher than the window counts when night passage rates exceeded re-ascension rates, and lower when the re-ascension rate was higher than the night passage rate.
+For weeks when very few fish were caught in the trap or there was more uncertainty about the trap rate, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \@ref(fig:jags-fit-fig).
+That year also shows the utility of STADEM in dealing with missing data, as the trap was shut down for several weeks in July and August.
+The model's uncertainty is always smaller than the uncertainty from the trap estimates alone, whereas the window counts alone provide no estimate of uncertainty.
+
+STADEM is an improvement on window counts alone for several reasons.
+The first is that by accounting for night passage and re-ascension rates that are sometimes unequal, it corrects the bias in window counts.
+That bias ranged from -9.9% to 11% across all species and year combinations we fit STADEM to (Table \@ref(tab:lgr-tab)).
+Whether that amount of potential bias is of concern to management agencies is a decision for them, but we have presented a tool to correct that bias.
+The other justification for STADEM over window counts is that it provides an estimate of observer error for the window counts, leading to estimates of precision for unique fish which are unobtainable with window counts alone.
+At the same time, STADEM's precision is still quite good, with CV's near 2-3%.
+@Hatch1994
 
@@
 Certainly, estimating an unbiased total return to the entire Columbia River basin (i.e., Bonneville Dam) and Upper Columbia River with uncertainty would benefit managers and decision making.
+
+With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall-run Chinook Salmon, Pacific lamprey *Lampetra tridentata*), or elsewhere, provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run and tag observation or detection infrastructure (e.g., a PIT tag detection array or similar).
 
@@
         linesep = "",
-        caption = "Summary of simulation scenarios including varying adult trapping, fallback and re-ascension, nighttime passage, and window count error rates used to evaluate the performance of STADEM.") %>%
+        caption = "Summary of different simulation scenarios including varying adult trapping, fallback and re-ascension, nighttime passage, and window count error rates used to evaluate the performance of STADEM.") %>%
   kable_styling(latex_options = c('scale_down',
@@
         linesep = "",
-        caption = "Summary statistics, including relative bias, mean coefficient of variation (CV), root mean squared error (RMSE) and 95\\% credible interval coverage for results from each of the twelve simulation scenarios.") %>%
+        caption = "Summary statistics of unique wild fish estimates, including relative bias, mean coefficient of variation (CV), root mean squared error (RMSE) and 95\\% credible interval coverage for results from each of the twelve simulation scenarios.") %>%
   kable_styling(latex_options = c('scale_down',
@@
 
-```{r jags-fit-fig, fig.cap = "Time-series plot showing estimates of total escapement for spring/summer-run Chinook Salmon in 2014, including window counts, trap estimates and STADEM estimates.
+```{r jags-fit-fig, fig.cap = "Time-series plot showing estimates of escapement for spring/summer-run Chinook Salmon in 2014, including window counts, trap estimates and STADEM estimates of unique fish.
 The dark gray ribbon represents the 95% credible intervals for STADEM estimates, while the light gray ribbon represents the 95% confidence intervals for the trap estimates."}
@@
                                      param = str_extract_all(var, "[:alpha:]+", simplify = T)[,3],
-                                     param = recode(param, "tot" = "all")) %>%
-                                     # param = ifelse(param == '', 'all', param)) %>%
+                                     param = ifelse(param == '', 'all', param),
+                                     param = recode(param, 
+                                                    "tot" = "all")) %>%
                               select(var, param, week, everything()) %>%
@@
   labs(x = 'Date',
-       y = 'Total Escapement (x 10,000)')
+       y = 'Escapement (x 10,000)')
 
@@
 
-```{r child = 'appendices/Appendix_STADEM_descrip.Rmd'}
-```
-
-\newpage
-
 ```{r child = 'appendices/Appendix_simulation_details.Rmd'}

### 2021-03-17 analysis/paper/paper.Rmd
added a little code to statistically test whether night-passage and re-ascension rates are different each week.
Made a few other edits.

@@
 
+```{r test-rates-equality}
+rate_test = rate_comp %>%
+  mutate(night_tags = tot_tags - day_tags) %>%
+  filter(tot_tags > 0,
+         reascent_tags > 0,
+         night_tags > 0,
+         reascent_tags < tot_tags,
+         night_tags < tot_tags) %>%
+  tidyr::nest(tag_data = -c(Species:win_cnt)) %>%
+  mutate(tot_tags = map_dbl(tag_data,
+                            "tot_tags"),
+         prop_test = map(tag_data,
+                         .f = function(df) {
+                           prop.test(x = c(df$night_tags,
+                                           df$reascent_tags),
+                                     n = rep(df$tot_tags, 2))
+                         }),
+         p_value = map_dbl(prop_test,
+                           "p.value"),
+         sig = if_else(p_value < 0.05, T, F))
+
+# rate_test %>%
+#   janitor::tabyl(sig) %>%
+#   janitor::adorn_totals() %>%
+#   janitor::adorn_pct_formatting()
+
+```
+
 # Introduction
@@
 One additional biological process, the fallback and re-ascension of adult migrants at the dam, should also be accounted for to estimate the true escapement at Lower Granite Dam.
-Adult fish that migrate through the ladder and past the dam may fallback over the dam (e.g., over a spillway, through navigation locks) and later may re-ascend the fish ladder again [@Boogs2004] and be double-counted at the window (re-ascension).
+Adult fish that migrate through the ladder and past the dam may fallback over the dam (e.g., over a spillway, through navigation locks) and later may re-ascend the fish ladder again [@Boggs2004] and be double-counted at the window (re-ascension).
 Whereas unaccounted nighttime passage can lead to an underestimate in escapement, both fallback without and with re-ascension instead potentially lead to an overestimate of escapement [@Dauble2000].
@@
 
-<!-- A trap rate estimate was derived using Lincoln-Peterson mark-recapture methods [@Seber2002] and PIT tag observations of both sp/sum Chinook Salmon and steelhead at Lower Granite Dam adult detection sites.
-The "mark" group included all tags (both species) detected in the adult trap and the "capture" group included tags observed to cross the weir at the upstream end of the fish ladder as adults left the passage system.
-The proportion of tags detected leaving the passage system that were also caught in the trap each week was assumed to reflect the trap rate experienced by target species as they migrated through the ladder.
-The sample size of previously tagged fish detected at Lower Granite Dam influences the uncertainty in that trap rate, which is also informed by the number of adults caught in the trap and the window counts.
-The set trapping rate (i.e., the recorded time that the trap is open to trap adults) does not always reflect the true trapping rate or proportion of fish that are actually captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error, among others.
-Therefore, we use the mark-recapture approach to estimate a "true" trapping rate.
--->
-
-<!-- We used PIT tag observation data from the adult fish ladder to estimate both nighttime passage and re-ascension rates; each estimated on a weekly basis.
-The nighttime passage rate was based on the count of PIT tags that migrated through the fish ladder during non-window observation hours and the total number of tags passing the fish ladder.
-The re-ascension rate incorporated the count of tags observed passing the upstream most detection sites in the adult fish ladder (i.e., passing the dam) and later detected re-entering the downstream end of the fish ladder at a later time and the total number of tags leaving the fish ladder.
-Previously, we looked for differences in nighttime passage and re-ascension rates estimated using wild fish only, versus combining hatchery and wild fish together, and found no difference.
-Therefore, we combine wild and hatchery PIT tagged fish observations to estimate common nighttime passage and re-ascension rates to increase sample sizes.
--->
-
 ## Model Framework
@@
 
-Estimates of weekly nighttime passage and re-ascension rates were not equal in most cases (Figure \@ref(fig:night-reasc-diff-fig)).
+Estimates of weekly nighttime passage and re-ascension rates were significantly different (*p* < 0.05) in nearly 46% of the weeks with sufficient PIT tag data to test.
+These differences were also apparent visually (Figure \@ref(fig:night-reasc-diff-fig)).
 In particular, there were several weeks when the window counts were quite large, but the nighttime passage and re-ascension rates differed by as much as 10%, in which case assuming them equal would lead to a biased estimate of escapement.
@@
 STADEM estimates were often slightly lower than those based on the number of fish caught in the adult trap because the trap estimates do not account for fallback and re-ascension.
-STADEM estimates were higher than the window counts when night passage rates exceeded re-ascension rates, and lower when the re-ascension rate was higher than the night passage rate.
-For weeks when very few fish were caught in the trap or there was more uncertainty about the trap rate, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \@ref(fig:jags-fit-fig).
-That year also shows the utility of STADEM in dealing with missing data, as the trap was shut down for several weeks in July and August.
+However, in comparison to window counts, STADEM estimates were higher when night passage rates exceeded re-ascension rates, and lower when the re-ascension rate was higher than the night passage rate.
+***For weeks when very few fish were caught in the trap or there was more uncertainty about the trap rate, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \@ref(fig:jags-fit-fig).*** That year also shows the utility of STADEM in dealing with missing data, as the trap was shut down for several weeks in July and August.
 The model's uncertainty is always smaller than the uncertainty from the trap estimates alone, whereas the window counts alone provide no estimate of uncertainty.
@@
 STADEM is an improvement on window counts alone for several reasons.
-The first is that by accounting for night passage and re-ascension rates that are sometimes unequal, it corrects the bias in window counts.
+The first is that by accounting for night passage and re-ascension rates that are sometimes unequal, it corrects any bias in window counts.
 That bias ranged from -9.9% to 11% across all species and year combinations we fit STADEM to (Table \@ref(tab:lgr-tab)).
@@
 The other justification for STADEM over window counts is that it provides an estimate of observer error for the window counts, leading to estimates of precision for unique fish which are unobtainable with window counts alone.
-At the same time, STADEM's precision is still quite good, with CV's near 2-3%.
-@Hatch1994
-
-Combining data from the adult fish trap with live and video window counts provides several benefits.
-First, it allows us to model observer error in the window counts, which is typically unknown.
-If estimates rely on window counts alone, quantifying observer error is impossible.
-Capturing and accounting for known sources of error is prudent to minimize management decision risk.
-Second, by incorporating both sources of information in a state-space framework, STADEM incorporates missing data at either the observation window or adult trap seamlessly.
+Capturing and accounting for known sources of error is prudent to minimize management decision risk [@Harwood2003; @Ascough-Ii2008].
+At the same time, STADEM's precision is still quite good, with CV's near 2-3%, well below the goal of 15% set forth by NOAA [@Crawford2011].
+
+By incorporating data from the adult fish trap with live and video window counts in a state-space framework, STADEM incorporates missing data at either the observation window or adult trap seamlessly.
 At Lower Granite Dam, the adult trap has been closed for brief or extended periods of time (i.e., days, weeks) intermittently in five of the last ten years, often during peak run times [@USACE2010; @USACE2011; @USACE2012; @USACE2013; @Ogden2016a; @USACE2015; @USACE2016; @USACE2017; @USACE2018; @USACE2019].
@@
 
-Although STADEM was developed with salmonid escapement at Lower Granite Dam in mind, it could be applied to any migratory fish species at locations with similar monitoring infrastructure.
+With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall-run Chinook Salmon, Pacific lamprey *Lampetra tridentata*), or elsewhere, provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run and tag observation or detection infrastructure (e.g., a PIT tag detection array or similar).
 Justification and infrastructure exist for applying a modified STADEM framework for fish passing Bonneville Dam, the lowest dam on the Columbia River, or Priest Rapids Dam in the upper Columbia River.
@@
 Certainly, estimating an unbiased total return to the entire Columbia River basin (i.e., Bonneville Dam) and Upper Columbia River with uncertainty would benefit managers and decision making.
-
-With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall-run Chinook Salmon, Pacific lamprey *Lampetra tridentata*), or elsewhere, provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run and tag observation or detection infrastructure (e.g., a PIT tag detection array or similar).
 

### 2021-03-18 analysis/paper/paper.Rmd
responded to many reviewer comments, both in the review comment word doc, and within the manuscript.

@@
          night_tags > 0,
+         # reascent_tags >= 10,
+         # night_tags >= 10,
+         # tot_tags >= 20,
          reascent_tags < tot_tags,
@@
                                            df$reascent_tags),
-                                     n = rep(df$tot_tags, 2))
+                                     n = rep(df$tot_tags, 2),
+                                     correct = F)
                          }),
@@
 mykiss* in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric projects on the Snake and Columbia rivers) and tributary habitats [@Nehlsen1991; @McClure2003; @NWFSC2015].
-As a result, Snake River spring/summer-run Chinook Salmon (hereafter sp/sum Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937).
+As a result, most Snake River spring/summer-run Chinook Salmon (hereafter sp/sum Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937).
 Snake River sp/sum Chinook Salmon and steelhead have substantial cultural, recreational, commercial and subsistence value within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries.
 The aggregate escapement of Snake River sp/sum Chinook Salmon and steelhead populations, with the exception of Tucannon River (a tributary of the Snake River whose confluence is downstream of Lower Granite Dam), is monitored at Lower Granite Dam located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations.
-Many fisheries management and conservation actions are made based on estimates of escapement at Lower Granite Dam parsed by species and origin [@NMFS2019; @NWFSC2015; @NPCC2014].
+Many fisheries management and conservation actions are made based on estimates of escapement by species and origin at Lower Granite Dam [@NMFS2019; @NWFSC2015; @NPCC2014].
 Additionally, harvest openings and closures, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia rivers fisheries, are predicated on escapements at Lower Granite Dam.
@@
 Window counts were accessed through the Columbia Basin Research Data Access in Real Time (DART) website, [www.cbr.washington.edu/dart/query/adult_daily](www.cbr.washington.edu/dart/query/adult_daily), using their window count query.
-Counts were provided for each day the fish ladder was open to passage.
+Counts were provided for each day the fish ladder was open to passage, and had already been expanded by 1.2 (to account for the counting during 50 minutes of each hour).
 
@@
 The last source of data were observations of PIT tagged adult sp/sum Chinook Salmon and steelhead at detection sites located in the Lower Granite Dam fish ladder.
-These observations provide daily estimates of 1) a trapping rate, 2) a nighttime passage rate, and 3) a re-ascension rate.
+These observations provide daily estimates of 1) a trap rate for the adult fish trap, 2) a nighttime passage rate, and 3) a re-ascension rate.
 Detections used in the model included all fish that were previously PIT tagged as juveniles or adults prior to arriving at Lower Granite Dam (i.e., does not include newly tagged adults at the dam) and detected at adult detection sites in the dam passage system.
@@
 The proportion of fish passing the window during non-operational hours, nighttime passage rate, is just the complement of the rate of fish passing during the day when the window is operating.
-The daytime passage rate for week $t$, $\theta_t$, is modeled as a random walk process and estimated from a binomial distribution based on the number of PIT tags observed to cross the dam during operational hours, $y^{day}_t$, and the total number of PIT tags observed to cross the dam at any point that week, $N_t$ [@Shumway2010].
+The daytime passage rate for week $t$, $\theta_t$, is modeled as a random walk process (with deviations, $g_t$, drawn from a normal distribution with variance $\sigma^2_theta$) and estimated from a binomial distribution based on the number of PIT tags observed to cross the dam during operational hours, $y^{day}_t$, and the total number of PIT tags observed to cross the dam at any point that week, $N_t$ [@Shumway2010].
 
@@
 The number of tags known to be re-ascending the dam each week, $y^{reasc}_t$, is modeled as a binomial process based on the estimated re-ascension rate, $\eta_t$, and the total number of tags crossing the dam that week, $N_t$.
-The logit of the re-ascension rate is modeled as a random walk process similar to day-time passage [@Shumway2010].
+The logit of the re-ascension rate is modeled as a random walk process similar to day-time passage [@Shumway2010], with deviations, $f_t$, drawn from a normal distribution with variance $\sigma^2_\eta$.
 
@@
 We chose to use a weekly time-step to ensure sufficient sample sizes of PIT tag detections within each time strata, but this choice could be easily modified.
-The log of the true number of fish crossing ($X_t$), is modeled as a random walk process [@Shumway2010].
+The log of the true number of fish crossing during time-step $t$, ($X_t$), is modeled as a random walk process [@Shumway2010], with $e_t$ representing the process errors drawn from a normal distributrion with variance $\sigma^2_X$.
 
@@
 The number of fish caught in the trap, $Y^T_t$, for week $t$ is modeled as a binomial process based on the unknown true trap rate that week, $\nu_t$, and the unknown true number of fish crossing the dam that week, $X_t$.
+
+$$
+Y^T_t \sim \text{Bin} \left( \nu_t, X_t \right) \\
+$$
 The set trapping rate (i.e., the recorded time that the trap is open to trap adults) does not always reflect the true trapping rate or proportion of fish that are actually captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error, among others.
-Therefore, we use a mark-recapture approach to estimate a "true" trapping rate.
+Therefore, we use a mark-recapture approach to estimate a "true" trapping rate, $\nu_t$.
 The estimate of the true weekly trap rate is derived based on previously PIT-tagged sp/sum Chinook Salmon and steelhead that are crossing LGD that week, using a Lincoln-Peterson mark-recapture model [@Seber2002].
+
+$$
+m_t \sim \text{Bin} \left( \nu_t, M_t \right) \\
+$$
 The fish, from both species, caught in the trap that week are considered the "mark" group ($m_t$), and all the previously PIT tagged fish who are detected at the upper end of the LGD fish ladder that week are considered the second "capture" group, $M_t$ (which includes recaptures of the the "marked" fish).
@@
 The more tagged fish crossing the dam in a particular week, the more certain we can be of the true trap rate.
-
-$$
-\begin{aligned}
-  m_t &\sim \text{Bin} \left( \nu_t, M_t \right) \\
-  Y^T_t &\sim \text{Bin} \left( \nu_t, X_t \right) \\
-\end{aligned}
-$$
 
@@
 The observed number of wild, $Y^T_{w,t}$, hatchery clipped, $Y^T_{hc,t}$, and hatchery no-clip, $Y^T_{hnc,t}$, fish caught in the trap that week is assumed to come from a multinomial distribution with probability vector $\omega_t$.
-The log-odds ratio of the proportions in $\omega_t$, in relation to the proportion of clipped hatchery fish, $\omega_{hc,t}$ is modeled as a random walk, so it can change through time, while smoothing over large weekly flucuations in the proportions due to small sample size.
+The log-odds ratio of the proportions in $\omega_t$, in relation to the proportion of clipped hatchery fish, $\omega_{hc,t}$ is modeled as a random walk, so it can change through time, while smoothing over large weekly fluctuations in the proportions due to small sample size.
 This allows the proportions of wild, hatchery clipped and hatchery no-clip fish to shift throughout the season, based on the data available from the fish trap.
+The random walk deviations, $d_{w,t}$ and $d_{hnc,t}$, are drawn from the same normal distribution with variance $\sigma^2_\omega$.
 
@@
 
-Our objective was to examine STADEM model estimates of origin-specific (wild, hatchery, hatchery no-clip) escapement from the combinations of two trap rate scenarios (constant and shut down for 3 weeks), two fallback/re-ascension and nighttime passage combinations (equal and offsetting, and unequal), and three window count error rates (none, low and high); resulting in twelve different scenarios (Table \@ref(tab:lgr-sim-tab)).
+Our objective was to assess STADEM model estimates of origin-specific (wild, hatchery, hatchery no-clip) escapement for accuracy and precision, given a known simulated "truth", under different possible conditions.
+We developed those conditions from the combinations of two trap rate scenarios (constant and shut down for 3 weeks), two fallback/re-ascension and nighttime passage combinations (equal and offsetting, and unequal), and three window count error rates (none, low and high); resulting in twelve different scenarios (Table \@ref(tab:lgr-sim-tab)).
 The simulation parameters such as proportion of origin, run-timing, nighttime passage rates, fallback and re-ascension rates and trap rates were based on observed values at Lower Granite Dam between 2010-2015.
@@
 
-We grouped the simulation results by fish origin and evaluated them by several measures.
-Relative bias is the difference between the simulated "truth" and the STADEM estimate, divided by the simulated "truth".
+To assess the performance of the model, we evaluated the estimates of origin-specific escapement by several measures.
+Accuracy was captured by the relative bias, the difference between the simulated "truth" and the STADEM estimate, divided by the simulated "truth".
 We estimated the precision by examining the average coefficient of variation (CV) of the estimates.
 We calculated the root mean squared error (RMSE) as the square root of the mean of the squared bias in the estimate, which is equivalent to the square root of the variance of the estimator plus the square of the expected bias.
-Finally, we evaluated coverage probabilities by determining what proportion of the model results generated an estimated 95% credible interval that contained the simulated true value.
-We also compared the model estimates of total unique fish with the simulated truth and the window counts, as an evaluation of how well window counts captured this metric.
+Finally, we evaluated whether estimates of uncertainty were appropriate by calculating coverage probabilities by determining what proportion of the model results generated an estimated 95% credible interval that contained the simulated true value.
+To compare the STADEM results to strictly window counts, we compared the model estimates of total unique fish and the window counts with the simulated truth.
 
@@
 
-Weekly estimates of total escapement over Lower Granite Dam tracked the window counts and trap estimates (One example: Figure \@ref(fig:jags-fit-fig).
-Similar plots for all model runs [each species and year combination] are available on the manuscript GitHub page: [https://www.github.com/KevinSee/ManuscriptSTADEM]([https://www.github.com/KevinSee/ManuscriptSTADEM])).
-
-Estimates of weekly nighttime passage and re-ascension rates were significantly different (*p* < 0.05) in nearly 46% of the weeks with sufficient PIT tag data to test.
-These differences were also apparent visually (Figure \@ref(fig:night-reasc-diff-fig)).
+Weekly estimates of total escapement over Lower Granite Dam tracked the window counts and trap estimates (One example: Figure \@ref(fig:jags-fit-fig), with smaller credible intervals than the trap estimates alone.
+Similar plots for all model runs (each species and year combination) are available on the manuscript GitHub page: [https://www.github.com/KevinSee/ManuscriptSTADEM]([https://www.github.com/KevinSee/ManuscriptSTADEM])).
+
+Weekly nighttime passage and re-ascension proportions were significantly different (2-sample test of proportions, *p* < 0.05) in approximately 50% of the weeks with sufficient PIT tag data to test.
+The differences in the estimates were also apparent visually (Figure \@ref(fig:night-reasc-diff-fig)).
 In particular, there were several weeks when the window counts were quite large, but the nighttime passage and re-ascension rates differed by as much as 10%, in which case assuming them equal would lead to a biased estimate of escapement.
@@
 Our state-space model combined multiple imperfect sources of data to reduce bias in adult escapement estimates and provided quantitative estimates of uncertainty.
-Accurate population or group abundance estimates and uncertainty accounting for observation and process error can be particularly important when estimates are used for management and conservation decisions such as population viability analyses.
+Accurate population or group abundance estimates and uncertainty accounting for observation and process error can be particularly important when estimates are used for management and conservation decisions such as population viability analyses [@Paulsen2007].
 
@@
 
-With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall-run Chinook Salmon, Pacific lamprey *Lampetra tridentata*), or elsewhere, provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run and tag observation or detection infrastructure (e.g., a PIT tag detection array or similar).
+With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall-run Chinook Salmon, Pacific lamprey *Lampetra tridentata*), or elsewhere, provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run and tag detection infrastructure (e.g., a PIT tag detection array or similar).
 Justification and infrastructure exist for applying a modified STADEM framework for fish passing Bonneville Dam, the lowest dam on the Columbia River, or Priest Rapids Dam in the upper Columbia River.
@@
             RMSE = sqrt(mean(bias^2)),
-            Coverage = inCI/nsims) %>%
+            Coverage = inCI/nsims,
+            .groups = "drop") %>%
   mutate(Scenario = str_replace_all(Scenario,

### 2021-03-18 analysis/paper/paper.Rmd
made further revisions based on reviewer comments

@@
 However, using window counts as a census of sp/sum Chinook Salmon and steelhead passing Lower Granite Dam can be problematic as it fails to account for multiple sources of uncertainty and disregards known biological processes.
-In-person window counts, where observers look directly into the fish ladders to identify and count all passing fish, by species, occur for 50 minutes per hour, 16 hours a day from April through October (which corresponds with peak run timing for several species).
-Counts are then expanded to provide an estimate for the entire hour [@USACE2015].
-For the remainder of the year (November through March), fish passage is video-taped for 10 hours each day; fish counters then read the video tapes and submit daily fish counts.
 Typically, the observational error rates of live and video window counts are unknown, and sampling error rates are ignored.
@@
 We also test whether observed nighttime passage and fallback/re-ascension rates are typically equal.
-Our method for estimating species-specific escapement past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of fish previously tagged with passive integrated transponder (PIT) tags in the adult ladder to explicitly model nighttime passage, re-ascension, and observation error using a state-space approach [@Royle2008] which separates process variance (e.g., week to week variance in true escapment) from observation error variance (e.g., observation error at the window, or sampling variance at the trap).
+Our method for estimating species-specific escapement past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of fish previously tagged with passive integrated transponder (PIT) tags in the adult ladder to explicitly model nighttime passage, re-ascension, and observation error using a state-space approach [@Royle2008] which separates process variance (e.g., week to week variance in true escapement) from observation error variance (e.g., observation error at the window, or sampling variance at the trap).
 To meet desired management and conservation objectives, modeled escapement includes estimates of uncertainty and is parsed into weekly strata.
@@
 Our model is implemented in the **ST**ate-space **A**dult **D**am **E**scapement **M**odel (STADEM) package for the statistical software R [@R-Core-Team2020], and is available for download or installation from [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM).
-To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates.
+To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates and compared STADEM results to the simulated "truth".
 We then applied this model to sp/sum Chinook Salmon and steelhead returns at Lower Granite Dam for spawn years 2010 - 2019.
-The STADEM model combines multiple imperfect sources of data to reduce bias in escapement estimates and provides improved estimates of uncertainty.
 
@@
 Some subset of hatchery fish are either intentionally or unintentionally released without a clipped adipose fin, and these are referred to as unclipped hatchery fish, or hatchery no-clips (HNC).
-For adults with intact adipose fins (unclipped), which includes wild and hatchery no-clip individuals, either a portion or all of fish trapped (depending on the year) had scale and genetic tissue samples taken.
+For adults with intact adipose fins (unclipped), which includes wild and hatchery no-clip individuals, either a portion or all of the fish trapped (depending on the year) had scale and genetic tissue samples taken.
 Scale samples were used to estimate age [@Wright2015] and genetic tissue samples were used to determine sex [@Campbell2012] and estimate the location of origin of wild fish using genetic stock identification [e.g., @Hargrove2019].
@@
 The last source of data were observations of PIT tagged adult sp/sum Chinook Salmon and steelhead at detection sites located in the Lower Granite Dam fish ladder.
-These observations provide daily estimates of 1) a trap rate for the adult fish trap, 2) a nighttime passage rate, and 3) a re-ascension rate.
+These observations provide data related to 1) a trap rate for the adult fish trap, 2) a nighttime passage rate, and 3) a re-ascension rate.
 Detections used in the model included all fish that were previously PIT tagged as juveniles or adults prior to arriving at Lower Granite Dam (i.e., does not include newly tagged adults at the dam) and detected at adult detection sites in the dam passage system.
@@
 
+### Total and Weekly Escapement
+
+Escapement at Lower Granite Dam (LGD) is estimated by combining two independent observations (trap catches and window counts) of the true number of fish crossing the dam in a state-space model [@Royle2008] with a weekly time-step.
+Both are assumed to be corrupted observations of the true unknown number of fish crossing LGD each week.
+We chose to use a weekly time-step to ensure sufficient sample sizes of PIT tag detections within each time strata, but this choice could be easily modified.
+The log of the true number of fish crossing during time-step $t$, ($X_t$), is modeled as a random walk process [@Shumway2010], with $e_t$ representing the process errors drawn from a normal distribution with variance $\sigma^2_X$.
+
+$$
+\begin{aligned}
+  \ln \left( X_{t} \right) &= \ln \left( X_{t-1} \right) + e_t \\
+  e_t &\sim \mathcal{N}(0, \sigma^2_X)
+\end{aligned}
+$$
+
+The number of fish caught in the trap, $Y^T_t$, for week $t$ is modeled as a binomial process based on the unknown true trap rate that week, $\nu_t$, and the unknown true number of fish crossing the dam that week, $X_t$.
+
+$$
+Y^T_t \sim \text{Bin} \left( \nu_t, X_t \right) \\
+$$
+
+The set trapping rate (i.e., the recorded time that the trap is open to trap adults) does not always reflect the true trapping rate or proportion of fish that are actually captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error, among others.
+Therefore, we use a mark-recapture approach to estimate a "true" trapping rate, $\nu_t$.
+The estimate of the true weekly trap rate is derived based on previously PIT-tagged sp/sum Chinook Salmon and steelhead that are crossing LGD that week, using a Lincoln-Peterson mark-recapture model [@Seber2002].
+
+$$
+m_t \sim \text{Bin} \left( \nu_t, M_t \right) \\
+$$
+
+The fish, from both species, caught in the trap that week are considered the "mark" group ($m_t$), and all the previously PIT tagged fish who are detected at the upper end of the LGD fish ladder that week are considered the second "capture" group, $M_t$ (which includes recaptures of the the "marked" fish).
+The proportion of previously marked fish that are caught in the adult trap, $m/M$ can be modeled with a binomial distribution using the same trap rate parameter, $\nu_t$.
+Although the group of previously PIT tagged fish is not assumed to be representative of the overall run, the rate at which they are caught in the trap should be the same rate that the overall run experiences.
+The more tagged fish crossing the dam in a particular week, the more certain we can be of the true trap rate.
+
+The number of fish counted at the window, $Y^W_t$, is modeled as a (potentially) over-dispersed negative binomial process, with an expected value of $X_t^{day}$, the number of fish crossing the dam while the window is open.
+This is simply the total number of fish crossing that week, $X_t$, multiplied by the proportion of fish crossing while the window is open for counting, $\theta_t$.
+In the formula below, $p_t$ is the proportion of fish observed at the window and $r$ is the shape parameter.
+If $r$ is estimated to be small it provides evidence for over-dispersion, and as it grows very large, the negative binomial distribution behaves like a Poisson distribution.
+
+$$
+\begin{aligned}
+  X_t^{day} &= X_t * \theta_t \\
+  p_t &= \frac{r}{\left( r + X_t^{day} \right)} \\
+  Y^W_t &\sim \text{NegBin} \left( p_t, r \right) \\
+\end{aligned}
+$$
+
+Thus, the unknown true number of fish crossing LGD each week, $X_t$, is estimated from two different data sources: window counts and fish sampled in the trap.
+The window counts provide an estimate (with some potential observer error) of the fish crossing during daytime hours, while the fish in the trap, when expanded by the estimated true trap rate, provide an estimate of the total fish crossing that week.
+The use of both estimates allows for an estimation of observer error in the window counts which has previously been inestimable.
+For weeks when the trap is down, STADEM relies exclusively on the window counts and nighttime passage data, but there will be more uncertainty in the estimates.
+
 ### Day-time Passage and Re-ascension Rates
@@
 The proportion of fish passing the window during non-operational hours, nighttime passage rate, is just the complement of the rate of fish passing during the day when the window is operating.
-The daytime passage rate for week $t$, $\theta_t$, is modeled as a random walk process (with deviations, $g_t$, drawn from a normal distribution with variance $\sigma^2_theta$) and estimated from a binomial distribution based on the number of PIT tags observed to cross the dam during operational hours, $y^{day}_t$, and the total number of PIT tags observed to cross the dam at any point that week, $N_t$ [@Shumway2010].
+The daytime passage rate for week $t$, $\theta_t$, is modeled as a random walk process (with deviations, $g_t$, drawn from a normal distribution with variance $\sigma^2_{\theta}$) and estimated from a binomial distribution based on the number of PIT tags observed to cross the dam during operational hours, $y^{day}_t$, and the total number of PIT tags observed to cross the dam at any point that week, $N_t$ [@Shumway2010].
 
@@
 Therefore, we combine wild and hatchery PIT tagged fish observations to estimate common nighttime passage and re-ascension rates to increase sample sizes.
-
-### Total and Weekly Escapement
-
-Escapement at Lower Granite Dam (LGD) is estimated by combining two independent observations (trap catches and window counts) of the true number of fish crossing the dam in a state-space model [@Royle2008] with a weekly time-step.
-Both are assumed to be corrupted observations of the true unknown number of fish crossing LGD each week.
-We chose to use a weekly time-step to ensure sufficient sample sizes of PIT tag detections within each time strata, but this choice could be easily modified.
-The log of the true number of fish crossing during time-step $t$, ($X_t$), is modeled as a random walk process [@Shumway2010], with $e_t$ representing the process errors drawn from a normal distributrion with variance $\sigma^2_X$.
-
-$$
-\begin{aligned}
-  \ln \left( X_{t} \right) &= \ln \left( X_{t-1} \right) + e_t \\
-  e_t &\sim \mathcal{N}(0, \sigma^2_X)
-\end{aligned}
-$$
-
-The number of fish caught in the trap, $Y^T_t$, for week $t$ is modeled as a binomial process based on the unknown true trap rate that week, $\nu_t$, and the unknown true number of fish crossing the dam that week, $X_t$.
-
-$$
-Y^T_t \sim \text{Bin} \left( \nu_t, X_t \right) \\
-$$
-The set trapping rate (i.e., the recorded time that the trap is open to trap adults) does not always reflect the true trapping rate or proportion of fish that are actually captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error, among others.
-Therefore, we use a mark-recapture approach to estimate a "true" trapping rate, $\nu_t$.
-The estimate of the true weekly trap rate is derived based on previously PIT-tagged sp/sum Chinook Salmon and steelhead that are crossing LGD that week, using a Lincoln-Peterson mark-recapture model [@Seber2002].
-
-$$
-m_t \sim \text{Bin} \left( \nu_t, M_t \right) \\
-$$
-The fish, from both species, caught in the trap that week are considered the "mark" group ($m_t$), and all the previously PIT tagged fish who are detected at the upper end of the LGD fish ladder that week are considered the second "capture" group, $M_t$ (which includes recaptures of the the "marked" fish).
-The proportion of previously marked fish that are caught in the adult trap, $m/M$ can be modeled with a binomial distribution using the same trap rate parameter, $\nu_t$.
-Although the group of previously PIT tagged fish is not assumed to be representative of the overall run, the rate at which they are caught in the trap should be the same rate that the overall run experiences.
-The more tagged fish crossing the dam in a particular week, the more certain we can be of the true trap rate.
-
-The number of fish counted at the window, $Y^W_t$, is modeled as a (potentially) over-dispersed negative binomial process, with an expected value of $X_t^{day}$, the number of fish crossing the dam while the window is open.
-This is simply the total number of fish crossing that week, $X_t$, multiplied by the proportion of fish crossing while the window is open for counting, $\theta_t$.
-In the formula below, $p_t$ is the proportion of fish observed at the window and $r$ is the shape parameter.
-If $r$ is estimated to be small it provides evidence for over-dispersion, and as it grows very large, the negative binomial distribution behaves like a Poisson distribution.
-
-$$
-\begin{aligned}
-  X_t^{day} &= X_t * \theta_t \\
-  p_t &= \frac{r}{\left( r + X_t^{day} \right)} \\
-  Y^W_t &\sim \text{NegBin} \left( p_t, r \right) \\
-\end{aligned}
-$$
-
-Thus, the unknown true number of fish crossing LGD each week, $X_t$, is estimated from two different data sources: window counts and fish sampled in the trap.
-The window counts provide an estimate (with some potential observer error) of the fish crossing during daytime hours, while the fish in the trap, when expanded by the estimated true trap rate, provide an estimate of the total fish crossing that week.
-The use of both estimates allows for an estimation of observer error in the window counts which has previously been inestimable.
-For weeks when the trap is down, STADEM relies exclusively on the window counts and nighttime passage data, but there will be more uncertainty in the estimates.
 
@@
 The initial day-time passage and re-ascension rates, $\theta_1$ and $\eta_1$ were given Uniform(0,1) priors.
-Finally, $\phi_{w,1}$ and $\phi_{hnc,1}$ were given priors of Uniform(-3,3), in an effort to make $\omega_1$ as uniformative as possible.
+Finally, $\phi_{w,1}$ and $\phi_{hnc,1}$ were given priors of Uniform(-3,3), in an effort to make $\omega_1$ as uninformative as possible.
 
@@
 
-All data used in this manuscript, as well as associated code for both the simulations and the Lower Granite appllication, can be found at [https://www.github.com/KevinSee/ManuscriptSTADEM]([https://www.github.com/KevinSee/ManuscriptSTADEM]).
+All data used in this manuscript, as well as associated code for both the simulations and the Lower Granite application, can be found at [https://www.github.com/KevinSee/ManuscriptSTADEM]([https://www.github.com/KevinSee/ManuscriptSTADEM]).
 
@@
 
-Weekly estimates of total escapement over Lower Granite Dam tracked the window counts and trap estimates (One example: Figure \@ref(fig:jags-fit-fig), with smaller credible intervals than the trap estimates alone.
+Weekly estimates of escapement over Lower Granite Dam tracked the window counts and trap estimates, with smaller credible intervals than the trap estimates alone.
+(One example: Figure \@ref(fig:jags-fit-fig).
 Similar plots for all model runs (each species and year combination) are available on the manuscript GitHub page: [https://www.github.com/KevinSee/ManuscriptSTADEM]([https://www.github.com/KevinSee/ManuscriptSTADEM])).
@@
 The differences in the estimates were also apparent visually (Figure \@ref(fig:night-reasc-diff-fig)).
-In particular, there were several weeks when the window counts were quite large, but the nighttime passage and re-ascension rates differed by as much as 10%, in which case assuming them equal would lead to a biased estimate of escapement.
-When nighttime passage is larger than re-ascension, the window counts will be biased low, and vice versa.
+In particular, there were several weeks when the window counts were quite large, but the nighttime passage and re-ascension rates differed by as much as 10%.
 
@@
 
-STADEM estimates were often slightly lower than those based on the number of fish caught in the adult trap because the trap estimates do not account for fallback and re-ascension.
-However, in comparison to window counts, STADEM estimates were higher when night passage rates exceeded re-ascension rates, and lower when the re-ascension rate was higher than the night passage rate.
-***For weeks when very few fish were caught in the trap or there was more uncertainty about the trap rate, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \@ref(fig:jags-fit-fig).*** That year also shows the utility of STADEM in dealing with missing data, as the trap was shut down for several weeks in July and August.
-The model's uncertainty is always smaller than the uncertainty from the trap estimates alone, whereas the window counts alone provide no estimate of uncertainty.
+The results from our simulations demonstrate that STADEM provides consistent, unbiased estimates of escapement across a range of scenarios.
+Regardless of whether observer error at the window is negligible, low or high, whether nighttime passage and re-ascension rates are equal or not, whether the trap is shut down for several weeks in a season or not, STADEM produced estimates of unique escapement that were unbiased, precise (CVs in the 2-3% range) and contained appropriate credible intervals.
+Although the average bias was often positive across our simulation scenarios (Table \@ref(tab:summ-tab)), Figure \@ref(fig:rel-bias-lgd) shows that some simulations resulted in negative biases, while some were positive.
+We posit that conducting further simulations would bring the average bias closer to zero, but run times constrained how many simulations we were able to conduct.
+
+As for application at Lower Granite Dam, STADEM weekly estimates were often slightly lower than those based on the number of fish caught in the adult trap because the trap estimates do not account for fallback and re-ascension (Figure \@ref(fig:jags-fit-fig)).
+However, the model's uncertainty is always smaller than the uncertainty from the trap estimates alone, whereas the window counts by themselves provide no estimate of uncertainty.
+When comparing the total estimates of each season with the window counts (Table \@ref(tab:lgr-tab)), STADEM estimates were higher when night passage rates exceeded re-ascension rates, and lower when the re-ascension rate was higher than the night passage rate.
+Given that there were several weeks with large window counts when the rates differed by as much as 10% (Figure \@ref(fig:night-reasc-diff-fig)), the counts for those weeks should be considered biased.
+***For weeks when very few fish were caught in the trap or there was more uncertainty about the trap rate, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \@ref(fig:jags-fit-fig).***
+
+<!-- Do we need that last sentence, or is it confusing about how or why STADEM "tracks" the window counts or trap estimates more closely in some weeks vs. others? -->
 
@@
 
-By incorporating data from the adult fish trap with live and video window counts in a state-space framework, STADEM incorporates missing data at either the observation window or adult trap seamlessly.
+By incorporating data from the adult fish trap with live and video window counts in a state-space framework, STADEM incorporates missing data at either the observation window or adult trap seamlessly (for example, see Figure \@ref(fig:jags-fit-fig) when the trap was shut down for several weeks in July and August).
 At Lower Granite Dam, the adult trap has been closed for brief or extended periods of time (i.e., days, weeks) intermittently in five of the last ten years, often during peak run times [@USACE2010; @USACE2011; @USACE2012; @USACE2013; @Ogden2016a; @USACE2015; @USACE2016; @USACE2017; @USACE2018; @USACE2019].

### 2021-03-22 analysis/paper/paper.Rmd
made a few more revisions to address reveiewer comments and clean up some confusing language.
Clarified some responses to reviewer comments.

Re-knit to html and pdf versions

@@
 Further, total and weekly estimates are parsed into three origin groups: wild fish, hatchery fish with a clipped adipose fin, and unclipped hatchery fish.
-Estimates of escapement account for fish that migrate through the ladder outside of observation hours (nighttime passage) and those that ascend the ladder multiple times (re-ascension).
 Our model is implemented in the **ST**ate-space **A**dult **D**am **E**scapement **M**odel (STADEM) package for the statistical software R [@R-Core-Team2020], and is available for download or installation from [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM).
@@
 
-We used STADEM and three sources of data to estimate sp/sum Chinook Salmon and steelhead escapement at Lower Granite Dam from 2010-2019.
-Data sources included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of fish PIT tagged prior to arriving at Lower Granite Dam and detected in the adult fish ladder.
+Data sources for STADEM included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of fish PIT tagged prior to arriving at Lower Granite Dam and detected in the adult fish ladder.
 Below, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; similar data could likely be obtained from other fish passage facilities.
@@
 
-Daily counts of adult sp/sum Chinook Salmon and steelhead passing the observation window located on the Lower Granite Dam fish ladder were estimated and provided by the US Army Corps of Engineers.
+Daily counts of adult sp/sum Chinook Salmon and steelhead passing the observation window located on the Lower Granite Dam fish ladder were compiled and provided by the US Army Corps of Engineers.
 When summed, they provide an estimate of the number of fish ascending and passing Lower Granite Dam each season.
@@
 
-All captured fish were anesthetized, identified to species, examined for existing marks or tags, measured for fork length and visually identified as wild or hatchery.
+All captured fish were anesthetized, identified to species, examined for existing marks or tags, measured for fork length, and visually identified as wild or hatchery.
 The most widely used marking of hatchery fish is a clipped (removed) adipose fin, although coded wire tags are used in less than 10% of the hatchery releases.
@@
 Prior to release, all non-PIT tagged fish with an intact adipose fin (i.e., putatively wild) received a PIT tag.
-Final determination of wild, clipped hatchery, or unclipped hatchery origins were assigned using a post-hoc analysis of marks and tags, including parentage-based tagging [@Steele2013; @Steele2019].
+Final determination of wild, clipped hatchery, or unclipped hatchery origins were assigned using a post-hoc analysis of marks and tags, including parentage-based tagging based on the genetic samples taken [@Steele2013; @Steele2019].
 Data from the adult trap were collected and managed by multiple agencies and were made available by the Idaho Department of Fish and Game [@Camacho2018].
@@
 This is simply the total number of fish crossing that week, $X_t$, multiplied by the proportion of fish crossing while the window is open for counting, $\theta_t$.
-In the formula below, $p_t$ is the proportion of fish observed at the window and $r$ is the shape parameter.
+In the formula below, $p_t$ is the probability parameter and $r$ is the size parameter of a negative binomial distribution (as defined in @R-rjags).
 If $r$ is estimated to be small it provides evidence for over-dispersion, and as it grows very large, the negative binomial distribution behaves like a Poisson distribution.
@@
 Our objective was to assess STADEM model estimates of origin-specific (wild, hatchery, hatchery no-clip) escapement for accuracy and precision, given a known simulated "truth", under different possible conditions.
-We developed those conditions from the combinations of two trap rate scenarios (constant and shut down for 3 weeks), two fallback/re-ascension and nighttime passage combinations (equal and offsetting, and unequal), and three window count error rates (none, low and high); resulting in twelve different scenarios (Table \@ref(tab:lgr-sim-tab)).
+We developed those conditions from the combinations of two trap rate scenarios (constant and shut down for 3 weeks), two fallback/re-ascension and nighttime passage combinations (equal and unequal), and three window count error rates (none, low and high); resulting in twelve different scenarios (Table \@ref(tab:lgr-sim-tab)).
 The simulation parameters such as proportion of origin, run-timing, nighttime passage rates, fallback and re-ascension rates and trap rates were based on observed values at Lower Granite Dam between 2010-2015.
 We generated 99 simulations for each scenario, and ran STADEM on each one, resulting in a total of 1,188 model runs.
-Further details about simulation procedures can be found in [Appendix A](#append1).
+Further details about the simulation procedures can be found in [Appendix A](#append1).
 
 To assess the performance of the model, we evaluated the estimates of origin-specific escapement by several measures.
-Accuracy was captured by the relative bias, the difference between the simulated "truth" and the STADEM estimate, divided by the simulated "truth".
+Accuracy was captured by the relative bias: the difference between the simulated "truth" and the STADEM estimate, divided by the simulated "truth".
 We estimated the precision by examining the average coefficient of variation (CV) of the estimates.
@@
 Finally, we evaluated whether estimates of uncertainty were appropriate by calculating coverage probabilities by determining what proportion of the model results generated an estimated 95% credible interval that contained the simulated true value.
-To compare the STADEM results to strictly window counts, we compared the model estimates of total unique fish and the window counts with the simulated truth.
+To compare the STADEM results directly to window counts, we compared the model estimates of total unique fish and the simulated window counts to the simulated number of unique fish.
 
@@
 
-STADEM results showed similar patterns for estimates of unique total fish and unique wild fish, in that both were unbiased with similar variance in the estimator across all scenarios (Figure \@ref(fig:rel-bias-lgd)).
+STADEM results showed similar patterns for estimates of unique total fish and unique wild fish, in that both were unbiased with similar variance in the estimator across all scenarios as seen in the consistent interquartile distance in the boxplots of Figure \@ref(fig:rel-bias-lgd).
 For unique total fish, window counts were unbiased, except in scenarios when night passage and re-ascension rates were not equal (N-R).
-In those scenarios, they were biased by nearly 5%, the difference between night passage and re-ascension rates (Table \@ref(tab:lgr-sim-tab)).
+In those scenarios, they were biased by nearly 5% (Figure \@ref(fig:rel-bias-lgd)), the difference between night passage and re-ascension rates (Table \@ref(tab:lgr-sim-tab)).
 The variance of the window counts also grew as the simulated observer error grew (no error vs.
@@
 
-Estimates of wild escapement were unbiased, with an average relative bias of 0.2–0.3%.
+Estimates of wild escapement were unbiased, with an average relative bias of 0.2–0.3%, although the bias appeared equally distributed between positive and negative values across the simulations (Figure \@ref(fig:rel-bias-lgd)).
 The CV of the estimates averaged 2.0–3.0%, with higher CV's being associated with scenarios when the trap was closed for 3 weeks.
@@
 Weekly nighttime passage and re-ascension proportions were significantly different (2-sample test of proportions, *p* < 0.05) in approximately 50% of the weeks with sufficient PIT tag data to test.
-The differences in the estimates were also apparent visually (Figure \@ref(fig:night-reasc-diff-fig)).
+The differences in the estimated rates were also apparent visually (Figure \@ref(fig:night-reasc-diff-fig)).
 In particular, there were several weeks when the window counts were quite large, but the nighttime passage and re-ascension rates differed by as much as 10%.
@@
 Although the average bias was often positive across our simulation scenarios (Table \@ref(tab:summ-tab)), Figure \@ref(fig:rel-bias-lgd) shows that some simulations resulted in negative biases, while some were positive.
-We posit that conducting further simulations would bring the average bias closer to zero, but run times constrained how many simulations we were able to conduct.
-
-As for application at Lower Granite Dam, STADEM weekly estimates were often slightly lower than those based on the number of fish caught in the adult trap because the trap estimates do not account for fallback and re-ascension (Figure \@ref(fig:jags-fit-fig)).
+We posit that conducting further simulations would bring the average bias closer to zero, but run times constrained how many simulations we were able to conduct for each scenario.
+
+As for application at Lower Granite Dam, the STADEM weekly estimates were always close to the window counts and the trap estimates, indicating that both sources of data were providing information about the same phenomenon.
+The STADEM weekly estimates were often slightly lower than those based on the number of fish caught in the adult trap because the trap estimates do not account for fallback and re-ascension (Figure \@ref(fig:jags-fit-fig)).
 However, the model's uncertainty is always smaller than the uncertainty from the trap estimates alone, whereas the window counts by themselves provide no estimate of uncertainty.
@@
 The first is that by accounting for night passage and re-ascension rates that are sometimes unequal, it corrects any bias in window counts.
-That bias ranged from -9.9% to 11% across all species and year combinations we fit STADEM to (Table \@ref(tab:lgr-tab)).
+That bias ranged from -9.9% to 11% across all species and year combinations we fit STADEM to at Lower Granite Dam (Table \@ref(tab:lgr-tab)).
 Whether that amount of potential bias is of concern to management agencies is a decision for them, but we have presented a tool to correct that bias.
@@
 Trap closures are typically associated with elevated water temperatures resulting in potential fish handling stress and/or trap malfunctions [@Ogden2016b].
-Given predicted Pacific Northwest climate change scenarios [@Zhang2019] trap closures from high water temperatures may become more commonplace in the future, amplifying the need for a modeling framework that accounts for periods of missing data while still capturing estimate uncertainty.
+Given predicted Pacific Northwest climate change scenarios [@Zhang2019] trap closures from high water temperatures may become more commonplace in the future, amplifying the need for a modeling framework that accounts for periods of missing data while still capturing uncertainty in the estimates.
 Additionally, having a framework in place that accounts for missing periods of data will allow for increased logistic flexibility if, for example, maintenance or construction is needed at the observational window or adult trap.
@@
 However, if in-season management decisions do not require this correction or could accept the potential bias, origin calls at the trap could be used in-season as a first approximation to escapement.
-Final post-hoc estimates parsed by origin could then be finalized at season’s end.
+Final post-hoc estimates parsed by origin could then be completed at season’s end.
 All other data included in this model (e.g., window counts and PIT observations) are otherwise provided in near real-time by DART.
@@
 Estimates of species and origin-specific escapement at Lower Granite Dam, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates [e.g., @Camacho2018; @Schrader2013] that are important for fisheries management and productivity monitoring of wild populations.
-As an example, STADEM is being applied at Lower Granite Dam to estimate the total unique wild fish migrating past the dam.
-Estimates of fish passing the dam are then combined with estimated movement or transition probabilities based on PIT tag observations at instream PIT tag detection systems throughout the Snake River basin, similar to @Waterhouse2020, to estimate escapement to Snake River populations and locations throughout the basin [@Orme2019].
+As an example, STADEM estimates of unique wild fish at Lower Granite are combined with estimated movement or transition probabilities based on PIT tag observations at instream PIT tag detection systems throughout the Snake River basin, similar to @Waterhouse2020, to estimate escapement to Snake River populations and locations throughout the basin [@Orme2019].
 Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations.
@@
 
-With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall-run Chinook Salmon, Pacific lamprey *Lampetra tridentata*), or elsewhere, provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run and tag detection infrastructure (e.g., a PIT tag detection array or similar).
+With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall-run Chinook Salmon, Pacific lamprey *Lampetra tridentata*), or elsewhere, provided there exists a fish counting mechanism, a trap that can be used to sample a portion of the run, and tag detection infrastructure (e.g., a PIT tag detection array or similar).
 Justification and infrastructure exist for applying a modified STADEM framework for fish passing Bonneville Dam, the lowest dam on the Columbia River, or Priest Rapids Dam in the upper Columbia River.
@@
 
-```{r dag-fig, fig.height = 5, fig.cap = "Directed acyclic graph showing the STADEM model framework."}
+```{r dag-fig, out.width = "100%", fig.cap = "Directed acyclic graph showing the STADEM model framework."}
 include_graphics("../figures/ModelSchematic.png")

### 2021-03-26 analysis/paper/appendices/Appendix_simulation_details.Rmd
started to address reviewer comments
changed the title and tag of simulation appendix since model framework one is being dropped
round of edits from Mike

@@
-# Appendix B - Simulation Details {#append2}
+---
+output:
+  pdf_document: default
+  html_document: default
+---
+# Appendix A - Simulation Details {#append1}
 
-To simulate fish passing a dam, an $\mathbb{R}$ software function was developed [@R-Core-Team2020].
+To simulate fish passing a dam, we developed an $\mathbb{R}$ software function [@R-Core-Team2020].
 The function randomly samples observations from assumed probability distribution functions (pdf) with known parameters.
@@
 
-All code for simulating data and fitting STADEM to that data can be found at [https://www.github.com/KevinSee/ManuscriptSTADEM]([https://www.github.com/KevinSee/ManuscriptSTADEM]).
+All code for simulating data and fitting STADEM to that data can be found at [https://www.github.com/KevinSee/ManuscriptSTADEM](https://www.github.com/KevinSee/ManuscriptSTADEM).

### 2021-03-26 analysis/paper/paper.Rmd
round of edits from Mike

@@
 Our results demonstrate that the model produced unbiased estimates across all tested scenarios.
-We also applied this model to data from the Lower Granite Dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring/summer-run Chinook Salmon and steelhead from spawn years 2010-2019.
+We also applied this model to empirical data from the Lower Granite Dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring/summer-run Chinook Salmon and steelhead from spawn years 2010-2019.
 keywords: |
@@
 Fish escapement refers to the number of adults that survive juvenile and subadult rearing, escape harvest, and achieve a size and age to potentially spawn [e.g., @Bue1998].
-For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their natal habitat to spawn.
+For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their natal habitat to spawn [e.g., @Woody2004; @Jasper2018].
 Escapement estimates facilitate effective fisheries management, particularly estimates of escapement for component groups [e.g., by stock, population, age, origin [wild, hatchery], @Hess2014; @Steinhorst2017; @Camacho2018] which provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations.
@@
 Snake River sp/sum Chinook Salmon and steelhead have substantial cultural, recreational, commercial and subsistence value within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries.
-The aggregate escapement of Snake River sp/sum Chinook Salmon and steelhead populations, with the exception of Tucannon River (a tributary of the Snake River whose confluence is downstream of Lower Granite Dam), is monitored at Lower Granite Dam located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations.
-Many fisheries management and conservation actions are made based on estimates of escapement by species and origin at Lower Granite Dam [@NMFS2019; @NWFSC2015; @NPCC2014].
-Additionally, harvest openings and closures, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia rivers fisheries, are predicated on escapements at Lower Granite Dam.
-
-The majority of sp/sum Chinook Salmon and steelhead returning to the Snake River must ascend a fish ladder on Lower Granite Dam before migrating upriver.
+The aggregate escapement of Snake River sp/sum Chinook Salmon and steelhead populations, with the exception of Tucannon River (a tributary of the Snake River whose confluence is downstream of Lower Granite Dam [LGR]), is monitored at LGR located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations.
+Many fisheries management and conservation actions are made based on estimates of escapement by species and origin at LGR [@NMFS2019; @NWFSC2015; @NPCC2014].
+As an example, tribal ceremonial and subsistence harvest rates in the region are informed by escapement at LGR; additionally, the openings, closures and bag limits of fisheries targeting hatchery fish, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia rivers, can be informed by escapements at LGR.
+
+The majority of sp/sum Chinook Salmon and steelhead returning to the Snake River must ascend a fish ladder on LGR before migrating upriver.
 Fish management agencies previously used counts of fish, by species, passing an observation window within the fish ladder as a census of fish escaping past the dam.
@@
 
-However, using window counts as a census of sp/sum Chinook Salmon and steelhead passing Lower Granite Dam can be problematic as it fails to account for multiple sources of uncertainty and disregards known biological processes.
-Typically, the observational error rates of live and video window counts are unknown, and sampling error rates are ignored.
-@Hatch1994 conducted a study of the counting system at Lower Granite Dam in 1992, and found that sp/sum Chinook Salmon were undercounted at the window by in-person window counters compared to video counts.
-There was a significant difference between daily standard counts (16 or 10 hours per day) and daily 24-hour video counts; they found a significant nighttime passage rate (crossing while the counting window was closed) with 3.5% of sp/sum Chinook Salmon and 6.6% of steelhead adults passing during "nighttime" hours.
+However, using window counts as a census of fish passing LGR can be problematic as it fails to account for multiple sources of uncertainty and disregards known biological processes.
+Typically, the observational error rates of sp/sum Chinook Salmon and steelhead counted passing a window on the fish ladder, using both in-person and video methods, are unknown, and sampling error rates are ignored.
+@Hatch1994 conducted a study of the counting system at LGR, and found that sp/sum Chinook Salmon were undercounted at the window by in-person window counters when compared to video counts.
+There was also a significant difference between daily standard counts (16 or 10 hours per day) and daily 24-hour video counts; the undercounting was contributed to a significant nighttime passage rate (crossing while the counting window was closed) with an estimated 3.5% of sp/sum Chinook Salmon and 6.6% of steelhead adults passing during "nighttime" hours.
 They also found some species misidentification issues, which could result in either under- or over-counting of a particular species.
-While that study identified potential issues with the current window counting procedure, it was only conducted in a single year, nearly 30 years ago, at a time when sp/sum Chinook Salmon and steelhead total returns were lower than in the past decade.
-
-One additional biological process, the fallback and re-ascension of adult migrants at the dam, should also be accounted for to estimate the true escapement at Lower Granite Dam.
-Adult fish that migrate through the ladder and past the dam may fallback over the dam (e.g., over a spillway, through navigation locks) and later may re-ascend the fish ladder again [@Boggs2004] and be double-counted at the window (re-ascension).
+While that study identified potential issues with the current window counting procedure, it was only conducted in a single year, nearly 30 years ago, and at a time when sp/sum Chinook Salmon and steelhead total returns were typically lower than in the past decade.
+
+One additional biological process, the fallback and re-ascension of adult migrants at the dam, should also be accounted for to estimate the true escapement of these species at LGR.
+Adult fish that migrate through the ladder and past the dam may fallback over the dam (e.g., over a spillway, through navigation locks) and later may re-ascend the fish ladder again [@Boggs2004] and be counted multiple times at the window (re-ascension).
 Whereas unaccounted nighttime passage can lead to an underestimate in escapement, both fallback without and with re-ascension instead potentially lead to an overestimate of escapement [@Dauble2000].
-Previously, it was often assumed that these two biological processes canceled each other out, resulting in window counts providing an unbiased estimate of escapement [@Camacho2018].
+Previously, it has often been assumed that these two biological processes essentially canceled each other out, resulting in window counts providing an unbiased estimate of escapement [@Camacho2018], which may be true in many or most years, but perhaps not always.
 
@@
 We also test whether observed nighttime passage and fallback/re-ascension rates are typically equal.
-Our method for estimating species-specific escapement past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of fish previously tagged with passive integrated transponder (PIT) tags in the adult ladder to explicitly model nighttime passage, re-ascension, and observation error using a state-space approach [@Royle2008] which separates process variance (e.g., week to week variance in true escapement) from observation error variance (e.g., observation error at the window, or sampling variance at the trap).
+Our method for estimating species-specific escapement past LGR incorporates window counts, data from the adult fish trap located on the ladder, and passive detection of adults in the ladder from fish previously tagged with passive integrated transponder (PIT) tags to explicitly model nighttime passage, re-ascension, and observation error using a state-space approach [@Royle2008] which separates process variance (e.g., week to week variance in true escapement) from observation error variance (e.g., observation error at the window, or sampling variance at the trap).
 To meet desired management and conservation objectives, modeled escapement includes estimates of uncertainty and is parsed into weekly strata.
@@
 To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates and compared STADEM results to the simulated "truth".
-We then applied this model to sp/sum Chinook Salmon and steelhead returns at Lower Granite Dam for spawn years 2010 - 2019.
+We then applied this model to sp/sum Chinook Salmon and steelhead returns at LGR for spawn years 2010 - 2019.
 
@@
 
-Data sources for STADEM included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of fish PIT tagged prior to arriving at Lower Granite Dam and detected in the adult fish ladder.
-Below, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; similar data could likely be obtained from other fish passage facilities.
+Data sources for STADEM included 1) counts of fish migrating past the observation window located on the adult fish ladder at LGR, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of fish PIT tagged prior to arriving at LGR and detected at arrays in the adult fish ladder.
+Below, we describe each of the data sources in more detail as they pertain to LGR; similar data could likely be obtained from other fish passage facilities.
 
@@
 
-Daily counts of adult sp/sum Chinook Salmon and steelhead passing the observation window located on the Lower Granite Dam fish ladder were compiled and provided by the US Army Corps of Engineers.
-When summed, they provide an estimate of the number of fish ascending and passing Lower Granite Dam each season.
+Daily counts of adult sp/sum Chinook Salmon and steelhead passing the observation window located on the LGR fish ladder were compiled and provided by the US Army Corps of Engineers.
+When summed, they provide an estimate of the number of fish ascending and passing LGR each season.
 Window counts were made for each species using direct in-person visual monitoring and video monitoring during daytime hours [@Hatch1994].
 Direct visual monitoring occurred during peak run times (April 1 – October 31) for 16 hours per day (0400 – 2000 hours).
-Video monitoring occurred during the beginning and tail ends of the adult runs (March 1 – March 31 and November 1 – December) for 10 hours per day (0600 – 1600 hours) [@USACE2015].
+Video monitoring occurred during the beginning and tail ends of the adult runs (March 1 – March 31 and November 1 – December 31) for 10 hours per day (0600 – 1600 hours) [@USACE2015].
 During direct visual monitoring, observers recorded each adult (≥ 30cm), by species, passing the window for 50 minutes of each hour of operation.
-The sum of the daily 50-minute counts were then multiplied by 1.2 to account for the 10 minutes when fish were not counted.
+The sum of the daily 50-minute counts were then multiplied by 1.2 to account for the 10 minutes each hour when fish were not counted.
 Daytime window counts were not expanded for fish that may have ascended the ladder outside of operational hours (i.e., nighttime passage) [@USACE2015].
 Window counts were accessed through the Columbia Basin Research Data Access in Real Time (DART) website, [www.cbr.washington.edu/dart/query/adult_daily](www.cbr.washington.edu/dart/query/adult_daily), using their window count query.
-Counts were provided for each day the fish ladder was open to passage, and had already been expanded by 1.2 (to account for the counting during 50 minutes of each hour).
+Counts were provided for each day the fish ladder was open to passage, and had already been expanded by 1.2.
 
@@
 
-The second source of data are from a sample of fish collected in the adult trap as they migrated past Lower Granite Dam [@Ogden2016b].
-The trap, also located within the adult fish ladder and upstream of the observation window, provides an opportunity to collect biological data (e.g., origin [wild, hatchery], genetic stock, length, age, sex) from captured adults which allows for decomposition of the escapement into specific groups [e.g., @Camacho2018; @Steinhorst2017].
+The second source of data are from a sample of fish collected in the adult trap as they migrated past LGR [@Ogden2016b].
+The trap, also located within the adult fish ladder and upstream of the observation window, provides an opportunity to collect biological data (e.g., phenotypic origin [wild, hatchery], tissue sample for genetic data, length, scale sample for age, sex) from captured adults which allows for decomposition of the escapement into specific groups [e.g., @Camacho2018; @Steinhorst2017].
 The trap was operational for 24 hours per day and randomly sampled the run by opening four times per hour for a length of time determined by a set daily trapping rate.
@@
 All captured fish were anesthetized, identified to species, examined for existing marks or tags, measured for fork length, and visually identified as wild or hatchery.
-The most widely used marking of hatchery fish is a clipped (removed) adipose fin, although coded wire tags are used in less than 10% of the hatchery releases.
-Some subset of hatchery fish are either intentionally or unintentionally released without a clipped adipose fin, and these are referred to as unclipped hatchery fish, or hatchery no-clips (HNC).
+The most widely used marking of hatchery fish is a clipped (removed) adipose fin, although coded wire tags are also used in less than 10% of the hatchery releases.
+Some subset of hatchery fish are either intentionally or unintentionally released without a clipped adipose fin, and these are referred to as unclipped hatchery fish, or hatchery no-clips.
 For adults with intact adipose fins (unclipped), which includes wild and hatchery no-clip individuals, either a portion or all of the fish trapped (depending on the year) had scale and genetic tissue samples taken.
@@
 Prior to 2013, only fish determined to be wild in origin at the trap were sampled for scale and genetics.
-Starting in 2013, every unclipped sp/sum Chinook Salmon and steelhead trapped at Lower Granite Dam was genotyped to simplify collaborative logistics and better estimate the proportion of unclipped hatchery fish that appear phenotypically wild.
+Starting in 2013, every unclipped sp/sum Chinook Salmon and steelhead trapped at LGR was genotyped to simplify collaborative logistics and to better estimate the true proportion of unclipped hatchery fish that can otherwise appear phenotypically wild.
 @Camacho2018 provide further details on trap sample rates and valid sample selection.
@@
 
-The last source of data were observations of PIT tagged adult sp/sum Chinook Salmon and steelhead at detection sites located in the Lower Granite Dam fish ladder.
-These observations provide data related to 1) a trap rate for the adult fish trap, 2) a nighttime passage rate, and 3) a re-ascension rate.
-Detections used in the model included all fish that were previously PIT tagged as juveniles or adults prior to arriving at Lower Granite Dam (i.e., does not include newly tagged adults at the dam) and detected at adult detection sites in the dam passage system.
+The last source of data were observations of PIT tagged adult sp/sum Chinook Salmon and steelhead at PIT tag detection sites located in the LGR fish ladder.
+These observations provide data related to 1) an estimate of the "true" trap rate for the adult fish trap, 2) a nighttime passage rate, and 3) a re-ascension rate.
+Detections used in the model included all fish that were previously PIT tagged as juveniles or adults prior to arriving at LGR (i.e., does not include newly tagged adults at the dam) and detected at adult detection sites in the dam passage system.
 PIT tag data were provided through DART and the adult ladder PIT tag query; [http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail](http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail).
@@
 
-Escapement at Lower Granite Dam (LGD) is estimated by combining two independent observations (trap catches and window counts) of the true number of fish crossing the dam in a state-space model [@Royle2008] with a weekly time-step.
-Both are assumed to be corrupted observations of the true unknown number of fish crossing LGD each week.
+Escapement at LGR is estimated by combining two independent observations, trap catches and window counts, of the true number of fish crossing the dam in a state-space model [@Royle2008], with a weekly time-step.
+Both are assumed to be corrupted observations of the true unknown number of fish crossing LGR each week.
 We chose to use a weekly time-step to ensure sufficient sample sizes of PIT tag detections within each time strata, but this choice could be easily modified.
@@
 
-The set trapping rate (i.e., the recorded time that the trap is open to trap adults) does not always reflect the true trapping rate or proportion of fish that are actually captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error, among others.
+The set trapping rate (i.e., the pre-determined time that the trap is open each hour to trap adults) does not always reflect the true trapping rate or proportion of fish that are actually captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error (pulses or lack thereof of passing adults), among others.
 Therefore, we use a mark-recapture approach to estimate a "true" trapping rate, $\nu_t$.
-The estimate of the true weekly trap rate is derived based on previously PIT-tagged sp/sum Chinook Salmon and steelhead that are crossing LGD that week, using a Lincoln-Peterson mark-recapture model [@Seber2002].
+The estimate of the true weekly trap rate is derived based on previously PIT-tagged sp/sum Chinook Salmon and steelhead that are crossing LGR that week, and using a Lincoln-Peterson mark-recapture model [@Seber2002].
 
@@
 
-The fish, from both species, caught in the trap that week are considered the "mark" group ($m_t$), and all the previously PIT tagged fish who are detected at the upper end of the LGD fish ladder that week are considered the second "capture" group, $M_t$ (which includes recaptures of the the "marked" fish).
+The fish, from both species, caught in the trap that week are considered the "mark" group ($m_t$), and all the previously PIT tagged fish who are detected at the upper end of the LGR fish ladder that week are considered the second "capture" group, $M_t$ (which includes recaptures of the "marked" fish).
 The proportion of previously marked fish that are caught in the adult trap, $m/M$ can be modeled with a binomial distribution using the same trap rate parameter, $\nu_t$.
@@
 
-Thus, the unknown true number of fish crossing LGD each week, $X_t$, is estimated from two different data sources: window counts and fish sampled in the trap.
+Thus, the unknown true number of fish crossing LGR each week, $X_t$, is estimated from two different data sources: window counts and fish sampled in the trap.
 The window counts provide an estimate (with some potential observer error) of the fish crossing during daytime hours, while the fish in the trap, when expanded by the estimated true trap rate, provide an estimate of the total fish crossing that week.
@@
 
-### Day-time Passage and Re-ascension Rates
-
-There are two processes that must be accounted for to correct any bias in the window counts: first, the proportion of fish that cross the dam while the window is closed for counting (nighttime passage rate), and second, the proportion of fish that are crossing the dam multiple times (re-ascension rate) and therefore potentially double-counted.
+### Daytime Passage and Re-ascension Rates
+
+There are two processes that must be accounted for to correct any bias in the window counts: first, the proportion of fish that cross the dam while the window is closed for counting (nighttime passage rate), and second, the proportion of fish that cross the dam multiple times (re-ascension rate) and therefore potentially counted multiple times.
 Both rates can be estimated from previously PIT tagged fish that are crossing the dam each week.
 
-The proportion of fish passing the window during non-operational hours, nighttime passage rate, is just the complement of the rate of fish passing during the day when the window is operating.
+The nighttime passage rate is just the complement of the rate of fish passing during the day when the window is operating.
 The daytime passage rate for week $t$, $\theta_t$, is modeled as a random walk process (with deviations, $g_t$, drawn from a normal distribution with variance $\sigma^2_{\theta}$) and estimated from a binomial distribution based on the number of PIT tags observed to cross the dam during operational hours, $y^{day}_t$, and the total number of PIT tags observed to cross the dam at any point that week, $N_t$ [@Shumway2010].
@@
 
-The number of total fish crossing Lower Granite differs from the number of unique fish crossing Lower Granite because some fish fallback and re-ascend the dam.
-These fish are potentially double-counted at the window, and have the potential to be caught in the fish trap more than once.
+The number of total fish crossing LGR differs from the number of unique fish crossing LGR because some fish fallback and re-ascend the dam.
+These fish are potentially double-counted (or more) at the window, and have the potential to be caught in the fish trap more than once.
 The number of tags known to be re-ascending the dam each week, $y^{reasc}_t$, is modeled as a binomial process based on the estimated re-ascension rate, $\eta_t$, and the total number of tags crossing the dam that week, $N_t$.
-The logit of the re-ascension rate is modeled as a random walk process similar to day-time passage [@Shumway2010], with deviations, $f_t$, drawn from a normal distribution with variance $\sigma^2_\eta$.
+The logit of the re-ascension rate is modeled as a random walk process similar to daytime passage [@Shumway2010], with deviations, $f_t$, drawn from a normal distribution with variance $\sigma^2_\eta$.
 
@@
 
-After estimating the total number of fish to have crossed Lower Granite each week, $X_t$, the total must be further refined into the number of wild fish, $X_{w,t}$, hatchery clipped fish, $X_{hc,t}$ and hatchery no-clip fish, $X_{hnc, t}$.
+After estimating the total number of fish to have crossed LGR each week, $X_t$, the total must be further refined into the number of wild fish, $X_{w,t}$, hatchery clipped fish, $X_{hc,t}$ and hatchery no-clip fish, $X_{hnc, t}$.
 This is done by estimating a weekly origin proportion vector, $\omega_t$ based on the random sample of fish caught in trap that week, $Y^T_t$.
@@
 
-Finally, the number of unique fish by origin crossing Lower Granite each week ($X_{w,t}$, $X_{hc,t}$, and $X_{hnc,t}$) is the product of the total fish crossing that week, $X_t$, multiplied by one minus the re-ascension rate, $(1 - \eta_t)$, and the origin proportion vector, $\omega_t$.
+Finally, the number of unique fish by origin crossing LGR each week ($X_{w,t}$, $X_{hc,t}$, and $X_{hnc,t}$) is the product of the total fish crossing that week, $X_t$, multiplied by one minus the re-ascension rate, $(1 - \eta_t)$, and the origin proportion vector, $\omega_t$.
 
@@
 Variance parameters $\sigma_X, \sigma_\eta, \sigma_\theta$, and $\sigma_\omega$, as well as the initial abundance, $X_1$, and the overdispersion parameter of the negative binomial, $r$, were given half-Cauchy priors with mean of 0 and scale of 100.
-The initial day-time passage and re-ascension rates, $\theta_1$ and $\eta_1$ were given Uniform(0,1) priors.
+The initial daytime passage and re-ascension rates, $\theta_1$ and $\eta_1$ were given Uniform(0,1) priors.
 Finally, $\phi_{w,1}$ and $\phi_{hnc,1}$ were given priors of Uniform(-3,3), in an effort to make $\omega_1$ as uninformative as possible.
@@
 
-We tested the STADEM model on a variety of simulated data sets, constructed to mimic the dynamics at Lower Granite Dam.
-These simulated data sets contained a fixed number of unique adult fish of known origin crossing a dam, from a total of 25 fictional populations with differential run-timing (i.e., date of passage at Lower Granite Dam).
+We tested the STADEM model on a variety of simulated data sets, constructed to mimic some dynamics of interest at LGR.
+These simulated data sets each contained a fixed number of unique adult fish of known origin crossing a dam, from a total of 25 fictional populations with differential run-timing (i.e., date of passage at LGR).
 Each simulated fish was given a date of ladder ascension, based on its population and the range of observed run timing for that population.
@@
 Our objective was to assess STADEM model estimates of origin-specific (wild, hatchery, hatchery no-clip) escapement for accuracy and precision, given a known simulated "truth", under different possible conditions.
-We developed those conditions from the combinations of two trap rate scenarios (constant and shut down for 3 weeks), two fallback/re-ascension and nighttime passage combinations (equal and unequal), and three window count error rates (none, low and high); resulting in twelve different scenarios (Table \@ref(tab:lgr-sim-tab)).
-The simulation parameters such as proportion of origin, run-timing, nighttime passage rates, fallback and re-ascension rates and trap rates were based on observed values at Lower Granite Dam between 2010-2015.
+We developed those conditions from the combinations of two trap rate scenarios (constant and shut down for 3 weeks), two fallback/re-ascension and nighttime passage combinations (equal and unequal), and three window count error rates (none, low and high); resulting in twelve different scenarios overall (Table \@ref(tab:lgr-sim-tab)).
+The simulation parameters such as proportion of origin, run-timing, nighttime passage rates, fallback and re-ascension rates and trap rates were based on observed values at LGR between 2010-2015.
 We generated 99 simulations for each scenario, and ran STADEM on each one, resulting in a total of 1,188 model runs.
@@
 
-## Lower Granite Application
-
-We applied STADEM to empirical data from Lower Granite Dam for sp/sum Chinook Salmon and steelhead returning to the Snake River during spawn years 2010 to 2019.
+## Lower Granite Dam Application
+
+We applied STADEM to empirical data from LGR for sp/sum Chinook Salmon and steelhead returning to the Snake River during spawn years 2010 to 2019.
 Window counts for both species were accessed from DART via functions within STADEM.
 For sp/sum Chinook Salmon, a spawn year refers to adults that migrate past the dam prior to August 17 each year and spawn that late summer and fall.
-For steelhead, the spawn year is defined as steelhead that migrate past Lower Granite Dam starting July 1 the previous year and prior to June 30 of the given year (e.g., spawn year 2017 steelhead migrate past Lower Granite between July 1, 2016 and June 30, 2017).
-Data from the adult trap were made available by Idaho Department of Fish and Game, and adult PIT tag detection data within the fish passage ladder at Lower Granite Dam were accessed from the PTAGIS regional database ([https://www.ptagis.org/](https://www.ptagis.org/)).
-
-All data used in this manuscript, as well as associated code for both the simulations and the Lower Granite application, can be found at [https://www.github.com/KevinSee/ManuscriptSTADEM]([https://www.github.com/KevinSee/ManuscriptSTADEM]).
+For steelhead, the spawn year is defined as steelhead that migrate past LGR starting July 1 the previous year and prior to June 30 of the given year (e.g., spawn year 2017 steelhead migrate past LGR between July 1, 2016 and June 30, 2017).
+Data from the adult trap were made available by Idaho Department of Fish and Game, and adult PIT tag detection data within the fish passage ladder at LGR were accessed from the PTAGIS regional database ([https://www.ptagis.org/](https://www.ptagis.org/)).
+
+All data used in this manuscript, as well as associated code for both the simulations and the LGR application, can be found at [https://www.github.com/KevinSee/ManuscriptSTADEM]([https://www.github.com/KevinSee/ManuscriptSTADEM]).
 
@@
 
-The STADEM performance evaluation statistics were very similar for hatchery clipped (N = 70,000), hatchery no-clip (N = 5,000) and wild fish (N = 25,000).
-In the interest of brevity, we only present the results from wild fish, which corresponds to a medium sized escapement level.
+The evaluation statistics (relative bias, CV, RMSE, coverage probabilities) to assess performance of the STADEM model were very similar among hatchery clipped (N = 70,000), hatchery no-clip (N = 5,000) and wild fish (N = 25,000).
+In the interest of brevity, we only present the results from wild fish here, which corresponds to a moderate sized escapement relative to the hatchery clipped and hatchery no-clip groups.
 
@@
 The variance of the window counts also grew as the simulated observer error grew (no error vs.
-Err_L vs.
-Err_H), whereas the variance in the STADEM estimates remained fairly constant, albeit often larger than the window count variance.
+Err L vs.
+Err H), whereas the variance in the STADEM estimates remained fairly constant, albeit often larger than the window count variance.
 
 Estimates of wild escapement were unbiased, with an average relative bias of 0.2–0.3%, although the bias appeared equally distributed between positive and negative values across the simulations (Figure \@ref(fig:rel-bias-lgd)).
-The CV of the estimates averaged 2.0–3.0%, with higher CV's being associated with scenarios when the trap was closed for 3 weeks.
+The CV of the estimates averaged 2.0–3.0%, with higher CV's being associated with scenarios when the trap was simulated closed for 3 weeks.
 The coverage probabilities always exceeded 95% across all scenarios.
@@
 
-## Lower Granite Application
-
-We applied STADEM to data from Lower Granite Dam for sp/sum Chinook Salmon and steelhead for spawn years 2010–2019.
+## Lower Granite Dam Application
+
+We applied STADEM to data from LGR for sp/sum Chinook Salmon and steelhead for spawn years 2010–2019.
 Estimates of total escapement, as well as estimates of wild, hatchery clipped, and hatchery no-clip estimates are presented in Table \@ref(tab:lgr-tab).
-Estimates of total unique fish escaping past Lower Granite Dam were sometimes higher and sometimes lower than the raw window counts, indicating that the relative strength of nighttime passage and re-ascension rates differed across years.
-Window count bias ranged from -3% to 11% for sp/sum Chinook Salmon and -9.9% to 6.6% for steelhead.
-Coefficients of variation ranged from 2.5 to 7.1% for wild fish, 2.3 to 5.1% for clipped hatchery fish, 3.4 to 8.8% for hatchery no-clip fish and 2.2 to 4.7% for total unique fish past Lower Granite Dam.
-
-Weekly estimates of escapement over Lower Granite Dam tracked the window counts and trap estimates, with smaller credible intervals than the trap estimates alone.
+Estimates of total unique fish escaping past LGR were sometimes higher and sometimes lower than the raw window counts, indicating that the relative strength of nighttime passage and re-ascension rates differed across years.
+Window count bias ranged from -3.0% to 11.0% for sp/sum Chinook Salmon and -9.9% to 6.6% for steelhead.
+Coefficients of variation ranged from 2.5 to 7.1% for wild fish, 2.3 to 5.1% for clipped hatchery fish, 3.4 to 8.8% for hatchery no-clip fish and 2.2 to 4.7% for total unique fish past LGR.
+
+Weekly estimates of escapement over LGR tracked the window counts and trap estimates, with smaller credible intervals than the trap estimates alone.
 (One example: Figure \@ref(fig:jags-fit-fig).
-Similar plots for all model runs (each species and year combination) are available on the manuscript GitHub page: [https://www.github.com/KevinSee/ManuscriptSTADEM]([https://www.github.com/KevinSee/ManuscriptSTADEM])).
+Similar plots for all model runs [each species and year combination] are available on the manuscript GitHub page: [https://www.github.com/KevinSee/ManuscriptSTADEM]([https://www.github.com/KevinSee/ManuscriptSTADEM])).
 
 Weekly nighttime passage and re-ascension proportions were significantly different (2-sample test of proportions, *p* < 0.05) in approximately 50% of the weeks with sufficient PIT tag data to test.
-The differences in the estimated rates were also apparent visually (Figure \@ref(fig:night-reasc-diff-fig)).
+The differences in the estimated weekly rates were also apparent visually (Figure \@ref(fig:night-reasc-diff-fig)).
 In particular, there were several weeks when the window counts were quite large, but the nighttime passage and re-ascension rates differed by as much as 10%.
@@
 
-We have presented a novel method for estimating adult salmonid escapement past a large hydroelectric facility (e.g., Lower Granite Dam) that incorporates data from window counts, a fish trap, and observations of PIT tagged fish in the adult passage ladder.
+We have presented a novel method for estimating adult salmonid escapement past a large hydroelectric facility (e.g., LGR) that incorporates data from window counts, a fish trap, and observations of PIT tagged fish in the adult passage ladder.
 Our model explicitly accounts for nighttime passage, re-ascension, and potential error in both window and trap estimates.
-In doing so, we demonstrated that at Lower Granite Dam, nighttime passage and re-ascension rates do not always offset each other, and assuming they do will lead to biased escapement estimates in some years.
+In doing so, we demonstrated that at LGR, nighttime passage and re-ascension rates do not always offset each other, and assuming they do will lead to biased escapement estimates in some years.
 Our state-space model combined multiple imperfect sources of data to reduce bias in adult escapement estimates and provided quantitative estimates of uncertainty.
@@
 Regardless of whether observer error at the window is negligible, low or high, whether nighttime passage and re-ascension rates are equal or not, whether the trap is shut down for several weeks in a season or not, STADEM produced estimates of unique escapement that were unbiased, precise (CVs in the 2-3% range) and contained appropriate credible intervals.
-Although the average bias was often positive across our simulation scenarios (Table \@ref(tab:summ-tab)), Figure \@ref(fig:rel-bias-lgd) shows that some simulations resulted in negative biases, while some were positive.
-We posit that conducting further simulations would bring the average bias closer to zero, but run times constrained how many simulations we were able to conduct for each scenario.
-
-As for application at Lower Granite Dam, the STADEM weekly estimates were always close to the window counts and the trap estimates, indicating that both sources of data were providing information about the same phenomenon.
+Although the average bias among runs within a given simulation scenario was often positive (Table \@ref(tab:summ-tab)), Figure \@ref(fig:rel-bias-lgd) shows that for each scenario that some simulations resulted in a negative bias, while some were positive.
+We posit that conducting additional simulations would bring the average bias closer to zero, but run times constrained how many simulations we were able to conduct for each scenario.
+
+As for the application of STADEM to empirical data at LGR, the weekly estimates were always close to the window counts and the trap estimates, indicating that both sources of data were providing information about the same phenomenon.
 The STADEM weekly estimates were often slightly lower than those based on the number of fish caught in the adult trap because the trap estimates do not account for fallback and re-ascension (Figure \@ref(fig:jags-fit-fig)).
@@
 Given that there were several weeks with large window counts when the rates differed by as much as 10% (Figure \@ref(fig:night-reasc-diff-fig)), the counts for those weeks should be considered biased.
-***For weeks when very few fish were caught in the trap or there was more uncertainty about the trap rate, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \@ref(fig:jags-fit-fig).***
-
-<!-- Do we need that last sentence, or is it confusing about how or why STADEM "tracks" the window counts or trap estimates more closely in some weeks vs. others? -->
-
-STADEM is an improvement on window counts alone for several reasons.
-The first is that by accounting for night passage and re-ascension rates that are sometimes unequal, it corrects any bias in window counts.
-That bias ranged from -9.9% to 11% across all species and year combinations we fit STADEM to at Lower Granite Dam (Table \@ref(tab:lgr-tab)).
-Whether that amount of potential bias is of concern to management agencies is a decision for them, but we have presented a tool to correct that bias.
+For those weeks when very few fish were caught in the trap or there was more uncertainty about the trap rate (e.g., few previously tagged fish detected in the ladder), STADEM estimates more closely tracked the window counts, as seen in the second week of July 2014, in Figure \@ref(fig:jags-fit-fig).
+Vice-versa, for weeks when the estimate of the trap rate was more precise, STADEM estimates more closely follow the trap estimates, as seen in the first half of May (Figure \@ref(fig:jags-fit-fig)).
+In either case, the state-space nature of STADEM tends to put more emphasis on the data stream with less observation error.
+
+Using a state-space approach is an improvement on window counts alone for several reasons.
+The first is that by accounting for both night passage and re-ascension rates, that are sometimes unequal, it corrects any potential bias in window counts.
+That bias ranged from -9.9% to 11.0% across all species and year combinations we fit STADEM to empirical data from LGR (Table \@ref(tab:lgr-tab)).
+Whether that amount of potential bias is of concern to management agencies is a decision for them to make, but we have provided a tool to correct that bias if it is of concern.
 The other justification for STADEM over window counts is that it provides an estimate of observer error for the window counts, leading to estimates of precision for unique fish which are unobtainable with window counts alone.
 Capturing and accounting for known sources of error is prudent to minimize management decision risk [@Harwood2003; @Ascough-Ii2008].
-At the same time, STADEM's precision is still quite good, with CV's near 2-3%, well below the goal of 15% set forth by NOAA [@Crawford2011].
-
-By incorporating data from the adult fish trap with live and video window counts in a state-space framework, STADEM incorporates missing data at either the observation window or adult trap seamlessly (for example, see Figure \@ref(fig:jags-fit-fig) when the trap was shut down for several weeks in July and August).
-At Lower Granite Dam, the adult trap has been closed for brief or extended periods of time (i.e., days, weeks) intermittently in five of the last ten years, often during peak run times [@USACE2010; @USACE2011; @USACE2012; @USACE2013; @Ogden2016a; @USACE2015; @USACE2016; @USACE2017; @USACE2018; @USACE2019].
+At the same time, the precision among STADEMs estimates is still quite good, with CVs near 2-3%, well below the goal of 15% set forth by NOAA [@Crawford2011].
+
+By combining live and video window counts with data from the adult fish trap in a state-space framework, STADEM incorporates missing data at either the observation window or adult trap seamlessly (e.g., see Figure \@ref(fig:jags-fit-fig) when the trap was shut down for several weeks in July and August).
+At LGR, the adult trap has been closed for brief or extended periods of time (i.e., days, weeks) intermittently in five of the last ten years, and often during peak run times [@USACE2010; @USACE2011; @USACE2012; @USACE2013; @Ogden2016a; @USACE2015; @USACE2016; @USACE2017; @USACE2018; @USACE2019].
 Trap closures are typically associated with elevated water temperatures resulting in potential fish handling stress and/or trap malfunctions [@Ogden2016b].
-Given predicted Pacific Northwest climate change scenarios [@Zhang2019] trap closures from high water temperatures may become more commonplace in the future, amplifying the need for a modeling framework that accounts for periods of missing data while still capturing uncertainty in the estimates.
-Additionally, having a framework in place that accounts for missing periods of data will allow for increased logistic flexibility if, for example, maintenance or construction is needed at the observational window or adult trap.
+Given predicted Pacific Northwest climate change scenarios [@Zhang2019], trap closures from high water temperatures may become more commonplace in the future, amplifying the need for a modeling framework that accounts for periods of missing data while still capturing uncertainty in the estimates.
+Additionally, having a framework in place that accounts for missing periods of data will allow for increased logistical flexibility if, for example, maintenance or construction is needed at the observational window or adult trap.
 
 STADEM could be modified and run on a weekly basis or in near real-time to provide better in-season estimates for fisheries managers.
-Currently, the only roadblock to this at Lower Granite Dam is the identification of hatchery no-clip fish using genetic tissue samples [@Steele2013; @Steele2019] collected at the adult trap, which currently is completed post-hoc after the trapping season.
-The inclusion of genetic information typically results in a reduction in wild escapement estimates and an associated increase in hatchery no-clip escapement.
+Currently, the only roadblock to this at LGR is the identification of hatchery no-clip fish, which is aided by genetic tissue samples [@Steele2013; @Steele2019] collected at the adult trap, and parentage-based tagging [@Steele2013; @Steele2019] which currently is completed post-hoc after the trapping season.
+The inclusion of this genetic information typically results in a reduction in wild escapement estimates and an associated increase in hatchery no-clip escapement, as phenotypically wild fish are identified as the progeny of hatchery broodstock.
 However, if in-season management decisions do not require this correction or could accept the potential bias, origin calls at the trap could be used in-season as a first approximation to escapement.
@@
 All other data included in this model (e.g., window counts and PIT observations) are otherwise provided in near real-time by DART.
-Provided the Lower Granite Dam adult fish trap database was updated and available in near real-time, there are minimal obstacles for adapting the STADEM framework to provide in-season estimates of escapement.
-
-Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate escapement of sp/sum Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary or population specific spawning areas [@IPTDSW2020; @Kinzer2020].
-Estimates of species and origin-specific escapement at Lower Granite Dam, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates [e.g., @Camacho2018; @Schrader2013] that are important for fisheries management and productivity monitoring of wild populations.
-As an example, STADEM estimates of unique wild fish at Lower Granite are combined with estimated movement or transition probabilities based on PIT tag observations at instream PIT tag detection systems throughout the Snake River basin, similar to @Waterhouse2020, to estimate escapement to Snake River populations and locations throughout the basin [@Orme2019].
+Provided the LGR adult fish trap database was updated and available in near real-time, there are minimal obstacles for adapting the STADEM framework to provide in-season estimates of escapement.
+
+Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate escapement of sp/sum Chinook Salmon and steelhead past LGR, and returning to tributary or population specific spawning areas [@IPTDSW2020; @Kinzer2020].
+Estimates of species and origin-specific escapement at LGR, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates [e.g., @Camacho2018; @Schrader2013] that are important for fisheries management and productivity monitoring of wild populations.
+As an example, STADEM estimates of unique wild fish at LGR are combined with estimated movement or transition probabilities based on PIT tag observations at instream PIT tag detection systems throughout the Snake River basin, similar to @Waterhouse2020, to estimate escapement to Snake River populations and locations throughout the basin [@Orme2019].
 Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations.
@@
 
-With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall-run Chinook Salmon, Pacific lamprey *Lampetra tridentata*), or elsewhere, provided there exists a fish counting mechanism, a trap that can be used to sample a portion of the run, and tag detection infrastructure (e.g., a PIT tag detection array or similar).
+With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at LGR (e.g., fall-run Chinook Salmon, Pacific lamprey *Lampetra tridentata*), or elsewhere, provided there exists a fish counting mechanism, a trap that can be used to sample a portion of the run, and tag detection infrastructure (e.g., a PIT tag detection array or similar).
 Justification and infrastructure exist for applying a modified STADEM framework for fish passing Bonneville Dam, the lowest dam on the Columbia River, or Priest Rapids Dam in the upper Columbia River.
 Both locations currently trap a sub-sample of passing Chinook Salmon and/or steelhead for biological information and use window counts as a surrogate of true escapement.
-However, each has at least some similar problems to those observed at Lower Granite Dam such as unaccounted observer and sampling error, nighttime passage, and/or re-ascension.
-Certainly, estimating an unbiased total return to the entire Columbia River basin (i.e., Bonneville Dam) and Upper Columbia River with uncertainty would benefit managers and decision making.
+However, each has at least some similar problems to those observed at LGR such as unaccounted observer and sampling error, nighttime passage, and/or re-ascension.
+Certainly, estimating an unbiased total return to the vast majority of the Columbia River basin (i.e., Bonneville Dam) and Upper Columbia River with uncertainty would benefit managers and decision making.
 
@@
 Funding for this study and development of the STADEM model was partially provided by the Bonneville Power Administration under project 2003-017-00.
-Special thanks to Darren Ogden and staff at the Lower Granite Dam adult trap for their hard work and diligent data collection.
+Special thanks to Darren Ogden and staff at the LGR adult trap for their hard work and diligent data collection.
 Thank you to personnel at the Idaho Department of Fish and Game, particularly Paul Bunn, Tim Copeland, and Bill Schrader, for providing access to data from the adult fish trap and conceptualizing methods, and to Matthew Campbell and staff at the Eagle Fish Genetics Laboratory for analyzing genetic samples.
 Special thanks to Columbia Basin Research staff and the Columbia River Data Access in Real Time (DART) application, and Susannah Iltis in particular.
-And finally, thank you to Rick Orme and Brice Semmens for their contributions to the development of STADEM, as well as Eric Buhle, Sarah Hoffman, John Hargrove and all the other people who have contributed through productive critique and conversations.
+And finally, thank you to Rick Orme and Brice Semmens for their contributions to the development of STADEM, as well as Eric Buhle, Sarah Hoffmann, John Hargrove and all the other people who have contributed through productive critique and conversations.
 

### 2021-03-29 analysis/paper/paper.Rmd
rk revisions to .Rmd and reviewer comments

@@
         in_header: ../templates/header_manuscript.tex
+    bookdown::html_document2:
+      fig_caption: yes
+      number_sections: FALSE
+      toc: no
+      fig_width: 7
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks2.lua
+      - --lua-filter=../templates/pagebreak.lua
     bookdown::word_document2:
@@
 Our results demonstrate that the model produced unbiased estimates across all tested scenarios.
-We also applied this model to empirical data from the Lower Granite Dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring/summer-run Chinook Salmon and steelhead from spawn years 2010-2019.
+We also applied this model to empirical data from the Lower Granite Dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring/summer-run Chinook Salmon *Oncorhynchus tshawytscha* and steelhead *O.
+mykiss* from spawn years 2010-2019.
 keywords: |
@@
 
-Populations of Chinook Salmon *Oncorhynchus tshawytscha* and steelhead trout *O.
+Populations of Chinook Salmon *Oncorhynchus tshawytscha* and anadromous steelhead trout *O.
 mykiss* in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric projects on the Snake and Columbia rivers) and tributary habitats [@Nehlsen1991; @McClure2003; @NWFSC2015].
@@
 Many fisheries management and conservation actions are made based on estimates of escapement by species and origin at LGR [@NMFS2019; @NWFSC2015; @NPCC2014].
-As an example, tribal ceremonial and subsistence harvest rates in the region are informed by escapement at LGR; additionally, the openings, closures and bag limits of fisheries targeting hatchery fish, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia rivers, can be informed by escapements at LGR.
+As an example, tribal ceremonial and subsistence harvest rates in the region are informed by escapement at LGR; additionally, the openings, closures and bag limits of fisheries targeting hatchery fish, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia rivers, are informed by escapements at LGR[@UnitedStatesOregon].
 
@@
 We also test whether observed nighttime passage and fallback/re-ascension rates are typically equal.
-Our method for estimating species-specific escapement past LGR incorporates window counts, data from the adult fish trap located on the ladder, and passive detection of adults in the ladder from fish previously tagged with passive integrated transponder (PIT) tags to explicitly model nighttime passage, re-ascension, and observation error using a state-space approach [@Royle2008] which separates process variance (e.g., week to week variance in true escapement) from observation error variance (e.g., observation error at the window, or sampling variance at the trap).
+Our method for estimating species-specific escapement past LGR incorporates window counts, data from the adult fish trap located on the ladder, and passive detection of adults in the ladder from fish previously tagged with passive integrated transponder (PIT) tags to explicitly model nighttime passage, re-ascension, and observation error using a state-space approach [@Royle2008] which separates process variance (e.g., week to week variance in true escapement) from observational variance (e.g., observation error at the window, or sampling variance at the trap).
 To meet desired management and conservation objectives, modeled escapement includes estimates of uncertainty and is parsed into weekly strata.
@@
 To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates and compared STADEM results to the simulated "truth".
-We then applied this model to sp/sum Chinook Salmon and steelhead returns at LGR for spawn years 2010 - 2019.
+Then, as a case study, we applied this model to sp/sum Chinook Salmon and steelhead returns at LGR for spawn years 2010 - 2019.
 
@@
 Data sources for STADEM included 1) counts of fish migrating past the observation window located on the adult fish ladder at LGR, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of fish PIT tagged prior to arriving at LGR and detected at arrays in the adult fish ladder.
-Below, we describe each of the data sources in more detail as they pertain to LGR; similar data could likely be obtained from other fish passage facilities.
+Below, we describe each of the data sources in more detail as they pertain to LGR; similar data could likely be obtained from other species passing LGR (e.g., fall Chinook Salmon) or other fish passage facilities (e.g., Bonneville Dam).
 
@@
 $$
-Y^T_t \sim \text{Bin} \left( \nu_t, X_t \right) \\
-$$
-
-The set trapping rate (i.e., the pre-determined time that the trap is open each hour to trap adults) does not always reflect the true trapping rate or proportion of fish that are actually captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error (pulses or lack thereof of passing adults), among others.
+Y^T_t \sim \text{Bin} \left( \nu_t, X_t \right)
+$$
+
+The set trapping rate (i.e., the pre-determined time that the trap is open each hour to trap adults) does not always reflect the true trapping rate or proportion of fish actually captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error (pulses or lack thereof of passing adults), among others.
 Therefore, we use a mark-recapture approach to estimate a "true" trapping rate, $\nu_t$.
@@
 $$
-m_t \sim \text{Bin} \left( \nu_t, M_t \right) \\
+m_t \sim \text{Bin} \left( \nu_t, M_t \right)
 $$
@@
 
-We looked for differences in nighttime passage and re-ascension rates estimated using wild fish only, versus combining hatchery and wild fish together, and found no difference.
+We looked for differences in weekly nighttime passage and re-ascension rates estimated using wild fish only, versus combining hatchery and wild fish together, and found no difference.
+Additionally, sample sizes of observed tags during most weeks are small, and may prohibit reliable estimation of rates for individual wild or hatchery stocks.
 Therefore, we combine wild and hatchery PIT tagged fish observations to estimate common nighttime passage and re-ascension rates to increase sample sizes.
@@
 $$
+In practice, the total number of fish crossing LGR each week, $X_t$, could be split into other discrete groups if desired; for example, individual wild stocks using genetic stock identification or hatchery stocks with parentage based tagging [@Steele2019].
 
@@
 The simulation parameters such as proportion of origin, run-timing, nighttime passage rates, fallback and re-ascension rates and trap rates were based on observed values at LGR between 2010-2015.
-We generated 99 simulations for each scenario, and ran STADEM on each one, resulting in a total of 1,188 model runs.
+We generated 500 simulations for each scenario, and ran STADEM on each one, resulting in a total of 6,000 model runs.
 Further details about the simulation procedures can be found in [Appendix A](#append1).
@@
 To assess the performance of the model, we evaluated the estimates of origin-specific escapement by several measures.
-Accuracy was captured by the relative bias: the difference between the simulated "truth" and the STADEM estimate, divided by the simulated "truth".
+Accuracy was captured by the relative bias: the difference between the simulated "truth" and the mean STADEM estimate for each scenario, divided by the simulated "truth".
 We estimated the precision by examining the average coefficient of variation (CV) of the estimates.
@@
 For steelhead, the spawn year is defined as steelhead that migrate past LGR starting July 1 the previous year and prior to June 30 of the given year (e.g., spawn year 2017 steelhead migrate past LGR between July 1, 2016 and June 30, 2017).
-Data from the adult trap were made available by Idaho Department of Fish and Game, and adult PIT tag detection data within the fish passage ladder at LGR were accessed from the PTAGIS regional database ([https://www.ptagis.org/](https://www.ptagis.org/)).
+Data from the adult trap were made available by Idaho Department of Fish and Game, and adult PIT tag detection data within the fish passage ladder at LGR were accessed through DART and the adult ladder PIT tag query ([http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail](http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail)).
 
@@
 
-STADEM results showed similar patterns for estimates of unique total fish and unique wild fish, in that both were unbiased with similar variance in the estimator across all scenarios as seen in the consistent interquartile distance in the boxplots of Figure \@ref(fig:rel-bias-lgd).
+STADEM results showed similar patterns for estimates of unique total fish and unique wild fish, in that both were unbiased with similar variance in the estimator across all scenarios as seen in the consistent interquartile range in the boxplots of Figure \@ref(fig:rel-bias-lgd).
 For unique total fish, window counts were unbiased, except in scenarios when night passage and re-ascension rates were not equal (N-R).
-In those scenarios, they were biased by nearly 5% (Figure \@ref(fig:rel-bias-lgd)), the difference between night passage and re-ascension rates (Table \@ref(tab:lgr-sim-tab)).
+In those scenarios, they were biased by nearly 5% (Figure \@ref(fig:rel-bias-lgd)), the designed simulated difference between night passage and re-ascension rates (Table \@ref(tab:lgr-sim-tab)).
 The variance of the window counts also grew as the simulated observer error grew (no error vs.
@@
 Window count bias ranged from -3.0% to 11.0% for sp/sum Chinook Salmon and -9.9% to 6.6% for steelhead.
-Coefficients of variation ranged from 2.5 to 7.1% for wild fish, 2.3 to 5.1% for clipped hatchery fish, 3.4 to 8.8% for hatchery no-clip fish and 2.2 to 4.7% for total unique fish past LGR.
-
-Weekly estimates of escapement over LGR tracked the window counts and trap estimates, with smaller credible intervals than the trap estimates alone.
+STADEM coefficients of variation ranged from 2.5 to 7.1% for wild fish, 2.3 to 5.1% for clipped hatchery fish, 3.4 to 8.8% for hatchery no-clip fish and 2.2 to 4.7% for total unique fish past LGR.
+
+Weekly STADEM estimates of escapement over LGR tracked the window counts and trap estimates, with smaller credible intervals than the trap estimates alone.
 (One example: Figure \@ref(fig:jags-fit-fig).
@@
 The differences in the estimated weekly rates were also apparent visually (Figure \@ref(fig:night-reasc-diff-fig)).
-In particular, there were several weeks when the window counts were quite large, but the nighttime passage and re-ascension rates differed by as much as 10%.
+In particular, there were several weeks when the window counts were quite large, but the nighttime passage and re-ascension rates differed by as much as 10%; leading to larger window count bias in some years \@ref(tab:lgr-tab).
 
@@
 Our state-space model combined multiple imperfect sources of data to reduce bias in adult escapement estimates and provided quantitative estimates of uncertainty.
+Using both data streams proved beneficial, and protected estimates from unforeseen problems or periodic biases in observations (e.g., trap closures, unequal night time and re-ascension rates).
 Accurate population or group abundance estimates and uncertainty accounting for observation and process error can be particularly important when estimates are used for management and conservation decisions such as population viability analyses [@Paulsen2007].
@@
 Estimates of species and origin-specific escapement at LGR, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates [e.g., @Camacho2018; @Schrader2013] that are important for fisheries management and productivity monitoring of wild populations.
-As an example, STADEM estimates of unique wild fish at LGR are combined with estimated movement or transition probabilities based on PIT tag observations at instream PIT tag detection systems throughout the Snake River basin, similar to @Waterhouse2020, to estimate escapement to Snake River populations and locations throughout the basin [@Orme2019].
+As an example, STADEM estimates of unique wild fish at LGR are combined with estimated movement or transition probabilities based on PIT tag observations at instream PIT tag detection systems throughout the Snake River basin, similar to @Waterhouse2020, to estimate escapement to Snake River populations and locations throughout the basin [@IPTDSW2020; @Kinzer2020; @Orme2019].
 Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations.
@@
 Special thanks to Columbia Basin Research staff and the Columbia River Data Access in Real Time (DART) application, and Susannah Iltis in particular.
+Thank you to three anonymous reviewers that helped us improve the manuscript with their thoughtful comments and suggestions.
 And finally, thank you to Rick Orme and Brice Semmens for their contributions to the development of STADEM, as well as Eric Buhle, Sarah Hoffmann, John Hargrove and all the other people who have contributed through productive critique and conversations.
@@
 The size of each point is proportional to the window count that week.
-The dashed line is the 1-1 line."}
+Points falling below the 1-1 dashed line indicate weekly periods of higher nighttime passage as compared to re-ascension, conversly, points falling above indicate a lower nighttime passage than re-ascension rate."}
 

### 2021-03-30 analysis/paper/paper.Rmd
minor wording changes

@@
 output:
-    bookdown::pdf_document2:
-      fig_caption: yes
-      number_sections: FALSE
-      toc: no
-      geometry: margin=1in
-      latex_engine: "xelatex"
-      fig_width: 7
-      # fig_height: 6
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks2.lua
-      - --lua-filter=../templates/pagebreak.lua
-      includes:
-        in_header: ../templates/header_manuscript.tex
     bookdown::html_document2:
@@
       - --lua-filter=../templates/pagebreak.lua
+    bookdown::pdf_document2:
+      fig_caption: yes
+      number_sections: FALSE
+      toc: no
+      geometry: margin=1in
+      latex_engine: "xelatex"
+      fig_width: 7
+      # fig_height: 6
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks2.lua
+      - --lua-filter=../templates/pagebreak.lua
+      includes:
+        in_header: ../templates/header_manuscript.tex
 fontsize: 12pt
@@
                            levels = paste0(rep(c('Baseline', 'TrapDown', 'N-R', 'N-R_TrapDown'), each = 3), 
-                                           c('', '_Err_L', '_Err_H'))))
+                                           c('', '_Err_L', '_Err_H')))) %>%
+  mutate(across(sim,
+                as.integer))
 
@@
 Many fisheries management and conservation actions are made based on estimates of escapement by species and origin at LGR [@NMFS2019; @NWFSC2015; @NPCC2014].
-As an example, tribal ceremonial and subsistence harvest rates in the region are informed by escapement at LGR; additionally, the openings, closures and bag limits of fisheries targeting hatchery fish, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia rivers, are informed by escapements at LGR[@UnitedStatesOregon].
+As an example, tribal ceremonial and subsistence harvest rates in the region are informed by escapement at LGR; additionally, the openings, closures and bag limits of fisheries targeting hatchery fish, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia rivers, are informed by escapements at LGR [@USOregon2018].
 
@@
 @Hatch1994 conducted a study of the counting system at LGR, and found that sp/sum Chinook Salmon were undercounted at the window by in-person window counters when compared to video counts.
-There was also a significant difference between daily standard counts (16 or 10 hours per day) and daily 24-hour video counts; the undercounting was contributed to a significant nighttime passage rate (crossing while the counting window was closed) with an estimated 3.5% of sp/sum Chinook Salmon and 6.6% of steelhead adults passing during "nighttime" hours.
+There was also a significant difference between daily standard counts (16 or 10 hours per day) and daily 24-hour video counts; the undercounting was attributed to a significant nighttime passage rate (crossing while the counting window was closed) with an estimated 3.5% of sp/sum Chinook Salmon and 6.6% of steelhead adults passing during "nighttime" hours.
 They also found some species misidentification issues, which could result in either under- or over-counting of a particular species.
@@
 Whereas unaccounted nighttime passage can lead to an underestimate in escapement, both fallback without and with re-ascension instead potentially lead to an overestimate of escapement [@Dauble2000].
-Previously, it has often been assumed that these two biological processes essentially canceled each other out, resulting in window counts providing an unbiased estimate of escapement [@Camacho2018], which may be true in many or most years, but perhaps not always.
+Previously, it has often been assumed that these two biological processes essentially cancel each other out, resulting in window counts that provide an unbiased estimate of escapement [@Camacho2018], which may be true in many or most years, but perhaps not always.
 
@@
 We also test whether observed nighttime passage and fallback/re-ascension rates are typically equal.
-Our method for estimating species-specific escapement past LGR incorporates window counts, data from the adult fish trap located on the ladder, and passive detection of adults in the ladder from fish previously tagged with passive integrated transponder (PIT) tags to explicitly model nighttime passage, re-ascension, and observation error using a state-space approach [@Royle2008] which separates process variance (e.g., week to week variance in true escapement) from observational variance (e.g., observation error at the window, or sampling variance at the trap).
+Our method for estimating species-specific escapement past LGR incorporates window counts, data from the adult fish trap located on the ladder, and passive detections of adults in the ladder from fish previously tagged with passive integrated transponder (PIT) tags to explicitly model nighttime passage, re-ascension, and observation error using a state-space approach [@Royle2008] which separates process variance (e.g., week to week variance in true escapement) from observational variance (e.g., observation error at the window, or sampling variance at the trap).
 To meet desired management and conservation objectives, modeled escapement includes estimates of uncertainty and is parsed into weekly strata.
-Further, total and weekly estimates are parsed into three origin groups: wild fish, hatchery fish with a clipped adipose fin, and unclipped hatchery fish.
+Further, total and weekly estimates are divided into three origin groups: wild fish, hatchery fish with a clipped adipose fin, and unclipped hatchery fish.
 Our model is implemented in the **ST**ate-space **A**dult **D**am **E**scapement **M**odel (STADEM) package for the statistical software R [@R-Core-Team2020], and is available for download or installation from [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM).
@@
 Therefore, we use a mark-recapture approach to estimate a "true" trapping rate, $\nu_t$.
-The estimate of the true weekly trap rate is derived based on previously PIT-tagged sp/sum Chinook Salmon and steelhead that are crossing LGR that week, and using a Lincoln-Peterson mark-recapture model [@Seber2002].
+The estimate of the true weekly trap rate is derived based on previously PIT-tagged sp/sum Chinook Salmon and steelhead that are crossing LGR that week, and uses a Lincoln-Peterson mark-recapture model [@Seber2002].
 
@@
 To assess the performance of the model, we evaluated the estimates of origin-specific escapement by several measures.
-Accuracy was captured by the relative bias: the difference between the simulated "truth" and the mean STADEM estimate for each scenario, divided by the simulated "truth".
+Accuracy was captured by the mean relative bias: the difference between the simulated "truth" and the STADEM estimate for each simulation, divided by the simulated "truth".
 We estimated the precision by examining the average coefficient of variation (CV) of the estimates.
 We calculated the root mean squared error (RMSE) as the square root of the mean of the squared bias in the estimate, which is equivalent to the square root of the variance of the estimator plus the square of the expected bias.
-Finally, we evaluated whether estimates of uncertainty were appropriate by calculating coverage probabilities by determining what proportion of the model results generated an estimated 95% credible interval that contained the simulated true value.
+Finally, we evaluated whether estimates of uncertainty were appropriate by calculating coverage probabilities to determine what proportion of the model results generated an estimated 95% credible interval that contained the simulated true value.
 To compare the STADEM results directly to window counts, we compared the model estimates of total unique fish and the simulated window counts to the simulated number of unique fish.
@@
 Estimates of wild escapement were unbiased, with an average relative bias of 0.2–0.3%, although the bias appeared equally distributed between positive and negative values across the simulations (Figure \@ref(fig:rel-bias-lgd)).
-The CV of the estimates averaged 2.0–3.0%, with higher CV's being associated with scenarios when the trap was simulated closed for 3 weeks.
+The CV of the estimates averaged 2.0–3.0%, with higher CV's being associated with scenarios with a simulated trap closure for 3 weeks.
 The coverage probabilities always exceeded 95% across all scenarios.
@@
 The differences in the estimated weekly rates were also apparent visually (Figure \@ref(fig:night-reasc-diff-fig)).
-In particular, there were several weeks when the window counts were quite large, but the nighttime passage and re-ascension rates differed by as much as 10%; leading to larger window count bias in some years \@ref(tab:lgr-tab).
+In particular, there were several weeks when the window counts were quite large, but the nighttime passage and re-ascension rates differed by as much as 10%; leading to larger window count bias in some years (Table \@ref(tab:lgr-tab)).
 
@@
 Regardless of whether observer error at the window is negligible, low or high, whether nighttime passage and re-ascension rates are equal or not, whether the trap is shut down for several weeks in a season or not, STADEM produced estimates of unique escapement that were unbiased, precise (CVs in the 2-3% range) and contained appropriate credible intervals.
-Although the average bias among runs within a given simulation scenario was often positive (Table \@ref(tab:summ-tab)), Figure \@ref(fig:rel-bias-lgd) shows that for each scenario that some simulations resulted in a negative bias, while some were positive.
+Although the average bias among runs within a given simulation scenario was often positive (Table \@ref(tab:summ-tab)), Figure \@ref(fig:rel-bias-lgd) shows that for each scenario some simulations resulted in a negative bias, while some had a positive one.
 We posit that conducting additional simulations would bring the average bias closer to zero, but run times constrained how many simulations we were able to conduct for each scenario.
@@
 The first is that by accounting for both night passage and re-ascension rates, that are sometimes unequal, it corrects any potential bias in window counts.
-That bias ranged from -9.9% to 11.0% across all species and year combinations we fit STADEM to empirical data from LGR (Table \@ref(tab:lgr-tab)).
+Across all combinations of years and species with empirical data from LGR that we fit STADEM to, that bias ranged from -9.9% to 11.0% (Table \@ref(tab:lgr-tab)).
 Whether that amount of potential bias is of concern to management agencies is a decision for them to make, but we have provided a tool to correct that bias if it is of concern.
@@
 Capturing and accounting for known sources of error is prudent to minimize management decision risk [@Harwood2003; @Ascough-Ii2008].
-At the same time, the precision among STADEMs estimates is still quite good, with CVs near 2-3%, well below the goal of 15% set forth by NOAA [@Crawford2011].
+At the same time, the precision among STADEM's estimates is still quite good, with CVs near 2-3%, well below the goal of 15% set forth by NOAA [@Crawford2011].
 

### 2021-03-31 analysis/paper/paper.Rmd
A few minor revisions, to the manuscript and the response to reviewers

@@
 Our results demonstrate that the model produced unbiased estimates across all tested scenarios.
-We also applied this model to empirical data from the Lower Granite Dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring/summer-run Chinook Salmon *Oncorhynchus tshawytscha* and steelhead *O.
-mykiss* from spawn years 2010-2019.
+We also applied this model to empirical data from the Lower Granite Dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring/summer-run Chinook Salmon, *Oncorhynchus tshawytscha*, and steelhead, *O.
+mykiss*, from spawn years 2010-2019.
 keywords: |
@@
 The inclusion of this genetic information typically results in a reduction in wild escapement estimates and an associated increase in hatchery no-clip escapement, as phenotypically wild fish are identified as the progeny of hatchery broodstock.
+Accepting a potential bias due to unclipped hatchery fish can create a greater risk during status assessments of these ESA listed wild populations.
 However, if in-season management decisions do not require this correction or could accept the potential bias, origin calls at the trap could be used in-season as a first approximation to escapement.

### 2021-05-06 analysis/paper/paper.Rmd
changed some latex formatting so equations are numbered. responded to a few very minor requested revisions from associate editor

@@
 
-$$
+\begin{equation}
   \begin{aligned}
@@
   \end{aligned}
-$$
+\end{equation}
 
@@
 
-$$
+\begin{equation}
   Y^T_t \sim \text{Bin} \left( \nu_t, X_t \right)
-$$
+\end{equation}
 
@@
 
-$$
+\begin{equation}
   m_t \sim \text{Bin} \left( \nu_t, M_t \right)
-$$
+\end{equation}
 
@@
 The observed number of wild, $Y^T_{w,t}$, hatchery clipped, $Y^T_{hc,t}$, and hatchery no-clip, $Y^T_{hnc,t}$, fish caught in the trap that week is assumed to come from a multinomial distribution with probability vector $\omega_t$.
-The log-odds ratio of the proportions in $\omega_t$, in relation to the proportion of clipped hatchery fish, $\omega_{hc,t}$ is modeled as a random walk, so it can change through time, while smoothing over large weekly fluctuations in the proportions due to small sample size.
+The log-odds ratio of the proportions in $\omega_t$, in relation to the proportion of clipped hatchery fish $\omega_{hc,t}$ ($\phi_{w,t}$ and $\phi_{hnc,t}$), is modeled as a random walk, so it can change through time, while smoothing over large weekly fluctuations in the proportions due to small sample size.
 This allows the proportions of wild, hatchery clipped and hatchery no-clip fish to shift throughout the season, based on the data available from the fish trap.
@@
   \end{aligned}
-$$
+\end{equation}
+
 In practice, the total number of fish crossing LGR each week, $X_t$, could be split into other discrete groups if desired; for example, individual wild stocks using genetic stock identification or hatchery stocks with parentage based tagging [@Steele2019].
+A graphical depiction (directed acyclic graph) of the model framework described above is found in Figure \@ref(fig:dag-fig).
 

### 2021-05-06 analysis/paper/paper.Rmd
acknowledged associate editor's comments
replaced "N" with "Normal" in all equations

@@
     \ln \left( X_{t} \right) &= \ln \left( X_{t-1} \right) + e_t \\
-    e_t &\sim \mathcal{N}(0, \sigma^2_X)
+    e_t &\sim \text{Normal}(0, \sigma^2_X)
   \end{aligned}
@@
     \text{logit} (\theta_{t}) &= \text{logit} \left( \theta_{t-1} \right) + g_t \\
-    g_t &\sim \mathcal{N}(0, \sigma^2_\theta) \\
+    g_t &\sim \text{Normal}(0, \sigma^2_\theta) \\
   \end{aligned}
@@
     \text{logit} \left( \eta_{t} \right) &= \text{logit} \left(\eta_{t-1} \right) + f_t \\
-    f_t &\sim \mathcal{N}(0, \sigma^2_\eta)
+    f_t &\sim \text{Normal}(0, \sigma^2_\eta)
   \end{aligned}
@@
     \phi_{hnc,t} &= \phi_{hnc, t-1} + d_{hnc,t} \\
-    d_{w,t} &\sim \mathcal{N}(0, \sigma^2_\omega) \\
-    d_{hnc,t} &\sim \mathcal{N}(0, \sigma^2_\omega) \\
+    d_{w,t} &\sim \text{Normal}(0, \sigma^2_\omega) \\
+    d_{hnc,t} &\sim \text{Normal}(0, \sigma^2_\omega) \\
   \end{aligned}
@@
 Special thanks to Columbia Basin Research staff and the Columbia River Data Access in Real Time (DART) application, and Susannah Iltis in particular.
-Thank you to three anonymous reviewers that helped us improve the manuscript with their thoughtful comments and suggestions.
+Thank you to three anonymous reviewers and an associate journal editor that helped us improve the manuscript with their thoughtful comments and suggestions.
 And finally, thank you to Rick Orme and Brice Semmens for their contributions to the development of STADEM, as well as Eric Buhle, Sarah Hoffmann, John Hargrove and all the other people who have contributed through productive critique and conversations.

### 2021-05-10 analysis/paper/paper.Rmd
reknit word and html versions, and pdf in order to keep the .tex file

@@
       # fig_height: 6
+      keep_tex: true
       pandoc_args:

### 2021-05-10 analysis/paper/paper.tex
reknit word and html versions, and pdf in order to keep the .tex file

@@
-\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
 \PassOptionsToPackage{hyphens}{url}
@@
   \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provides euro and other symbols
-\else % if luatex or xelatex
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
   \usepackage{unicode-math}
@@
 \fi
-% use upquote if available, for straight quotes in verbatim environments
+% Use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
   pdftitle={State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources},
-  pdfauthor={Kevin E. See1,, Ryan N. Kinzer2, and Mike W. Ackerman1},
-  pdfkeywords={escapement; Chinook; steelhead; Lower Granite Dam; state-space model; Bayesian model},
-  pdfborder={0 0 0},
-  breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
+  pdfauthor={Kevin E. See1,, Ryan N. Kinzer2, and Michael W. Ackerman1},
+  pdfkeywords={escapement; Chinook; steelhead; salmon; Lower Granite Dam; state-space model; Bayesian model},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
 \usepackage[margin=1in]{geometry}
 \usepackage{longtable,booktabs}
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
 % Allow footnotes in longtable head/foot
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-2}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-% set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 % double spacing
@@
 \title{State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources}
-\author{Kevin E. See\textsuperscript{1,*}, Ryan N. Kinzer\textsuperscript{2}, and Mike W. Ackerman\textsuperscript{1}}
-\date{August 18, 2020}
+\author{Kevin E. See\textsuperscript{1,*}, Ryan N. Kinzer\textsuperscript{2}, and Michael W. Ackerman\textsuperscript{1}}
+\date{May 10, 2021}
 
@@
 \begin{abstract}
-This is the abstract
+Accurate estimates of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successful management of fisheries in the Snake River basin.
+Here we describe a state-space model that estimates such escapement past a dam by using window counts, passive integrated transponder (PIT) tag observations and data from an adult fish trap, accounting for issues such as nighttime passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate.
+We tested the approach using a simulation framework that mimicked several levels of observation error, differences between nighttime passage and re-ascension rates and the possibility of the adult trap being closed for some period of time.
+Our results demonstrate that the model produced unbiased estimates across all tested scenarios.
+We also applied this model to empirical data from the Lower Granite Dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring/summer-run Chinook Salmon, \emph{Oncorhynchus tshawytscha}, and steelhead, \emph{O.
+mykiss}, from spawn years 2010-2019.
 \end{abstract}
 
-\textsuperscript{1} Biomark, Inc.~705 South 8th St., Boise, Idaho, 83702, USA\\
+\textsuperscript{1} Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA\\
 \textsuperscript{2} Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
@@
 \textsuperscript{*} Correspondence: \href{mailto:Kevin.See@merck.com}{Kevin E. See \textless{}\href{mailto:Kevin.See@merck.com}{\nolinkurl{Kevin.See@merck.com}}\textgreater{}}
-
-\textbf{Possible journals:}
-
-First Choice:
-
-\begin{itemize}
-\tightlist
-\item
-  \textbf{\href{https://afspubs.onlinelibrary.wiley.com/journal/15488675}{North American Journal of Fisheries Management}}
-\end{itemize}
-
-Second Choice:
-
-\begin{itemize}
-\tightlist
-\item
-  \href{https://afspubs.onlinelibrary.wiley.com/journal/15488659}{Transactions of the American Fisheries Society}
-\end{itemize}
-
-Others:
-
-\begin{itemize}
-\item
-  \href{http://www.nrcresearchpress.com/page/cjfas/editors}{Canadian Journal of Fisheries and Aquatic Sciences}
-\item
-  \href{https://www.journals.elsevier.com/fisheries-research}{Fisheries Research}
-\item
-  \href{http://fishbull.noaa.gov/}{Fishery Bulletin}
-\item
-  \href{http://onlinelibrary.wiley.com/journal/10.1111/(ISSN)1467-2979/homepage/ProductInformation.html}{Fish and Fisheries}
-\end{itemize}
 
@@
 
-Fish escapement is the number of adults that survive juvenile and adult rearing, escape harvest and return to their natal habitat to potentially spawn (e.g.~Bue et al.
-(1998)).
-For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their spawning area.
-Escapement estimates facilitate effective fisheries management, and particularly, estimates of escapement for each component group (e.g., stock, population, age, origin {[}wild, hatchery{]}; Hess et al.
-(2014); Steinhorst et al.
-(2017)) provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations.
-Accurate escapement estimates are increasingly important for depleted populations as they inform viable salmonid metrics (McElhany et al.
-2000), and facilitate assessments of population viability and extinction risk (Ford et al.
-2015; Williams et al.
-2016) and climate change vulerability (Crozier et al.
+Fish escapement refers to the number of adults that survive juvenile and subadult rearing, escape harvest, and achieve a size and age to potentially spawn (e.g., Bue et al.
+1998).
+For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their natal habitat to spawn (e.g., Woody 2004; Jasper et al.
+2018).
+Escapement estimates facilitate effective fisheries management, particularly estimates of escapement for component groups (e.g., by stock, population, age, origin {[}wild, hatchery{]}, Hess et al.
+2014; Steinhorst et al.
+2017; Camacho et al.
+2018) which provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations.
+Accurate escapement estimates are increasingly important for depleted populations as they inform status metrics (McElhany et al.
+2000) and facilitate assessments of population viability, extinction risk (Northwest Fisheries Science Center 2015; Williams et al.
+2016) and climate change vulnerability (Crozier et al.
 2019).
 
-Populations of Chinook Salmon Oncorhynchus tshawytscha and steelhead trout O.
-mykiss in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric dams on the Snake and Columbia rivers) and tributary habitats (Nehlsen et al.
+Populations of Chinook Salmon \emph{Oncorhynchus tshawytscha} and anadromous steelhead trout \emph{O.
+mykiss} in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric projects on the Snake and Columbia rivers) and tributary habitats (Nehlsen et al.
 1991; McClure et al.
-2003; Ford et al.
-2015).
-As a result, Snake River spring-summer run Chinook Salmon (hereafter Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937).
-Snake River Chinook Salmon and steelhead have substantial recreational, commercial, cultural and subsistence value both within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries.
-The aggregate escapement of Snake River Chinook Salmon and steelhead populations, with the exception of Tucannon River, is measured at Lower Granite Dam located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations.
-Many fisheries management and conservation actions are made based on estimates of escapement at Lower Granite Dam parsed by species and origin (\textbf{Snake Basin Harvest Plans?}).
-Additionally, harvest openings and closures, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia River fisheries, are predicated on escapements at Lower Granite Dam (\textbf{could cite US v OR Management Agreement and Snake Basin Harvest Plans?}).
-
-Chinook Salmon and steelhead returning to a majority of populations in the Snake River must ascend a fish ladder on Lower Granite Dam before migrating to their natal tributary spawning locations.
-Fish management agencies previously used counts of fish, by species, passing an observation window located on the side of the fish ladder as a census of fish escaping past the dam.
-For both species, total escapement was then parsed into groups (e.g., wild, hatchery) using observed marks and genetic data from a sample of fish captured at an adult trap located on the fish ladder upstream of the observation window (Camacho et al.
-2017 p.
-@Steinhorst2017).
-Window counts as a census proved beneficial as being an easy, straight-forward method that was ascertained in near real-time.
-Moreover, downriver fisheries management arenas have used window counts at lower-river dams as escapement estimates for the past several years, and consistency in methods is often desirable for management decision making (\textbf{citation?}).
-
-Using window counts, however, as a census of Chinook Salmon and steelhead passing Lower Granite Dam can be problematic as it avoids multiple sources of uncertainty and ignores known biological processes.
-First, live (i.e., in-person) window counts only occur from April through October each year.
-Additionally, live window counts only occur for 16-hours each day, and fish counters working at the observation window look directly into the fish ladders to identify and count all passing fish, by species, for only 50 minutes of each hour.
-Counts are then expanded to provide an estimate for the hour (USACE 2015).
-From November through March, the remainder of the year, video tape fish counting is used and only occurs for 10 hours each day; fish counters then read the video tapes and submit daily fish counts (Hatch et al.
-1994).
-Typically, the observational error rates of live and video window counts are unknown, and sampling error rates are ignored.
-Additionally, two biological processes are unaccounted for: 1) fish that cross the dam during the 8-hours when the window is closed for counting (i.e., nighttime passage) which may result in an underestimate of escapement, and 2) fish that cross the dam multiple times (i.e., re-ascension), and therefore potentially double-counted.
-Finally, Chinook Salmon and steelhead that migrate through the ladder and past the dam may fallback over the dam (e.g., over spillway, through navigation locks), and later, may or may not re-ascend the fish ladder again (Boggs et al.
-2004).
-Both fallback with no re-ascension and fallback with re-ascension potentially result in an overestimate of escapement (Dauble and Mueller 2000).
-Previously, it was assumed that nighttime passage rates and fallback/re-ascension rates cancelled each other out resulting in window counts providing an unbiased estimate of escapement (\textbf{citation?})
-
-In this study, we describe a novel method to estimate aggregate and group escapement which incorporates all sources of known uncertainty, we demonstrate the estimation method is essentially unbiased, thus improving effective conservation and management decision making, and show night-time passage and fallback/re-ascension rates are unequal.
-Our method for estimating escapement, by species, past Lower Granite Dam incorporates window counts, data from the adult fish trap, and observations of PIT tagged fish in the adult ladder to explicitly model nighttime passage, re-ascension, and error from both window and trap estimates using a state-space approach (Royle and Dorazio 2008).
+2003; Northwest Fisheries Science Center 2015).
+As a result, most Snake River spring/summer-run Chinook Salmon (hereafter sp/sum Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937).
+Snake River sp/sum Chinook Salmon and steelhead have substantial cultural, recreational, commercial and subsistence value within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries.
+The aggregate escapement of Snake River sp/sum Chinook Salmon and steelhead populations, with the exception of Tucannon River (a tributary of the Snake River whose confluence is downstream of Lower Granite Dam {[}LGR{]}), is monitored at LGR located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations.
+Many fisheries management and conservation actions are made based on estimates of escapement by species and origin at LGR (NPCC 2014; Northwest Fisheries Science Center 2015; National Marine Fisheries Service 2019).
+As an example, tribal ceremonial and subsistence harvest rates in the region are informed by escapement at LGR; additionally, the openings, closures and bag limits of fisheries targeting hatchery fish, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia rivers, are informed by escapements at LGR (U.S.
+v Oregon 2018).
+
+The majority of sp/sum Chinook Salmon and steelhead returning to the Snake River must ascend a fish ladder on LGR before migrating upriver.
+Fish management agencies previously used counts of fish, by species, passing an observation window within the fish ladder as a census of fish escaping past the dam.
+For both species, total escapement was then parsed into groups (e.g., wild, hatchery) using observed marks and genetic data from a sample of fish captured at an adult trap located on the fish ladder upstream of the observation window (Steinhorst et al.
+2017; Camacho et al.
+2018; Steele et al.
+2019).
+Treating window counts as a census proved beneficial as being an easy, straight-forward method ascertained in near real-time.
+Moreover, downriver fisheries management arenas have used window counts at lower-river dams as escapement estimates for the past several years, and consistency in methods is often desirable for management decision making (Joint Columbia River Management Staff 2019).
+
+However, using window counts as a census of fish passing LGR can be problematic as it fails to account for multiple sources of uncertainty and disregards known biological processes.
+Typically, the observational error rates of sp/sum Chinook Salmon and steelhead counted passing a window on the fish ladder, using both in-person and video methods, are unknown, and sampling error rates are ignored.
+Hatch et al.
+(1994) conducted a study of the counting system at LGR, and found that sp/sum Chinook Salmon were undercounted at the window by in-person window counters when compared to video counts.
+There was also a significant difference between daily standard counts (16 or 10 hours per day) and daily 24-hour video counts; the undercounting was attributed to a significant nighttime passage rate (crossing while the counting window was closed) with an estimated 3.5\% of sp/sum Chinook Salmon and 6.6\% of steelhead adults passing during ``nighttime'' hours.
+They also found some species misidentification issues, which could result in either under- or over-counting of a particular species.
+While that study identified potential issues with the current window counting procedure, it was only conducted in a single year, nearly 30 years ago, and at a time when sp/sum Chinook Salmon and steelhead total returns were typically lower than in the past decade.
+
+One additional biological process, the fallback and re-ascension of adult migrants at the dam, should also be accounted for to estimate the true escapement of these species at LGR.
+Adult fish that migrate through the ladder and past the dam may fallback over the dam (e.g., over a spillway, through navigation locks) and later may re-ascend the fish ladder again (Boggs et al.
+2004) and be counted multiple times at the window (re-ascension).
+Whereas unaccounted nighttime passage can lead to an underestimate in escapement, both fallback without and with re-ascension instead potentially lead to an overestimate of escapement (Dauble and Mueller 2000).
+Previously, it has often been assumed that these two biological processes essentially cancel each other out, resulting in window counts that provide an unbiased estimate of escapement (Camacho et al.
+2018), which may be true in many or most years, but perhaps not always.
+
+In this study, we describe a novel method to estimate aggregate and group escapements which incorporate all sources of known uncertainty and demonstrate that the estimation method is essentially unbiased, thus better informing conservation and management decision making.
+We also test whether observed nighttime passage and fallback/re-ascension rates are typically equal.
+Our method for estimating species-specific escapement past LGR incorporates window counts, data from the adult fish trap located on the ladder, and passive detections of adults in the ladder from fish previously tagged with passive integrated transponder (PIT) tags to explicitly model nighttime passage, re-ascension, and observation error using a state-space approach (Royle and Dorazio 2008) which separates process variance (e.g., week to week variance in true escapement) from observational variance (e.g., observation error at the window, or sampling variance at the trap).
 To meet desired management and conservation objectives, modeled escapement includes estimates of uncertainty and is parsed into weekly strata.
-Further, total and weekly estimates are parsed into three origin groups: wild, hatchery, and hatchery no-clip.
-Estimates of escapement account for fish that migrate through the ladder at night outside of observation hours and account for fish that may ascend the ladder multiple times due to fallback and re-ascension.
-Our model is implemented in the \textbf{ST}ate-space \textbf{A}dult \textbf{D}am \textbf{E}scapement \textbf{M}odel (STADEM) package for the statistical software R (R Core Team 2020), and is available for downloaded at \url{https://github.com/KevinSee/STADEM}.
-To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates.
-The STADEM model combines multiple imperfect sources of data to reduce bias in escapement estimates and provide more reasonable estimates of uncertainty.
-Additionally, the STADEM model framework can be applied elsewhere to estimate aggregate and group escapement, provided a passage barrier containing multiple counting and observational mechanisms exist, such as, a counting window, an adult trap sampling a portion of the run, and PIT tag antenna infrastructure.
+Further, total and weekly estimates are divided into three origin groups: wild fish, hatchery fish with a clipped adipose fin, and unclipped hatchery fish.
+Our model is implemented in the \textbf{ST}ate-space \textbf{A}dult \textbf{D}am \textbf{E}scapement \textbf{M}odel (STADEM) package for the statistical software R (R Core Team 2020), and is available for download or installation from \url{https://github.com/KevinSee/STADEM}.
+To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates and compared STADEM results to the simulated ``truth''.
+Then, as a case study, we applied this model to sp/sum Chinook Salmon and steelhead returns at LGR for spawn years 2010 - 2019.
 
@@
 
-We used STADEM and three sources of data to estimate Chinook Salmon and steelhead escapement at Lower Granite Dam from 2010-2019.
-Data sources included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of previously PIT tagged fish detected in the adult fish ladder.
-Following, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; however, similar data could likely be obtained from other fish passage facilities.
+Data sources for STADEM included 1) counts of fish migrating past the observation window located on the adult fish ladder at LGR, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of fish PIT tagged prior to arriving at LGR and detected at arrays in the adult fish ladder.
+Below, we describe each of the data sources in more detail as they pertain to LGR; similar data could likely be obtained from other species passing LGR (e.g., fall Chinook Salmon) or other fish passage facilities (e.g., Bonneville Dam).
 
@@
 
-Daily counts of adult Chinook Salmon and steelhead passing an observation window located on the Lower Granite fish ladder were the first source of data.
-Daily counts were made and provided by the US Army Corps of Engineers, and when summed, provide an estimate of the number of fish ascending and passing (i.e., escapement) Lower Granite Dam each season.
-Window counts were made for each species using video monitoring and direct in-person visual monitoring during daytime hours (Hatch et al.
+Daily counts of adult sp/sum Chinook Salmon and steelhead passing the observation window located on the LGR fish ladder were compiled and provided by the US Army Corps of Engineers.
+When summed, they provide an estimate of the number of fish ascending and passing LGR each season.
+Window counts were made for each species using direct in-person visual monitoring and video monitoring during daytime hours (Hatch et al.
 1994).
-Video monitoring occured during the beginning and tail ends of the adult runs (March 1 -- March 31 and November 1 -- December) for 10 hours per day (0600 -- 1600 hours).
-Direct visual monitoring occured during peak run times (April 1 -- October 31) for 16 hours per day (0400 -- 2000 hours) (USACE 2015).
+Direct visual monitoring occurred during peak run times (April 1 -- October 31) for 16 hours per day (0400 -- 2000 hours).
+Video monitoring occurred during the beginning and tail ends of the adult runs (March 1 -- March 31 and November 1 -- December 31) for 10 hours per day (0600 -- 1600 hours) (USACE 2015).
 During direct visual monitoring, observers recorded each adult (≥ 30cm), by species, passing the window for 50 minutes of each hour of operation.
-Salmonids under 30cm in length were not identified to species.
-The sum of the daily 50-minute counts were then multiplied by 1.2 to account for the 10 minutes when fish were not counted.
-Daytime window counts were not expanded for fish that may have ascended the ladder outside of operational hours (i.e., nighttime) (USACE 2015).
+The sum of the daily 50-minute counts were then multiplied by 1.2 to account for the 10 minutes each hour when fish were not counted.
+Daytime window counts were not expanded for fish that may have ascended the ladder outside of operational hours (i.e., nighttime passage) (USACE 2015).
 Window counts were accessed through the Columbia Basin Research Data Access in Real Time (DART) website, \url{www.cbr.washington.edu/dart/query/adult_daily}, using their window count query.
-Counts were provided for each day the fish ladder was open to passage.
-Although window counts were assumed to be a census of every fish passing Lower Granite Dam, corrections were not applied for nighttime passage or re-ascending fish.
-Further, there was no estimation of daily or seasonal observation or sampling error.
+Counts were provided for each day the fish ladder was open to passage, and had already been expanded by 1.2.
 
@@
 
-The second source of data came from a sample of fish collected in the adult trap as they migrated past Lower Granite Dam (Ogden 2016a).
-The trap, also located within the adult fish ladder and upstream of the observation window, provided biological data (e.g., origin {[}wild, hatchery{]}, genetic stock, length, age, sex) for captured adults that allowed decomposition of the escapement into specific groups (e.g., Camacho et al.
-(2019), Steinhorst et al.
-(2017)).
-The trap was operational for 24 hours per day, and randomly sampled the daily run by opening four times per hour for a length of time determined by a set daily trapping rate.
-The trap rate was determined by a committee of collaborating management agencies with a goal of capturing a target number of wild fish, but also balancing fish handling concerns.
-Trap sample rates were typically 10-25\%, but fluctuated throughout the season due to high water temperatures, decreased flows, trap malfunctions and/or closures, fish handling logistics, etc.
-All captured fish were anesthetized, identified to species, examined for existing marks/tags, and measured for fork length.
-Additionally, each fish had scale and genetic tissue samples taken; scale samples were used to estimate age (Wright et al.
-2015), tissue samples were used to determine sex and for genetic stock identification analysis (e.g., Hargrove et al.
-(2019)).
+The second source of data are from a sample of fish collected in the adult trap as they migrated past LGR (Ogden 2016a).
+The trap, also located within the adult fish ladder and upstream of the observation window, provides an opportunity to collect biological data (e.g., phenotypic origin {[}wild, hatchery{]}, tissue sample for genetic data, length, scale sample for age, sex) from captured adults which allows for decomposition of the escapement into specific groups (e.g., Camacho et al.
+2018; Steinhorst et al.
+2017).
+The trap was operational for 24 hours per day and randomly sampled the run by opening four times per hour for a length of time determined by a set daily trapping rate.
+The trap rate was determined by a committee of collaborating management agencies with a goal of capturing a target number of wild fish while also balancing fish handling concerns.
+Trap sample rates were typically 10-25\%, but fluctuated throughout the season due to, for example, high water temperatures, decreased flows, trap malfunctions and/or closures, fish handling logistics, in-season forecast adjustments, or adjustments due to shifting species composition throughout the year.
+
+All captured fish were anesthetized, identified to species, examined for existing marks or tags, measured for fork length, and visually identified as wild or hatchery.
+The most widely used marking of hatchery fish is a clipped (removed) adipose fin, although coded wire tags are also used in less than 10\% of the hatchery releases.
+Some subset of hatchery fish are either intentionally or unintentionally released without a clipped adipose fin, and these are referred to as unclipped hatchery fish, or hatchery no-clips.
+For adults with intact adipose fins (unclipped), which includes wild and hatchery no-clip individuals, either a portion or all of the fish trapped (depending on the year) had scale and genetic tissue samples taken.
+Scale samples were used to estimate age (Wright et al.
+2015) and genetic tissue samples were used to determine sex (Campbell et al.
+2012) and estimate the location of origin of wild fish using genetic stock identification (e.g., Hargrove et al.
+2019).
+Prior to 2013, only fish determined to be wild in origin at the trap were sampled for scale and genetics.
+Starting in 2013, every unclipped sp/sum Chinook Salmon and steelhead trapped at LGR was genotyped to simplify collaborative logistics and to better estimate the true proportion of unclipped hatchery fish that can otherwise appear phenotypically wild.
+Camacho et al.
+(2018) provide further details on trap sample rates and valid sample selection.
 Prior to release, all non-PIT tagged fish with an intact adipose fin (i.e., putatively wild) received a PIT tag.
-Final determination of wild, hatchery, and hatchery no-clip origins were assigned using a post-hoc analysis of marks, tags, and genetic information.
+Final determination of wild, clipped hatchery, or unclipped hatchery origins were assigned using a post-hoc analysis of marks and tags, including parentage-based tagging based on the genetic samples taken (Steele et al.
+2013, 2019).
 Data from the adult trap were collected and managed by multiple agencies and were made available by the Idaho Department of Fish and Game (Camacho et al.
-2019).
+2018).
 
 \hypertarget{pit-tag-data}{%
-\subsubsection{PIT tag Data}\label{pit-tag-data}}
-
-The last source of data was observations of PIT tagged adult Chinook Salmon and steelhead at detection sites located in the Lower Granite Dam fish ladder.
-These observations provide estimates of 1) a trapping rate, 2) the proportion of fish passing during nighttime hours (i.e., outside of window observation hours), and 3) the proportion of fish that ascend the fish ladder multiple times (i.e., re-ascension rate).
-Detections used in the model include all fish that were previously PIT tagged as juveniles or adults prior to reaching Lower Granite Dam (i.e., does not include newly tagged adults at the dam) and detected at adult detection sites in the dam passage system.
-PIT tag data was provided through DART and the adult ladder PIT tag query; \url{http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail}.
-
-A trap rate estimate was derived using mark-recapture methods and PIT tag observations of both Chinook Salmon and steelhead at Lower Granite Dam adult detection sites.
-The ``mark'' group included all tags detected in the adult trap and the ``capture'' group included tags observed to cross the weir at the upstream end of the fish ladder as adults left the passage system.
-Using a mark-recapture model with differing capture probabilities, we estimated the trap rate on a weekly basis.
-Those estimates, with associated uncertainties, were then incorporated into the model as informed priors, while the model estimated the true trap rate based on all the data, including trap and window counts.
-
-The nighttime passage rate was estimated as the count of tags that migrated through the fish ladder during non-window observation hours divided by the total number of tags passing the fish ladder, and was calculated on a weekly basis.
-The re-ascension rate was the count of tags observed passing the upstream most detection sites in the adult fish ladder (i.e., passing the dam) and later detected re-entering the downstream end of the fish ladder at a later time divided by the total number of tags leaving the fish ladder.
-Previously, we examined for differences in nighttime passage and re-ascension rates as estimated using wild fish only, versus combining hatchery and wild fish together, and found no difference.
-Therefore, we combine wild and hatchery PIT tagged fish observations to estimate common nighttime passage and re-ascension rates to increase sample sizes.
+\subsubsection{PIT Tag Data}\label{pit-tag-data}}
+
+The last source of data were observations of PIT tagged adult sp/sum Chinook Salmon and steelhead at PIT tag detection sites located in the LGR fish ladder.
+These observations provide data related to 1) an estimate of the ``true'' trap rate for the adult fish trap, 2) a nighttime passage rate, and 3) a re-ascension rate.
+Detections used in the model included all fish that were previously PIT tagged as juveniles or adults prior to arriving at LGR (i.e., does not include newly tagged adults at the dam) and detected at adult detection sites in the dam passage system.
+PIT tag data were provided through DART and the adult ladder PIT tag query; \url{http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail}.
 
@@
 
-We estimated the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for nighttime passage, and fallback/re-ascension rates using a state-space modeling approach (Royle and Dorazio 2008) implemented in the STADEM package for the R statistical software (R Core Team 2020).
-We assumed that the window counts and the estimates from the trap (fish in the trap divided by trap rate that week) were generated by processes with observation error.
-In the case of the trap for example, we assumed there was sampling variation and uncertainty around our estimates of the true unknown trap rate.
-STADEM further accounted for the proportion of fish that ascended the ladder while the counting window was closed (i.e., night), as well as for fish potentially double-counted (or more) after falling back below the dam and later re-ascending the fish ladder.
-Last, adult sampling data from the trap (wild, hatchery, hatchery no-clip) were used to partition the total escapement estimate by origin (Figure \ref{fig:stadem-examp-fig}.
-Additional model details can be found in \protect\hyperlink{append1}{Appendix A}.
-The STADEM package is available from the primary author at a \url{https://github.com/KevinSee/STADEM}, and requires the use of the JAGS software (Plummer 2019) for Bayesian inference.
+\hypertarget{total-and-weekly-escapement}{%
+\subsubsection{Total and Weekly Escapement}\label{total-and-weekly-escapement}}
+
+Escapement at LGR is estimated by combining two independent observations, trap catches and window counts, of the true number of fish crossing the dam in a state-space model (Royle and Dorazio 2008), with a weekly time-step.
+Both are assumed to be corrupted observations of the true unknown number of fish crossing LGR each week.
+We chose to use a weekly time-step to ensure sufficient sample sizes of PIT tag detections within each time strata, but this choice could be easily modified.
+The log of the true number of fish crossing during time-step \(t\), (\(X_t\)), is modeled as a random walk process (Shumway and Stoffer 2010), with \(e_t\) representing the process errors drawn from a normal distribution with variance \(\sigma^2_X\).
+
+\begin{equation}
+  \begin{aligned}
+    \ln \left( X_{t} \right) &= \ln \left( X_{t-1} \right) + e_t \\
+    e_t &\sim \text{Normal}(0, \sigma^2_X)
+  \end{aligned}
+\end{equation}
+
+The number of fish caught in the trap, \(Y^T_t\), for week \(t\) is modeled as a binomial process based on the unknown true trap rate that week, \(\nu_t\), and the unknown true number of fish crossing the dam that week, \(X_t\).
+
+\begin{equation}
+  Y^T_t \sim \text{Bin} \left( \nu_t, X_t \right)
+\end{equation}
+
+The set trapping rate (i.e., the pre-determined time that the trap is open each hour to trap adults) does not always reflect the true trapping rate or proportion of fish actually captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error (pulses or lack thereof of passing adults), among others.
+Therefore, we use a mark-recapture approach to estimate a ``true'' trapping rate, \(\nu_t\).
+The estimate of the true weekly trap rate is derived based on previously PIT-tagged sp/sum Chinook Salmon and steelhead that are crossing LGR that week, and uses a Lincoln-Peterson mark-recapture model (Seber 2002).
+
+\begin{equation}
+  m_t \sim \text{Bin} \left( \nu_t, M_t \right)
+\end{equation}
+
+The fish, from both species, caught in the trap that week are considered the ``mark'' group (\(m_t\)), and all the previously PIT tagged fish who are detected at the upper end of the LGR fish ladder that week are considered the second ``capture'' group, \(M_t\) (which includes recaptures of the ``marked'' fish).
+The proportion of previously marked fish that are caught in the adult trap, \(m/M\) can be modeled with a binomial distribution using the same trap rate parameter, \(\nu_t\).
+Although the group of previously PIT tagged fish is not assumed to be representative of the overall run, the rate at which they are caught in the trap should be the same rate that the overall run experiences.
+The more tagged fish crossing the dam in a particular week, the more certain we can be of the true trap rate.
+
+The number of fish counted at the window, \(Y^W_t\), is modeled as a (potentially) over-dispersed negative binomial process, with an expected value of \(X_t^{day}\), the number of fish crossing the dam while the window is open.
+This is simply the total number of fish crossing that week, \(X_t\), multiplied by the proportion of fish crossing while the window is open for counting, \(\theta_t\).
+In the formula below, \(p_t\) is the probability parameter and \(r\) is the size parameter of a negative binomial distribution (as defined in Plummer (2019)).
+If \(r\) is estimated to be small it provides evidence for over-dispersion, and as it grows very large, the negative binomial distribution behaves like a Poisson distribution.
+
+\begin{equation}
+  \begin{aligned}
+    X_t^{day} &= X_t * \theta_t \\
+    p_t &= \frac{r}{\left( r + X_t^{day} \right)} \\  
+    Y^W_t &\sim \text{NegBin} \left( p_t, r \right) \\
+  \end{aligned}
+\end{equation}
+
+Thus, the unknown true number of fish crossing LGR each week, \(X_t\), is estimated from two different data sources: window counts and fish sampled in the trap.
+The window counts provide an estimate (with some potential observer error) of the fish crossing during daytime hours, while the fish in the trap, when expanded by the estimated true trap rate, provide an estimate of the total fish crossing that week.
+The use of both estimates allows for an estimation of observer error in the window counts which has previously been inestimable.
+For weeks when the trap is down, STADEM relies exclusively on the window counts and nighttime passage data, but there will be more uncertainty in the estimates.
+
+\hypertarget{daytime-passage-and-re-ascension-rates}{%
+\subsubsection{Daytime Passage and Re-ascension Rates}\label{daytime-passage-and-re-ascension-rates}}
+
+There are two processes that must be accounted for to correct any bias in the window counts: first, the proportion of fish that cross the dam while the window is closed for counting (nighttime passage rate), and second, the proportion of fish that cross the dam multiple times (re-ascension rate) and therefore potentially counted multiple times.
+Both rates can be estimated from previously PIT tagged fish that are crossing the dam each week.
+
+The nighttime passage rate is just the complement of the rate of fish passing during the day when the window is operating.
+The daytime passage rate for week \(t\), \(\theta_t\), is modeled as a random walk process (with deviations, \(g_t\), drawn from a normal distribution with variance \(\sigma^2_{\theta}\)) and estimated from a binomial distribution based on the number of PIT tags observed to cross the dam during operational hours, \(y^{day}_t\), and the total number of PIT tags observed to cross the dam at any point that week, \(N_t\) (Shumway and Stoffer 2010).
+
+\begin{equation}
+  \begin{aligned}
+    y^{day}_t &\sim \text{Bin} \left(\theta_t, N_t \right) \\
+    \text{logit} (\theta_{t}) &= \text{logit} \left( \theta_{t-1} \right) + g_t \\
+    g_t &\sim \text{Normal}(0, \sigma^2_\theta) \\
+  \end{aligned}
+\end{equation}
+
+The number of total fish crossing LGR differs from the number of unique fish crossing LGR because some fish fallback and re-ascend the dam.
+These fish are potentially double-counted (or more) at the window, and have the potential to be caught in the fish trap more than once.
+The number of tags known to be re-ascending the dam each week, \(y^{reasc}_t\), is modeled as a binomial process based on the estimated re-ascension rate, \(\eta_t\), and the total number of tags crossing the dam that week, \(N_t\).
+The logit of the re-ascension rate is modeled as a random walk process similar to daytime passage (Shumway and Stoffer 2010), with deviations, \(f_t\), drawn from a normal distribution with variance \(\sigma^2_\eta\).
+
+\begin{equation}
+  \begin{aligned}
+    y^{reasc}_t &\sim \text{Bin} \left(\eta_t, N_t \right) \\
+    \text{logit} \left( \eta_{t} \right) &= \text{logit} \left(\eta_{t-1} \right) + f_t \\
+    f_t &\sim \text{Normal}(0, \sigma^2_\eta)
+  \end{aligned}
+\end{equation}
+
+We looked for differences in weekly nighttime passage and re-ascension rates estimated using wild fish only, versus combining hatchery and wild fish together, and found no difference.
+Additionally, sample sizes of observed tags during most weeks are small, and may prohibit reliable estimation of rates for individual wild or hatchery stocks.
+Therefore, we combine wild and hatchery PIT tagged fish observations to estimate common nighttime passage and re-ascension rates to increase sample sizes.
+
+\hypertarget{origin-proportions}{%
+\subsubsection{Origin Proportions}\label{origin-proportions}}
+
+After estimating the total number of fish to have crossed LGR each week, \(X_t\), the total must be further refined into the number of wild fish, \(X_{w,t}\), hatchery clipped fish, \(X_{hc,t}\) and hatchery no-clip fish, \(X_{hnc, t}\).
+This is done by estimating a weekly origin proportion vector, \(\omega_t\) based on the random sample of fish caught in trap that week, \(Y^T_t\).
+The observed number of wild, \(Y^T_{w,t}\), hatchery clipped, \(Y^T_{hc,t}\), and hatchery no-clip, \(Y^T_{hnc,t}\), fish caught in the trap that week is assumed to come from a multinomial distribution with probability vector \(\omega_t\).
+The log-odds ratio of the proportions in \(\omega_t\), in relation to the proportion of clipped hatchery fish \(\omega_{hc,t}\) (\(\phi_{w,t}\) and \(\phi_{hnc,t}\)), is modeled as a random walk, so it can change through time, while smoothing over large weekly fluctuations in the proportions due to small sample size.
+This allows the proportions of wild, hatchery clipped and hatchery no-clip fish to shift throughout the season, based on the data available from the fish trap.
+The random walk deviations, \(d_{w,t}\) and \(d_{hnc,t}\), are drawn from the same normal distribution with variance \(\sigma^2_\omega\).
+
+\begin{equation}
+  \begin{aligned}
+    \left( Y^T_{w,t}, Y^T_{hc,t}, Y^T_{hnc,t} \right) &\sim \text{Multinom} \left( \omega_t, Y^T_t \right) \\
+    \omega_t &= \frac{\exp(\phi_t)}{\sum{\exp(\phi_t})} \\
+    \phi_{hc,t} &= 0 \\
+    \phi_{w,t} &= \ln \left( \frac{\omega_{w,t}}{\omega_{hc,t}} \right) \\
+    \phi_{hnc,t} &= \ln \left( \frac{\omega_{hnc,t}}{\omega_{hc,t}} \right) \\
+    \phi_{w,t} &= \phi_{w, t-1} + d_{w,t} \\
+    \phi_{hnc,t} &= \phi_{hnc, t-1} + d_{hnc,t} \\
+    d_{w,t} &\sim \text{Normal}(0, \sigma^2_\omega) \\
+    d_{hnc,t} &\sim \text{Normal}(0, \sigma^2_\omega) \\
+  \end{aligned}
+\end{equation}
+
+Finally, the number of unique fish by origin crossing LGR each week (\(X_{w,t}\), \(X_{hc,t}\), and \(X_{hnc,t}\)) is the product of the total fish crossing that week, \(X_t\), multiplied by one minus the re-ascension rate, \((1 - \eta_t)\), and the origin proportion vector, \(\omega_t\).
+
+\begin{equation}
+  \begin{aligned}
+    \begin{bmatrix} X_{w,t} \\ X_{hc,t} \\ X_{hnc,t} \end{bmatrix}
+    &= X_t * (1 - \eta_t) * \begin{bmatrix} \omega_{w,t} \\ \omega_{hc,t} \\ \omega_{hnc,t} \end{bmatrix}
+  \end{aligned}
+\end{equation}
+
+In practice, the total number of fish crossing LGR each week, \(X_t\), could be split into other discrete groups if desired; for example, individual wild stocks using genetic stock identification or hatchery stocks with parentage based tagging (Steele et al.
+2019).
+A graphical depiction (directed acyclic graph) of the model framework described above is found in Figure \ref{fig:dag-fig}.
+
+\hypertarget{model-fitting}{%
+\subsubsection{Model Fitting}\label{model-fitting}}
+
+The model was fit using the JAGS program (Plummer 2019), run with R software (R Core Team 2020).
+The STADEM package is available from the primary author at \url{https://github.com/KevinSee/STADEM}, and requires the use of the JAGS software (Plummer 2019) for Bayesian inference.
+Variance parameters \(\sigma_X, \sigma_\eta, \sigma_\theta\), and \(\sigma_\omega\), as well as the initial abundance, \(X_1\), and the overdispersion parameter of the negative binomial, \(r\), were given half-Cauchy priors with mean of 0 and scale of 100.
+The initial daytime passage and re-ascension rates, \(\theta_1\) and \(\eta_1\) were given Uniform(0,1) priors.
+Finally, \(\phi_{w,1}\) and \(\phi_{hnc,1}\) were given priors of Uniform(-3,3), in an effort to make \(\omega_1\) as uninformative as possible.
 
@@
 
-We tested the STADEM model on a variety of simulated data sets.
-These simulated data sets contained a fixed number of unique adult fish of known origin crossing a dam, from a total of 25 fictional populations with differential run-timing (i.e., date of passage at Lower Granite Dam).
-Each simulated fish was given a date of ladder ascension, based on its population and the range of run timing for that population.
+We tested the STADEM model on a variety of simulated data sets, constructed to mimic some dynamics of interest at LGR.
+These simulated data sets each contained a fixed number of unique adult fish of known origin crossing a dam, from a total of 25 fictional populations with differential run-timing (i.e., date of passage at LGR).
+Each simulated fish was given a date of ladder ascension, based on its population and the range of observed run timing for that population.
 Each fish was also simulated to cross the dam either while the window was open for counting, or not, and was given the chance to be ``caught'' in the simulated fish trap given the week when it ascended the dam, and the known trap rate that week.
@@
 
-Our objective was to examine STADEM model estimates of origin-specific (wild, hatchery, hatchery no-clip) escapement from the combinations of two separate trapping rates, two fallback/re-ascension and nighttime passage combinations, and three window count error rates; resulting in twelve different scenarios (Table \ref{tab:lgr-sim-tab}).
-The simulation parameters such as proportion of origin, run-timing, nighttime passage rates, fallback and re-ascension rates and trap rates were based on observed values at Lower Granite Dam between 2010-2015.
-Further details about simulation procedures can be found in \protect\hyperlink{append2}{Appendix B}.
-
-\hypertarget{lower-granite-application}{%
-\subsection{Lower Granite Application}\label{lower-granite-application}}
-
-Finally, we applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead returning to the Snake River to spawn during 2010 to 2019.
+Our objective was to assess STADEM model estimates of origin-specific (wild, hatchery, hatchery no-clip) escapement for accuracy and precision, given a known simulated ``truth'', under different possible conditions.
+We developed those conditions from the combinations of two trap rate scenarios (constant and shut down for 3 weeks), two fallback/re-ascension and nighttime passage combinations (equal and unequal), and three window count error rates (none, low and high); resulting in twelve different scenarios overall (Table \ref{tab:lgr-sim-tab}).
+The simulation parameters such as proportion of origin, run-timing, nighttime passage rates, fallback and re-ascension rates and trap rates were based on observed values at LGR between 2010-2015.
+We generated 500 simulations for each scenario, and ran STADEM on each one, resulting in a total of 6,000 model runs.
+Further details about the simulation procedures can be found in \protect\hyperlink{append1}{Appendix A}.
+
+To assess the performance of the model, we evaluated the estimates of origin-specific escapement by several measures.
+Accuracy was captured by the mean relative bias: the difference between the simulated ``truth'' and the STADEM estimate for each simulation, divided by the simulated ``truth''.
+We estimated the precision by examining the average coefficient of variation (CV) of the estimates.
+We calculated the root mean squared error (RMSE) as the square root of the mean of the squared bias in the estimate, which is equivalent to the square root of the variance of the estimator plus the square of the expected bias.
+Finally, we evaluated whether estimates of uncertainty were appropriate by calculating coverage probabilities to determine what proportion of the model results generated an estimated 95\% credible interval that contained the simulated true value.
+To compare the STADEM results directly to window counts, we compared the model estimates of total unique fish and the simulated window counts to the simulated number of unique fish.
+
+\hypertarget{lower-granite-dam-application}{%
+\subsection{Lower Granite Dam Application}\label{lower-granite-dam-application}}
+
+We applied STADEM to empirical data from LGR for sp/sum Chinook Salmon and steelhead returning to the Snake River during spawn years 2010 to 2019.
 Window counts for both species were accessed from DART via functions within STADEM.
-For Chinook Salmon, a spawn year refers to adults that migrate past the dam prior to August 17 each year and spawn that late summer and fall.
-For steelhead, the spawn year is defined as steelhead that migrate past Lower Granite Dam starting July 1 the previous year and prior to June 30 of the given year (e.g., spawn year 2017 steelhead migrate past Lower Granite between July 1, 2016 and June 30, 2017).
-Data from the adult trap was made available by Idaho Department of Fish and Game, and adult PIT tag detection data within the fish passage ladder at Lower Granite Dam was accessed from the PTAGIS regional database (\url{https://www.ptagis.org/}).
+For sp/sum Chinook Salmon, a spawn year refers to adults that migrate past the dam prior to August 17 each year and spawn that late summer and fall.
+For steelhead, the spawn year is defined as steelhead that migrate past LGR starting July 1 the previous year and prior to June 30 of the given year (e.g., spawn year 2017 steelhead migrate past LGR between July 1, 2016 and June 30, 2017).
+Data from the adult trap were made available by Idaho Department of Fish and Game, and adult PIT tag detection data within the fish passage ladder at LGR were accessed through DART and the adult ladder PIT tag query (\url{http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail}).
+
+All data used in this manuscript, as well as associated code for both the simulations and the LGR application, can be found at \href{\%5Bhttps://www.github.com/KevinSee/ManuscriptSTADEM\%5D}{https://www.github.com/KevinSee/ManuscriptSTADEM}.
 
@@
 
-Simulation results of observed bias, sampling variation, precision, root mean squared deviation (RMSD), and coverage probabilities were qualitatively similar for hatchery (N = 70,000) and hatchery no-clip (N = 5,000) origins as observed in the wild origin (N = 25,000) comparisons.
-As such, only diagnostic measures of Lower Granite Dam model fits to a medium sized escapement level (e.g., wild origin escapement) are presented.
-
-STADEM results were very similar across all scenarios (Figure \ref{fig:rel-bias-lgd}).
-Estimates were unbiased, with an average relative bias of 0.2--0.3\%.
-The CV of the estimates average 2.0--3.0\%, with coverage probabilities that always exceeded 95\%.
-We calculated RMSD as the square root of the sum of the variance of the estimate and the squared expected bias, which accounts for the size of the uncertainty in the estimator as well as its bias.
-The RMSD was near 500 for each scenario, representing an estimate within 2\% of the true value (Table \ref{tab:summ-tab}).
-
-\hypertarget{lower-granite-application-1}{%
-\subsection{Lower Granite Application}\label{lower-granite-application-1}}
-
-We applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead for spawn years 2010--2019.
-Estimates of total escapement, as well as estimates of wild, hatchery, and hatchery no-clip estimates are presented in Table \ref{tab:lgr-tab}.
-Coefficients of variation ranged from 3.5-7.5\% for wild fish, 3.0-6.2\% for hatchery fish, 4.0-8.4\% for hatchery no-clip fish and 2.9-5.6\% for total unique fish past Lower Granite Dam.
-
-Weekly estimates of total escapement over Lower Granite Dam tracked the window counts and trap estimates (Figure \ref{fig:jags-fit-fig}).
-STADEM point estimates were often between estimates based on window counts, and those based on the number of fish caught in the adult trap.
-However, when very few fish were caught in the trap, or there was more uncertainty about the trap rate that week, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \ref{fig:jags-fit-fig}.
-That year shows the utility of STADEM in dealing with missing data, as the trap was shut down for several weeks in July and August.
-
-Examining the estimates of nighttime passage and re-ascension rates based on the observed PIT tags crossing over Lower Granite Dam, the two rates did not match up (Figure \ref{fig:night-reasc-diff-fig}).
-In particular, there are several weeks when the window counts are quite high, and the rates differ by as much as 10\%.
-Clearly, those two biological processes do not cancel each other out, and thus, employing a model that accounts for both will result in more accurate estimates of escapement.
+The evaluation statistics (relative bias, CV, RMSE, coverage probabilities) to assess performance of the STADEM model were very similar among hatchery clipped (N = 70,000), hatchery no-clip (N = 5,000) and wild fish (N = 25,000).
+In the interest of brevity, we only present the results from wild fish here, which corresponds to a moderate sized escapement relative to the hatchery clipped and hatchery no-clip groups.
+
+STADEM results showed similar patterns for estimates of unique total fish and unique wild fish, in that both were unbiased with similar variance in the estimator across all scenarios as seen in the consistent interquartile range in the boxplots of Figure \ref{fig:rel-bias-lgd}.
+For unique total fish, window counts were unbiased, except in scenarios when night passage and re-ascension rates were not equal (N-R).
+In those scenarios, they were biased by nearly 5\% (Figure \ref{fig:rel-bias-lgd}), the designed simulated difference between night passage and re-ascension rates (Table \ref{tab:lgr-sim-tab}).
+The variance of the window counts also grew as the simulated observer error grew (no error vs.~Err L vs.~Err H), whereas the variance in the STADEM estimates remained fairly constant, albeit often larger than the window count variance.
+
+Estimates of wild escapement were unbiased, with an average relative bias of 0.2--0.3\%, although the bias appeared equally distributed between positive and negative values across the simulations (Figure \ref{fig:rel-bias-lgd}).
+The CV of the estimates averaged 2.0--3.0\%, with higher CV's being associated with scenarios with a simulated trap closure for 3 weeks.
+The coverage probabilities always exceeded 95\% across all scenarios.
+The RMSE was near 500 for each scenario, representing an estimate within 2\% of the true value, demonstrating the accuracy and precision of STADEM (Table \ref{tab:summ-tab}).
+
+\hypertarget{lower-granite-dam-application-1}{%
+\subsection{Lower Granite Dam Application}\label{lower-granite-dam-application-1}}
+
+We applied STADEM to data from LGR for sp/sum Chinook Salmon and steelhead for spawn years 2010--2019.
+Estimates of total escapement, as well as estimates of wild, hatchery clipped, and hatchery no-clip estimates are presented in Table \ref{tab:lgr-tab}.
+Estimates of total unique fish escaping past LGR were sometimes higher and sometimes lower than the raw window counts, indicating that the relative strength of nighttime passage and re-ascension rates differed across years.
+Window count bias ranged from -3.0\% to 11.0\% for sp/sum Chinook Salmon and -9.9\% to 6.6\% for steelhead.
+STADEM coefficients of variation ranged from 2.5 to 7.1\% for wild fish, 2.3 to 5.1\% for clipped hatchery fish, 3.4 to 8.8\% for hatchery no-clip fish and 2.2 to 4.7\% for total unique fish past LGR.
+
+Weekly STADEM estimates of escapement over LGR tracked the window counts and trap estimates, with smaller credible intervals than the trap estimates alone.
+(One example: Figure \ref{fig:jags-fit-fig}.
+Similar plots for all model runs {[}each species and year combination{]} are available on the manuscript GitHub page: \href{\%5Bhttps://www.github.com/KevinSee/ManuscriptSTADEM\%5D}{https://www.github.com/KevinSee/ManuscriptSTADEM}).
+
+Weekly nighttime passage and re-ascension proportions were significantly different (2-sample test of proportions, \emph{p} \textless{} 0.05) in approximately 50\% of the weeks with sufficient PIT tag data to test.
+The differences in the estimated weekly rates were also apparent visually (Figure \ref{fig:night-reasc-diff-fig}).
+In particular, there were several weeks when the window counts were quite large, but the nighttime passage and re-ascension rates differed by as much as 10\%; leading to larger window count bias in some years (Table \ref{tab:lgr-tab}).
 
@@
 
-We have presented a novel method for estimating adult salmonid escapement past a large hydroelectric facility (e.g., Lower Granite Dam) that incorporates data from window counts, fish trap, and observations of PIT tagged fish in the adult passage ladder.
-Our model explicitly models nighttime passage, re-ascension, and potential error in both window and trap estimates.
-In doing so, we demonstrated that at Lower Granite Dam, nighttime passage and re-ascension rates do not always offset each other, and assuming they do, will lead to biased escapement estimates in some years.
-With minor adjustments this modelling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall Chinook Salmon, Pacific lamprey Lampetra tridentata), or elsewhere; provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run, and a tag observation or detection infrastructure (e.g., a PIT tag detection array or similar) exists.
-Our state-space model combined multiple imperfect sources of data to reduce bias in adult escapement estimates and provided more reasonable estimates of uncertainty.
-Accurate population or stock abundance estimates and uncertainty accounting for observation and process error can be particularly important when estimates are used or leveraged for management and conservation decisions such as population viability analyses.
-
-Combining data from the adult fish trap with live and video window counts provides several benefits.
-First, it allows us to model observer error in the window counts, which is typically unknown.
-If estimates rely on window counts alone, quantifying observer error is impossible, and we believe it's prudent to capture and account for known sources of error to minimize management decision risk.
-Second, by incorporating both sources of information in a state-space framework, STADEM incorporates missing data at either the observation window or adult trap seamlessly.
-At Lower Granite Dam, the adult trap has been closed for brief or extended periods of time (i.e., days, weeks) intermittently over the past several years, often during peak run times.
-Trap closures are typically associated with elevated water temperatures resulting in potential fish handling stress and/or trap malfunctions.
-Given predicted Pacific Northwest climate change scenarios trap closures from high water temperatures may become more commonplace in the future; amplifying the need for a modeling framework that accounts for periods of missing data while still capturing estimate uncertainty.
-Additionally, having a framework in place that accounts for missing periods of data will allow for increased logistic flexibility if, for example, maintenance or construction is needed at the observational window or adult trap.
-
-Although not currently set up for this, STADEM could be modified and run on a weekly basis or in near real-time to provide better in-season estimates for fisheries managers.
-Currently, the only roadblock to this, at Lower Granite Dam, is the identification of hatchery origin fish from phenotypically wild fish (i.e., hatchery no-clip) using genetic tissues samples (\textbf{\emph{Steele et al.~2013}}) collected at the adult trap, which currently is completed post hoc after the trapping season.
-The inclusion of genetic information typically results in a reduction in wild escapement estimates and an associated increase in hatchery no-clip escapement.
+We have presented a novel method for estimating adult salmonid escapement past a large hydroelectric facility (e.g., LGR) that incorporates data from window counts, a fish trap, and observations of PIT tagged fish in the adult passage ladder.
+Our model explicitly accounts for nighttime passage, re-ascension, and potential error in both window and trap estimates.
+In doing so, we demonstrated that at LGR, nighttime passage and re-ascension rates do not always offset each other, and assuming they do will lead to biased escapement estimates in some years.
+Our state-space model combined multiple imperfect sources of data to reduce bias in adult escapement estimates and provided quantitative estimates of uncertainty.
+Using both data streams proved beneficial, and protected estimates from unforeseen problems or periodic biases in observations (e.g., trap closures, unequal night time and re-ascension rates).
+Accurate population or group abundance estimates and uncertainty accounting for observation and process error can be particularly important when estimates are used for management and conservation decisions such as population viability analyses (Paulsen et al.
+2007).
+
+The results from our simulations demonstrate that STADEM provides consistent, unbiased estimates of escapement across a range of scenarios.
+Regardless of whether observer error at the window is negligible, low or high, whether nighttime passage and re-ascension rates are equal or not, whether the trap is shut down for several weeks in a season or not, STADEM produced estimates of unique escapement that were unbiased, precise (CVs in the 2-3\% range) and contained appropriate credible intervals.
+Although the average bias among runs within a given simulation scenario was often positive (Table \ref{tab:summ-tab}), Figure \ref{fig:rel-bias-lgd} shows that for each scenario some simulations resulted in a negative bias, while some had a positive one.
+We posit that conducting additional simulations would bring the average bias closer to zero, but run times constrained how many simulations we were able to conduct for each scenario.
+
+As for the application of STADEM to empirical data at LGR, the weekly estimates were always close to the window counts and the trap estimates, indicating that both sources of data were providing information about the same phenomenon.
+The STADEM weekly estimates were often slightly lower than those based on the number of fish caught in the adult trap because the trap estimates do not account for fallback and re-ascension (Figure \ref{fig:jags-fit-fig}).
+However, the model's uncertainty is always smaller than the uncertainty from the trap estimates alone, whereas the window counts by themselves provide no estimate of uncertainty.
+When comparing the total estimates of each season with the window counts (Table \ref{tab:lgr-tab}), STADEM estimates were higher when night passage rates exceeded re-ascension rates, and lower when the re-ascension rate was higher than the night passage rate.
+Given that there were several weeks with large window counts when the rates differed by as much as 10\% (Figure \ref{fig:night-reasc-diff-fig}), the counts for those weeks should be considered biased.
+For those weeks when very few fish were caught in the trap or there was more uncertainty about the trap rate (e.g., few previously tagged fish detected in the ladder), STADEM estimates more closely tracked the window counts, as seen in the second week of July 2014, in Figure \ref{fig:jags-fit-fig}.
+Vice-versa, for weeks when the estimate of the trap rate was more precise, STADEM estimates more closely follow the trap estimates, as seen in the first half of May (Figure \ref{fig:jags-fit-fig}).
+In either case, the state-space nature of STADEM tends to put more emphasis on the data stream with less observation error.
+
+Using a state-space approach is an improvement on window counts alone for several reasons.
+The first is that by accounting for both night passage and re-ascension rates, that are sometimes unequal, it corrects any potential bias in window counts.
+Across all combinations of years and species with empirical data from LGR that we fit STADEM to, that bias ranged from -9.9\% to 11.0\% (Table \ref{tab:lgr-tab}).
+Whether that amount of potential bias is of concern to management agencies is a decision for them to make, but we have provided a tool to correct that bias if it is of concern.
+The other justification for STADEM over window counts is that it provides an estimate of observer error for the window counts, leading to estimates of precision for unique fish which are unobtainable with window counts alone.
+Capturing and accounting for known sources of error is prudent to minimize management decision risk (Harwood and Stokes 2003; Ascough Ii et al.
+2008).
+At the same time, the precision among STADEM's estimates is still quite good, with CVs near 2-3\%, well below the goal of 15\% set forth by NOAA (Crawford and Rumsey 2011).
+
+By combining live and video window counts with data from the adult fish trap in a state-space framework, STADEM incorporates missing data at either the observation window or adult trap seamlessly (e.g., see Figure \ref{fig:jags-fit-fig} when the trap was shut down for several weeks in July and August).
+At LGR, the adult trap has been closed for brief or extended periods of time (i.e., days, weeks) intermittently in five of the last ten years, and often during peak run times (USACE 2010, 2011, 2012, 2013, 2015, 2016, 2017, 2018, 2019; Ogden 2016b).
+Trap closures are typically associated with elevated water temperatures resulting in potential fish handling stress and/or trap malfunctions (Ogden 2016a).
+Given predicted Pacific Northwest climate change scenarios (Zhang et al.
+2019), trap closures from high water temperatures may become more commonplace in the future, amplifying the need for a modeling framework that accounts for periods of missing data while still capturing uncertainty in the estimates.
+Additionally, having a framework in place that accounts for missing periods of data will allow for increased logistical flexibility if, for example, maintenance or construction is needed at the observational window or adult trap.
+
+STADEM could be modified and run on a weekly basis or in near real-time to provide better in-season estimates for fisheries managers.
+Currently, the only roadblock to this at LGR is the identification of hatchery no-clip fish, which is aided by genetic tissue samples (Steele et al.
+2013, 2019) collected at the adult trap, and parentage-based tagging (Steele et al.
+2013, 2019) which currently is completed post-hoc after the trapping season.
+The inclusion of this genetic information typically results in a reduction in wild escapement estimates and an associated increase in hatchery no-clip escapement, as phenotypically wild fish are identified as the progeny of hatchery broodstock.
+Accepting a potential bias due to unclipped hatchery fish can create a greater risk during status assessments of these ESA listed wild populations.
 However, if in-season management decisions do not require this correction or could accept the potential bias, origin calls at the trap could be used in-season as a first approximation to escapement.
-Final post-hoc estimates, parsed by origin, could then be finalized at season's end.
-All other data included in this model (e.g., window counts and PIT observations) are otherwise provided in almost real-time by DART.
-Provided the Lower Granite Dam adult fish trap database was updated and made available frequently, there are minimal obstacles for adapting the STADEM framework to provide in-season estimates of escapement.
-
-Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary spawning areas(Kinzer et al.
-2020a, 2020b).
-Estimates of escapement at Lower Granite Dam, by species and origin, include known uncertainty, and are available to further parse into sex- or age-structured escapement estimates (e.g., Camacho et al.
-(2017), Schrader et al.
-(2013)) that are important for fisheries management and productivity monitoring of wild populations.
-As an example, STADEM is being applied at Lower Granite to estimate the total unique wild fish migrating past the dam.
-Estimates of fish passing the dam are then combined with PIT tag observations at instream PIT tag detection systems throughout the Snake River basin and estimated movement or transition probabilities, simialar to Waterhouse et al.
+Final post-hoc estimates parsed by origin could then be completed at season's end.
+All other data included in this model (e.g., window counts and PIT observations) are otherwise provided in near real-time by DART.
+Provided the LGR adult fish trap database was updated and available in near real-time, there are minimal obstacles for adapting the STADEM framework to provide in-season estimates of escapement.
+
+Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate escapement of sp/sum Chinook Salmon and steelhead past LGR, and returning to tributary or population specific spawning areas (IPTDSW 2020; Kinzer et al.
+2020).
+Estimates of species and origin-specific escapement at LGR, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates (e.g., Camacho et al.
+2018; Schrader et al.
+2013) that are important for fisheries management and productivity monitoring of wild populations.
+As an example, STADEM estimates of unique wild fish at LGR are combined with estimated movement or transition probabilities based on PIT tag observations at instream PIT tag detection systems throughout the Snake River basin, similar to Waterhouse et al.
 (2020), to estimate escapement to Snake River populations and locations throughout the basin (Orme et al.
-2018).
-Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations that, joined with sex and age data collected at the adult fish trap (Hargrove et al.
-2019), provides necessary information to evaluate productivity and population viability for select Snake River Chinook Salmon and steelhead groups (Kinzer et al.
-2020b).
-
-Although STADEM was developed with salmonid escapement at Lower Granite Dam in mind, it could be applied to any migratory fish species at locations with similar monitoring infrastructure.
+2019; IPTDSW 2020; Kinzer et al.
+2020).
+Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations.
+With sex and age data collected at the adult fish trap (Hargrove et al.
+2019), this approach provides necessary information to evaluate productivity and population viability for select Snake River sp/sum Chinook Salmon and steelhead groups (IPTDSW 2020).
+
+With minor adjustments this modeling framework and the STADEM package could be applied to similar migratory species at LGR (e.g., fall-run Chinook Salmon, Pacific lamprey \emph{Lampetra tridentata}), or elsewhere, provided there exists a fish counting mechanism, a trap that can be used to sample a portion of the run, and tag detection infrastructure (e.g., a PIT tag detection array or similar).
 Justification and infrastructure exist for applying a modified STADEM framework for fish passing Bonneville Dam, the lowest dam on the Columbia River, or Priest Rapids Dam in the upper Columbia River.
-Both locations currently trap a sub-sample of passing Chinook Salmon and steelhead for biological information, and use window counts as a surrogate of true escapement, however, similar problems to those observed at Lower Granite Dam, such as, unaccounted observer and sampling error, night-time passage, and re-ascension exists at the sites.
-Certainly, estimating an unbiased total return to the entire Columbia River basin (i.e., Bonneville Dam) and Upper Columbia River with uncertainty would benefit managers and decision making.
+Both locations currently trap a sub-sample of passing Chinook Salmon and/or steelhead for biological information and use window counts as a surrogate of true escapement.
+However, each has at least some similar problems to those observed at LGR such as unaccounted observer and sampling error, nighttime passage, and/or re-ascension.
+Certainly, estimating an unbiased total return to the vast majority of the Columbia River basin (i.e., Bonneville Dam) and Upper Columbia River with uncertainty would benefit managers and decision making.
 
@@
 Funding for this study and development of the STADEM model was partially provided by the Bonneville Power Administration under project 2003-017-00.
-Special thanks to Darren Ogden and staff at the Lower Granite Dam adult trap for their hard work and diligent data collection.
-Thank you to personnel at the Idaho Department of Fish and Game, particularly Paul Bunn, Tim Copeland, and Bill Schrader, for providing access to data from the adult fish trap and conceptualizing methods, and the staff at the Eagle Fish Lab for analyzing genetic samples.
+Special thanks to Darren Ogden and staff at the LGR adult trap for their hard work and diligent data collection.
+Thank you to personnel at the Idaho Department of Fish and Game, particularly Paul Bunn, Tim Copeland, and Bill Schrader, for providing access to data from the adult fish trap and conceptualizing methods, and to Matthew Campbell and staff at the Eagle Fish Genetics Laboratory for analyzing genetic samples.
 Special thanks to Columbia Basin Research staff and the Columbia River Data Access in Real Time (DART) application, and Susannah Iltis in particular.
-And finally, thank you to Rick Orme for his contributions to the development of STADEM, and all the other folks who have contributed through productive critique and conversations.
+Thank you to three anonymous reviewers and an associate journal editor that helped us improve the manuscript with their thoughtful comments and suggestions.
+And finally, thank you to Rick Orme and Brice Semmens for their contributions to the development of STADEM, as well as Eric Buhle, Sarah Hoffmann, John Hargrove and all the other people who have contributed through productive critique and conversations.
 
@@
 \hypertarget{refs}{}
+\leavevmode\hypertarget{ref-Ascough-Ii2008}{}%
+Ascough Ii, J., H.
+Maier, J.
+Ravalico, and M.
+Strudley.
+2008.
+Future research challenges for incorporation of uncertainty in environmental and ecological decision-making.
+Ecological modelling 219(3-4):383--399.
+Elsevier.
+
 \leavevmode\hypertarget{ref-Boggs2004}{}%
@@
 2004.
-Fallback, reascension, and adjusted fishway escapement estimates for adult chinook salmon and steelhead at columbia and snake river dams.
+Fallback, reascension, and adjusted fishway escapement estimates for adult Chinook Salmon and steelhead at Columbia and Snake River dams.
 Transactions of the American Fisheries Society 133(4):932--949.
-Taylor \& Francis.
 
 \leavevmode\hypertarget{ref-Bue1998}{}%
-Bue, B., S.
+Bue, B.
+G., S.
+M.
 Fried, S.
 Sharr, D.
+G.
 Sharp, J.
+A.
 Wilcock, and H.
+J.
 Geiger.
 1998.
-Estimating salmon escapement using area-under-the-curve, aerial observer efficiency, and stream-life estimates: The prince william sound pink salmon example.
+Estimating salmon escapement using area-under-the-curve, aerial observer efficiency, and stream-life estimates: The Prince William Sound pink salmon example.
 North Pacific Anadromous Fish Commission Bulletin 1:240--250.
 
-\leavevmode\hypertarget{ref-Camacho2019}{}%
+\leavevmode\hypertarget{ref-Camacho2018}{}%
 Camacho, C.
-A., T.
-Delomas, M.
+A., J.
+Powell, M.
 Davison, M.
@@
 C.
+Schrader, T.
+Copeland, and M.
+Campbell.
+2018.
+Wild adult steelhead and Chinook Salmon abundance and composition at Lower Granite Dam, spawn year 2017.
+Idaho Department of Fish and Game, Fishery Research, Annual Progress Report 19-09.
+
+\leavevmode\hypertarget{ref-Campbell2012}{}%
+Campbell, M.
+R., C.
+C.
+Kozfkay, T.
+Copeland, W.
+C.
 Schrader, M.
+W.
+Ackerman, and S.
 R.
-Campbell, and T.
-Copeland.
-2019.
-Wild juvenile steelhead and Chinook Salmon abundance and composition at Lower Granite Dam, migratory year 2018.
-Idaho Department of Fish and Game, 19-12.
-
-\leavevmode\hypertarget{ref-Camacho2017}{}%
-Camacho, C., K.
-Wright, J.
-Powell, W.
-C.
-Schrader, T.
-Copeland, M.
-Ackerman, M.
-Dobos, M.
-Corsi, and M.
-R.
-Campbell.
-2017.
-Wild adult steelhead and Chinook salmon abundance and composition at Lower Granite Dam, spawn years 2009-2016.
-Idaho Department of Fish and Game, 17-06.
-
-\leavevmode\hypertarget{ref-crozier2019climate}{}%
+Narum.
+2012.
+Estimating abundance and life history characteristics of threatened wild Snake River steelhead stocks by using genetic stock identification.
+Transactions of the American Fisheries Society 141(5):1310--1327.
+
+\leavevmode\hypertarget{ref-Crawford2011}{}%
+Crawford, B., and S.
+Rumsey.
+2011.
+Guidance for monitoring recovery of pacific northwest Salmon \& steelhead listed under the federal Endangered Species Act.
+National Oceanic and Atmospheric Administration, National Marine Fisheries Service, Northwest Region, Portland, Oregon.
+Available: www.
+pnamp.
+org/sites/default/files/noaa\_rme\_guidanceappendices2011.
+pdf.(November 2014).
+NOAA National Marine Fisheries Service.
+
+\leavevmode\hypertarget{ref-Crozier2019}{}%
 Crozier, L.
@@
 2019.
-Climate vulnerability assessment for pacific salmon and steelhead in the california current large marine ecosystem.
+Climate vulnerability assessment for Pacific salmon and steelhead in the California current large marine ecosystem.
 PloS one 14(7):e0217711.
-Public Library of Science San Francisco, CA USA.
 
 \leavevmode\hypertarget{ref-Dauble2000}{}%
-Dauble, D.
-D., and R.
-P.
-Mueller.
-2000.
-Upstream passage monitoring: Difficulties in estimating survival for adult chinook salmon in the columbia and snake rivers.
-Fisheries 25(8):24--34.
-Taylor \& Francis.
-
-\leavevmode\hypertarget{ref-nwfsc2015}{}%
-Ford, M., K.
-Barnas, T.
-Cooney, L.
-Crozier, M.
-Diaz, J.J.
-Hard, E.
-Holmes, D.
-Holzer, R.
-Kope, P.
-Lawson, M.
-Liermann, J.
-Myers, M.
-Rowse, D.
-Teel, D.
-V.
-Doornik, T.
-Wainwright, L.
-Weitkamp, and M.
-Williams.
-2015.
-Status review update for pacific salmon and steelhead listed under the endangered species act: Pacific northwest.
-National Marine Fisheries Service, Northwest Fisheries Science Center.
+Dauble, D. D., and R. P. Mueller. 2000. Upstream passage monitoring: Difficulties in estimating survival for adult Chinook Salmon in the Columbia and Snake rivers. Fisheries 25(8):24--34.
 
 \leavevmode\hypertarget{ref-Hargrove2019}{}%
-Hargrove, J. S., T. A. Delomas, and M. Davison. 2019. Chinook and steelhead genotyping for genetic stock identification at Lower Granite Dam. Idaho Department of Fish and Game, 19-08.
+Hargrove, J.
+S., T.
+A.
+Delomas, and M.
+Davison.
+2019.
+Chinook Salmon and steelhead genotyping for genetic stock identification at Lower Granite Dam.
+Idaho Department of Fish and Game, Fishery Research, Annual Progress Report 19-08.
+
+\leavevmode\hypertarget{ref-Harwood2003}{}%
+Harwood, J., and K. Stokes. 2003. Coping with uncertainty in ecological advice: Lessons from fisheries. Trends in Ecology \& Evolution 18(12):617--622. Elsevier.
 
@@
 1994.
-Feasibility of documenting and estimating adult fish passage at large hydroelectric facilities in the snake river using video technology; 1993 final report.
+Feasibility of documenting and estimating adult fish passage at large hydroelectric facilities in the Snake River using video technology; 1993 final report.
 Columbia River Inter-Tribal Fish Commission.
 
-\leavevmode\hypertarget{ref-hess2014monitoring}{}%
+\leavevmode\hypertarget{ref-Hess2014}{}%
 Hess, J.
@@
 2014.
-Monitoring stock-specific abundance, run timing, and straying of chinook salmon in the columbia river using genetic stock identification (gsi).
+Monitoring stock-specific abundance, run timing, and straying of Chinook Salmon in the Columbia River using genetic stock identification (GSI).
 North American Journal of Fisheries Management 34(1):184--201.
-Taylor \& Francis.
-
-\leavevmode\hypertarget{ref-kinzer2020report}{}%
+
+\leavevmode\hypertarget{ref-IPTDSW2020}{}%
+IPTDSW.
+2020.
+Report to NOAA Fisheries for 5-year ESA status review: Snake River basin steelhead and Chinook Salmon population abundance, life history, and diversity metrics calculated from in-stream PIT-tag observations (SY2010-SY2019).
+IPTDSW (In-stream PIT-tag detection systems workgroup).
+
+\leavevmode\hypertarget{ref-Jasper2018}{}%
+Jasper, J.
+R., M.
+Short, C.
+Shelden, and W.
+S.
+Grant.
+2018.
+Hierarchical Bayesian estimation of unobserved salmon passage through weirs.
+Canadian Journal of Fisheries and Aquatic Sciences 75(7):1151--1159.
+
+\leavevmode\hypertarget{ref-WDFW2019}{}%
+Joint Columbia River Management Staff.
+2019.
+2019 Joint Staff Report: Stock status and fisheries for spring Chinook, summer Chinook, sockeye, steelhead, and other species.
+Oregon Department of Fish \& Wildlife; Washington Department of Fish \& Wildlife.
+
+\leavevmode\hypertarget{ref-Kinzer2020}{}%
 Kinzer, R.
-N., H.
-Arnsberg Bill, A.
+N., B.
+Arnsberg, J.
+Harbeck, A.
 Maxwell, R.
@@
 Vatland.
-2020a.
-Snake river basin adult chinook salmon and steelhead monitoring 2019 annual report.
+2020.
+Snake River basin adult Chinook Salmon and steelhead monitoring, 2019 annual report.
 Nez Perce Tribe, Department of Fisheries Resources Management, Research Division, Lapwai, ID.
 
-\leavevmode\hypertarget{ref-iptds2020report}{}%
-Kinzer, R.
-N., Orme Rick, M.
-Campbell, J.
-Hargrove, and K.
-See.
-2020b.
-REPORT to noaa fisheries for 5-year esa status review: SNAKE river basin steelhead and chinook salmon population abundance, life history, and diversity metrics calculated from in-stream pit-tag observations (sy2010-sy2019).
-
-\leavevmode\hypertarget{ref-mcclure2003large}{}%
+\leavevmode\hypertarget{ref-McClure2003}{}%
 McClure, M.
@@
 2003.
-A large-scale, multispecies status assessment: Anadromous salmonids in the columbia river basin.
+A large-scale, multispecies status assessment: Anadromous salmonids in the Columbia River basin.
 Ecological Applications 13(4):964--989.
-Wiley Online Library.
-
-\leavevmode\hypertarget{ref-mcelhany2000viable}{}%
-McElhany, P., M. H. Rucklelshaus, M. J. Ford, T. C. Wainwright, and E. P. Bjorkstedt. 2000. Viable salmonid populations and the recovery of evolutionarily significant units.
-
-\leavevmode\hypertarget{ref-nehlsen1991pacific}{}%
-Nehlsen, W., J. E. Williams, and J. A. Lichatowich. 1991. Pacific salmon at the crossroads: Stocks at risk from california, oregon, idaho, and washington. Fisheries 16(2):4--21. Wiley Online Library.
+
+\leavevmode\hypertarget{ref-McElhany2000}{}%
+McElhany, P., M.
+H.
+Rucklelshaus, M.
+J.
+Ford, T.
+C.
+Wainwright, and E.
+P.
+Bjorkstedt.
+2000.
+Viable salmonid populations and the recovery of evolutionarily significant units.
+U.S.
+Department of Commerce, NOAA Technical Memorandum NMFS-NWFSC-42.
+
+\leavevmode\hypertarget{ref-NMFS2019}{}%
+National Marine Fisheries Service.
+2019.
+Endangered Species Act (ESA) Section 7(a)(2) biological opinion and Magnuson-Stevens Fishery Conservation and Management Act essential fish habitat response.
+Continued operation and maintenance of the Columbia River system.
+National Marine Fisheries Service, NMFS Consultation WCRO-2019-02395.
+
+\leavevmode\hypertarget{ref-Nehlsen1991}{}%
+Nehlsen, W., J. E. Williams, and J. A. Lichatowich. 1991. Pacific salmon at the crossroads: Stocks at risk from California, Oregon, Idaho, and Washington. Fisheries 16(2):4--21.
+
+\leavevmode\hypertarget{ref-NWFSC2015}{}%
+Northwest Fisheries Science Center. 2015. Status review update for Pacific salmon and steelhead listed under the Endangered Species Act: Pacific Northwest.
+
+\leavevmode\hypertarget{ref-NPCC2014}{}%
+NPCC. 2014. Columbia River basin fish and wildlife program. Northwest Power \& Conservation Council.
 
 \leavevmode\hypertarget{ref-Ogden2014}{}%
-Ogden, D. A. 2014. Operation of the adult trap at lower granite dam 2013.
+Ogden, D. A. 2014. Operation of the adult trap at Lower Granite Dam 2013. NOAA National Marine Fisheries Service.
 
 \leavevmode\hypertarget{ref-Ogden2016b}{}%
-Ogden, D. A. 2016a. Operation of the adult trap at Lower Granite Dam 2015.
-
-\leavevmode\hypertarget{ref-Ogden2016}{}%
-Ogden, D. A. 2016b. Operation of the adult trap at Lower Granite Dam 2014.
-
-\leavevmode\hypertarget{ref-orme2018population}{}%
-Orme, R., N.
-Kinzer Ryan, and C.
+Ogden, D. A. 2016a. Operation of the adult trap at Lower Granite Dam 2015. NOAA National Marine Fisheries Service.
+
+\leavevmode\hypertarget{ref-Ogden2016a}{}%
+Ogden, D. A. 2016b. Operation of the adult trap at Lower Granite Dam 2014. NOAA National Marine Fisheries Service.
+
+\leavevmode\hypertarget{ref-Orme2019}{}%
+Orme, R., R.
+Kinzer, and C.
 Albee.
-2018.
-Population and tributary level escapement estimates of snake river natural-origin spring/summer chinook salmon and steelhead from in-stream pit tag detection systems-2019 annual report.
+2019.
+Population and tributary level escapement estimates of Snake River natural-origin spring/summer Chinook Salmon and steelhead from in-stream PIT tag detection systems - 2019 annual report.
 Nez Perce Tribe, Department of Fisheries Resources Management, Research Division, Lapwai, ID.
 
+\leavevmode\hypertarget{ref-Paulsen2007}{}%
+Paulsen, C.
+M., R.
+A.
+Hinrichsen, and T.
+R.
+Fisher.
+2007.
+Measure twice, estimate once: Pacific Salmon population viability analysis for highly variable populations.
+Transactions of the American Fisheries Society 136(2):346--364.
+Taylor \& Francis.
+
 \leavevmode\hypertarget{ref-R-rjags}{}%
-Plummer, M. 2019. Rjags: Bayesian graphical models using mcmc.
+Plummer, M. 2019. Rjags: Bayesian graphical models using MCMC.
 
 \leavevmode\hypertarget{ref-R-Core-Team2020}{}%
-R Core Team. 2020. R: A Language and Environment for Statistical Computing. R Foundation for Statistical Computing, Vienna, Austria.
-
-\leavevmode\hypertarget{ref-royle2008hierarchical}{}%
+R Core Team. 2020. R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria.
+
+\leavevmode\hypertarget{ref-Royle2008}{}%
 Royle, J. A., and R. M. Dorazio. 2008. Hierarchical modeling and inference in ecology: The analysis of data from populations, metapopulations and communities. Elsevier.
@@
 2013.
-Wild adult steelhead and Chinook salmon abundance and composition at Lower Granite dam, spawn year 2011.
-Idaho Department of Fish; Game, Report 13-15.
+Wild adult steelhead and Chinook Salmon abundance and composition at Lower Granite Dam, spawn year 2011.
+Idaho Department of Fish and Game, Fishery Research, Annual Report 13-15.
+
+\leavevmode\hypertarget{ref-Seber2002}{}%
+Seber, G. A. F. 2002. The estimation of animal abundance and related parameters. Blackburn Press Caldwell, New Jersey.
 
@@
 Shumway, R. H., and D. S. Stoffer. 2010. Time series analysis and its applications: With R examples. Springer.
+
+\leavevmode\hypertarget{ref-Steele2013}{}%
+Steele, C.
+A., E.
+C.
+Anderson, M.
+W.
+Ackerman, M.
+A.
+Hess, N.
+R.
+Campbell, S.
+R.
+Narum, and M.
+R.
+Campbell.
+2013.
+A validation of parentage-based tagging using hatchery steelhead in the Snake River basin.
+Canadian Journal of Fisheries and Aquatic Sciences 70(7):1046--1054.
+
+\leavevmode\hypertarget{ref-Steele2019}{}%
+Steele, C. A., M. Hess, S. Narum, and M. Campbell. 2019. Parentage-based tagging: Reviewing the implementation of a new tool for an old problem. Fisheries:1--11.
 
@@
 
+\leavevmode\hypertarget{ref-USACE2010}{}%
+USACE. 2010. Annual fish passage report, Columbia and Snake river projects. U.S. Army Corps of Engineers.
+
+\leavevmode\hypertarget{ref-USACE2011}{}%
+USACE. 2011. Annual fish passage report, Columbia and Snake river projects. U.S. Army Corps of Engineers.
+
+\leavevmode\hypertarget{ref-USACE2012}{}%
+USACE. 2012. Annual fish passage report, Columbia and Snake river projects. U.S. Army Corps of Engineers.
+
+\leavevmode\hypertarget{ref-USACE2013}{}%
+USACE. 2013. Annual fish passage report, Columbia and Snake river projects. U.S. Army Corps of Engineers.
+
 \leavevmode\hypertarget{ref-USACE2015}{}%
-USACE. 2015. Annual fish passage report Columbia and Snake rivers. U.S. Army Corps of Engineers.
-
-\leavevmode\hypertarget{ref-waterhouse2020bayesian}{}%
-Waterhouse, L., J.
-White, K.
-See, A.
-Murdoch, and B.
-X.
-Semmens.
-2020.
-A bayesian nested patch occupancy model to estimate steelhead movement and abundance.
-Ecological Applications:e02202.
-Wiley Online Library.
-
-\leavevmode\hypertarget{ref-williams2016viability}{}%
+USACE. 2015. Annual fish passage report, Columbia and Snake river projects. U.S. Army Corps of Engineers.
+
+\leavevmode\hypertarget{ref-USACE2016}{}%
+USACE. 2016. Annual fish passage report, Columbia and Snake river projects. U.S. Army Corps of Engineers.
+
+\leavevmode\hypertarget{ref-USACE2017}{}%
+USACE. 2017. Annual fish passage report, Columbia and Snake river projects. U.S. Army Corps of Engineers.
+
+\leavevmode\hypertarget{ref-USACE2018}{}%
+USACE. 2018. Annual fish passage report, Columbia and Snake river projects. U.S. Army Corps of Engineers.
+
+\leavevmode\hypertarget{ref-USACE2019}{}%
+USACE. 2019. Annual fish passage report, Columbia and Snake river projects. U.S. Army Corps of Engineers.
+
+\leavevmode\hypertarget{ref-USOregon2018}{}%
+U.S. v Oregon. 2018. 2018-2027 United States v. Oregon, Case No. 3:68-cv-00513-MO.
+
+\leavevmode\hypertarget{ref-Waterhouse2020}{}%
+Waterhouse, L., J. White, K. See, A. Murdoch, and B. X. Semmens. 2020. A Bayesian nested patch occupancy model to estimate steelhead movement and abundance. Ecological Applications:e02202.
+
+\leavevmode\hypertarget{ref-Williams2016}{}%
 Williams, T.
@@
 2016.
-Viability assessment for pacific salmon and steelhead listed under the endangered species act: Southwest.
+Viability assessment for Pacific salmon and steelhead listed under the Endangered Species Act: Southwest.
+U.S.
+Department of Commerce, NOAA Technical Memorandum NMFS-SWFSC-564.
+
+\leavevmode\hypertarget{ref-Woody2004}{}%
+Woody, C. A. 2004. Population monitoring of Lake Clark and Tazimina River sockeye salmon, Kvichak River watershed, Bristol Bay, Alaska, 2000-2003. Final Report for Study 01-095.:50.
 
@@
 Process and methods for assigning ages to anadromous salmonids from scale samples.
-Idaho Department of Fish; Game, 15-03.
+Idaho Department of Fish and Game, Fishery Research Report 15-03.
+
+\leavevmode\hypertarget{ref-Zhang2019}{}%
+Zhang, X., H. Li, Z. D. Deng, L. R. Leung, J. R. Skalski, and S. J. Cooke. 2019. On the variable effects of climate change on Pacific salmon. Ecological Modelling 397:95--106. Elsevier.
 
@@
 
-\begin{table}
-
-\caption{\label{tab:lgr-sim-tab}Summary of simulation scenarios including varying adult trapping, fallback rates, re-ascension, nighttime passage, and window count error rates use to evaluate the performance of STADEM.}
+\begin{table}[!h]
+
+\caption{\label{tab:lgr-sim-tab}Summary of different simulation scenarios including varying adult trapping, fallback and re-ascension, nighttime passage, and window count error rates used to evaluate the performance of STADEM.}
 \centering
-\begin{tabular}[t]{llrrrl}
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{l>{\raggedright\arraybackslash}p{1in}>{\raggedleft\arraybackslash}p{1.2in}>{\raggedleft\arraybackslash}p{1.2in}>{\raggedright\arraybackslash}p{1.2in}}
 \toprule
-Scenario & Trap rate & Fallback rate & Re-ascension rate & Nighttime passage rate & Window count error\\
+Scenario & Trap rate & Fallback \& Re-ascension rate & Nighttime passage rate & Window count error\\
 \midrule
-Baseline & 0.15 & 0.06 & 1 & 0.06 & No Error\\
-Baseline Err H & 0.15 & 0.06 & 1 & 0.06 & 10\% Error\\
-Baseline Err L & 0.15 & 0.06 & 1 & 0.06 & 5\% Error\\
-N-R & 0.15 & 0.10 & 1 & 0.05 & No Error\\
-N-R Err H & 0.15 & 0.10 & 1 & 0.05 & 10\% Error\\
-\addlinespace
-N-R Err L & 0.15 & 0.10 & 1 & 0.05 & 5\% Error\\
-N-R trap down & 0.15 and 0.00 3 weeks & 0.06 & 1 & 0.06 & No Error\\
-N-R trap down Err H & 0.15 and 0.00 3 weeks & 0.06 & 1 & 0.06 & 10\% Error\\
-N-R trap down Err L & 0.15 and 0.00 3 weeks & 0.06 & 1 & 0.06 & 5\% Error\\
-Trap down & 0.15 and 0.00 3 weeks & 0.10 & 1 & 0.05 & No Error\\
-\addlinespace
-Trap down Err H & 0.15 and 0.00 3 weeks & 0.10 & 1 & 0.05 & 10\% Error\\
-Trap down Err L & 0.15 and 0.00 3 weeks & 0.10 & 1 & 0.05 & 5\% Error\\
+Baseline & 0.15 & 0.06 & 0.06 & No Error\\
+Baseline Err L & 0.15 & 0.06 & 0.06 & 5\% Error\\
+Baseline Err H & 0.15 & 0.06 & 0.06 & 10\% Error\\
+N-R & 0.15 & 0.10 & 0.05 & No Error\\
+N-R Err L & 0.15 & 0.10 & 0.05 & 5\% Error\\
+N-R Err H & 0.15 & 0.10 & 0.05 & 10\% Error\\
+Trap down & 0.15 and 0.00 3 weeks & 0.06 & 0.06 & No Error\\
+Trap down Err L & 0.15 and 0.00 3 weeks & 0.06 & 0.06 & 5\% Error\\
+Trap down Err H & 0.15 and 0.00 3 weeks & 0.06 & 0.06 & 10\% Error\\
+N-R trap down & 0.15 and 0.00 3 weeks & 0.10 & 0.05 & No Error\\
+N-R trap down Err L & 0.15 and 0.00 3 weeks & 0.10 & 0.05 & 5\% Error\\
+N-R trap down Err H & 0.15 and 0.00 3 weeks & 0.10 & 0.05 & 10\% Error\\
 \bottomrule
-\end{tabular}
+\end{tabular}}
 \end{table}
 
-\begin{table}
-
-\caption{\label{tab:summ-tab}Summary statistics, including relative bias, mean coefficient of variation (CV), root mean squared deviation (RMSD) and coverage for results from each of the twelve simulation scenarios.}
+\newpage
+
+\begin{table}[!h]
+
+\caption{\label{tab:summ-tab}Summary statistics of unique wild fish estimates, including relative bias, mean coefficient of variation (CV), root mean squared error (RMSE) and 95\% credible interval coverage for results from each of the twelve simulation scenarios.}
 \centering
-\begin{tabular}[t]{lrrrr}
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{lrrrrrr}
 \toprule
-Scenario & Relative bias & Mean CV & RMSD & Coverage\\
+Scenario & Relative bias & Mean bias & Mean SE & Mean CV & RMSE & Coverage\\
 \midrule
-Baseline & 0.002 & 0.024 & 551.228 & 0.978\\
-Baseline Err L & 0.002 & 0.024 & 525.404 & 0.984\\
-Baseline Err H & 0.002 & 0.024 & 553.310 & 0.978\\
-Trap Down & 0.003 & 0.030 & 534.096 & 0.996\\
-Trap Down Err L & 0.002 & 0.030 & 546.582 & 0.998\\
-\addlinespace
-Trap Down Err H & 0.003 & 0.030 & 566.949 & 0.994\\
-N-R & -0.001 & 0.023 & 503.934 & 0.988\\
-N-R Err L & 0.002 & 0.024 & 478.226 & 0.990\\
-N-R Err H & 0.002 & 0.024 & 508.481 & 0.984\\
-N-R Trap Down & 0.002 & 0.030 & 523.817 & 0.994\\
-\addlinespace
-N-R Trap Down Err L & 0.003 & 0.030 & 568.434 & 0.996\\
-N-R Trap Down Err H & 0.003 & 0.030 & 580.082 & 0.994\\
+Baseline & 0.002 & 58.5 & 584 & 0.024 & 495 & 0.978\\
+Baseline Err L & 0.002 & 54.3 & 585 & 0.024 & 473 & 0.984\\
+Baseline Err H & 0.002 & 54.0 & 591 & 0.024 & 502 & 0.978\\
+Trap Down & 0.003 & 77.0 & 741 & 0.030 & 495 & 0.996\\
+Trap Down Err L & 0.002 & 60.9 & 735 & 0.030 & 503 & 0.998\\
+Trap Down Err H & 0.003 & 68.2 & 741 & 0.030 & 527 & 0.994\\
+N-R & -0.001 & -12.4 & 574 & 0.023 & 459 & 0.988\\
+N-R Err L & 0.002 & 37.3 & 580 & 0.024 & 428 & 0.990\\
+N-R Err H & 0.002 & 41.0 & 585 & 0.024 & 462 & 0.984\\
+N-R Trap Down & 0.002 & 40.9 & 741 & 0.030 & 501 & 0.994\\
+N-R Trap Down Err L & 0.003 & 64.5 & 732 & 0.030 & 515 & 0.996\\
+N-R Trap Down Err H & 0.003 & 69.4 & 738 & 0.030 & 527 & 0.994\\
 \bottomrule
-\end{tabular}
+\end{tabular}}
 \end{table}
 
-\begin{table}
-
-\caption{\label{tab:lgr-tab}Window counts, and estimates of total, wild, hatchery and hatchery no-clip escapement, with coefficients of variation in parenthesis, for Chinook Salmon and steelhead from spawn years 2010 to 2019.}
+\newpage
+
+\begin{table}[!h]
+
+\caption{\label{tab:lgr-tab}Window counts, the relative bias of window counts compared to estimates of total escapement, and estimates of total, wild, hatchery and hatchery no-clip escapement, with coefficients of variation in parenthesis, for spring/summer-run Chinook Salmon and steelhead from spawn years 2010 to 2019.}
 \centering
-\begin{tabular}[t]{lrlllll}
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{lrllllll}
 \toprule
-Species & Year & Window counts & Total & Wild & Hatchery & Hatchery no-clip\\
+Species & Year & Window counts & Window bias & Total & Wild & Hatchery & Hatchery no-clip\\
 \midrule
-Chinook & 2010 & 134,684 & 133,706 (0.046) & 26,994 (0.053) & 99,609 (0.047) & 7,103 (0.079)\\
-Chinook & 2011 & 134,594 & 123,610 (0.021) & 24,692 (0.027) & 93,845 (0.022) & 5,072 (0.046)\\
-Chinook & 2012 & 84,771 & 83,875 (0.052) & 21,424 (0.048) & 57,929 (0.056) & 4,522 (0.069)\\
-Chinook & 2013 & 70,966 & 69,500 (0.023) & 19,086 (0.033) & 44,159 (0.027) & 6,255 (0.048)\\
-Chinook & 2014 & 114,673 & 106,741 (0.034) & 28,468 (0.037) & 68,895 (0.038) & 9,377 (0.048)\\
-\addlinespace
-Chinook & 2015 & 132,432 & 133,330 (0.03) & 23,887 (0.044) & 98,902 (0.033) & 10,540 (0.059)\\
-Chinook & 2016 & 81,753 & 84,810 (0.031) & 17,300 (0.031) & 59,471 (0.034) & 8,039 (0.036)\\
-Chinook & 2017 & 48,192 & 43,151 (0.039) & 5,162 (0.046) & 34,476 (0.041) & 3,513 (0.051)\\
-Chinook & 2018 & 42,232 & 39,621 (0.037) & 7,004 (0.045) & 28,996 (0.039) & 3,621 (0.052)\\
-Chinook & 2019 & 29,617 & 27,876 (0.104) & 4,758 (0.128) & 21,185 (0.105) & 1,932 (0.11)\\
-\addlinespace
-Steelhead & 2010 & 323,382 & 348,082 (0.035) & 45,320 (0.038) & 266,570 (0.036) & 36,193 (0.039)\\
-Steelhead & 2011 & 208,296 & 217,640 (0.037) & 45,917 (0.037) & 148,564 (0.038) & 23,159 (0.041)\\
-Steelhead & 2012 & 180,320 & 190,947 (0.023) & 40,476 (0.027) & 139,476 (0.024) & 10,995 (0.038)\\
-Steelhead & 2013 & 109,186 & 121,265 (0.037) & 25,160 (0.042) & 85,707 (0.038) & 10,398 (0.058)\\
-Steelhead & 2014 & 108,154 & 117,169 (0.037) & 28,222 (0.067) & 81,145 (0.042) & 7,803 (0.081)\\
-\addlinespace
-Steelhead & 2015 & 165,591 & 176,628 (0.029) & 47,959 (0.036) & 118,185 (0.03) & 10,483 (0.054)\\
-Steelhead & 2016 & 136,126 & 144,310 (0.031) & 36,141 (0.037) & 101,923 (0.031) & 6,245 (0.056)\\
-Steelhead & 2017 & 101,827 & 104,479 (0.035) & 15,472 (0.039) & 80,707 (0.035) & 8,299 (0.061)\\
-Steelhead & 2018 & 74,097 & 69,584 (0.033) & 10,105 (0.038) & 56,865 (0.033) & 2,614 (0.055)\\
-Steelhead & 2019 & 51,818 & 56,024 (0.104) & 11,073 (0.248) & 41,726 (0.079) & 3,224 (0.092)\\
+Chinook & 2010 & 134,684 & 2.3 \% & 131,565 (0.047) & 26,563 (0.054) & 97,995 (0.048) & 7,007 (0.078)\\
+Chinook & 2011 & 134,594 & 8.6 \% & 122,977 (0.024) & 24,566 (0.029) & 93,364 (0.025) & 5,047 (0.046)\\
+Chinook & 2012 & 84,771 & 2 \% & 83,081 (0.047) & 21,206 (0.043) & 57,379 (0.051) & 4,495 (0.065)\\
+Chinook & 2013 & 70,966 & 2.4 \% & 69,273 (0.023) & 19,023 (0.032) & 44,011 (0.027) & 6,238 (0.047)\\
+Chinook & 2014 & 114,673 & 7.2 \% & 106,460 (0.034) & 28,339 (0.036) & 68,770 (0.038) & 9,351 (0.048)\\
+Chinook & 2015 & 132,432 & -0.1 \% & 132,527 (0.029) & 23,752 (0.043) & 98,292 (0.031) & 10,483 (0.059)\\
+Chinook & 2016 & 81,753 & -3 \% & 84,216 (0.027) & 17,187 (0.029) & 59,031 (0.03) & 7,997 (0.035)\\
+Chinook & 2017 & 48,192 & 11 \% & 42,902 (0.038) & 5,137 (0.045) & 34,265 (0.04) & 3,500 (0.049)\\
+Chinook & 2018 & 42,232 & 6.7 \% & 39,416 (0.037) & 6,965 (0.044) & 28,853 (0.039) & 3,597 (0.052)\\
+Chinook & 2019 & 29,617 & 7.8 \% & 27,293 (0.041) & 4,771 (0.045) & 20,578 (0.043) & 1,943 (0.06)\\
+Steelhead & 2010 & 323,382 & -7.4 \% & 347,285 (0.035) & 45,240 (0.037) & 265,983 (0.036) & 36,063 (0.037)\\
+Steelhead & 2011 & 208,296 & -4 \% & 216,633 (0.036) & 45,677 (0.036) & 147,937 (0.037) & 23,019 (0.039)\\
+Steelhead & 2012 & 180,320 & -5.5 \% & 190,171 (0.022) & 40,278 (0.025) & 138,918 (0.023) & 10,975 (0.037)\\
+Steelhead & 2013 & 109,186 & -9.9 \% & 120,025 (0.035) & 24,884 (0.039) & 84,854 (0.036) & 10,286 (0.057)\\
+Steelhead & 2014 & 108,154 & -8.1 \% & 116,913 (0.039) & 28,043 (0.071) & 80,911 (0.043) & 7,959 (0.08)\\
+Steelhead & 2015 & 165,591 & -5.9 \% & 175,285 (0.032) & 47,606 (0.04) & 117,245 (0.032) & 10,433 (0.056)\\
+Steelhead & 2016 & 136,126 & -5.5 \% & 143,646 (0.033) & 35,962 (0.039) & 101,446 (0.033) & 6,237 (0.06)\\
+Steelhead & 2017 & 101,827 & -1.7 \% & 103,524 (0.035) & 15,311 (0.038) & 79,942 (0.035) & 8,270 (0.062)\\
+Steelhead & 2018 & 74,097 & 6.6 \% & 69,180 (0.033) & 10,043 (0.039) & 56,537 (0.034) & 2,601 (0.055)\\
+Steelhead & 2019 & 51,818 & -3.7 \% & 53,738 (0.036) & 8,639 (0.06) & 41,291 (0.033) & 3,808 (0.088)\\
 \bottomrule
-\end{tabular}
+\end{tabular}}
 \end{table}
+
+\newpage
+
+\listoffigures
 
@@
 \begin{figure}
-\centering
-\includegraphics{../figures/stadem-examp-fig-1.pdf}
-\caption{\label{fig:stadem-examp-fig}Schematic of how the STADEM model works.
-Panel A shows the posterior of the estimate of fish crossing the dam while the window is open (dashed line shows observed window counts).
-That estimate is divided by the nighttime passage rate (B).
-The total fish is then discounted by the reascension rate to estimate unique fish (C).
-Those unique fish are then multiplied by the proportion of wild fish (D), to estimate unique wild fish.}
+\includegraphics[width=1\linewidth]{../figures/ModelSchematic} \caption{Directed acyclic graph showing the STADEM model framework.}\label{fig:dag-fig}
 \end{figure}
+
+\newpage
 
@@
 \includegraphics{../figures/rel-bias-lgd-1.pdf}
-\caption{\label{fig:rel-bias-lgd}'Boxplots show relative bias of STADEM estimates for wild escapement across various scenarios.
-The boxes contain 50\% of the simulations, whiskers contain 95\% of the simulations, and points are outliers.}
+\caption{\label{fig:rel-bias-lgd}Boxplots of the relative bias of window counts and STADEM estimates for total unique fish and STADEM estimates of unique wild fish across various scenarios (See Table 1).}
 \end{figure}
+
+\newpage
 
@@
 \includegraphics{../figures/jags-fit-fig-1.pdf}
-\caption{\label{fig:jags-fit-fig}Time-series plot showing estimates of total escapement for Chinook in 2014, including raw window counts, window counts adjusted for night-time passage, trap estimates and STADEM estimates.
-Gray ribbon represents the 95\% credible interval for STADEM estimates.}
+\caption{\label{fig:jags-fit-fig}Time-series plot showing estimates of escapement for spring/summer-run Chinook Salmon in 2014, including window counts, trap estimates and STADEM estimates of unique fish.
+The dark gray ribbon represents the 95\% credible intervals for STADEM estimates, while the light gray ribbon represents the 95\% confidence intervals for the trap estimates.}
 \end{figure}
+
+\newpage
 
@@
 \includegraphics{../figures/night-reasc-diff-fig-1.pdf}
-\caption{\label{fig:night-reasc-diff-fig}Night-time passage rate plotted against re-ascension rate, calculated from observed PIT tags for each week of spawn years 2010-2019.
-Colors correspond to different spawn year, while the size of each point is proportional to the window count that week.
-The dashed line is the 1-1 line.}
+\caption{\label{fig:night-reasc-diff-fig}Nighttime passage rate plotted against re-ascension rate on the logit scale, calculated from observed PIT tags for each week of spawn years 2010-2019.
+The size of each point is proportional to the window count that week.
+Points falling below the 1-1 dashed line indicate weekly periods of higher nighttime passage as compared to re-ascension, conversly, points falling above indicate a lower nighttime passage than re-ascension rate.}
 \end{figure}
@@
 \hypertarget{append1}{%
-\section{Appendix A - STADEM Model Description}\label{append1}}
-
-\hypertarget{total-and-weekly-escapement}{%
-\subsection{Total and Weekly Escapement}\label{total-and-weekly-escapement}}
-
-Escapement at LGD is estimated by combining two independent observations, trap catches and window counts, of the true number of fish crossing the dam in a state-space model with a weekly time-step.
-Both are assumed to be corrupted observations of the true unknown number of fish crossing LGD each week.
-The log of the true number of fish crossing (\(X_t\)), is modeled as a random walk process (Shumway and Stoffer 2010).
-
-\[
-\begin{aligned}
-  \ln \left( X_{t} \right) &= \ln \left( X_{t-1} \right) + e_t \\
-  e_t &\sim \mathcal{N}(0, \sigma^2_X)
-\end{aligned}
-\]
-
-The number of fish caught in the trap, \(Y^T_t\), for week \(t\) is modeled as a binomial process based on the unknown true trap rate that week, \(\nu_t\), and the unknown true number of fish crossing the dam that week, \(X_t\).
-The true trap rate is estimated from a beta distribution with previously estimated parameters, \(\hat{\alpha_t}\) and \(\hat{\beta_t}\), informed by the mark-recapture estimate of the trap rate.
-
-\[
-\begin{aligned}
-  \nu_t &\sim \text{Beta} \left(\hat{\alpha_t}, \hat{\beta_t} \right) \\
-  Y^T_t &\sim \text{Bin} \left( \nu_t, X_t \right) \\
-\end{aligned}
-\]
-
-The estimate of the weekly trap rate is derived based on previously PIT-tagged spring/summer Chinook and steelhead who are crossing LGD that week.
-The fish caught in the trap that week are considered the ``mark'' group, and all the PIT tagged fish who are detected at the upper end of the LGD fish ladder that week are considered the second capture group (which includes recaptures of the the ``marked'' fish).
-From this data, we can estimate the total number of previoiusly PIT tagged fish crossing LGD that week, and the proportion of those that are caught in the trap is the weekly trap rate.
-The uncertainty in estimates of total fish is translated into uncertainty in the trap rate, and this is then summarised with a beta distribution (\(\text{Beta} \left(\hat{\alpha_t}, \hat{\beta_t} \right)\)) that becomes an input into STADEM.
-Although the group of previously PIT tagged fish is not assumed to be representative of the overall run, the rate at which they are caught in the trap should be the same rate that the overall run experiences.
-
-The number of fish counted at the window, \(Y^W_t\), is modeled as a (potentially) over-dispersed negative binomial process, with an expected value of \(X_t^{day}\), the number of fish crossing the dam while the window is open.
-This is simply the total number of fish crossing that week, \(X_t\), multiplied by the propoortion of fish crossing while the window is open for counting, \(\theta_t\), calculated on a weekly basis.
-In the formula below, \(p_t\) is the proportion of fish observed at the window and \(r\) is the shape parameter.
-If \(r\) is estimated to be small provides evidence for over-dispersion, and as it grows very large, the negative binomial distribution behaves like a Poisson distribution.
-
-\[
-\begin{aligned}
-  X_t^{day} &= X_t * \theta_t \\
-  p_t &= \frac{r}{\left( r + X_t^{day} \right)} \\
-  Y^W_t &\sim \text{NegBin} \left( p_t, r \right) \\
-\end{aligned}
-\]
-
-Thus, the unknown true number of fish crossing LGD each week, \(X_t\), is estimated from two different data source: window counts and fish sampled in the trap.
-The window counts provide an estimate (with some potential observer error) of the fish crossing during daytime hours, while the fish in the trap, when expanded by the estimated true trap rate, provide an estimate of the total fish crossing that week.
-For weeks when we have a more precise estimate of the trap rate (i.e.~weeks when lots of previously PIT tagged fish are crossing LGD), STADEM will tend to favor the estimate of total escapement based on the trap data, whereas when that trap rate is more uncertain (e.g.~fewer PIT tagged fish to use in estimating the trap rate), STADEM will rely more on the window counts to estimate total escapement.
-During peak run times, when lots of fish are crossing LGD, estimates based on trap data and trap rates will be more precise, while estimates from the window counts may have more observation error due to so many fish passing the window.
-For weeks when the trap is down, STADEM relies exclusively on the window counts and night-time passage data, but there will be more uncertainty in the estimates.
-
-\hypertarget{day-time-passage-and-re-ascension-rates}{%
-\subsection{Day-time Passage and Re-ascension Rates}\label{day-time-passage-and-re-ascension-rates}}
-
-There are two other processes that must be accounted for, first, the proportion of fish that cross the dam while the window is closed for counting (night-time passage rate), and the second, the proportion of fish that are crossing the dam multiple times (re-ascension rate) and therefore potentially double-counted.
-Both rates can be estimated from previously PIT tagged fish that are crossing the dam each week.
-
-The proportion of fish passing the window during non-operational hours, night-time passage rate, is just the complement of the rate of fish passing during the day when the window is operating.
-The daytime passage rate for week \(t\), \(\theta_t\), is modeled as a random walk process and estimated from a binomial distribution based on the number of PIT tags observed to cross the dam during operational hours, \(y^{day}_t\), and the total number of PIT tags observed to cross the dam at any point that week, \(N_t\) (Shumway and Stoffer 2010).
-
-\[
-\begin{aligned}
-  y^{day}_t &\sim \text{Bin} \left(\theta_t, N_t \right) \\
-  \text{logit} (\theta_{t}) &= \text{logit} \left( \theta_{t-1} \right) + g_t \\
-  g_t &\sim \mathcal{N}(0, \sigma^2_\theta) \\
-\end{aligned}
-\]
-
-The number of total fish crossing Lower Granite differs from the number of unique fish crossing Lower Granite because some fish fall back and re-ascend the dam.
-These fish are potentially double-counted at the window, and have the potential to be caught in the fish trap more than once.
-The number of tags known to be re-ascending the dam each week, \(y^{reasc}_t\), is modeled as a binomial process based on the estimated re-ascension rate, \(\eta_t\), and the total number of tags crossing the dam that week, \(N_t\).
-The logit of the re-ascension rate is modeled as a random walk process similar to day-time passage (Shumway and Stoffer 2010).
-
-\[
-\begin{aligned}
-  y^{reasc}_t &\sim \text{Bin} \left(\eta_t, N_t \right) \\
-  \text{logit} \left( \eta_{t} \right) &= \text{logit} \left(\eta_{t-1} \right) + f_t \\
-  f_t &\sim \mathcal{N}(0, \sigma^2_\eta)
-\end{aligned}
-\]
-
-\hypertarget{origin-proportions}{%
-\subsection{Origin Proportions}\label{origin-proportions}}
-
-After estimating the total number of fish to have crossed Lower Granite each week, \(X_t\), the total must be further refined into the number of wild fish, \(X_{w,t}\), hatchery fish, \(X_{h,t}\) and hatchery no-clip fish, \(X_{hnc, t}\).
-This is done by estimating a weekly origin proportion vector, \(\omega_t\) based on the random sample of fish caught in trap that week, \(Y^T_t\).
-The observed number of wild, \(Y^T_{w,t}\), hatchery, \(Y^T_{h,t}\), and hatchery no-clip, \(Y^T_{hnc,t}\), fish caught in the trap that week is assumed to come from a multinomial distribution with probability vector \(\omega_t\).
-The log-odds ratio of the proportions in \(\omega_t\), in relation to the proportion of hatchery fish, \(\omega_{h,t}\) is modeled as a random walk, so it can change through time.
-This allows the proportions of wild, hatchery and hatchery no-clip fish to shift throughout the season, based on the data available from the fish trap.
-
-\[
-\begin{aligned}
-  \left( Y^T_{w,t}, Y^T_{h,t}, Y^T_{hnc,t} \right) &\sim \text{Multinom} \left( \omega_t, Y^T_t \right) \\
-  \omega_t &= \frac{\exp(\phi_t)}{\sum{\exp(\phi_t})} \\
-  \phi_{w,t} &= \ln \left( \frac{\omega_{w,t}}{\omega_{h,t}} \right) \\
-  \phi_{hnc,t} &= \ln \left( \frac{\omega_{hnc,t}}{\omega_{h,t}} \right) \\
-  \phi_{h,t} &= 0 \\
-  \phi_{w,t} &= \phi_{w, t-1} + d_{w,t} \\
-  \phi_{hnc,t} &= \phi_{hnc, t-1} + d_{hnc,t} \\
-  d_t &\sim \mathcal{N}(0, \sigma^2_\omega) \\
-\end{aligned}
-\]
-
-Finally, the number of unique fish crossing Lower Granite each week, \(X_{w,t}\), is the product of the total fish crossing that week, \(X_t\) multiplied by one minus the re-ascension rate, \((1 - \eta_t)\), and the origin proportion vector, \(\omega_t\).
-
-\[
-\begin{aligned}
-  \begin{bmatrix} X_{w,t} \\ X_{h,t} \\ X_{hnc,t} \end{bmatrix}
-  &= X_t * (1 - \eta_t) * \begin{bmatrix} \omega_{w,t} \\ \omega_{h,t} \\ \omega_{hnc,t} \end{bmatrix}
-\end{aligned}
-\]
-
-\hypertarget{model-fitting}{%
-\subsection{Model Fitting}\label{model-fitting}}
-
-The model was fit using the JAGS program (Plummer 2019), run with R software (R Core Team 2020).
-Uninformative priors were used for \(\sigma_X, \sigma_\eta, \sigma_\theta, \sigma_\omega\) and \(\ln(X_1)\) (Uniform(0,10)), as well as logit(\(\eta_1\)) and logit(\(\theta_1\)) (\(\mathcal{N}(0, 1000)\)), and finally \(\phi_{w,1}\) and \(\phi_{hnc,1}\) (\(\mathcal{N}(0, 100)\)).
-
-\newpage
-
-\hypertarget{append2}{%
-\section{Appendix B - Simulation Details}\label{append2}}
-
-To simulate fish passing a dam, an \(\mathbb{R}\) software function was developed (R Core Team 2020).
-The function randomly samples observations from assumed probability distribution functions (\(\mathcal{PDF}\)) with known parameters.
+\section{Appendix A - Simulation Details}\label{append1}}
+
+To simulate fish passing a dam, we developed an \(\mathbb{R}\) software function (R Core Team 2020).
+The function randomly samples observations from assumed probability distribution functions (pdf) with known parameters.
 Total unique fish, \(N\), and a vector, \(\omega\), containing the proportions of wild (\(w\)), hatchery (\(h\)) and hatchery no-clip (\(hnc\)) fish passing the dam is set to establish known ``truths'' of escapement by origin.
@@
 
-Escapement is then randomly divided across a set number of populations, \(n\), by randomly drawing proportions, \(\phi_{j,p}\), of origin group \(j\) in each poplulation \(p\) using a dirichlet \(\mathcal{PDF}\).
-The dirichlet function is parameterized from a vector, \(\eta_j\), containing 1's and 0's designating populations with origin \(j\) fish returning.
-\(\eta_j\) originates from a bernoulli \(\mathcal{PDF}\) and the proportion of populations with each origin, \(\tau_j\).
+Escapement of each origin is then randomly divided across a set number of populations, \(n\), by randomly drawing proportions, \(\phi_{j,p}\), of origin group \(j\) in each population \(p\) using a Dirichlet pdf.
+The Dirichlet function is parameterized from a vector, \(\zeta_j\), containing 1's and 0's designating populations with origin \(j\) fish returning.
+For each population \(p\), \(\zeta_{j,p}\) is drawn from a Bernoulli pdf using the proportion of populations that contain each origin, \(\tau_j\).
 Wild fish are assumed to be in all populations; \(\tau_w = 1.0\).
@@
 \begin{aligned}
-  \left[\eta_{j,p=1},  ... , \eta_{j,p=n}\right] &\sim \text{Bern}(\tau_j) \\
-  \left[\phi_{j,p=1},  ... , \phi_{j,p=n}\right] &\sim \text{Dir}(\eta_{j,p=1},  ... , \eta_{j,p=n}) \\
+  \zeta_{j,p} &\sim \text{Bernoulli}(\tau_j) \\
+  \left[\phi_{j,p=1},  ... , \phi_{j,p=n}\right] &\sim \text{Dir} \left(\zeta_{j,p=1},  ... , \zeta_{j,p=n} \right) \\
   N_{j,p} &= N_j * \phi_{j,p} \\
-  N_p &= \sum_{j=w, h, hnc} N_{j,p}
+  N_p &= \sum_{j \in w, h, hnc} N_{j,p}
 \end{aligned}
@@
 
-Mean arrival date, \(\bar{a}_p\), for each population returning to the dam is drawn from a normal \(\mathcal{PDF}\) with hyper-parameters \(\mu_a\) and \(\sigma^2_a\).
-Similarly, the variance or spread in run-timing within populations is the absolute value of random variables drawn from a normal \(\mathcal{PDF}\) with hyper-parameters \(\mu_s\) and \(\sigma^2_s\).
+Mean arrival date, \(\bar{a}_p\), for each population returning to the dam is drawn from a normal pdf with hyper-parameters \(\mu_a\) and \(\sigma^2_a\).
+Similarly, the variance or spread in run-timing within populations is the absolute value of random variables drawn from a normal pdf with hyper-parameters \(\mu_s\) and \(\sigma^2_s\).
 
@@
 \begin{aligned}
- \left[\bar{a}_p, ..., \bar{a}_n\right] &\sim \text{Norm}(\mu_a, \sigma^2_a) \\
- \left[s_p, ..., s_n\right] &\sim \text{Norm}(\mu_s, \sigma^2_s)
+ \left[\bar{a}_p, ..., \bar{a}_n\right] &\sim \mathcal{N}(\mu_a, \sigma^2_a) \\
+ \left[s_p, ..., s_n\right] &\sim \left | \mathcal{N}(\mu_s, \sigma^2_s) \right |
 \end{aligned}
@@
 
-After sampling the mean date of arrival and variances for each population, the date of arrival, \(a_{i,p}\), for individual fish, \(i\), within each population are drawn from a normal \(\mathcal{PDF}\) with population parameters \(\bar{a}_p\) and \(s^2_p\).
+After sampling the mean date of arrival and variances for each population, the date of arrival, \(a_{i,p}\), for individual fish, \(i\), within each population are drawn from a normal pdf with population parameters \(\bar{a}_p\) and \(s^2_p\).
 This simulates a random arrival day that is similar for all fish returning to the same population, regardless of origin.
@@
 
-To examine the sensitivites of models to different fish behavior and dam operational scenarios, seven addtional attributes are randomly assigned to each individual fish.
-Each attribute is randomly assigned a TRUE/FALSE using a bernoulli \(\mathcal{PDF}\) and a fixed probability parameter.
+To model different fish behavior and dam operational scenarios, seven additional attributes are randomly assigned to each individual fish.
+Each attribute is randomly assigned a TRUE/FALSE using a Bernoulli pdf and a fixed probability parameter.
 Fish passage during the day-time (i.e., during periods of window operation) is modeled using one minus the night-time passage rate (\(1 - \nu\)).
@@
 Re-ascension occurs with probability \(\rho\), conditioned on fish \(i\) falling back.
-If fish \(i\) falls back and re-ascends, the entire process described above is repeated, with some time-lag between initial ascension and re-ascension that is governed by a Poisson \(\mathcal{PDF}\) with mean = 2 days.
-Fish may fall-back and re-ascend up to 3 times, allowing for the possibility of the same fish being counted or trapped multiple times.
+If fish \(i\) falls back and re-ascends, the entire process described above is repeated, with some time-lag between initial ascension and re-ascension that is governed by a Poisson pdf with mean = 2 days.
+Fish may fallback and re-ascend up to 3 times, allowing for the possibility of the same fish being counted or trapped multiple times.
 
@@
 
-Simulation parameters for model evaluations were set to mimic typical ecapement of spring/summer Chinook to LGD with similar origin proportions, marking rates and run timing as those observed from return years 2010 - 2015.
+Simulation parameters for model evaluations were set to mimic typical escapement of spring/summer Chinook Salmon to LGD with similar origin proportions, marking rates and run timing as those observed from return years 2010 - 2015.
 Escapement of each origin (\(N_j\)) was set at 25,000 wild, 70,000 hatchery and 5,000 hatchery no-clips spread randomly across 25 populations (\(n\)).
@@
 
-For the specific simulated scenarios, we were interested in STADEM model estimates of origin specific escapment from the combinations of two seperate trapping rates, two fallback, re-ascension and night-passage combinations and three window count error rates; resulting in twelve different scenarios.
-First, trapping rates were set static at 0.15 across all weeks for six scenarios to mimic an optimium trap operation for an expected return of 25,000 wild fish (i.e., trap \(\approx\) 4,000 wild fish).
+For the specific simulated scenarios, we were interested in STADEM model estimates of origin specific escapement from the combinations of two separate trapping rates, two fallback, re-ascension and night-passage combinations and three window count error rates; resulting in twelve different scenarios.
+First, trapping rates were set static at 0.15 across all weeks for six scenarios to mimic an optimum trap operation for an expected return of 25,000 wild fish (i.e., trap \(\approx\) 4,000 wild fish).
 For the remaining six scenarios, trapping rates for weeks 30, 31 and 32 (i.e., July \(22^{nd}\) to August \(11^{th}\)) were changed to 0.00 to test STADEM sensitivities to potential trap shut downs similar to those observed in 2013, 2014 and 2015 (Ogden 2014, 2016b, 2016a).
@@
 A potential 5.0\% weekly bias was determined from PIT-tag data and within the range of observed weekly difference for return years 2010 - 2015 (Figure \ref{fig:night-reasc-diff-fig}).
-Finally, we desired to test the sensitivites of STADEM to potential rates of window count error; 0\%, 5\% and 10\% (Hatch et al.
+Finally, we desired to test the sensitivities of STADEM to potential rates of window count error; 0\%, 5\% and 10\% (Hatch et al.
 1994).
-Error at the window was simulated as unbiased (i.e., expect varying high and low counts to cancel each other) and is introduced to the sum of daily window counts using a normal \(\mathcal{PDF}\).
+To simulate window count error, we assumed the observed daily count was a random variable from a normal distribution with a mean equal to the true daily count, and a standard deviation equal to the applied error rate (i.e., 0\%, 5\%, 10\%) multiplied by the true daily count.
+This method simulated observed counts as unbiased, and allowed for possible under and overcounts at the window.
+
+All code for simulating data and fitting STADEM to that data can be found at \url{https://www.github.com/KevinSee/ManuscriptSTADEM}.
 

### 2021-05-24 README.Rmd
updated README with correct DOI and journal information, removed possible journal sections

--- 2020-05-14 README.Rmd
@@
 # Please put your title here to include it in the file below.
-Title <- "Estimating total escapment past Lower Granite Dam using all possible data"
+Title <- "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
 ```
@@
 
-> See, K.E., R.N. Kinzer and M.W. Ackerman, (2020). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
+> See, K.E., R.N. Kinzer and M.W. Ackerman (2021). _`r Title`_. North American Journal of Fisheries Management <https://doi.org/10.1002/nafm.10649>
 
@@
 
-> See, K.E., R.N. Kinzer and M.W. Ackerman, (2020). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> See, K.E., R.N.
+Kinzer and M.W.
+Ackerman (2021).
+_`r Title`_.
+North American Journal of Fisheries Management, Accessed `r format(Sys.Date(), "%d %b %Y")`.
+Online at <https://doi.org/10.1002/nafm.10649>
 
@@
 Kinzer and M.W.
-Ackerman, (`r format(Sys.Date(), "%Y")`).
+Ackerman (`r format(Sys.Date(), "%Y")`).
 _Compendium of R code and data for `r Title`_.
 Accessed `r format(Sys.Date(), "%d %b %Y")`.
-Online at <https://doi.org/xxx/xxx>
+Online at <https://doi.org/10.1002/nafm.10649>
 
